### 2020-05-27 text/article_vedrovice.Rmd
add Makefile edit paths and names

---
title           : "Supplement: Vedrovice burial analysis"
author          : "Petr Pajdla"
    # name          : "Petr Pajdla"
    # affiliation   : "Department of Archaeology and Museology,
    #                  Masaryk University, Brno, Czech Republic"
    # email         : "petr.pajdla@gmail.com"
date            : "`r format(Sys.time(), '%d %B, %Y')`"
abstract        : "Fuck me."
output :
    pdf_document:
        latex: pdflatex
        fig_caption: true
        number_sections: false
        keep_tex: false
bibliography    : "bib_vedrovice.bib"
csl             : "european-journal-of-archaeology.csl"
---

<!-- note: length - approx. 10 - 12 pages (no bibliography, tables, figures) -->
<!-- i.e. 18 000 - 22 000 characters -->

<!-- ======================================================================= -->

# Introduction

### Questions

* Is there evidence that certain artefacts covary,
    either positively or negatively?
    Do certain artefacts tend to occur together, or are certain combinations
    absent or rare?
* How can the presence and absence data best display which burials
    contain similar artefacts,
    and which artefacts have similar distributions over burials?
* *How can the data be analysed so as to show most clearly the relationships*
    *between the different* **burials** in terms of their **contents**?
* *How can the data be analysed so as to show most clearly the relationship*
    *between the different* **artefacts** in terms of their
    **distributions in graves**?
* How are grave good assemblages related to age and sex?
(... variations in burial rite coinciding with age or sex of the deceased ...)

<!-- ======================================================================= -->

# Material and methods

The material analyzed here comes from Vedrovice – Široká u lesa cemetery.
Originally, the burial ground consisted of at least 108 burials.
Some of the burials were disturbed by later settlement and burial pits in the
area of the cemetery, amateur excavations during 19th century or later
agricultural activities.
Thus some of the burials are excluded from the analysis due to missing and/or
incomplete data and only a sample of 78 well-preserved burials is analyzed here.
Most of the data was obtained from the original publication [@podborsky2002],
anthropological determination of sex and age were corrected based on
more recent analysis by @dockalova2008. Carbon, nitrogen and strontium isotope
ratios were collected in @whittle2013 and @richards2008.

Each burial was originally characterized by 55 variables but only 35 are
selected for further analysis.
Some of  the original variables are omitted due to their rare occurrence,
for instance spondylus arm ring is present in a single burial only,
thus being useless in establishing relationships between pairs of burials.
Other variables are removed because they are highly dependent one on each other.

The variables fall into several categories.
Firstly there are biological and anthropological characteristics derived from
retrieved human skeletal remains including sex, age simplified into age groups,
isotope ratios or body height.

Second large category of variables consists of data on artifacts present in the
burials, the grave goods.
These include variables noting quantities of various types of pottery,
positions of the pottery vessels within the graves, lithics and most common
raw materials of lithics, polished stone axes or adzes, grinding slabs,
pebbles, bone tools and different kinds of personal ornaments including beads
made of spondylus shell, pendants or oval shaped spondylus buckles and presence
of ochre pigment.
All the variables in this category are simplified by binarization (presence vs
absence of a trait) in order to minimize effects of only several numerous
cases on the overall pattern, for instance 20 beads in a grave <!-- 70 -->
would have much larger effect then 1 pottery vessels in the same grave.

Last category describes physical characteristics of the grave pit and the
position of the body.
I. e. size of the pit, its orientation, position of the body on one of
the sides and direction of the head.

Data is analyzed in R environment for statistical computing and graphics
[@rcore]  using functions of readr [@wickham2017a], dplyr [@wickham2016] and
reshape2 [@wickham2007] packages to manipulate the data.
Function daisy [@kaufman1990] from package cluster [@maechler2016] is used to
count distances between observations used in further analyses.

The input variables are of mixed data types, thus Gower's similarity
coefficient [@gower1971] is used in order to obtain a square distance matrix.
Gower's coefficient treats differently symmetric and asymmetric dichotomous
data, nominal and ordinal data and interval and ratio scaled data.
Advantage of this approach is that different kinds of variables can be analyzed
together and it is not necessary to transform the variables into same data
types, as it would be if other distance measure was used.
In our case, most of the variables concerning artifacts deposited within the
burials are asymmetric binary data, i. e. the presence of a trait means
similarity but absence does not imply similarity.
Continuous quantitative variables are treated as interval scaled.
Age categories are treated as ordinal variable and the rest of categorical
variables is treated as nominal data.

## Exceptionality index

Exceptionality index ($EI$) is used in order to induce relative exceptionality
of individual burials based on their attributes.
This approach is independent on a prior selection of attributes of importance
or *prestige*.
The concept was introduced by @demjan2015.

It is defined by @demjan2015[367] as a Root Mean Square (*quadratic mean*) of
PCA factor scores:
$$ EI_{attr_g}=\sqrt{\frac{1}{n}\sum_{i=1}^{n}s_i^2} $$
where $s_i$ are factor scores of individual graves. The mean value is
normalized to range $[0-1]$:

$$ EI_{attr_g}=\frac{EI_{attr_g}}{max~EI_{attr_g}} $$

## Distance between burials

<!-- ======================================================================= -->

# Results

<!-- ======================================================================= -->

# Discussion

<!-- ======================================================================= -->

# References

### 2020-05-27 code/knit_text.R

@@ -1,0 +1,10 @@
+# Project "Vedrovice"
+# Script nr. 00
+# KNIT OUTPUT FILE
+# author: Petr Pajdla
+# Create the text of the article...
+
+# knitr::knit(input = here::here("text", "article-vedrovice.Rmd"),  
+#             output = here::here("text", "article-vedrovice.md"))
+
+rmarkdown::render(input = here::here("text", "article_vedrovice.Rmd"))

### 2020-05-27 text/article_vedrovice.Rmd
text - permutation

@@
 Some of the burials were disturbed by later settlement and burial pits in the
-area of the cemetery, amateur excavations during 19th century or later
+area of the cemetery, amateur excavations during the 19th century or later
 agricultural activities.
@@
 thus being useless in establishing relationships between pairs of burials.
-Other variables are removed because they are highly dependent one on each other.
+Other variables are removed because they are highly dependent on each other.
 
@@
 Second large category of variables consists of data on artifacts present in the
-burials, the grave goods.
-These include variables noting quantities of various types of pottery,
-positions of the pottery vessels within the graves, lithics and most common
-raw materials of lithics, polished stone axes or adzes, grinding slabs,
-pebbles, bone tools and different kinds of personal ornaments including beads
-made of spondylus shell, pendants or oval shaped spondylus buckles and presence
-of ochre pigment.
+burials.
+These include variables noting quantities of various types of pottery and
+positions of the pottery vessels within the graves,
+counts of various raw materials of chipped stone tools,
+presence of polished stone axes or adzes, grinding slabs, pebbles,
+bone tools and different kinds of personal ornaments.
+The personal ornaments include beads made of spondylus shell or marble,
+sponylus pendants or oval shaped buckles.
+Presence of ochre pigment is also noted.
 All the variables in this category are simplified by binarization (presence vs
@@
 cases on the overall pattern, for instance 20 beads in a grave <!-- 70 -->
-would have much larger effect then 1 pottery vessels in the same grave.
+would have much larger effect on the analysis then 1 pottery vessel in the same grave.
 
@@
 
-Data is analyzed in R environment for statistical computing and graphics
+Data is analyzed in an R environment for statistical computing and graphics
 [@rcore]  using functions of readr [@wickham2017a], dplyr [@wickham2016] and
@@
 count distances between observations used in further analyses.
+<!-- ADD METHODS AND PACKAGES USED IN R -->
 
-The input variables are of mixed data types, thus Gower's similarity
-coefficient [@gower1971] is used in order to obtain a square distance matrix.
-Gower's coefficient treats differently symmetric and asymmetric dichotomous
-data, nominal and ordinal data and interval and ratio scaled data.
-Advantage of this approach is that different kinds of variables can be analyzed
-together and it is not necessary to transform the variables into same data
-types, as it would be if other distance measure was used.
-In our case, most of the variables concerning artifacts deposited within the
-burials are asymmetric binary data, i. e. the presence of a trait means
-similarity but absence does not imply similarity.
-Continuous quantitative variables are treated as interval scaled.
-Age categories are treated as ordinal variable and the rest of categorical
-variables is treated as nominal data.
+## Randomness of Artefact Combinations
+
+To determine the non-randomness of various artefact combinations we conduct a
+permutation test on the presence/absence data matrix.
+The method based on a theory of generalized Monte Carlo tests was first
+introduced to archaeology by @manly1996.
+Original (observed) co-occurence matrix is permutated $n$ times with constant
+row and column totals, i. e. both the count of artifacts in individual graves
+and sum of the given type of artifacts are fixed.
+
+To examine which associations of variables are infrequent or even not present
+and which tend to appear together we conduct a randomization test
 
@@
 
+The input variables are of mixed data types, thus Gower's similarity
+coefficient [@gower1971] is used in order to obtain a square distance matrix.
+Gower's coefficient treats differently symmetric and asymmetric dichotomous
+data, nominal and ordinal data and interval and ratio scaled data.
+Advantage of this approach is that different kinds of variables can be analyzed
+together and it is not necessary to transform the variables into same data
+types, as it would be if other distance measure was used.
+In our case, most of the variables concerning artifacts deposited within the
+burials are asymmetric binary data, i. e. the presence of a trait means
+similarity but absence does not imply similarity.
+Continuous quantitative variables are treated as interval scaled.
+Age categories are treated as ordinal variable and the rest of categorical
+variables is treated as nominal data.
+
 <!-- ======================================================================= -->

### 2020-05-29 code/dataset.R
add long names for variables

@@ -28,7 +28,8 @@ ved <- list(bin_vars = list(count_mat = NA,
             metadata = NA,
             layout = NA,
             var_names = list(full = NA,
-                             short = NA),
+                             short = NA,
+                             long = NA),
             id_burials = input$id_burial,
             orig_dataset = input)
 
@@ -87,22 +88,38 @@ ved$layout <- input %>% select(layout_x, layout_y) %>%
 rownames(ved$layout) <- input$id_burial
 
 ved$var_names$full <- tibble(vnames = names(input),
-                        abbrv = c("id", "row", "id.unified", "id.full", "undist",
-                                  "x", "y", "sex", "sex.d", "sex.p",
-                                  "age.cat", "age.int", "pit.l", "pit.w", "pit.d",
-                                  "pit.o", "pit.o.cat", "head.o", "head.o.cat",
-                                  "body.side",
-                                  "pigm", "pol.sum", "adz", "axe", "grd", "peb",
-                                  "l.sum", "l.kl", "l.skj", "l.oth", "l.rad", "l.sgs",
-                                  "p.sum", "p.bow", "p.glo", "p.bot", "p.spec", "p.frag",
-                                  "p.top", "p.mid", "p.leg","p.fill",
-                                  "bon.t", "bon.awl", "pen", "pen.L", "pen.U", "pen.I", 
-                                  "buck", "brac", "neck", "b.sp", "b.mar", "b.div",
-                                  "b.cyl", "b.circ", "b.olive", "d.tooth", "ant", "shell",
-                                  "dat", "d13C", "d15N", "Sr", "Sr.ppm", "body.h"))
+                             abbrv = c("id", "row", "id.unified", "id.full", "undist",
+                                       "x", "y", "sex", "sex.d", "sex.p",
+                                       "age.cat", "age.int", "pit.l", "pit.w", "pit.d",
+                                       "pit.o", "pit.o.cat", "head.o", "head.o.cat",
+                                       "body.side",
+                                       "pigm", "pol.sum", "adz", "axe", "grd", "peb",
+                                       "l.sum", "l.kl", "l.skj", "l.oth", "l.rad", "l.sgs",
+                                       "p.sum", "p.bow", "p.glo", "p.bot", "p.spec", "p.frag",
+                                       "p.top", "p.mid", "p.leg","p.fill",
+                                       "bon.t", "bon.awl", "pen", "pen.L", "pen.U", "pen.I", 
[…]

### 2020-05-29 code/randomization.R
add long varnames to v-vals

@@ -131,7 +131,8 @@ v_statistic <- tibble(variable = names(v_observed),
                       v = t(unname(v_observed))[, 1],
                       p = unname(v_p_values),
                       signif = unname(v_p_values < (5 / length(v_p_values)))) %>% 
-  mutate(abbrv = unname(ved$var_names$short[variable]))
+  mutate(abbrv = unname(ved$var_names$short[variable]),
+         full = unname(ved$var_names$long[variable]))
 
 readr::write_csv(v_statistic, here("data/temp", "v_statistics.csv"))
 

### 2020-05-29 text/article_vedrovice.Rmd
text and SI - methods

@@
 ---
-title           : "Supplement: Vedrovice burial analysis"
+title           : "Vedrovice burial analysis"
 author          : "Petr Pajdla"
@@
 abstract        : "Fuck me."
-output :
-    pdf_document:
+output:
+    bookdown::pdf_document2:
+    # bookdown::word_document2:
         latex: pdflatex
@@
 <!-- i.e. 18 000 - 22 000 characters -->
+
+```{r init, include=FALSE}
+knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
+library(tidyverse)
+```
 
@@
 
-## Randomness of Artefact Combinations
+## Randomness of artefact combinations
 
-To determine the non-randomness of various artefact combinations we conduct a
+To determine non-randomness of various artefact combinations we conduct a
 permutation test on the presence/absence data matrix.
-The method based on a theory of generalized Monte Carlo tests was first
+The method based on generalized Monte Carlo tests was first
 introduced to archaeology by @manly1996.
@@
 row and column totals, i. e. both the count of artifacts in individual graves
-and sum of the given type of artifacts are fixed.
+and sum of the given type of artifacts across the burials are fixed.
 
 To examine which associations of variables are infrequent or even not present
-and which tend to appear together we conduct a randomization test
+and which tend to appear together co-occurences of observed and simulated data
+are summarized using two statistics suggested by @manly1996.
+The $S$ statistic is a measure of overall deviation between expected and observed
+number of artefact occurences while
+the $v$ statistic is a measure of difference between co-occurences of
+individual artefact with other artifacts
+[@manly1996, 475, for details see SI].
 
@@
 or *prestige*.
-The concept was introduced by @demjan2015.
-
-It is defined by @demjan2015[367] as a Root Mean Square (*quadratic mean*) of
+It was defined by @demjan2015[367] based on a Root Mean Square (*quadratic mean*) of
 PCA factor scores:
@@
 
+## Randomness of artefact combinations
+
+```{r vtable}
+vstat <- read_csv(here::here("data/temp", "v_statistics.csv")) %>% 
+    select(full, v, p) %>% 
+    mutate(v = round(v, 2), p = round(p, 2)) %>% 
+    rename("variable" = full, "v statistic" = v, "p value" = p)
+
+knitr::kable(
+  vstat,
+  caption = 'Fuuuuu'
+)
+```
+
 <!-- ======================================================================= -->

### 2020-05-29 text/si_vedrovice.Rmd
text and SI - methods

---
title           : "Supplementary Information: Vedrovice burial analysis"
author          : "Petr Pajdla"
    # name          : "Petr Pajdla"
    # affiliation   : "Department of Archaeology and Museology,
    #                  Masaryk University, Brno, Czech Republic"
    # email         : "petr.pajdla@gmail.com"
date            : "`r format(Sys.time(), '%d %B, %Y')`"
output:
    bookdown::pdf_document2:
        latex: pdflatex
        fig_caption: true
        number_sections: false
        keep_tex: false
bibliography    : "bib_vedrovice.bib"
csl             : "european-journal-of-archaeology.csl"
---

# Randomness of Artefact Combinations

For pair of artefacts $i$ and $j$, the observed and expected number of co-occurences are
denoted $o_{ij}$ and $e_{ij}$ respectively.
$R$ is the number of variables.
The $S$ statistic, the overall deviation from expected number of co-occurences for
all artefacts is defined as the mean square of deviations:

\begin{equation}
S = \sum_{i=1}^{R}\sum_{j=1}^{R}\frac{(o_{ij} - e_{ij})^2}{R^2} (\#eq:sstat)
\end{equation}

also:\@ref(eq:sstat)

$$ S = \sum_{i=1}^{R} \frac{v_{i}}{R}$$

where

$$ v_i = \sum_{j=1}^{R} \frac{(o_{ij} - e_{ij})^2}{R}$$

## Implementation

### 2020-06-09 text/article_vedrovice.Rmd
continue analysis spatial analysis

@@
 library(tidyverse)
+library(here)
 ```
@@
 count distances between observations used in further analyses.
-<!-- ADD METHODS AND PACKAGES USED IN R -->
+<!-- ADD METHODS AND PACKAGES USED IN R, or into SI?-->
+
+## Spatial randomness
 
@@
 [@manly1996, 475, for details see SI].
+
+With $n = 999$ permutations of the original data matrix the results obtained
+are robust
 
@@
 
-```{r vtable}
-vstat <- read_csv(here::here("data/temp", "v_statistics.csv")) %>% 
+By counting co-occurences of analyzed variables on simulated data matrices and comparing
+the distributions of the two studied statistics with the values calculated for
+the original dataset we can establish whether the overall pattern of artefacts
+distribution in the graves is non-random and further identify artefacts that deviate
+from the random pattern.
+The value of $S$ statistic is deemed significantly large at the 0.05 alpha level with
+$S \approx 14 $ and $ p \approx $ (Fig. \@ref(fig:f-sstat)).
+This suggests that there is enough evidence that the overall distribution of
+artefacts in the graves is not random.
+
+```{r t-vstat}
+vstat <- read_csv(here("data/temp", "v_statistics.csv")) %>%
     select(full, v, p) %>%
@@
 
+```{r f-sstat, fig.cap = "Distribution of simulated and observed S statistic"}
+knitr::include_graphics(here("plots", "s_statistic.pdf"))
+```
+
+```{r f-vstat, fig.cap = "Simulated and observed v statistics for different variables"}
+knitr::include_graphics(here("plots", "v_values.pdf"))
+```
+
+```{r f-cooc, fig.cap = "Networks of variables frequently present together and exluding each other"}
+knitr::include_graphics(here("plots", "cooc_networks.pdf"))
+```
+
 <!-- ======================================================================= -->

### 2020-06-09 code/common_art_cat.R

@@ -1,0 +1,74 @@
+# Project "Vedrovice"
+# Script nr. 6
+# CREATE GROUPS BASED ON COMMON ARTEFACTS
+# author: Petr Pajdla
+# Create groups based on common artefacts
+
+# pckg
+library(tidyverse)
+
+# data
+ved <- read_rds("./data/vedrovice_dataset.RDS")
+
+non_rand_vars <- read_lines("./data/output/non_random_vars.txt")
+
+ei <- read_csv("./data/output/exceptionality.csv")
+
+# experiment
+
+ved_bin_dist_nr <- dist(ved$bin_vars$bin_mat[, non_rand_vars], method = "binary")
+ved_bin_dist <- dist(ved$bin_vars$bin_mat, method = "binary")
+
+ved_hclust_nr <- hclust(ved_bin_dist_nr, method = "ward.D2")
+ved_hclust <- hclust(ved_bin_dist, method = "ward.D2")
+
+par(mfcol = c(1, 2))
+plot(ved_hclust)
+plot(ved_hclust_nr)
+
+groups_common_items <- cutree(ved_hclust_nr, k = 6)
+
+ved_gathered <- bind_cols(id = ved$metadata$id_burial, 
+                          as_tibble(ved$bin_vars$bin_mat), 
+                          gr_item = groups_common_items,
+                          gr_ei = ei$group) %>% 
+  gather(-gr_item, -gr_ei, -id, key = "item", value = "presence")
+
+# ad hoc function
+prop_item <- function(df) {
+  temp_total <- df %>% group_by(item) %>% 
+    summarise(tot = sum(presence))
+  temp_groups <- df %>% group_by(gr_item, item) %>% 
+    summarise(sum = sum(presence)) %>% 
+    group_map(~ mutate(.x, prop = sum/temp_total$tot))
+  return(temp_groups)
+}
+
+item_prop <- prop_item(ved_gathered) %>% 
+  bind_rows(.id = "gr_item") %>% 
+  mutate(gr_item = str_c("group", gr_item))
+
+# randomly distributed items
+ggplot(item_prop, aes(y = prop, x = item)) +
+  geom_col(fill = "white", color = "black") +
+  facet_wrap(~ gr_item, nrow = 1) +
+  coord_flip() +
+  labs(x = "artefact category", y = "proportion")
+
+item_prop %>% group_by(item) %>% summarise(sumx = sum(prop))
+
+# non-randomly distributed items
[…]

### 2020-06-09 code/dataset.R

@@ -64,7 +64,7 @@ ved$cat_vars$cat_vars <- input %>%
 rownames(ved$cat_vars$cat_vars) <- ved$id_burials
 
 # metadata
-lookup_sex <- c("C" = "gold", "F" = "tomato", "M" = "steelblue")
+lookup_sex <- c("n. a." = "gold", "F" = "tomato", "M" = "steelblue")
 
 ved$metadata <- input %>% select(id_burial, sex, age_cat, dat) %>% 
   mutate(sex = factor(sex),
@@ -93,9 +93,9 @@ ved$var_names$full <- tibble(vnames = names(input),
                                        "age.cat", "age.int", "pit.l", "pit.w", "pit.d",
                                        "pit.o", "pit.o.cat", "head.o", "head.o.cat",
                                        "body.side",
-                                       "pigm", "pol.sum", "adz", "axe", "grd", "peb",
+                                       "pigm", "pol.sum", "adz", "axe", "grnd", "peb",
                                        "l.sum", "l.kl", "l.skj", "l.oth", "l.rad", "l.sgs",
-                                       "p.sum", "p.bow", "p.glo", "p.bot", "p.spec", "p.frag",
+                                       "p.sum", "p.bow", "p.glo", "p.bot", "p.oth", "p.frag",
                                        "p.top", "p.mid", "p.leg","p.fill",
                                        "bon.t", "bon.awl", "pen", "pen.L", "pen.U", "pen.I", 
                                        "buck", "brac", "neck", "b.sp", "b.mar", "b.div",
@@ -107,7 +107,7 @@ ved$var_names$full <- tibble(vnames = names(input),
                                        "Pit orientation", "Pit orientation", "Head orientation", "Head orientation",
                                        "Side",
                                        "Pigment", "PST(sum)", "Adze", "Axe", "Grinding tool", "Pebble",
-                                       "Lith. (sum)", "Lith. (KF)", "Lith (Krak)", "Lith. (other)", "Lith. (Rad.)", "Lith. (SGS)",
+                                       "Lithics (sum)", "Lithics (KF)", "Lithics (SKJ)", "Lithics (other)", "Lithics (Rad.)", "Lithics (SGS)",
                                        "Pottery (sum)", "Bowl", "Globular pot", "Bottle", "Other pottery", "Pottery fragment",
                                        "Pot around head", "Pot in the middle ", "Pot around legs","Pot in fill",
[…]

### 2020-06-09 code/distances_vedro.R

@@ -1,0 +1,439 @@
+# Project "Vedrovice"
+# Script nr. 4
+# DISTANCES BETWEEN BURIALS
+# author: Petr Pajdla
+# what is here?
+
+set.seed(42)
+
+# packages =====================================================================
+library(dplyr)
+library(cluster)
+library(igraph)
+
+# functions ====================================================================
+# continuous vector to dichotomous (binarization)
+binarize <- function(x, threshold = 0) {
+  x[x > threshold] <- 1
+  x[x <= threshold] <- 0
+  return(x)
+}
+
+# replace missing (NA) data with mean for the vector
+replace_na <- function(x) {
+  avg <- mean(x, na.rm = TRUE)
+  x[is.na(x)] <- avg
+  return(x)
+}
+
+# perform chi-squre test on multiple columns of a data frame df
+# colpairs contains pairs of column name combinations that chi-squared test is 
+# performed on, created by grid.expand() on a vector of column names
+multiple.chisq.test <- function(df, colpairs) {
+  p_val <- data.frame(colpairs[, 1:2], 
+                      rep(NA, nrow(colpairs)),
+                      rep(NA, nrow(colpairs)))
+  colnames(p_val) <- c("var1", "var2", "p.val", "independent")
+  for (i in 1:nrow(colpairs)) {
+    cont_tab <- table(df[, colpairs[i, 1:2]])
+    p_val[i, "p.val"] <- round(chisq.test(cont_tab)$p.value, 3)
+    if (p_val[i, "p.val"] <= 0.05) {
+      p_val[i, "independent"] <- FALSE
+    } else {
+      p_val[i, "independent"] <- TRUE
+    }
+  }
+  return(p_val)
+}
+
+# count v values for a data frame (based on articles by Manly)
+# where: R...nr of variables, df...data frame of coocurences, 
+# expected = df of expected  coocurences (med.coocs)
+get_v <- function(df, expected = cooc_exp, R = length(vars_binary)) {
+  v_vals <- rep(NA, R)
+  w_df <- left_join(df, expected, by = c("var1", "var2"))
+  v_vals <- w_df %>% group_by(var1) %>% 
+    summarise(v = sum((nr.cooc - mean.cooc)^2 / R))
+  return(v_vals)
+}
+
+# expand grid function to contain only unique combinations of vector values
+# expand.grid.unique <- function(x, y, include.equals=FALSE) {
[…]

### 2020-06-09 code/exceptionality_index.R

@@ -1,0 +1,54 @@
+# Project "Vedrovice"
+# Script nr. 3
+# COUNT EXCEPTIONAITY INDEX OF BURIALS
+# author: Petr Pajdla
+# Exceptionality index of individual burials in order to create groups...
+
+# packages =====================================================================
+library(dplyr)
+
+# functions ====================================================================
+# root mean square (quadratic mean) for vector x
+rms <- function(x) {
+  x <- sqrt(mean(x^2))
+  return(x)
+}
+
+# data =========================================================================
+ved <- readRDS("./data/vedrovice_dataset.RDS")
+non_rand_vars <- readLines("./data/output/non_random_vars.txt")
+# exceptionality index =========================================================
+# non random variables
+art_pca <- prcomp(ved$bin_vars$bin_mat[, non_rand_vars], 
+                  scale. = FALSE, center = FALSE)
+summary(art_pca)
+ggbiplot::ggbiplot(art_pca)
+
+# # for_pca <- ved %>% select(c("pit_len", "pit_wid", "pit_dep", vars_binary)) %>% 
+# #   mutate_at(c("pit_len", "pit_wid", "pit_dep"), scale) %>% 
+# #   mutate_all(replace_na)
+# 
+# # pca object
+# pca <- princomp(for_pca, scores = T)
+# # summary(pca)
+
+# counting ei for individual burials
+ei <- apply(art_pca$x[, 1:9], 1, rms) / max(apply(art_pca$x[, 1:9], 1, rms))
+hist(as.matrix(ei))
+
+ei_tbl <- tibble(burial = names(ei), ei)
+ei_tbl %>% arrange(desc(ei))
+
+ggplot(ei_tbl, aes(x = forcats::fct_reorder(burial, ei), y = ei)) +
+  geom_point() +
+  coord_flip() +
+  xlab("burial nr.")
+
+ei_tbl %>% arrange(desc(ei)) %>% View()
+
+output <- ei_tbl %>% mutate(group = if_else(ei > 0.6, "group1",
+                                  if_else((ei < 0.6 & ei > 0.5), "group2", 
+                                          if_else((ei < 0.5 & ei > 0.4), "group3",
+                                                  if_else((ei < 0.4 & ei > 0.2), "group4", "group5")))))
+
+readr::write_csv(output, "./data/output/exceptionality.csv")

### 2020-06-09 code/exp_mixsiar.R

@@ -1,0 +1,47 @@
+library(tidyverse)
+library(MixSIAR)
+
+x <- read_rds("./data/vedrovice_dataset.RDS")
+
+write_csv(bind_cols(x$metadata[, c("id_burial", "sex")],
+                    x$cont_vars$cont_vars[, c("d13c", "d15n")]), 
+            "./data/output/mixing.csv")
+          
+
+source_data <- tribble(
+  ~spec, ~Meand13c, ~SDd13c, ~Meand15n, ~SDd15n, ~n,
+  "cattle",-20.2,  0.29, 6.2,    1.04, 3,
+  "sheep", -19.8,  0.18, 5.9,    0.38, 3,
+  "pig",   -20.4,  0.53, 8.2,    0.86, 2,
+  "fishy", -18, 0.3, 13, 2.5, 2
+)
+
+write_csv(source_data, "./data/output/source_data.csv")
+
+discr <- tribble(
+  ~spec, ~Meand13c, ~SDd13c, ~Meand15n, ~SDd15n,
+  "cattle", 0, 0, 0, 0,
+  "sheep", 0, 0, 0, 0,
+  "pig", 0, 0, 0, 0,
+  "fishy", 0, 0, 0, 0
+)
+
+write_csv(discr, "./data/output/discr_data.csv")
+
+mix <- load_mix_data(filename = "./data/output/mixing.csv", 
+              iso_names = c("d13c", "d15n"), factors = c("sex"),
+              fac_random = FALSE, fac_nested = FALSE, cont_effects = NULL)
+
+srcs <- load_source_data("./data/output/source_data.csv", source_factors = NULL,
+                        conc_dep = FALSE, data_type = "means", mix = mix)
+
+discr <- load_discr_data("./data/output/discr_data.csv", mix = mix)
+
+plot_data(mix = mix, source = srcs, discr = discr)
+calc_area(srcs, mix, discr)
+plot_prior(alpha.prior = 1, srcs)
+write_JAGS_model("./data/output/mixsiar_model.txt", resid_err = TRUE, process_err = FALSE, mix = mix, source = srcs)
+
+jags.1 <- run_model(run="test", mix, srcs, discr, "./data/output/mixsiar_model.txt", 
+                    alpha.prior = 1, TRUE, FALSE)
+

### 2020-06-09 code/exp_simmr.R

@@ -1,0 +1,79 @@
+library(tidyverse)
+library(simmr)
+
+# vignette("quick_start", package = "simmr")
+# vignette("simmr", package = "simmr")
+
+# load data - vedrovice
+x <- read_rds("./data/vedrovice_dataset.RDS")
+
+# prepare data (matrix, grouping factor)
+mix_in <- cbind(x$metadata$sex, x$cont_vars$cont_vars[, c("d13c", "d15n")]) %>% 
+  drop_na()
+
+mix <- as.matrix(mix_in[, 2:3])
+grp <- mix_in$`x$metadata$sex`
+
+# create sources data (Bickle, Whittle 2013, p. 121)
+source_data <- tribble(
+  ~spec, ~Meand13c, ~SDd13c, ~Meand15n, ~SDd15n, ~n,
+  "cattle",-20.2,  0.29, 6.2,    1.04, 3,
+  "sheep", -19.8,  0.18, 5.9,    0.38, 3,
+  "pig",   -20.4,  0.53, 8.2,    0.86, 2
+)
+
+s_names <- source_data$spec
+s_means <- source_data[, c("Meand13c", "Meand15n")]
+s_sd <- source_data[, c("SDd13c", "SDd15n")]
+
+# create simmr object
+simmr_obj <- simmr_load(mix, s_names, s_means, s_sd, group = grp)
+
+# overall plot
+plot(simmr_obj, group = 1:3)
+
+# fit model
+simmr_out <- simmr_mcmc(simmr_obj)
+
+# validate - values should be close to 1
+summary(simmr_out, type='diagnostics', group = 1:3)
+
+# blue y values should fall within light blue y_rep intervals
+posterior_predictive(simmr_out, group = 1)
+posterior_predictive(simmr_out, group = 2)
+posterior_predictive(simmr_out, group = 3)
+
+# prior prob. distr. viz
+prior_viz(simmr_out, group = 1)
+prior_viz(simmr_out, group = 2)
+prior_viz(simmr_out, group = 3)
+
+summary(simmr_out, type='statistics', group = 1:3)
+summary(simmr_out, type='quantiles', group = 1:3)
+
+# plots
+plot(simmr_out, type = 'density', group = 1)
+plot(simmr_out, type = 'density', group = 2)
+plot(simmr_out, type = 'density', group = 3)
+
+# matrices
+# large  negative correlations - model is not able to differentiate the two groups
+plot(simmr_out, type = 'matrix', group = 1)
[…]

### 2020-06-09 code/explore_exceptionality.R

@@ -1,0 +1,99 @@
+# Project "Vedrovice"
+# Script nr. 5
+# EXPLORE EXCEPTIONALITY
+# author: Petr Pajdla
+# Explore exceptionality of burials, what does it cause etc...
+
+# libraries
+library(tidyverse)
+
+# data 
+base <- read_rds("./data/vedrovice_dataset.RDS")
+
+ei <- read_csv("./data/output/exceptionality.csv")
+
+non_random_vars <- read_lines("./data/output/non_random_vars.txt")
+
+# explore...
+# spatial distribution
+spatial_clustering <- hclust(dist(base$layout, method = "euclidean"))
+spatial_groups <- cutree(spatial_clustering, k = 8)
+
+
+ggplot(as_tibble(base$layout), aes(layout_x, layout_y, color = factor(spatial_groups))) +
+  geom_point() + 
+  ggforce::geom_mark_hull(expand = unit(2.2, "mm"))
+
+ggplot(as_tibble(base$layout), aes(layout_x, layout_y, color = factor(spatial_groups))) +
+  geom_point() + 
+  ggforce::geom_mark_ellipse(expand = unit(2, "mm"))
+
+ved_spatial <- bind_cols(as_tibble(base$layout), ei = ei$group, spat_gr = spatial_groups)
+
+ggplot(ved_spatial, aes(layout_x, layout_y, color = ei)) +
+  geom_point() + 
+  facet_wrap(~factor(spat_gr))
+
+ggplot() +
+  geom_point(ved_spatial[, c("layout_x", "layout_y")], mapping = aes(layout_x, layout_y), alpha = 0.1) +
+  geom_point(ved_spatial, mapping = aes(layout_x, layout_y, color = ei)) +
+  # stat_density2d() +
+  facet_wrap(~ei)
+
+# spat distr. using kmeans (experiment)
+spat_kmeans <- kmeans(base$layout, centers = 8, nstart = 200)
+
+scree1 <- function(df) {
+  x <- vector(mode = "double", length = 12)
+   for (i in 1:12) {
+     x[i] <- kmeans(df, centers = i, nstart = 20)$tot.withinss
+   }
+  return(x)
+}
+
+plot(scree1(base$layout), type = "b")
+
+ggplot(ved_spatial, aes(layout_x, layout_y, color = factor(spat_kmeans$cluster))) +
+  geom_point() +
+  ggforce::geom_mark_ellipse(expand = unit(2, "mm"))
+
+
+# groups items...
[…]

### 2020-06-09 code/randomization.R

@@ -62,8 +62,8 @@ cooc_obs <- cooc_obs %>% filter(var1 != var2)
 # cooc_obs %>% tidyr::spread(var2, nr.cooc)
 
 # randomization of co-occurrence matrix - list of many matrices ----------------
+n_permutations <-  999
 # # ======
-# n_permutations <-  999
 # rand_mat <- vegan::permatfull(ved$bin_vars$bin_mat,
 #                               fixedmar = "both",
 #                               mtype = "prab",
@@ -110,31 +110,40 @@ v_experimental <- bind_rows(parallel::parLapply(cl, v_experimental,
                                                 tidyr::spread, "var1", "v"))
 parallel::stopCluster(cl)
 
-# visualizing of v values by small multiples -----------------------------------
-v_exp_g <- v_experimental %>% tidyr::gather(variable, value) %>% 
-  mutate(abbrv = forcats::as_factor(unname(ved$var_names$short[variable])))
-v_obs_g <- v_observed %>% tidyr::gather(variable, value) %>% 
-  mutate(abbrv = forcats::as_factor(unname(ved$var_names$short[variable])))
-
-ggplot() +
-  geom_density(data = v_exp_g, mapping = aes(x = value), fill = "white") +
-  geom_vline(data = v_obs_g, mapping = aes(xintercept = value)) +
-  facet_wrap(~abbrv, scales = "free", nrow = 4) +
-  xlab("v statistic")
-
-ggsave(here("plots", "v_values.pdf"), width = 12, height = 8)
-
 # P-val: percentage of experimental v values that are larger than observed v ---
-v_p_values <- purrr::map2(v_experimental, v_observed, get_p, 1000) %>% unlist()
+v_p_values <- purrr::map2(v_experimental, v_observed, get_p, n_permutations) %>% unlist()
 
 v_statistic <- tibble(variable = names(v_observed),
                       v = t(unname(v_observed))[, 1],
                       p = unname(v_p_values),
                       signif = unname(v_p_values < (5 / length(v_p_values)))) %>% 
   mutate(abbrv = unname(ved$var_names$short[variable]),
-         full = unname(ved$var_names$long[variable]))
+         long = unname(ved$var_names$long[variable]))
 
 readr::write_csv(v_statistic, here("data/temp", "v_statistics.csv"))
+
[…]

### 2020-06-09 code/recoding_vars.R

@@ -1,0 +1,26 @@
+# recoding data, recode and reimport to spreadsheet...
+
+input <- read_csv("./input/data_vedrovice.csv")
+
+# recoding head_orient =========================================================
+# output <- input %>% transmute(id_burial, 
+#                     head_orient_new = ifelse(input$head_orient > 90 - (45/2) & input$head_orient < 90 + (45/2), "E", 
+#                                           ifelse(input$head_orient > 135 - (45/2) & input$head_orient < 135 + (45/2), "SE", 
+#                                                  ifelse(input$head_orient > 180 - (45/2) & input$head_orient < 180 + (45/2), "S", 
+#                                                         ifelse(input$head_orient > 315 - (45/2) - 2 & input$head_orient < 315 + (45/2), "NW", "NOPE")))))
+# 
+# write_csv(output, "./output/head_orient.csv")
+
+# recoding pit_orient ==========================================================
+# # CHECK DATA !!! 
+# input %>% group_by(pit_orient_cat) %>% summarise(min = min(pit_orient, na.rm = T), max = max(pit_orient, na.rm = T), n())
+# hist(input$pit_orient)
+# ggplot(input, aes(pit_orient, factor(pit_orient_cat))) + geom_jitter()
+# ggplot(input, aes(pit_orient)) + geom_histogram(binwidth = 2)
+
+# binarizing values ============================================================
+
+# binary <- biclust::binarize(input[, 14:36], threshold = 0)
+# 
+# write_csv(binary, "./output/binary.csv")
+

### 2020-06-09 code/spatial_org.R

@@ -1,0 +1,161 @@
+# Project "Vedrovice"
+# Script nr. ?
+# SPATIAL ARRANGEMENT
+# author: Petr Pajdla
+# Spatial organization within the cemetery
+
+library(here)
+library(tidyverse)
+library(broom)
+library(spatstat)
+library(sf)
+
+# read data
+ved <- read_rds(here("data/temp", "vedrovice_dataset.RDS"))
+
+# data prep ====================================================================
+# solving problems with negative coordinaes of polygon window
+ved_layout <- as_tibble(ved$layout + 10)
+
+# create window ----------------------------------------------------------------
+# not a rectangle, but a polygon closely bounding the burials using convex hull
+ved_sf <- st_as_sf(bind_cols(ved_layout, ved$metadata), 
+                   coords = c("layout_x", "layout_y")) %>% 
+  mutate(sex = fct_relevel(sex, c("n. a.", "F", "M", "ind.")))
+# ved_sf_buffer <- st_buffer(ved_sf, dist = 1.6)
+# ved_sf_buffer <- st_simplify(ved_sf_buffer, dTolerance = 1)
+ved_sf_hull <- st_convex_hull(st_union(st_geometry(ved_sf)))
+
+# excavation polygon
+ved_exc <- matrix(c(c(0.5, 0.5, 0, 0, -1, -1, 1.6, 1.6, 7.6, 7.6, 9.1, 9.1, 
+                      12.3, 12.3, 13.5, 13.5, 12.3, 12.3, 8.6, 8.6),
+                    c(-1, 2.5, 2.5, 10.3, 10.3, 14.3, 14.3, 13.6, 13.6, 
+                      12.8, 12.8, 13.4, 13.4, 12.4, 12.4, 4.5, 4.5, 1.5, 1.5, -1)), 
+                  ncol = 2, byrow = FALSE)
+ved_exc <- st_cast(st_as_sf(st_geometry(st_multipoint(ved_exc + 10))), "POLYGON")
+
+# window_rect <- owin(xrange = c(10, 22), yrange = c(10, 22.7))
+window_poly <- as.owin(st_cast(ved_sf_hull, "POLYGON"))
+window_exc <- as.owin(ved_exc)
+
+ggplot(data = ved_sf) +
+  geom_sf(data = ved_exc, fill = "gray90", color = NA) +
[…]

### 2020-06-17 text/article_vedrovice.Rmd
spatial analysis

@@
 
-## Spatial randomness
+## Spatial organization
+
+Spatial arrangement of prehistoric burial grounds is often examined to find
+any internal structures, arrangements in lines, possible grouping and so forth.
+We approach the task as a point pattern analysis problem given that
+a burial (event) is simplified as a point location in geographic space.
+The excavated polygon is used as a study area (a window).
+Using a window closely approximating locations of the burials was also considered,
+e. g. buffer zone or convex/concave hull but simplified excavation polygon is
+deemed the most useful because it preserves the *natural* borders and empty areas
+in the burial ground.
+
+Overall structure of the burial ground is examined using both density based approach,
+and distance based methods [@osullivan2010, 121-154].
+The density of burials is examined using quadrat test and the $G$ function while
+the distance between neighbours usnig Ripley's $K$ function and its derivative, the
+$L$ function [@ripley1976].
+Given statistics for independent random process are estimated using Monte Carlo simulation
+with 99 iterations and the most extreme results for each of the statistics
+are used to construct envelopes encompassing the expected ranges
+
+```{r spat-gl, fig.cap = "G and L functions"}
+knitr::include_graphics(here("plots", "pointprocess_gl.pdf"))
+```
 
@@
 The value of $S$ statistic is deemed significantly large at the 0.05 alpha level with
-$S \approx 14 $ and $ p \approx $ (Fig. \@ref(fig:f-sstat)).
+$S \approx 14$ and $p \approx xy$ (Fig. \@ref(fig:f-sstat)).
 This suggests that there is enough evidence that the overall distribution of
@@
 
-```{r t-vstat}
-vstat <- read_csv(here("data/temp", "v_statistics.csv")) %>%
-    select(full, v, p) %>%
-    mutate(v = round(v, 2), p = round(p, 2)) %>%
-    rename("variable" = full, "v statistic" = v, "p value" = p)
+<!-- ```{r t-vstat} -->
+<!-- vstat <- read_csv(here("data/temp", "v_statistics.csv")) %>% -->
+<!--     select(full, v, p) %>% -->
+<!--     mutate(v = round(v, 2), p = round(p, 2)) %>% -->
+<!--     rename("variable" = full, "v statistic" = v, "p value" = p) -->
 
-knitr::kable(
-  vstat,
-  caption = 'Fuuuuu'
-)
-```
+<!-- knitr::kable( -->
+<!--   vstat, -->
+<!--   caption = 'Fuuuuu' -->
+<!-- ) -->
+<!-- ``` -->
 

### 2020-06-17 code/spatial_org.R
combine plots of functions

@@ -87,7 +87,7 @@ estimate_L <- function(x) {
 # @fun = function name
 plot_estimate <- function(x, fun) {
   stopifnot(verifyclass(x, "tbl_df"))
-  stopifnot(all(names(ved_l) %in% c("r", "obs", "theo", "lo", "hi")))
+  stopifnot(all(names(x) %in% c("r", "obs", "theo", "lo", "hi")))
   ylabel = paste0(fun, "(r)")
   p1 <- x %>% 
     ggplot() +
@@ -98,7 +98,7 @@ plot_estimate <- function(x, fun) {
     scale_x_continuous(expand = c(0, 0)) +
     scale_y_continuous(expand = c(0, 0)) +
     theme_minimal()
-  ggsave(here("plots", paste0("pointprocess_", fun, ".pdf")))
+  # ggsave(here("plots", paste0("pointprocess_", fun, ".pdf")))
   p1
 }
 
@@ -130,6 +130,15 @@ plot_estimate(ved_k, "K")
 ved_l <- estimate_L(ved_pp)
 
 plot_estimate(ved_l, "L")
+
+# export combined figure
+
+grid_gl <- gridExtra::grid.arrange(plot_estimate(ved_g, "G"), 
+                                   plot_estimate(ved_l, "L"),
+                                   nrow = 1)
+
+ggsave(plot = grid_gl, here("plots", "pointprocess_gl.pdf"), device = "pdf",
+       width = 8, height = 4)
 
 # for different marks ==========================================================
 ved_pp_split <- split(ved_pp)

### 2020-06-17 text/article_vedrovice.Rmd
continue spatial analysis

@@
 
-Overall structure of the burial ground is examined using both density based approach,
+Overall structure of the burial ground is examined using both density based approach
 and distance based methods [@osullivan2010, 121-154].
-The density of burials is examined using quadrat test and the $G$ function while
-the distance between neighbours usnig Ripley's $K$ function and its derivative, the
-$L$ function [@ripley1976].
+The density of burials is examined using quadrat test while
+the distance between neighbours using the $G$, Ripley's $K$ [@ripley1976] and $L$ functions.
+
+The quadrat test divides the area of interest into several cells of fixed sizes
+(quadrats) and compares the count of events per quadrat with the count
+expected under the null hypothesis that the underlying point process
+is completely spatially random (approximated by a Poisson distribution).
+The $G$ function is a metric of how close events are and summarises
+cummulative frequency distribution of the nearest neighbour distances.
+Because of its dependence on the intensity of events in the study area, it is
+subject to changes in the size of the study area.
+Ripley's $K$ function (and variance stabilized $L$ function) is more resilient
+to these changes and is widely used in archaeology [cf. @negre2018].
+
 Given statistics for independent random process are estimated using Monte Carlo simulation
@@
 are used to construct envelopes encompassing the expected ranges
-
-```{r spat-gl, fig.cap = "G and L functions"}
-knitr::include_graphics(here("plots", "pointprocess_gl.pdf"))
-```
 
@@
 
+## Spatial organization
+
+```{r spat-gl, fig.cap = "G and L functions"}
+knitr::include_graphics(here("plots", "pointprocess_fun.pdf"))
+```
+
 ## Randomness of artefact combinations

### 2020-06-17 code/spatial_org.R

@@ -40,7 +40,7 @@ window_exc <- as.owin(ved_exc)
 
 ggplot(data = ved_sf) +
   geom_sf(data = ved_exc, fill = "gray90", color = NA) +
-  geom_sf(data = ved_sf_hull, fill = NA, color = "gray40", linetype = 3) +
+  # geom_sf(data = ved_sf_hull, fill = NA, color = "gray40", linetype = 3) +
   geom_sf(aes(shape = sex)) + 
   scale_shape_manual(values = c(22, 21, 24, 4)) +
   # ggsflabel::geom_sf_text_repel(aes(label = id_burial)) +
@@ -132,12 +132,11 @@ ved_l <- estimate_L(ved_pp)
 plot_estimate(ved_l, "L")
 
 # export combined figure
-
 grid_gl <- gridExtra::grid.arrange(plot_estimate(ved_g, "G"), 
-                                   plot_estimate(ved_l, "L"),
+                                   plot_estimate(ved_k, "K"),
                                    nrow = 1)
 
-ggsave(plot = grid_gl, here("plots", "pointprocess_gl.pdf"), device = "pdf",
+ggsave(plot = grid_gl, here("plots", "pointprocess_fun.pdf"), device = "pdf",
        width = 8, height = 4)
 
 # for different marks ==========================================================

### 2020-06-19 code/spatial_org.R
add code for not distance scaled L L = sqrt(K/pi) but in O'Sullivan L = sqrt(K/pi) - r

@@ -72,8 +72,8 @@ plot(ved_pp, use.marks = FALSE)
 # @x = ppp object
 estimate_L <- function(x) {
   stopifnot(verifyclass(x, "ppp"))
-  l <- envelope(x, Kest) %>% 
-    fix_data_frame() %>% 
+  l <- envelope(x, Kest, nrank = 2, nsim = 99) %>%
+    fix_data_frame() %>%
     mutate(r = r,
            obs = sqrt(obs/pi) - r,
            theo = sqrt(theo/pi) - r,
@@ -81,6 +81,20 @@ estimate_L <- function(x) {
            hi = sqrt(hi/pi) - r)
   return(l)
 }
+
+# note to self: in most of the sources, the distance r is not subtracted, 
+# see Nakoinz & Knitter 2016, p. 138
+# estimate_L <- function(x) {
+#   stopifnot(verifyclass(x, "ppp"))
+#   l <- envelope(x, Kest, nrank = 2, nsim = 99) %>%
+#     fix_data_frame() %>%
+#     mutate(r = r,
+#            obs = sqrt(obs/pi),
+#            theo = sqrt(theo/pi),
+#            lo = sqrt(lo/pi),
+#            hi = sqrt(hi/pi))
+#   return(l)
+# }
 
 # plot function estimate
 # @x = tidy tibble of estimated values
@@ -111,32 +125,36 @@ quadrat.test(ved_pp, alternative = "regular")
 # quadrat test suggests clustering
 
 # estimating function G
-ved_g <- envelope(ved_pp, Gest) %>% 
+ved_g <- envelope(ved_pp, Gest, nrank = 2, nsim = 99) %>% 
   fix_data_frame()
-# ved_f <- envelope(ved_pp, Fest) %>% 
-#   fix_data_frame()
 
 plot_estimate(ved_g, "G")
+
+# # F function
+# ved_f <- envelope(ved_pp, Fest, nrank = 2, nsim = 99) %>%
+#   fix_data_frame()
+# 
+# plot_estimate(ved_f, "F")
 
 # distance based ---------------------------------------------------------------
 # estimating Ripley's K function
 # ved_k <- Kest(ved_pp)
-ved_k <- envelope(ved_pp, Kest) %>% 
+ved_k <- envelope(ved_pp, Kest, nrank = 2, nsim = 99) %>% 
   fix_data_frame()
 
 plot_estimate(ved_k, "K")
 
-# estimating L function
[…]

### 2020-06-19 text/article_vedrovice.Rmd
continue spatial analysis first/second-order effects

@@
 Spatial arrangement of prehistoric burial grounds is often examined to find
-any internal structures, arrangements in lines, possible grouping and so forth.
+any internal structure, arrangement or possible grouping.
 We approach the task as a point pattern analysis problem given that
 a burial (event) is simplified as a point location in geographic space.
-The excavated polygon is used as a study area (a window).
+A polygon of the excavation extent is used as a study area (a window).
 Using a window closely approximating locations of the burials was also considered,
@@
 and distance based methods [@osullivan2010, 121-154].
-The density of burials is examined using quadrat test while
-the distance between neighbours using the $G$, Ripley's $K$ [@ripley1976] and $L$ functions.
+It is argued that density based approaches, for instance quadrat tests,
+characterize the point pattern in terms of its *first-order properties* [@osullivan2010, 126].
+The assumption underlying the *first-order effects* is that any point location is
+independent on other point locations and different parameters influence the location
+[@nakoinz2016, 130-135; @osullivan2010, 107].
+On the other hand the methods based on distances between nearest neighbours
+characterize the point pattern in regards to *second-order effects*.
+In this case it is presumed that the point location depends on other
+point locations.
+Functions $G$, Ripley's $K$ [@ripley1976] and $L$ are explored as the most widely
+used distance-based point pattern measures [@nakoinz2016, 135-144; @osullivan2010, 130-139]
+
+We assume that in case of a burial ground *first-order properties* are
+of little to no importance at all.
+In case of spatial analysis of settlement structure, individual locations of
+settlements (points) can depend on a number of parameters, for instance
+elevation, soil type etc.
+<!-- It is possible that the position of burials was dependent on parameters that were not
+captured by the excavation, for example location of trees but that is impossible to test.-->
+In case of a grave placement we presume that the choice of a location is influenced
+solely by other (previous) grave locations.
+The *second-order properties*, and possibly *third-order properties* [@nakoinz2016, 130, 144]
+which take into account triplets of points, thus effectively the overall structure of the pattern,
+are considered major factors in locating the graves.
 

### 2020-06-19 text/article_vedrovice.Rmd
spatial patterning cont. T function

@@
 
-## Spatial organization
+## Spatial pattern
 
@@
 [@nakoinz2016, 130-135; @osullivan2010, 107].
-On the other hand the methods based on distances between nearest neighbours
-characterize the point pattern in regards to *second-order effects*.
+On the other hand the methods based on distances between nearest neighbours and triples
+characterize the point pattern based on *second- and third-order effects*.
 In this case it is presumed that the point location depends on other
 point locations.
-Functions $G$, Ripley's $K$ [@ripley1976] and $L$ are explored as the most widely
-used distance-based point pattern measures [@nakoinz2016, 135-144; @osullivan2010, 130-139]
+<!-- [@nakoinz2016, 135-144; @osullivan2010, 130-139] -->
 
@@
 of little to no importance at all.
-In case of spatial analysis of settlement structure, individual locations of
+In spatial analysis of settlement patterns individual locations of
 settlements (points) can depend on a number of parameters, for instance
@@
 captured by the excavation, for example location of trees but that is impossible to test.-->
-In case of a grave placement we presume that the choice of a location is influenced
-solely by other (previous) grave locations.
+In case of a grave placement in a burial ground we presume that the choice of
+a location is influenced predominantly by other (previous) grave locations.
 The *second-order properties*, and possibly *third-order properties* [@nakoinz2016, 130, 144]
-which take into account triplets of points, thus effectively the overall structure of the pattern,
-are considered major factors in locating the graves.
+are thus considered major factors in locating of the graves.
 
-The quadrat test divides the area of interest into several cells of fixed sizes
+To characterize the pattern based on density, the quadrat test is employed.
+It divides the area of interest into several cells of fixed sizes
 (quadrats) and compares the count of events per quadrat with the count
@@
 is completely spatially random (approximated by a Poisson distribution).
+Functions $G$, $F$, Ripley's $K$ [@ripley1976] and $T$ [@schladitz2000] are
+based on the nearest neighbour distances [@nakoinz2016, 135-144].
 The $G$ function is a metric of how close events are and summarises
@@
 subject to changes in the size of the study area.
-Ripley's $K$ function (and variance stabilized $L$ function) is more resilient
-to these changes and is widely used in archaeology [cf. @negre2018].
+The $F$ function, also empty-space function, is based on simulating a point pattern
+and calculating of distances between random locations and the observed locations.
+It is a measure of how far are observed events from any arbitrary location [@osullivan2010, 132-135].
+Ripley's $K$ function is in comparison with the $G$ function resilient to
+changes in the window size and works with all the proximal events up to a
+certain threshold, not only the nearest neighbours [@nakoinz2016, 138].
+It is widely used in archaeology [@negre2018].
+Last but not least, the $T$ function is a third-order parallel to the $K$ function [@schladitz2000].
 
-Given statistics for independent random process are estimated using Monte Carlo simulation
-with 99 iterations and the most extreme results for each of the statistics
-are used to construct envelopes encompassing the expected ranges
+Given statistics for independent random process are estimated using Monte Carlo
+simulation with 99 iterations.
+The most extreme results of simulations are discarded as outliers and the
+second most extreme results are used to construct envelopes encompassing the
+expected range for a given function [@baddeley2016 384-403]
 

### 2020-06-23 code/spatial_neighbors.R
correct neighbor sex randomization

@@ -48,27 +48,27 @@ ggplot() +
   geom_sf(data = ved_sf, shape = 21, fill = "white") +
   theme_void()
 
-ggsave(here("plots", "plan_vedrovice_gabriel.pdf"), scale = 2)
+# ggsave(here("plots", "plan_vedrovice_gabriel.pdf"), scale = 2)
 
 # get sex of neighbors
-g <- graph_from_edgelist(cbind(from = ved_gabriel$from, to = ved_gabriel$to), 
-                         directed = FALSE)
-V(g)$name <- ved$metadata$id_burial
+ved_g <- graph_from_edgelist(cbind(from = ved_gabriel$from, to = ved_gabriel$to), 
+                             directed = FALSE)
+V(ved_g)$name <- ved$metadata$id_burial
 
 ved_sex <- as.character(ved$metadata$sex)
 names(ved_sex) <- ved$metadata$id_burial
 
-get_sex_neighbors <- function() {
+neigh_sex <- function(g, sex_vector) {
   res <- vector(mode = "list", length = length(V(g)))
   names(res) <- V(g)$name
   for (i in seq_along(V(g))) {
     x <- attr(neighbors(g, v = i), "names")
-    res[[i]] <- unname(ved_sex[x])
+    res[[i]] <- unname(sex_vector[x])
   }
   return(res)
 }
 
-nb_sex <- get_sex_neighbors() %>% 
+nb_sex <- neigh_sex(ved_g, ved_sex) %>% 
   map(tibble) %>% 
   map(set_names, "to_sex") %>% 
   bind_rows(.id = "id") %>% 
@@ -79,45 +79,44 @@ nb_sex <- get_sex_neighbors() %>%
   mutate(from_sex = unname(ved_sex[id])) %>% 
   ungroup() %>% 
   group_by(from_sex, to_sex) %>% 
-  summarise(mean = mean(n), .groups = "drop")
+  summarise(mean = mean(n), sum = sum(n), .groups = "drop")
 
 # randomization of sex for neighbors
-randomize_sex <- function(n_sim) {
-  n_bur <- nrow(ved$metadata)
-  prob <- ved$metadata %>% 
-    select(id_burial, sex) %>% 
-    group_by(sex) %>% 
-    count() %>% 
+randomize_neigh_sex <- function(g, n_sim, metadata) {
+  n_sim <- n_sim
+  n_bur <- nrow(metadata)
+  prob <- metadata %>%
[…]

### 2020-06-23 code/spatial_org.R
add x axis text

@@ -117,7 +117,9 @@ plot_estimate <- function(x, fun) {
     theme(panel.border = element_rect(colour = "black", fill = NA, size = 0.8),
           panel.background = element_blank(),
           line = element_blank(),
-          text = element_blank())
+          axis.text.y = element_blank(),
+          axis.title.y = element_blank(),
+          axis.title.x = element_blank())
   # ggsave(here("plots", paste0("pointprocess_", fun, ".pdf")))
   p1
 }

### 2020-06-23 text/article_vedrovice.Rmd
results - spatial patterning

@@
 
-## Spatial pattern
+## Spatial patterning
 
-Spatial arrangement of prehistoric burial grounds is often examined to find
+Spatial patterning in prehistoric burial grounds is often examined to find
 any internal structure, arrangement or possible grouping.
 We approach the task as a point pattern analysis problem given that
-a burial (event) is simplified as a point location in geographic space.
+a burial (event) is simplified as a point location in a two dimensional space.
 A polygon of the excavation extent is used as a study area (a window).
@@
 The assumption underlying the *first-order effects* is that any point location is
-independent on other point locations and different parameters influence the location
+independent on other point locations and different parameters influence the location.
+On the other hand the methods based on distances between neighbours and triples of neighbors
+characterize the point pattern based on *second- and third-order effects*
 [@nakoinz2016, 130-135; @osullivan2010, 107].
-On the other hand the methods based on distances between nearest neighbours and triples
-characterize the point pattern based on *second- and third-order effects*.
-In this case it is presumed that the point location depends on other
-point locations.
 <!-- [@nakoinz2016, 135-144; @osullivan2010, 130-139] -->
 
-We assume that in case of a burial ground *first-order properties* are
-of little to no importance at all.
+We presume that certain burial location is determined by other burial locations
+and is not quite dependent on on other parameters.
+In other words *first-order properties* are of little to no importance at all in
+our point of view.
 In spatial analysis of settlement patterns individual locations of
@@
 second most extreme results are used to construct envelopes encompassing the
-expected range for a given function [@baddeley2016 384-403]
+expected range for a given function [@baddeley2016 384-403].
+
+To explore whether neighboring individuals are more often of the same sex we
+construct a Gabriel graph [@gabriel1969] with burial locations as nodes (Fig. \@ref(fig:gabriel)).
+For each node, mean sex of the neighbors is determined and recorded.
+To check how common the observed value is, we perform a permutation test
+where sex of burials is randomly assigned while preserving the overall
+proportion of given categories and the neighborhood struture determined by the
+Gabriel graph.
+The observed values of neighbors sex categories are compared against distributions
+derived from the simulated data.
+
+```{r gabriel, fig.cap = "Gabriel graph"}
+knitr::include_graphics(here("plots", "plan_vedrovice_gabriel.pdf"))
+```
 
@@
 
-## Spatial pattern
+## Spatial patterning
 
-```{r spat-gl, fig.cap = "G and L functions"}
+The quadrat test suggests both random and clustered processes with p values
+under the 0.05 alpha value but the $\chi^2$ approximation is inaccurate due to small
+expected counts (for the details see supplementary information).
+Fig. \@ref(fig:spat-funs) shows that the curves of $G$ and $F$ functions do not
+deviate too much from the curves expected under independent random process nor
+do they traverse the boundaries of the envelopes.
+This tells us that both the cummulative nearest-neighbor distance (*G*) and
+distance to any random location in the space of the window (*F*) are well
+approximated by random processes and do not show evidence for spatial clustering.
+On the other hand, both *K* and *T* curves have a steep rise and exceed the
+envelope boundaries.
+These functions thus show evidence for clustering process in the observed point
+pattern.
+
+```{r spat-funs, fig.cap = "Functions *G*, *F*, *K* and *T*. *Dashed line* -- theoretical curve under CSR, *solid line* -- empirical curve, *gray area* -- envelope."}
 knitr::include_graphics(here("plots", "pointprocess_fun.pdf"))
+```
+
+```{r neigh-sex, fig.cap = "Mean neighbors in a given sex category. *Gray area* -- density estimation for the simulated data, *vertical line* -- observed value."}
+knitr::include_graphics(here("plots", "prob_neighbour_sex.pdf"))
 ```

### 2020-06-23 code/spatial_neighbors.R
create theme universe

@@ -13,6 +13,15 @@ library(igraph)
 
 set.seed(42)
 
+# theme for ggplot graphics
+theme_universe <- theme(panel.border = element_rect(colour = "black", fill = NA, size = 0.8),
+                        panel.background = element_blank(),
+                        line = element_blank(),
+                        strip.background = element_blank(), 
+                        axis.text.y = element_blank(), 
+                        axis.title.y = element_blank(), 
+                        panel.spacing = unit(1, "lines"))
+
 # read data
 ved <- read_rds(here("data/temp", "vedrovice_dataset.RDS"))
 
@@ -124,12 +133,7 @@ ved_rand_sex %>%
   facet_grid(to_sex ~ from_sex, scales = "free_y") +
   scale_y_continuous(expand = c(0, 0)) +
   scale_x_continuous(expand = c(0, 0)) +
-  theme(panel.border = element_rect(colour = "black", fill = NA, size = 0.8),
-        panel.background = element_blank(),
-        line = element_blank(),
-        strip.background = element_blank(), 
-        axis.text.y = element_blank(), axis.title.y = element_blank(), 
-        panel.spacing = unit(1, "lines")) +
-  xlab("mean neighbours")
+  xlab("mean neighbours") +
+  theme_universe
 
 ggsave(here("plots", "prob_neighbour_sex.pdf"), width = 12, height = 6)

### 2020-06-23 code/spatial_org.R

@@ -10,6 +10,16 @@ library(broom)
 library(spatstat)
 library(sf)
 
+# theme for ggplot graphics
+theme_universe <- theme(panel.border = element_rect(colour = "black", fill = NA, size = 0.8),
+                        panel.background = element_blank(),
+                        line = element_blank(),
+                        strip.background = element_blank(), 
+                        strip.text = element_text(face = "italic"),
+                        axis.text.y = element_blank(), 
+                        axis.title.y = element_blank(), 
+                        panel.spacing = unit(1, "lines"))
+
 # read data
 ved <- read_rds(here("data/temp", "vedrovice_dataset.RDS"))
 
@@ -114,13 +124,24 @@ plot_estimate <- function(x, fun) {
     # labs(y = ylabel) + 
     scale_x_continuous(expand = c(0, 0)) +
     scale_y_continuous(expand = c(0, 0)) +
-    theme(panel.border = element_rect(colour = "black", fill = NA, size = 0.8),
-          panel.background = element_blank(),
-          line = element_blank(),
-          axis.text.y = element_blank(),
-          axis.title.y = element_blank(),
-          axis.title.x = element_blank())
+    theme_universe +
+    theme(axis.title.x = element_blank())
   # ggsave(here("plots", paste0("pointprocess_", fun, ".pdf")))
+  p1
+}
+
+plot_estimate_facet <- function(x) {
+  stopifnot(verifyclass(x, "tbl_df"))
+  stopifnot(all(names(x) %in% c("fun", "r", "obs", "theo", "lo", "hi")))
+  p1 <- x %>% 
+    ggplot() +
+    geom_ribbon(aes(r, ymin = lo, ymax = hi), fill = "gray80", alpha = 0.6) +
+    geom_line(aes(r, theo), linetype = 2) +
+    geom_line(aes(r, obs), size = 0.8) + 
+    facet_wrap(vars(fun), nrow = 1, scales = "free") +
+    scale_x_continuous(expand = c(0, 0)) +
+    scale_y_continuous(expand = c(0, 0)) +
+    theme_universe
   p1
 }
 
@@ -164,11 +185,21 @@ ved_t <- envelope(ved_pp, Tstat, nrank = 2, nsim = 99) %>%
 plot_estimate(ved_t, "T")
 
 # export combined figure
-grid_fns <- gridExtra::grid.arrange(plot_estimate(ved_g, "G"),
-                                    plot_estimate(ved_f, "F"),
-                                    plot_estimate(ved_k, "K"),
[…]

### 2020-06-23 text/article_vedrovice.Rmd
separate figure labels from code block titles

@@
 
-```{r spat-funs, fig.cap = "Functions *G*, *F*, *K* and *T*. *Dashed line* -- theoretical curve under CSR, *solid line* -- empirical curve, *gray area* -- envelope."}
+```{r label-spat-funs}
+label_spat_funs <- paste("Functions *G*, *F*, *K* and *T*.",
+"*Dashed line* -- theoretical curve under complete spatial randomness, *solid line* -- empirical curve, *gray area* -- envelope.")
+```
+```{r spat-funs, fig.cap = label_spat_funs}
 knitr::include_graphics(here("plots", "pointprocess_fun.pdf"))
@@
 
-```{r neigh-sex, fig.cap = "Mean neighbors in a given sex category. *Gray area* -- density estimation for the simulated data, *vertical line* -- observed value."}
+```{r label-neigh_sex}
+label_neigh_sex <- paste("Mean neighbors in a given sex category",
+"*Gray area* -- density estimation for the simulated data, *vertical line* -- observed value.",
+"*F* -- female, *M* -- male, *n. a.* -- non adult.")
+```
+```{r neigh-sex, fig.cap = label_neigh_sex}
 knitr::include_graphics(here("plots", "prob_neighbour_sex.pdf"))

### 2020-07-14 code/knit_text.R
knit both word and pdf

@@ -7,4 +7,7 @@
 # knitr::knit(input = here::here("text", "article-vedrovice.Rmd"),  
 #             output = here::here("text", "article-vedrovice.md"))
 
-rmarkdown::render(input = here::here("text", "article_vedrovice.Rmd"))
+rmarkdown::render(input = here::here("text", "article_vedrovice.Rmd"), 
+                  output_format = bookdown::pdf_document2())
+rmarkdown::render(input = here::here("text", "article_vedrovice.Rmd"), 
+                  output_format = bookdown::word_document2())

### 2020-07-14 code/spatial_neighbors.R
scale to meters and rotate to north

@@ -14,10 +14,14 @@ library(igraph)
 set.seed(42)
 
 # theme for ggplot graphics
-theme_universe <- theme(panel.border = element_rect(colour = "black", fill = NA, size = 0.8),
+# theme for ggplot graphics
+theme_universe <- theme(panel.border = element_rect(colour = "black", 
+                                                    fill = NA, 
+                                                    size = 0.8),
                         panel.background = element_blank(),
                         line = element_blank(),
                         strip.background = element_blank(), 
+                        strip.text = element_text(face = "italic"),
                         axis.text.y = element_blank(), 
                         axis.title.y = element_blank(), 
                         panel.spacing = unit(1, "lines"))
@@ -26,28 +30,15 @@ theme_universe <- theme(panel.border = element_rect(colour = "black", fill = NA,
 ved <- read_rds(here("data/temp", "vedrovice_dataset.RDS"))
 
 # data prep ====================================================================
-# solving problems with negative coordinaes of polygon window
-ved_layout <- ved$layout + 10
-
-# create window ----------------------------------------------------------------
-# not a rectangle, but a polygon closely bounding the burials using convex hull
-ved_sf <- st_as_sf(bind_cols(as_tibble(ved_layout), ved$metadata), 
-                   coords = c("layout_x", "layout_y")) %>% 
-  mutate(sex = fct_relevel(sex, c("n. a.", "F", "M", "ind.")))
-
-# excavation polygon
-ved_exc <- matrix(c(c(0.5, 0.5, 0, 0, -1, -1, 1.6, 1.6, 7.6, 7.6, 9.1, 9.1, 
-                      12.3, 12.3, 13.5, 13.5, 12.3, 12.3, 8.6, 8.6),
-                    c(-1, 2.5, 2.5, 10.3, 10.3, 14.3, 14.3, 13.6, 13.6, 
-                      12.8, 12.8, 13.4, 13.4, 12.4, 12.4, 4.5, 4.5, 1.5, 1.5, -1)), 
-                  ncol = 2, byrow = FALSE)
-ved_exc <- st_cast(st_as_sf(st_geometry(st_multipoint(ved_exc + 10))), "POLYGON")
+ved_sf <- read_sf(here("data/temp", "layout.shp"))
[…]

### 2020-07-14 code/spatial_org.R

@@ -11,7 +11,9 @@ library(spatstat)
 library(sf)
 
 # theme for ggplot graphics
-theme_universe <- theme(panel.border = element_rect(colour = "black", fill = NA, size = 0.8),
+theme_universe <- theme(panel.border = element_rect(colour = "black", 
+                                                    fill = NA, 
+                                                    size = 0.8),
                         panel.background = element_blank(),
                         line = element_blank(),
                         strip.background = element_blank(), 
@@ -24,8 +26,12 @@ theme_universe <- theme(panel.border = element_rect(colour = "black", fill = NA,
 ved <- read_rds(here("data/temp", "vedrovice_dataset.RDS"))
 
 # data prep ====================================================================
-# solving problems with negative coordinaes of polygon window
-ved_layout <- as_tibble(ved$layout + 10)
+scale_m <- 4.762 # scale by a factor so 1 unit is 1 meter
+angle <- 40 / 360 * 100 * 1/2 # rotate by an angle of 40° so north faces upwards
+
+ved_layout <- DescTools::Rotate(ved$layout * scale_m, 
+                                theta = angle, mx = 0, my = 0)
+ved_layout <- tibble(layout_x = ved_layout$x, layout_y = ved_layout$y)
 
 # create window ----------------------------------------------------------------
 # not a rectangle, but a polygon closely bounding the burials using convex hull
@@ -42,21 +48,39 @@ ved_exc <- matrix(c(c(0.5, 0.5, 0, 0, -1, -1, 1.6, 1.6, 7.6, 7.6, 9.1, 9.1,
                     c(-1, 2.5, 2.5, 10.3, 10.3, 14.3, 14.3, 13.6, 13.6, 
                       12.8, 12.8, 13.4, 13.4, 12.4, 12.4, 4.5, 4.5, 1.5, 1.5, -1)), 
                   ncol = 2, byrow = FALSE)
-ved_exc <- st_cast(st_as_sf(st_geometry(st_multipoint(ved_exc + 10))), "POLYGON")
+ved_exc <- DescTools::Rotate(ved_exc * scale_m, theta = angle, mx = 0, my = 0)
+ved_exc <- st_cast(
+  st_as_sf(
+    st_geometry(
+      st_multipoint(
+        matrix(c(ved_exc$x, ved_exc$y), ncol = 2)))), "POLYGON")
 
 # window_rect <- owin(xrange = c(10, 22), yrange = c(10, 22.7))
[…]

### 2020-07-14 text/article_vedrovice.Rmd
continue writitng article

@@
 ---
-title           : "Vedrovice burial analysis"
+title           : "A resampling approach to spatial arrangement and contents of burials in the Vedrovice cemetery"
 author          : "Petr Pajdla"
@@
 date            : "`r format(Sys.time(), '%d %B, %Y')`"
-abstract        : "Fuck me."
+abstract        :
+  "Spatial patterns in cemeteries and grave goods associations are often explored using various informal methods.
+  We adopt computationaly intensive methods utilizing resampling approaches both to spatial analysis and
+  distributions of objects from the Early Neolithic cemetery in Vedrovice.
+
+  As a result of the investigation we conclude that...
+  "
+
 output:
@@
 We approach the task as a point pattern analysis problem given that
-a burial (event) is simplified as a point location in a two dimensional space.
+a burial (event) is simplified as its centroid location in a two dimensional space.
 A polygon of the excavation extent is used as a study area (a window).
-Using a window closely approximating locations of the burials was also considered,
+Using a window closely approximating locations of the burials was considered,
 e. g. buffer zone or convex/concave hull but simplified excavation polygon is
 deemed the most useful because it preserves the *natural* borders and empty areas
-in the burial ground.
+of the original cemetery.
 
 Overall structure of the burial ground is examined using both density based approach
-and distance based methods [@osullivan2010, 121-154].
+and distance based methods <!--[@osullivan2010, 121-154]-->.
 It is argued that density based approaches, for instance quadrat tests,
@@
 
-We presume that certain burial location is determined by other burial locations
-and is not quite dependent on on other parameters.
-In other words *first-order properties* are of little to no importance at all in
-our point of view.
 In spatial analysis of settlement patterns individual locations of
@@
 elevation, soil type etc.
+In case of a grave placement in a burial ground we presume that the choice of
+a location is influenced predominantly by other (previous) grave locations and
+is largely independent on other parameters.
+In other words *first-order properties* are of little to no importance in
+our point of view.
+The *second-order properties*, and possibly *third-order properties* [@nakoinz2016, 130, 144]
+are thus considered major factors in locating of the graves.
 <!-- It is possible that the position of burials was dependent on parameters that were not
 captured by the excavation, for example location of trees but that is impossible to test.-->
-In case of a grave placement in a burial ground we presume that the choice of
-a location is influenced predominantly by other (previous) grave locations.
-The *second-order properties*, and possibly *third-order properties* [@nakoinz2016, 130, 144]
-are thus considered major factors in locating of the graves.
 
@@
 is completely spatially random (approximated by a Poisson distribution).
-Functions $G$, $F$, Ripley's $K$ [@ripley1976] and $T$ [@schladitz2000] are
+Functions $G$, $F$, Ripley's $K$ [@ripley1976; @sayer2013] and $T$ [@schladitz2000] are
 based on the nearest neighbour distances [@nakoinz2016, 135-144].
@@
 Last but not least, the $T$ function is a third-order parallel to the $K$ function [@schladitz2000].
+
+If $s_i$ is an event location, then 
+$G$ function is given by the formula
+\@ref(eq:gfun)
+
+\begin{equation}
+  \label{eq:gfun}
+  G(d) = \frac{\#[d_{min}(s_i)<d]}{n}
+\end{equation}
+
+$F$ function by 
+$$F(d) = \frac{\#[d_{min}(p_i, S)<d]}{m}$$
+where ${p_1 \cdots p_i \cdots p_m}$ is a set of $m$ randomly selected locations.
+Ripley's $K$ function
+$$K(d) = \frac{\alpha}{n}\frac{1}{n}\sum_{i=1}^n \#[S \in C(s_i, d)]$$
+$T$ function
+$$ $$
 
@@
 
+<!--
+This study takes advantage of computer-intensive resam-
+pling techniques that are based on simulations of the behavior
+of statistical models through the generation of random samples
+(Baxter, 2003; Efron and Tibshirani, 1993; Manly, 1991). The
+strength of these techniques lies in the potential to effectively
+distinguish between random and non-random patterns and to
+test differences between samples without making assumptions
+about the population from which these samples were drawn
+(Drennan and Peterson, 2004). These techniques proved to
+be very effective for the analysis of patterns in cemeteries
+(Manly, 1996).
+@sosna2008 -->
+
+@sosna2013
+
 To determine non-randomness of various artefact combinations we conduct a
@@
 The method based on generalized Monte Carlo tests was first
-introduced to archaeology by @manly1996.
+introduced to archaeology by @manly1996 and adapted by @sosna2008.
+Our approach is based on both of these works.
 Original (observed) co-occurence matrix is permutated $n$ times with constant

### 2020-08-04 text/article_vedrovice.Rmd
add spatial analysis: kde, neighbours by buffer

@@
 ---
-title           : "A resampling approach to spatial arrangement and contents of burials in the Vedrovice cemetery"
+title           : "Spatial Analysis and Grave Goods Distribution at the Cemetery of Vedrovice"
+# "A resampling approach to spatial arrangement and contents of burials in the Vedrovice cemetery"
+
 author          : "Petr Pajdla"

### 2020-08-04 code/dataset.R

@@ -30,13 +30,19 @@ ved <- list(bin_vars = list(count_mat = NA,
             var_names = list(full = NA,
                              short = NA,
                              long = NA),
-            id_burials = input$id_burial,
+            id_burials = NA,
             orig_dataset = input)
 
 # filling the data -------------------------------------------------------------
 
+# burial id
+ved$id_burials <- input %>% 
+  filter(undisturbed == TRUE) %>% 
+  pull(id_burial)
+
 # binary variables - artefact coocurences
 ved$bin_vars$count_mat <- input %>% 
+  filter(undisturbed == TRUE) %>% 
   select(pigment, pol_adze, pol_axe, grinding, pebble, 
          lit_local, lit_nonlocal, 
          pot_special, pot_bowl, pot_globular, pot_bottle, 
@@ -51,15 +57,18 @@ rownames(ved$bin_vars$count_mat) <- ved$id_burials
 ved$bin_vars$bin_mat <- apply(ved$bin_vars$count_mat, 2, binarize)
 
 # continuous variables
-ved$cont_vars$cont_vars <-  input %>% select(pit_len, pit_wid, pit_dep,
-                                             d13c, d15n, sr, sr_ppm, 
-                                             body_height) %>% 
+ved$cont_vars$cont_vars <-  input %>% 
+  filter(undisturbed == TRUE) %>% 
+  select(pit_len, pit_wid, pit_dep,
+         d13c, d15n, sr, sr_ppm, 
+         body_height) %>% 
   as.data.frame()
 
 rownames(ved$cont_vars$cont_vars) <- ved$id_burials
 
 # factor variables
 ved$cat_vars$cat_vars <- input %>% 
+  filter(undisturbed == TRUE) %>% 
   select(pit_orient_cat, head_orient_cat, body_side) %>%
   mutate_all(factor) %>% 
   as.data.frame()
@@ -90,7 +99,6 @@ ved$layout <- input %>% select(layout_x, layout_y) %>%
 
 rownames(ved$layout) <- input$id_burial
 
-
 # variable names ----------------------------------------------------------
 
 ved$var_names$full <- tibble(vnames = names(input),
@@ -108,19 +116,19 @@ ved$var_names$full <- tibble(vnames = names(input),
                                        "b.cyl", "b.circ", "b.olive", "d.tooth", "ant", "shell",
                                        "dat", "d13C", "d15N", "Sr", "Sr.ppm", "body.h"),
[…]

### 2020-08-04 code/exceptionality_index.R

@@ -1,11 +1,12 @@
 # Project "Vedrovice"
-# Script nr. 3
-# COUNT EXCEPTIONAITY INDEX OF BURIALS
+# Script nr. 2.2
+# EXCEPTIONAITY INDEX FOR BURIALS
 # author: Petr Pajdla
 # Exceptionality index of individual burials in order to create groups...
 
 # packages =====================================================================
-library(dplyr)
+library(here)
+library(tidyverse)
 
 # functions ====================================================================
 # root mean square (quadratic mean) for vector x
@@ -15,11 +16,29 @@ rms <- function(x) {
 }
 
 # data =========================================================================
-ved <- readRDS("./data/vedrovice_dataset.RDS")
-non_rand_vars <- readLines("./data/output/non_random_vars.txt")
+ved <- read_rds(here("data/temp", "vedrovice_dataset.RDS"))
+# non_rand_vars <- readLines(here("data/temp", "non_random_vars.txt"))
+var_weights <- read_csv(here("data/temp/v_statistics.csv"))
+
+normalize01 <- function(x) {
+  rg <- range(x, na.rm = TRUE)
+  (x - rg[1]) / (rg[2] - rg[1])
+}
+
+var_weights <- var_weights %>% mutate(v_norm = normalize01(v) + 1)
+
 # exceptionality index =========================================================
 # non random variables
-art_pca <- prcomp(ved$bin_vars$bin_mat[, non_rand_vars], 
+matr <- matrix(nrow = nrow(ved$bin_vars$bin_mat), ncol = ncol(ved$bin_vars$bin_mat))
+for (i in 1:ncol(ved$bin_vars$bin_mat)) {
+  matr[, i] <- ved$bin_vars$bin_mat[, i] * var_weights$v_norm[i]
+}
+colnames(matr) <- colnames(ved$bin_vars$bin_mat)
+
+art_pca <- prcomp(matr, 
+                  scale. = FALSE, center = FALSE)
+
+art_pca <- prcomp(ved$bin_vars$bin_mat[, ] * matrix(var_weights$v_norm, nrow = 1), 
                   scale. = FALSE, center = FALSE)
 summary(art_pca)
 ggbiplot::ggbiplot(art_pca)

### 2020-08-04 code/randomization.R

@@ -1,6 +1,6 @@
 # Project "Vedrovice"
-# Script nr. 2
-# ANALYSIS OF ARTEFACT CO-OCCURRENCES 
+# Script nr. 2.1
+# ANALYSIS OF ARTEFACT CO-OCCURRENCES RANDOMNESS
 # author: Petr Pajdla
 # Randomization test on the input binary matrix is performed in order to 
 # find structure in artefact coocurences
@@ -60,23 +60,25 @@ get_v <- function(df, expected, R) {
 # @var v_exp = expected values of statistic v (data frame)
 # @var v_obs = observed v values (data frame)
 # @var size = number of v values derived by randomization
-get_p <- function(v_exp, v_obs, size) {
-  res <- (sum(v_exp >= v_obs)/size)
-  return(res)
+get_p <- function(v_exp, v_obs) {
+  mean(v_exp >= v_obs)
 }
 
 # read data ====================================================================
 ved <- readRDS(here("data/temp", "vedrovice_dataset.RDS"))
 
 # counting co-occurrences in observed matrix ===================================
-cooc_obs <- count_cooccurrence(matrix = ved$bin_vars$bin_mat)
+input_matrix <- ved$bin_vars$bin_mat
+ncol_input <- ncol(input_matrix)
+
+cooc_obs <- count_cooccurrence(matrix = input_matrix)
 cooc_obs <- cooc_obs %>% filter(var1 != var2)
 # cooc_obs %>% tidyr::spread(var2, nr.cooc)
 
 # randomization of co-occurrence matrix - list of many matrices ----------------
-n_permutations <-  999
+n_permutations <-  9999
 # # ======
-# rand_mat <- vegan::permatfull(ved$bin_vars$bin_mat,
+# rand_mat <- vegan::permatfull(input_matrix,
 #                               fixedmar = "both",
 #                               mtype = "prab",
 #                               times = n_permutations)
@@ -109,7 +111,7 @@ cooc_exp <- cooc_rand %>% bind_rows() %>%
 # v values =====================================================================
 # observed v values ------------------------------------------------------------
 v_observed <- tidyr::spread(get_v(cooc_obs, cooc_exp, 
-                                  R = length(colnames(ved$bin_vars$bin_mat))),
+                                  R = ncol_input),
                             var1, v)
 
[…]

### 2020-08-04 code/spatial_buffer.R

@@ -1,0 +1,125 @@
+# Project "Vedrovice"
+# Script nr. 3.4
+# SPATIAL ARRANGEMENT: neighbors based on a buffer
+# author: Petr Pajdla
+# Prevalent sex at a certain distance (buffer)
+
+library(here)
+library(sf)
+library(tidyverse)
+
+# theme -------------------------------------------------------------------
+
+theme_universe <- theme(panel.border = element_rect(colour = "black", 
+                                                    fill = NA, 
+                                                    size = 0.8),
+                        panel.background = element_blank(),
+                        line = element_blank(),
+                        strip.background = element_blank(), 
+                        strip.text = element_text(face = "italic"),
+                        # axis.text.y = element_blank(), 
+                        # axis.title.y = element_blank(), 
+                        panel.spacing = unit(1, "lines"))
+
+# data --------------------------------------------------------------------
+
+ved_sf <- read_sf(here("data/temp", "layout.shp")) %>% 
+  mutate(sex = fct_relevel(sex, c("n. a.", "F", "M", "ind.")))
+ved_exc <- read_sf(here("data/temp", "window.shp"))
+
+# buffer zone -------------------------------------------------------------
+# 6 m (based on NN functions)
+distance <- 6
+
+mean_neighbors <- function(sf, variable, dist) {
+  variable <- as.symbol(variable)
+  lvls <- levels(factor(pull(sf, !!variable)))
+  res <- matrix(nrow = length(lvls), ncol = length(lvls))
+  colnames(res) <- lvls
+  rownames(res) <- lvls
+  # get buffer for each level of a variable
+  for (i in seq_along(lvls)) {
+    buffer <- sf %>% filter(!!variable == lvls[i]) %>% 
+      st_buffer(dist = dist) %>% 
+      st_geometry()
+    all_nbs <- sum(st_intersects(buffer, st_geometry(sf), sparse = FALSE)) - length(buffer)
+    for (j in seq_along(lvls)) {
+      points <- sf %>% filter(!!variable == lvls[j]) %>% 
+        st_geometry()
+      intersect <- sum(st_intersects(buffer, points, sparse = FALSE))
+      if (lvls[i] == lvls[j]) {
[…]

### 2020-08-04 code/spatial_kde.R

@@ -1,0 +1,97 @@
+# Project "Vedrovice"
+# Script nr. 3.2
+# SPATIAL ARRANGEMENT: Kernel density estimations
+# author: Petr Pajdla
+# KDE
+
+library(here)
+library(sf)
+# library(raster)
+library(tidyverse)
+
+# data --------------------------------------------------------------------
+
+ved_sf <- read_sf(here("data/temp/", "layout.shp")) %>% 
+  mutate(sex = fct_relevel(sex, c("n. a.", "F", "M", "ind.")))
+ved_exc <- read_sf(here("data/temp/", "window.shp"))
+
+# kernel density est. -----------------------------------------------------
+
+get_kde <- function(x, win = NA, variable = NA, value = NA) {
+  if (any(is.na(win))) {
+    win <- x
+  }
+  ncells <- nrow(x) * 2
+  if (!is.na(variable) & !is.na(value)) {
+    x <- filter(x, !!as.symbol(variable) == value)
+  }
+  coord <- st_coordinates(x)
+  bbox <- st_bbox(win)
+  bbox <- c(bbox[1], bbox[3], bbox[2], bbox[4])
+  kde <- MASS::kde2d(coord[, 1], coord[, 2], lims = bbox, n = ncells)
+  raster::raster(kde)
+}
+
+# plot(get_kde(ved_sf, ved_exc, "sex", "M"))
+# plot(get_kde(ved_sf, ved_exc, "sex", "F"))
+# plot(get_kde(ved_sf, ved_exc, "sex", "n. a."))
+# plot(get_kde(ved_sf, ved_exc, "sex", "ind."))
+# 
+# plot(cut(get_kde(ved_sf, ved_exc, "sex", "M"), breaks = 4))
+# 
+# contour(get_kde(ved_sf, ved_exc, "sex", "M"))
+# contour(get_kde(ved_sf, ved_exc, "sex", "F"))
+
+# raster to tibble for computations...
+# as_tibble(rasterToPoints(get_kde(ved_sf, ved_exc, "sex", "M")))
+
+# plot kde ----------------------------------------------------------------
+
+ved <- bind_cols(ved_sf, 
+               x = st_coordinates(ved_sf)[, 1],
+               y = st_coordinates(ved_sf)[, 2])
+
+ved_bg <- ved %>% transmute(s = sex, x, y)
+
+ved_fg <- ved %>% filter(sex != "ind.")
+
+ggplot(data = ved_fg) +
[…]

### 2020-08-04 code/spatial_neighbors.R

@@ -14,7 +14,6 @@ library(igraph)
 set.seed(42)
 
 # theme for ggplot graphics
-# theme for ggplot graphics
 theme_universe <- theme(panel.border = element_rect(colour = "black", 
                                                     fill = NA, 
                                                     size = 0.8),
@@ -30,7 +29,8 @@ theme_universe <- theme(panel.border = element_rect(colour = "black",
 ved <- read_rds(here("data/temp", "vedrovice_dataset.RDS"))
 
 # data prep ====================================================================
-ved_sf <- read_sf(here("data/temp", "layout.shp"))
+ved_sf <- read_sf(here("data/temp", "layout.shp")) %>% 
+  mutate(sex = fct_relevel(sex, c("n. a.", "F", "M", "ind.")))
 ved_layout <- st_coordinates(ved_sf)
 ved_exc <- read_sf(here("data/temp", "window.shp"))
 
@@ -43,10 +43,12 @@ ved_gabriel_lines <- ved_gabriel %>%
   st_as_sf()
 
 ggplot() +
-  geom_sf(data = ved_exc, fill = "gray90", color = NA) +
+  geom_sf(data = ved_exc, fill = NA, color = "gray90", size = 4) +
   geom_sf(data = ved_gabriel_lines, linetype = 3) +
-  geom_sf(data = ved_sf, shape = 21, fill = "white") +
-  theme_void() +
+  geom_sf(data = ved_sf, aes(shape = sex), fill = "white") +
+  scale_shape_manual(values = c(22, 21, 24, 4)) +
+  theme_void() + 
+  theme(legend.position = c(0.9, 0.8)) +
   ggspatial::annotation_north_arrow(style = ggspatial::north_arrow_minimal(),
                                     location = "br", 
                                     pad_y = unit(2, "cm")) +
@@ -54,7 +56,7 @@ ggplot() +
                               location = "br",
                               pad_y = unit(1, "cm"))
 
-ggsave(here("plots", "plan_vedrovice_gabriel.pdf"), scale = 2)
+ggsave(here("plots", "plan_gabriel.pdf"), scale = 2)
 
 # get sex of neighbors
 ved_g <- graph_from_edgelist(cbind(from = ved_gabriel$from, to = ved_gabriel$to), 
@@ -91,7 +93,6 @@ nb_sex <- neigh_sex(ved_g, ved_sex) %>%
[…]

### 2020-08-04 code/spatial_org.R

@@ -61,7 +61,8 @@ window_exc <- as.owin(ved_exc)
 
 # plot
 gg_baseplan <- ggplot(data = ved_sf) +
-  geom_sf(data = ved_exc, fill = "gray90", color = NA) +
+  geom_sf(data = ved_exc, fill = NA, color = "gray90", size = 4) +
+  # geom_sf(data = ved_exc, fill = "gray90", color = NA) +
   # geom_sf(data = ved_sf_hull, fill = NA, color = "gray40", linetype = 3) +
   geom_sf(aes(shape = sex), fill = "white") + 
   scale_shape_manual(values = c(22, 21, 24, 4)) +
@@ -76,15 +77,15 @@ gg_baseplan <- ggplot(data = ved_sf) +
 
 gg_baseplan
 
-gg_baseplan + ggsflabel::geom_sf_text_repel(aes(label = id_burial))
+gg_ids <- gg_baseplan + ggsflabel::geom_sf_text_repel(aes(label = id_burial))
 
-ggsave(here("plots", "plan_vedrovice.pdf"), gg_baseplan, scale = 2)
-ggsave(here("plots", "plan_vedrovice_ids.pdf"), scale = 2)
+ggsave(here("plots", "plan_base.pdf"), gg_baseplan, scale = 2)
+ggsave(here("plots", "plan_ids.pdf"), gg_ids, scale = 2)
 
 # write layouts -----------------------------------------------------------
 
-write_sf(ved_sf, here("data/temp/", "layout.shp"))
-write_sf(ved_exc, here("data/temp/", "window.shp"))
+write_sf(ved_sf, here("data/temp", "layout.shp"))
+write_sf(ved_exc, here("data/temp", "window.shp"))
 
 # marks ------------------------------------------------------------------------
 ved_marks <- ved$metadata %>% 
@@ -209,7 +210,7 @@ plot_estimate(ved_l, "L")
 
 # estimate T fun/stat
 ved_t <- envelope(ved_pp, Tstat, nrank = 2, nsim = 99) %>% 
-  fix_data_frame()
+  as_tibble()
 
 plot_estimate(ved_t, "T")
 
@@ -229,9 +230,11 @@ grid_fns <- bind_rows("G(r)" = ved_g,
   mutate(fun = as_factor(fun)) %>% 
   plot_estimate_facet()
 
-ggsave(plot = grid_fns, here("plots", "pointprocess_fun.pdf"), device = "pdf",
-       width = 12, height = 3)
+grid_fns
 
[…]

### 2020-08-11 code/dataset.R
add code on local vs non-local

@@ -23,8 +23,8 @@ input <- readr::read_csv(here("data", "data_vedrovice_v01.csv"), skip = 3) %>%
 # list to store the data
 ved <- list(bin_vars = list(count_mat = NA,
                             bin_mat = NA),
-            cont_vars = list(cont_vars = NA),
-            cat_vars = list(cat_vars = NA),
+            cont_vars = NA,
+            cat_vars = NA,
             metadata = NA,
             layout = NA,
             var_names = list(full = NA,
@@ -35,9 +35,9 @@ ved <- list(bin_vars = list(count_mat = NA,
 
 # filling the data -------------------------------------------------------------
 
-# burial id
-ved$id_burials <- input %>% 
-  filter(undisturbed == TRUE) %>% 
+# # burial id
+ved$id_burials <- input %>%
+  filter(undisturbed == TRUE) %>%
   pull(id_burial)
 
 # binary variables - artefact coocurences
@@ -57,29 +57,30 @@ rownames(ved$bin_vars$count_mat) <- ved$id_burials
 ved$bin_vars$bin_mat <- apply(ved$bin_vars$count_mat, 2, binarize)
 
 # continuous variables
-ved$cont_vars$cont_vars <-  input %>% 
+ved$cont_vars <- input %>% 
   filter(undisturbed == TRUE) %>% 
   select(pit_len, pit_wid, pit_dep,
          d13c, d15n, sr, sr_ppm, 
          body_height) %>% 
   as.data.frame()
 
-rownames(ved$cont_vars$cont_vars) <- ved$id_burials
+rownames(ved$cont_vars) <- ved$id_burials
 
 # factor variables
-ved$cat_vars$cat_vars <- input %>% 
+ved$cat_vars <- input %>% 
   filter(undisturbed == TRUE) %>% 
   select(pit_orient_cat, head_orient_cat, body_side) %>%
   mutate_all(factor) %>% 
   as.data.frame()
 
-rownames(ved$cat_vars$cat_vars) <- ved$id_burials
+rownames(ved$cat_vars) <- ved$id_burials
 
 # metadata
 lookup_sex <- c("n. a." = "gold", "F" = "tomato", "M" = "steelblue")
 
 ved$metadata <- input %>% select(id_burial, sex, age_cat, dat) %>% 
-  mutate(sex = factor(sex),
+  mutate(id_burial = as.character(id_burial),
+         sex = factor(sex),
          sex_col = unname(lookup_sex[input$sex]),
[…]

### 2020-08-11 code/local.R

@@ -1,0 +1,80 @@
+# Project "Vedrovice"
+# Script nr. 1.2
+# LOCAL AND NON-LOCAL Sr RANGES
+# author: Petr Pajdla
+# Assign bodies into local vs non-local Sr range
+
+library(tidyverse)
+library(here)
+
+
+# data --------------------------------------------------------------------
+
+ved <- read_rds(here("data/temp/vedrovice_dataset.RDS"))
+
+
+# edit age ----------------------------------------------------------------
+
+age <- ved$orig_dataset %>% 
+  select(id_burial, age_interval) %>% 
+  mutate(age = if_else(age_interval == "Adult", "20-40", age_interval),
+         age = if_else(age_interval == "±1", "0-2", age),
+         age = if_else(age_interval == "20+", "20-30", age),
+         age = if_else(age_interval == "50+", "50-60", age),
+         age = if_else(age_interval == "60+", "60-70", age),
+         age = if_else(age_interval == "9", "8-10", age),
+         age = if_else(age_interval == "30", "25-35", age),
+         age = if_else(age_interval == "20", "15-25", age),
+         age_min = as.numeric(str_extract(age, "^[:digit:]*(?=-|–)")),
+         age_max = as.numeric(str_extract(age, "(?<=-|–)[:digit:]*$")),
+         id_burial = as.character(id_burial)) %>% 
+  select(-age_interval, -age)
+
+
+# group by age ------------------------------------------------------------
+
+ved_sr <- ved$metadata %>% 
+  select(id_burial, sex, age_cat) %>% 
+  left_join(tibble(rownames_to_column(ved$cont_vars)), 
+            by = c("id_burial" = "rowname")) %>% 
+  left_join(age) %>% 
+  mutate(age_mean = (age_min + age_max) / 2,
+         group = if_else(age_mean < 18, "juvenile", as.character(sex))
+         ) %>% 
+  filter(!is.na(group), !group %in% c("n. a.", "ind."))
+
+local <- ved_sr %>% group_by(group) %>% 
+  summarise(n(), 
+            mean = mean(sr, na.rm = TRUE), 
+            sd = sd(sr, na.rm = TRUE)) %>% 
+  filter(group == "juvenile") %>%
+  mutate(min = mean - 2 * sd,
+         max = mean + 2 * sd)
+
+
+# local range of sr -------------------------------------------------------
+
+ved_local <- ved_sr %>% 
[…]

### 2020-08-11 code/spatial_buffer.R

@@ -1,6 +1,6 @@
 # Project "Vedrovice"
 # Script nr. 3.4
-# SPATIAL ARRANGEMENT: neighbors based on a buffer
+# SPATIAL NEIGHBORS: Buffer zone
 # author: Petr Pajdla
 # Prevalent sex at a certain distance (buffer)
 

### 2020-08-11 code/spatial_kde.R

@@ -48,8 +48,8 @@ get_kde <- function(x, win = NA, variable = NA, value = NA) {
 # plot kde ----------------------------------------------------------------
 
 ved <- bind_cols(ved_sf, 
-               x = st_coordinates(ved_sf)[, 1],
-               y = st_coordinates(ved_sf)[, 2])
+                 x = st_coordinates(ved_sf)[, 1],
+                 y = st_coordinates(ved_sf)[, 2])
 
 ved_bg <- ved %>% transmute(s = sex, x, y)
 
@@ -71,8 +71,9 @@ ggplot(data = ved_fg) +
   labs(shape = "sex") +
   facet_wrap(vars(sex)) +
   theme_void() +
-  theme(strip.text = element_text(face = "italic"),
-        legend.position = c(0.96, 0.8))
+  theme(
+    # strip.text = element_text(face = "italic"),
+    legend.position = c(0.96, 0.8))
 
 ggsave(here("plots", "plan_kde.pdf"), width = 16, height = 5)
 

### 2020-08-11 code/spatial_neighbors.R

@@ -1,8 +1,8 @@
 # Project "Vedrovice"
-# Script nr. 3.2
-# NEIGHBOURS
+# Script nr. 3.3
+# SPATIAL NEIGHBORS: Gabriel graph
 # author: Petr Pajdla
-# Mean number of neighbours with a given sex
+# Mean number of neighbours with a given sex based on Gabriel graph
 
 library(here)
 library(tidyverse)
@@ -13,7 +13,8 @@ library(igraph)
 
 set.seed(42)
 
-# theme for ggplot graphics
+# theme -------------------------------------------------------------------
+
 theme_universe <- theme(panel.border = element_rect(colour = "black", 
                                                     fill = NA, 
                                                     size = 0.8),
@@ -25,10 +26,10 @@ theme_universe <- theme(panel.border = element_rect(colour = "black",
                         # axis.title.y = element_blank(), 
                         panel.spacing = unit(1, "lines"))
 
-# read data
+# data --------------------------------------------------------------------
+
 # ved <- read_rds(here("data/temp", "vedrovice_dataset.RDS"))
 
-# data prep ====================================================================
 ved_sf <- read_sf(here("data/temp", "layout.shp")) %>% 
   mutate(sex = fct_relevel(sex, c("n. a.", "F", "M", "ind.")))
 ved_layout <- st_coordinates(ved_sf)

### 2020-08-11 code/spatial_org.R

@@ -24,6 +24,8 @@ theme_universe <- theme(panel.border = element_rect(colour = "black",
 
 # read data
 ved <- read_rds(here("data/temp", "vedrovice_dataset.RDS"))
+origin <- read_rds(here("data/temp/local.rds")) %>% 
+  select(id_burial, age_mean, origin)
 
 # data prep ====================================================================
 scale_m <- 4.762 # scale by a factor so 1 unit is 1 meter
@@ -37,6 +39,7 @@ ved_layout <- tibble(layout_x = ved_layout$x, layout_y = ved_layout$y)
 # not a rectangle, but a polygon closely bounding the burials using convex hull
 ved_sf <- st_as_sf(bind_cols(ved_layout, ved$metadata), 
                    coords = c("layout_x", "layout_y")) %>% 
+  left_join(origin) %>% 
   mutate(sex = fct_relevel(sex, c("n. a.", "F", "M", "ind.")))
 # ved_sf_buffer <- st_buffer(ved_sf, dist = 1.6)
 # ved_sf_buffer <- st_simplify(ved_sf_buffer, dTolerance = 1)
@@ -81,6 +84,26 @@ gg_ids <- gg_baseplan + ggsflabel::geom_sf_text_repel(aes(label = id_burial))
 
 ggsave(here("plots", "plan_base.pdf"), gg_baseplan, scale = 2)
 ggsave(here("plots", "plan_ids.pdf"), gg_ids, scale = 2)
+
+# plan local vs non-local
+ggplot(data = filter(ved_sf, !is.na(origin))) +
+  geom_sf(data = select(ved_sf, -origin), color = "gray90") +
+  geom_sf(data = ved_exc, fill = NA, color = "gray90", size = 4) +
+  stat_density2d(aes(st_coordinates(filter(ved_sf, !is.na(origin)))[, 1], 
+                     st_coordinates(filter(ved_sf, !is.na(origin)))[, 2]),
+                 color = "black", alpha = 0.4) +
+  geom_sf(shape = 21, fill = "white") + 
+  facet_wrap(~origin) +
+  theme_void() +
+  ggspatial::annotation_north_arrow(style = ggspatial::north_arrow_minimal(),
+                                    location = "br",
+                                    pad_y = unit(2, "cm")) +
+  ggspatial::annotation_scale(plot_unit = "m",
+                              location = "br",
+                              pad_y = unit(1, "cm")) +
+  theme(legend.position = c(0.9, 0.8))
+
[…]

### 2020-08-11 text/article_vedrovice.Rmd
text - introduction

@@
 
+The Cemetery of Vedrovice is one of the earliest and most extensively studied
+cemeteries of the *Linearbandkeramik* (LBK).
+Recent contributions use the evidence from the cemetery in a broader discussion
+of post-marital residence patterns [@hrncir2020] and
+sexual inequalities and gender in the Neolithic [@masclanslatorre2020; @bickle2020].
+Here we focus on detailed analysis of the distribution of the grave goods and
+spatial analysis of the Vedrovice cemetery in respect to gender relations.
+
+<!-- gender relations -->
+
+<!-- view of gender in the Neolithic, archaeological context -->
+@robb2017 [128] noted lack of gender archaeology debate for the
+European Neolithic in comparison with other periods, especially the Bronze Age.
+They argue that a the perception of gender in the Neolithic and the Bronze Age is
+not the same (ibid. 142).
+In response, @bickle2020 [14] shows that 'biological sex may not have defined gender
+during the LBK' although the 'sexed bodies did matter'.
+
+<!-- resampling and computer-intensive techniques-->
+The approach taken is a quantitative one embracing resampling techniques and
+computer-intensive methods in general.
+The resampling techniques were introduced to mortuary studies by
+@manly1996 but the application is still limited [@sosna2008, 343].
+As noted by Sosna (ibid.), the most important benefit of resampling is in its
+ability to distinguish random and non-random patterns.
+
 ### Questions
@@
 # Material and methods
+
+<!-- site -->
+
+<!-- sample (burials, artefacts) -->
 
@@
 
+<!-- methods -->
+
+<!-- spatial analysis -->
+
 Data is analyzed in an R environment for statistical computing and graphics
@@
 <!-- ADD METHODS AND PACKAGES USED IN R, or into SI?-->
+
+<!-- software packages -->
 
@@
 ```{r gabriel, fig.cap = "Gabriel graph"}
-knitr::include_graphics(here("plots", "plan_vedrovice_gabriel.pdf"))
+knitr::include_graphics(here("plots", "plan_gabriel.pdf"))
 ```
@@
 
+<!-- results -->
+
 # Results
@@
 
-```{r label-neigh_sex}
+```{r label-nb}
 label_neigh_sex <- paste("Mean neighbors in a given sex category",
@@
 ```
-```{r neigh-sex, fig.cap = label_neigh_sex}
-knitr::include_graphics(here("plots", "prob_neighbour_sex.pdf"))
+```{r nb-buffer, fig.cap = label_neigh_sex}
+knitr::include_graphics(here("plots", "nb_buffer_sex.pdf"))
+```
+```{r nb-gabriel, fig.cap = label_neigh_sex}
+knitr::include_graphics(here("plots", "nb_gabriel_sex.pdf"))
 ```

### 2020-08-17 code/knit_text.R
add knit for SI, count p-values

@@ -7,7 +7,13 @@
 # knitr::knit(input = here::here("text", "article-vedrovice.Rmd"),  
 #             output = here::here("text", "article-vedrovice.md"))
 
+# knit main text
 rmarkdown::render(input = here::here("text", "article_vedrovice.Rmd"), 
                   output_format = bookdown::pdf_document2())
 rmarkdown::render(input = here::here("text", "article_vedrovice.Rmd"), 
                   output_format = bookdown::word_document2())
+
+# knit si
+rmarkdown::render(input = here::here("text", "si_vedrovice.Rmd"), 
+                  output_format = bookdown::pdf_document2())
+

### 2020-08-17 code/spatial_buffer.R

@@ -99,7 +99,7 @@ expected %>% ggplot(aes(value)) +
   facet_grid(rows = vars(to), cols = vars(from), scales = "fixed") +
   scale_y_continuous(expand = c(0, 0, 0, 0)) +
   scale_x_continuous(expand = c(0, 0)) +
-  xlab("mean neighbors") +
+  labs(xlab = "mean neighbours", title = "Neighborhood based on buffer") +
   theme_universe
 
 ggsave(here("plots", "nb_buffer_sex.pdf"), width = 12, height = 6)
@@ -111,7 +111,7 @@ ggsave(here("plots", "nb_buffer_sex.pdf"), width = 12, height = 6)
 # ggplot(data = ved_sf) +
 #   geom_sf(data = ved_exc, fill = NA, color = "gray90", size = 4) +
 #   geom_sf(data = ved_buff, fill = "gray80", color = "gray80", alpha = 0.2) +
-#   geom_sf(aes(shape = sex), fill = "white", size = 2) + 
+#   geom_sf(aes(shape = sex), fill = "white", size = 2) +
 #   scale_shape_manual(values = c(22, 21, 24, 4)) +
 #   ggspatial::annotation_north_arrow(style = ggspatial::north_arrow_minimal(),
 #                                     location = "br",
@@ -123,3 +123,22 @@ ggsave(here("plots", "nb_buffer_sex.pdf"), width = 12, height = 6)
 #   facet_wrap(vars(sex)) +
 #   theme_void()
 
+
+# p-value -----------------------------------------------------------------
+pval <- function(obs, exp) {
+  lvls <- levels(obs$from)
+  res <- obs %>% mutate(p = 0)
+  for (i in seq_along(lvls)) {
+    for (j in seq_along(lvls)) {
+      obs_val <- obs %>% filter(from == lvls[i], to == lvls[j]) %>% 
+        pull(value)
+      exp_val <- exp %>% filter(from == lvls[i], to == lvls[j]) %>% 
+        pull(value)
+      res[(res[, "from"] == lvls[i]) & (res[, "to"] == lvls[j]), "p"] <- mean(exp_val >= obs_val)
+    }
+  }
+  res <- res %>% mutate(signif = if_else(p <= 0.05, TRUE, FALSE))
+  return(res)
+}
+
+pval(observed, expected) %>% write_csv(here("data/temp/nb_p_buffer.csv"))

### 2020-08-17 code/spatial_neighbors.R

@@ -35,39 +35,75 @@ ved_sf <- read_sf(here("data/temp", "layout.shp")) %>%
 ved_layout <- st_coordinates(ved_sf)
 ved_exc <- read_sf(here("data/temp", "window.shp"))
 
-# check neighbours =============================================================
-ved_gabriel <- gabrielneigh(ved_layout)
 
+# create delaunay + gabriel neighborhoods ---------------------------------
+ved_gabriel <- gabrielneigh(ved_layout)
+ved_delaunay <- tri2nb(ved_layout)
+
+# plot delaunay triangulation and gabriel graph ---------------------------
 ved_gabriel_lines <- ved_gabriel %>% 
   graph2nb(row.names = rownames(ved_sf)) %>% 
   nb2lines(coords = ved_layout) %>% 
   st_as_sf()
 
-ggplot() +
+ved_delaunay_lines <- ved_delaunay %>% 
+  nb2lines(coords = ved_layout) %>% 
+  st_as_sf()
+
+g_gabriel <- ggplot() +
   geom_sf(data = ved_exc, fill = NA, color = "gray90", size = 4) +
-  geom_sf(data = ved_gabriel_lines, linetype = 3) +
+  # geom_sf(data = ved_delaunay_lines, linetype = 1, size = 0.2) +
+  geom_sf(data = ved_gabriel_lines, linetype = 1, size = 0.2, color = "gray40") +
   geom_sf(data = ved_sf, aes(shape = sex), fill = "white") +
   scale_shape_manual(values = c(22, 21, 24, 4)) +
   theme_void() + 
-  theme(legend.position = c(0.9, 0.8)) +
+  theme(legend.position = c(0.9, 0.8),
+        plot.title = element_text(hjust = 0.5)) +
   ggspatial::annotation_north_arrow(style = ggspatial::north_arrow_minimal(),
                                     location = "br", 
                                     pad_y = unit(2, "cm")) +
   ggspatial::annotation_scale(plot_unit = "m", 
                               location = "br",
-                              pad_y = unit(1, "cm"))
+                              pad_y = unit(1, "cm")) +
+  labs(title = "Gabriel graph")
 
-ggsave(here("plots", "plan_gabriel.pdf"), scale = 2)
+g_delaunay <- ggplot() +
+  geom_sf(data = ved_exc, fill = NA, color = "gray90", size = 4) +
[…]

### 2020-08-17 text/article_vedrovice.Rmd
si + text - continue spatial analysis

@@
 
-The Cemetery of Vedrovice is one of the earliest and most extensively studied
-cemeteries of the *Linearbandkeramik* (LBK).
+The Cemetery of Vedrovice is one of the earliest and most intensively studied
+*Linearbandkeramik* (LBK) cemeteries.
 Recent contributions use the evidence from the cemetery in a broader discussion
@@
 sexual inequalities and gender in the Neolithic [@masclanslatorre2020; @bickle2020].
-Here we focus on detailed analysis of the distribution of the grave goods and
-spatial analysis of the Vedrovice cemetery in respect to gender relations.
-
-<!-- gender relations -->
+<!-- Here we focus on detailed analysis of the distribution of the grave goods and
+spatial analysis of the Vedrovice cemetery in respect to gender relations and
+social structure. -->
+Here we explore the relations between burials and bodies in terms of biologicall
+sex, age, localness, exceptionality of grave goods and position within the cemetery.
 
@@
 @robb2017 [128] noted lack of gender archaeology debate for the
-European Neolithic in comparison with other periods, especially the Bronze Age.
-They argue that a the perception of gender in the Neolithic and the Bronze Age is
-not the same (ibid. 142).
-In response, @bickle2020 [14] shows that 'biological sex may not have defined gender
+European Neolithic in comparison with other periods.
+This is due to the notion of gender in the Neolihic that is presented as being
+of a very different nature than gender in the later prehistory.
+Authors show that gender in the Neolithic is not clearly expressed and
+biological differences between sexes are not transformed into binary cultural
+categories (ibid. 140-142).
+Considering the Neolithic burial evidence across Europe, authors find that it
+'shows very little gender distinction; when it does, it is rarely a clear-cut
+binary categorization but a matter of degree and overlap, sometimes polythetically
+combined with factors such as age and localness' (ibid. 138).
+In response, @bickle2020 [14] argues that 'biological sex may not have defined gender
 during the LBK' although the 'sexed bodies did matter'.
+
+<!-- gender relations in LBK -->
+Although gender relations in the LBK might not be
 
@@
 <!-- site -->
+The material analyzed here comes from Vedrovice – Široká u lesa cemetery.
 
 <!-- sample (burials, artefacts) -->
-
-The material analyzed here comes from Vedrovice – Široká u lesa cemetery.
 Originally, the burial ground consisted of at least 108 burials.
@@
 agricultural activities.
-Thus some of the burials are excluded from the analysis due to missing and/or
+These burials are excluded from the analysis due to missing and/or
 incomplete data and only a sample of 78 well-preserved burials is analyzed here.
@@
 ratios were collected in @whittle2013 and @richards2008.
+Further details on use and function of the artefacts were gathered from
+@masclamasclanslatorre2020.
 
-Each burial was originally characterized by 55 variables but only 35 are
-selected for further analysis.
+Each burial was originally characterized by 55 variables but only subset of
+35 variables was selected for further analysis.
 Some of  the original variables are omitted due to their rare occurrence,
@@
 retrieved human skeletal remains including sex, age simplified into age groups,
-isotope ratios or body height.
+isotope ratios and body height.
 
-Second large category of variables consists of data on artifacts present in the
-burials.
+Second large category of variables consists of data on artifacts presence and
+absence in the burials.
 These include variables noting quantities of various types of pottery and
@@
 <!-- spatial analysis -->
-
-Data is analyzed in an R environment for statistical computing and graphics
-[@rcore]  using functions of readr [@wickham2017a], dplyr [@wickham2016] and
-reshape2 [@wickham2007] packages to manipulate the data.
-Function daisy [@kaufman1990] from package cluster [@maechler2016] is used to
-count distances between observations used in further analyses.
-<!-- ADD METHODS AND PACKAGES USED IN R, or into SI?-->
-
-<!-- software packages -->
-
 ## Spatial patterning
@@
 a burial (event) is simplified as its centroid location in a two dimensional space.
-A polygon of the excavation extent is used as a study area (a window).
+A polygon of idealized excavation extent is used as a study area (a window).
 Using a window closely approximating locations of the burials was considered,
 e. g. buffer zone or convex/concave hull but simplified excavation polygon is
-deemed the most useful because it preserves the *natural* borders and empty areas
+deemed the most useful because it preserves the borders and empty areas
 of the original cemetery.
@@
 Overall structure of the burial ground is examined using both density based approach
-and distance based methods <!--[@osullivan2010, 121-154]-->.
-It is argued that density based approaches, for instance quadrat tests,
+and distance based methods. <!--[@osullivan2010, 121-154]-->
+Density based approaches, for instance quadrat tests,
 characterize the point pattern in terms of its *first-order properties* [@osullivan2010, 126].
@@
 our point of view.
-The *second-order properties*, and possibly *third-order properties* [@nakoinz2016, 130, 144]
+The *second-order properties*, and possibly *third-order properties*
 are thus considered major factors in locating of the graves.
@@
 subject to changes in the size of the study area.
-The $F$ function, also empty-space function, is based on simulating a point pattern
+The $F$ function, an empty-space function, is based on simulating a point pattern
 and calculating of distances between random locations and the observed locations.
@@
 Last but not least, the $T$ function is a third-order parallel to the $K$ function [@schladitz2000].
-
-If $s_i$ is an event location, then
-$G$ function is given by the formula
-\@ref(eq:gfun)
-
-\begin{equation}
-  \label{eq:gfun}
-  G(d) = \frac{\#[d_{min}(s_i)<d]}{n}
-\end{equation}
-
-$F$ function by
-$$F(d) = \frac{\#[d_{min}(p_i, S)<d]}{m}$$
-where ${p_1 \cdots p_i \cdots p_m}$ is a set of $m$ randomly selected locations.
-Ripley's $K$ function
-$$K(d) = \frac{\alpha}{n}\frac{1}{n}\sum_{i=1}^n \#[S \in C(s_i, d)]$$
-$T$ function
-$$ $$
+For formulas of the individual functions see the supplement.
 
@@
 
-To explore whether neighboring individuals are more often of the same sex we
-construct a Gabriel graph [@gabriel1969] with burial locations as nodes (Fig. \@ref(fig:gabriel)).
+To explore whether neighboring individuals are more often of the same sex we use
+two methods to establish neighborhood.
+The first method is based on constructing a Gabriel graph [@gabriel1969]
+with burial locations as nodes (Fig. \@ref(fig:gabriel)).
+Gabriel graph is a subgraph of a Delaunay triangulation
+
 For each node, mean sex of the neighbors is determined and recorded.
-To check how common the observed value is, we perform a permutation test
+To check how common the observed value is, a permutation test
 where sex of burials is randomly assigned while preserving the overall
 proportion of given categories and the neighborhood struture determined by the
-Gabriel graph.
+Gabriel graph is performed.
 The observed values of neighbors sex categories are compared against distributions
@@
 ```
+
+The second approach of establishing neigborhood is by constructing buffer zones
+around each of the burials with 6m radius.
+The size of a radius is based on the results of nearest neighbors analysis, where
+at the given radius there is strong evidence of a clustering process (Fig. \ref(fig:spat-funs)).
+The randomness is again tested using a permutation test similar to the one
+described for the neighborhood derived from the Gabriel graph.
+
+<!-- software packages -->
+Data is analyzed in an R environment for statistical computing and graphics
+[@rcore]  using functions of many different packages, for the complete list
+see the supplementary information.
+<!-- readr [@wickham2017a], dplyr [@wickham2016] and
+reshape2 [@wickham2007] packages to manipulate the data.
+Function daisy [@kaufman1990] from package cluster [@maechler2016] is used to
+count distances between observations used in further analyses. -->
 

### 2020-08-17 text/si_vedrovice.Rmd
si + text - continue spatial analysis

@@
 ---
+
+```{r}
+library(here)
+library(tidyverse)
+```
+
+# Spatial analysis
+
+<!--
+If $s_i$ is an event location, then
+$G$ function is given by the formula
+\@ref(eq:gfun)
+
+\begin{equation}
+\label{eq:gfun}
+G(d) = \frac{\#[d_{min}(s_i)<d]}{n}
+\end{equation}
+
+$F$ function by
+$$F(d) = \frac{\#[d_{min}(p_i, S)<d]}{m}$$
+where ${p_1 \cdots p_i \cdots p_m}$ is a set of $m$ randomly selected locations.
+Ripley's $K$ function
+$$K(d) = \frac{\alpha}{n}\frac{1}{n}\sum_{i=1}^n \#[S \in C(s_i, d)]$$
+$T$ function
+$$ $$
+-->
+
+$s_i$ is an event location.
+
+## G function
+
+$$ G(d) = \frac{\sum[d_{min}(s_i)<d]}{n} $$
+
+## F function
+
+$$F(d) = \frac{\#[d_{min}(p_i, S)<d]}{m}$$
+
+where ${p_1 \cdots p_i \cdots p_m}$ is a set of $m$ randomly selected locations.
+
+## Ripley's K function
+
+$$K(d) = \frac{\alpha}{n}\frac{1}{n}\sum_{i=1}^n \#[S \in C(s_i, d)]$$
 
@@
 ## Implementation
+
+# Results
+
+```{r neighborhood-sex, echo = FALSE}
+knitr::include_graphics(here("plots", "nb_buffer_sex.pdf"))
+knitr::include_graphics(here("plots", "nb_delaunay_sex.pdf"))
+knitr::include_graphics(here("plots", "nb_gabriel_sex.pdf"))
+```
+
+```{r, echo = FALSE, message = FALSE, results = 'asis'}
+b <- read_csv(here("data/temp/nb_p_buffer.csv"))
+d <- read_csv(here("data/temp/nb_p_delaunay.csv"))
+g <- read_csv(here("data/temp/nb_p_gabriel.csv"))
+
+tab <- full_join(g, d, by = c("from", "to"), suffix = c(".gabriel", ".delaunay")) %>% 
+  full_join(b, by = c("from", "to")) %>% 
+  mutate(across(where(is.numeric), round, 2)) %>% 
+  transmute(relation = str_c(from, to, sep = " - "),
+            gabriel = mean.gabriel,
+            p.gabriel = if_else(signif.gabriel, str_c(p.gabriel, "*"), as.character(p.gabriel)),
+            delaunay = mean.delaunay,
+            p.delaunay = if_else(signif.delaunay, str_c(p.delaunay, "*"), as.character(p.delaunay)),
+            buffer = value,
+            p.buffer = if_else(signif, str_c(p, "*"), as.character(p)))
+
+knitr::kable(tab)
+```

### 2020-08-18 code/spatial_buffer.R
join code for neighbors based on networks and buffer

@@ -1,144 +1,0 @@
-# Project "Vedrovice"
-# Script nr. 3.4
-# SPATIAL NEIGHBORS: Buffer zone
-# author: Petr Pajdla
-# Prevalent sex at a certain distance (buffer)
-
-library(here)
-library(sf)
-library(tidyverse)
-
-# theme -------------------------------------------------------------------
-
-theme_universe <- theme(panel.border = element_rect(colour = "black", 
-                                                    fill = NA, 
-                                                    size = 0.8),
-                        panel.background = element_blank(),
-                        line = element_blank(),
-                        strip.background = element_blank(), 
-                        strip.text = element_text(face = "italic"),
-                        # axis.text.y = element_blank(), 
-                        # axis.title.y = element_blank(), 
-                        panel.spacing = unit(1, "lines"))
-
-# data --------------------------------------------------------------------
-
-ved_sf <- read_sf(here("data/temp", "layout.shp")) %>% 
-  mutate(sex = fct_relevel(sex, c("n. a.", "F", "M", "ind.")))
-ved_exc <- read_sf(here("data/temp", "window.shp"))
-
-# buffer zone -------------------------------------------------------------
-# 6 m (based on NN functions)
-distance <- 6
-
-mean_neighbors <- function(sf, variable, dist) {
-  variable <- as.symbol(variable)
-  lvls <- levels(factor(pull(sf, !!variable)))
-  res <- matrix(nrow = length(lvls), ncol = length(lvls))
-  colnames(res) <- lvls
-  rownames(res) <- lvls
-  # get buffer for each level of a variable
-  for (i in seq_along(lvls)) {
-    buffer <- sf %>% filter(!!variable == lvls[i]) %>% 
-      st_buffer(dist = dist) %>% 
-      st_geometry()
-    all_nbs <- sum(st_intersects(buffer, st_geometry(sf), sparse = FALSE)) - length(buffer)
-    for (j in seq_along(lvls)) {
-      points <- sf %>% filter(!!variable == lvls[j]) %>% 
-        st_geometry()
-      intersect <- sum(st_intersects(buffer, points, sparse = FALSE))
-      if (lvls[i] == lvls[j]) {
[…]

### 2020-08-18 code/spatial_neighbors.R

@@ -104,7 +104,7 @@ ved_sex <- ved_sf$sex
 names(ved_sex) <- ved_sf$id_burial
 
 # count mean neighbors ----------------------------------------------------
-mean_neighbors <- function(g, variable_vector) {
+mean_neighbors_networks <- function(g, variable_vector) {
   res <- vector("list", length(V(g)))
   names(res) <- V(g)$name
   for (i in seq_along(V(g))) {
@@ -124,12 +124,12 @@ mean_neighbors <- function(g, variable_vector) {
     summarise(mean = mean(n), sum = sum(n), .groups = "drop")
 }
 
-nb_sex_g <- mean_neighbors(ved_g, ved_sex)
-nb_sex_d <- mean_neighbors(ved_d, ved_sex)
+nb_sex_g <- mean_neighbors_networks(ved_g, ved_sex)
+nb_sex_d <- mean_neighbors_networks(ved_d, ved_sex)
 
 # randomization of sex for neighbors --------------------------------------
 # sf - the attribute table - first column is ID
-randomize_neighbors <- function(g, sf, variable, n_sim = 99) {
+randomize_neighbors_network <- function(g, sf, variable, n_sim = 99) {
   n_bur <- nrow(sf)
   variable <- as.symbol(variable)
   
@@ -149,7 +149,7 @@ randomize_neighbors <- function(g, sf, variable, n_sim = 99) {
                                       conditions = pull(prob, !!variable)) %>% 
       as.character()
     names(rand_vect) <- sf %>% st_drop_geometry() %>% pull(1)
-    res[[i]] <- mean_neighbors(g, rand_vect)
+    res[[i]] <- mean_neighbors_networks(g, rand_vect)
   }
   res %>% bind_rows %>% 
     mutate(across(c(from, to), fct_relevel, lvls))
@@ -157,8 +157,142 @@ randomize_neighbors <- function(g, sf, variable, n_sim = 99) {
 
 # simulation (200 iterations) ----------------------------------------------
 
-ved_rand_sex_g <- randomize_neighbors(ved_g, ved_sf, "sex", n_sim = 200)
-ved_rand_sex_d <- randomize_neighbors(ved_d, ved_sf, "sex", n_sim = 200)
+ved_rand_sex_g <- randomize_neighbors_network(ved_g, ved_sf, "sex", n_sim = 200)
+ved_rand_sex_d <- randomize_neighbors_network(ved_d, ved_sf, "sex", n_sim = 200)
+
[…]

### 2020-08-18 text/article_vedrovice.Rmd
improve figures and tables

@@
 
-  As a result of the investigation we conclude that...
-  "
+  As a result of the investigation we conclude that...  "
 
@@
 Further details on use and function of the artefacts were gathered from
-@masclamasclanslatorre2020.
+@masclanslatorre2020.
 
@@
 To explore whether neighboring individuals are more often of the same sex we use
-two methods to establish neighborhood.
-The first method is based on constructing a Gabriel graph [@gabriel1969]
-with burial locations as nodes (Fig. \@ref(fig:gabriel)).
-Gabriel graph is a subgraph of a Delaunay triangulation
-
-For each node, mean sex of the neighbors is determined and recorded.
+three methods to establish neighborhood relations.
+The first two methods are based on constructing neighborhood graphs.
+We use Delaunay triangulation and its subgraph, a Gabriel graph [@gabriel1969]
+with burial locations as nodes (Fig. \@ref(fig:plan-graphs)).
+The Delaunay triangulation can be considered a maximum number of neighbors case
+scenario while the Gabriel graph the exact opposite.
+For each node, mean number of neighbors of a given sex is determined and recorded.
 To check how common the observed value is, a permutation test
 where sex of burials is randomly assigned while preserving the overall
-proportion of given categories and the neighborhood struture determined by the
-Gabriel graph is performed.
+proportion of given categories and the neighborhood structure determined by the
+graphs is performed.
 The observed values of neighbors sex categories are compared against distributions
@@
 
-```{r gabriel, fig.cap = "Gabriel graph"}
-knitr::include_graphics(here("plots", "plan_gabriel.pdf"))
-```
-
-The second approach of establishing neigborhood is by constructing buffer zones
-around each of the burials with 6m radius.
+The third approach of establishing neigborhood is by constructing buffer zones
+around each of the burials with 6 meters radius.
 The size of a radius is based on the results of nearest neighbors analysis, where
-at the given radius there is strong evidence of a clustering process (Fig. \ref(fig:spat-funs)).
+at the given radius there is strong evidence of a clustering process (Fig. \@ref(fig:spat-funs)).
 The randomness is again tested using a permutation test similar to the one
-described for the neighborhood derived from the Gabriel graph.
+described for the neighborhood derived from the graphs.
 
@@
 see the supplementary information.
+
 <!-- readr [@wickham2017a], dplyr [@wickham2016] and
@@
 
-```{r label-spat-funs}
-label_spat_funs <- paste("Functions *G*, *F*, *K* and *T*.",
-"*Dashed line* -- theoretical curve under complete spatial randomness, *solid line* -- empirical curve, *gray area* -- envelope.")
-```
-```{r spat-funs, fig.cap = label_spat_funs}
-knitr::include_graphics(here("plots", "pointprocess_fun.pdf"))
-```
+## Neighborhood structure
 
-```{r label-nb}
-label_neigh_sex <- paste("Mean neighbors in a given sex category",
-"*Gray area* -- density estimation for the simulated data, *vertical line* -- observed value.",
-"*F* -- female, *M* -- male, *n. a.* -- non adult.")
-```
-```{r nb-buffer, fig.cap = label_neigh_sex}
-knitr::include_graphics(here("plots", "nb_buffer_sex.pdf"))
-```
-```{r nb-gabriel, fig.cap = label_neigh_sex}
-knitr::include_graphics(here("plots", "nb_gabriel_sex.pdf"))
-```
+Three methods of neighborhood relationship construction were examined.
+One is distance based where all the neighbors that fall into a given buffer are considered,
+the other two are based on neighborhood graphs, the Delaunay triangulation as a
+maximum number of neighbors scenario and a Gabriel graph as a minimal scenario.
+
+For relationships (see Fig. \@ref(fig:nb-sex) and Tab. \@ref(tab:nb-tab-sex)) of male-sexed bodies, none of the methods showed any significant
+differences between expected and observed number of neighbors of a given sex.
+The relationship with female-sexed bodies is rather extreme for all of the methods employed,
+but never significant ($p > 0.07$).
+In case of bodies unsexed due to early age (non-adults), the neigborhood based
+on buffer zones show a signficant realtionship with other non-adults ($p \approx 0.04$).
+Finally burials with female-sexed bodies show a significantly higher number of
+burials with male-sexed bodies in both the methods based on neighborhood networks
+(Delaunay triangulation $p \approx 0.02$ and Gabriel graph $p \approx 0.03$).
+Two further relationships showed non-random deviations.
+In case of Delaunay trinagulation for burials with bodies with indeterminate sex with
+non-adults ($p \approx 0.02$) and in case of Gabriel graph between non-adults and
+indeterminate category $p \approx 0.02$.
 
@@
 
+<!-- ======================================================================= -->
+
+# Discussion
+
+<!-- ======================================================================= -->
+<!-- FIGURES --><!-- FIGURES --><!-- FIGURES --><!-- FIGURES --><!-- FIGURES -->
+```{r plan-graphs, fig.cap = "Neighborhood graphs: Delaunay triangulation and Gabriel graph"}
+knitr::include_graphics(here("plots/plan_gabriel.pdf"))
+```
+
+```{r label-spat-funs}
+label_spat_funs <- paste("Functions *G*, *F*, *K* and *T*.",
+"*Dashed line* -- theoretical curve under complete spatial randomness, *solid line* -- empirical curve, *gray area* -- envelope.")
+```
+```{r spat-funs, fig.cap=label_spat_funs}
+knitr::include_graphics(here("plots/pointprocess_fun.pdf"))
+```
+
+```{r label-nb-sex}
+label_nb_sex <- paste("Mean neighbors in a given sex category.",
+"*Gray area* -- density estimation for the simulated data, *vertical line* -- observed value.",
+"*F* -- female-sexed body, *M* -- male-sexed body, *n. a.* -- non adult, *ind.* -- indetermined sex.")
+```
+```{r nb-sex, fig.cap=label_nb_sex}
+knitr::include_graphics(here("plots/nb_sex.pdf"))
+```
+
 ```{r f-sstat, fig.cap = "Distribution of simulated and observed S statistic"}
-knitr::include_graphics(here("plots", "s_statistic.pdf"))
+knitr::include_graphics(here("plots/s_statistic.pdf"))
 ```
@@
 <!-- ======================================================================= -->
+<!-- TABLES --><!-- TABLES --><!-- TABLES --><!-- TABLES --><!-- TABLES -->
+```{r nb-tab-sex}
+tab <- read_csv(here("data/temp/nb_sex_p.csv")) %>%
+         mutate(relation = str_c(from, to, sep = " - ")) %>%
+  select(relation, everything(), -from, -to)
 
-# Discussion
+tnames <- c("relation", rep(c("mean", "p"), 3))
+
+knitr::kable(tab, booktabs = TRUE, col.names = tnames, linesep = "", caption = "Results...") %>%
+  kableExtra::add_header_above(c(" ", "Buffer" = 2, "Delaunay" = 2, "Gabriel" = 2))
+```
 

### 2020-08-18 text/si_vedrovice.Rmd
improve figures and tables

--- 2020-05-29 text/si_vedrovice.Rmd
@@
 ---
+
+```{r}
+library(here)
+library(tidyverse)
+library(knitr)
+library(kableExtra)
+```
+
+# Spatial analysis
+
+<!--
+If $s_i$ is an event location, then
+$G$ function is given by the formula
+\@ref(eq:gfun)
+
+\begin{equation}
+\label{eq:gfun}
+G(d) = \frac{\#[d_{min}(s_i)<d]}{n}
+\end{equation}
+
+$F$ function by
+$$F(d) = \frac{\#[d_{min}(p_i, S)<d]}{m}$$
+where ${p_1 \cdots p_i \cdots p_m}$ is a set of $m$ randomly selected locations.
+Ripley's $K$ function
+$$K(d) = \frac{\alpha}{n}\frac{1}{n}\sum_{i=1}^n \#[S \in C(s_i, d)]$$
+$T$ function
+$$ $$
+-->
+
+$s_i$ is an event location.
+
+## G function
+
+$$ G(d) = \frac{\sum[d_{min}(s_i)<d]}{n} $$
+
+## F function
+
+$$F(d) = \frac{\#[d_{min}(p_i, S)<d]}{m}$$
+
+where ${p_1 \cdots p_i \cdots p_m}$ is a set of $m$ randomly selected locations.
+
+## Ripley's K function
+
+$$K(d) = \frac{\alpha}{n}\frac{1}{n}\sum_{i=1}^n \#[S \in C(s_i, d)]$$
 
@@
 ## Implementation
+
+# Results
+
+```{r, echo = FALSE, message = FALSE, results = 'asis'}
+tab <- read_csv(here("data/temp/nb_sex_p.csv")) %>% 
+         mutate(relation = str_c(from, to, sep = " - ")) %>% 
+  select(relation, everything(), -from, -to)
+
+tnames <- c("relation", rep(c("mean", "p"), 3))
+
+kable(tab, booktabs = TRUE, col.names = tnames) %>% 
+  add_header_above(c(" ", "Buffer" = 2, "Delaunay" = 2, "Gabriel" = 2))
+```

### 2020-08-20 code/spatial_functions.R
separate spatial functions from plans

@@ -1,0 +1,218 @@
+# Project "Vedrovice"
+# Script nr. 3.2
+# SPATIAL ARRANGEMENT
+# author: Petr Pajdla
+# Spatial organization within the cemetery
+
+library(here)
+library(spatstat)
+library(sf)
+library(tidyverse)
+
+# number of simulations ---------------------------------------------------
+nsim = 999
+nrank = 10
+
+# k...nrank
+# M...nsim
+# probability 2 * k / M + 1
+# thus for 999 iterations 2 * 25 / 999 + 1 = 0.05 (5% probability)
+
+# theme for ggplot graphics
+theme_universe <- theme(panel.border = element_rect(colour = "black", 
+                                                    fill = NA, 
+                                                    size = 0.8),
+                        panel.background = element_blank(),
+                        line = element_blank(),
+                        strip.background = element_blank(), 
+                        strip.text = element_text(face = "italic"),
+                        axis.text.y = element_blank(), 
+                        axis.title.y = element_blank(), 
+                        panel.spacing = unit(1, "lines"))
+
+# read data and layouts --------------------------------------------------------
+ved <- read_rds(here("data/temp", "vedrovice_dataset.RDS"))
+
+ved_sf <- read_sf(here("data/temp", "layout.shp"))
+ved_exc <- read_sf(here("data/temp", "window.shp")) %>% 
+  as.owin(ved_exc)
+
+ved_sf <- bind_cols(ved_sf, 
+                    x = st_coordinates(ved_sf)[, 1], 
+                    y = st_coordinates(ved_sf)[, 2])
+
+# marks ------------------------------------------------------------------------
+ved_marks <- ved$metadata %>% 
+  select(sex) %>% 
+  mutate(sex = if_else(is.na(sex), "ind.", as.character(sex)),
+         sex = as_factor(sex))
+
+# create ppp object ------------------------------------------------------------
+# ved_pp <- ppp(x = ved_layout$layout_x, y = ved_layout$layout_y,
+#               window = window_poly,
+#               marks = ved_marks)
+ved_pp <- ppp(x = ved_sf$x, y = ved_sf$y, 
[…]

### 2020-08-20 code/spatial_org.R

@@ -5,22 +5,9 @@
 # Spatial organization within the cemetery
 
 library(here)
-library(spatstat)
+# library(spatstat)
 library(sf)
 library(tidyverse)
-library(broom)
-
-# theme for ggplot graphics
-theme_universe <- theme(panel.border = element_rect(colour = "black", 
-                                                    fill = NA, 
-                                                    size = 0.8),
-                        panel.background = element_blank(),
-                        line = element_blank(),
-                        strip.background = element_blank(), 
-                        strip.text = element_text(face = "italic"),
-                        axis.text.y = element_blank(), 
-                        axis.title.y = element_blank(), 
-                        panel.spacing = unit(1, "lines"))
 
 # read data
 ved <- read_rds(here("data/temp", "vedrovice_dataset.RDS"))
@@ -110,175 +97,3 @@ ggsave(here("plots", "plan_origin.pdf"), width = 10.5, height = 5)
 write_sf(ved_sf, here("data/temp", "layout.shp"))
 write_sf(ved_exc, here("data/temp", "window.shp"))
 
-# marks ------------------------------------------------------------------------
-ved_marks <- ved$metadata %>% 
-  select(sex) %>% 
-  mutate(sex = if_else(is.na(sex), "ind.", as.character(sex)),
-         sex = as_factor(sex))
-
-# create ppp object ------------------------------------------------------------
-# ved_pp <- ppp(x = ved_layout$layout_x, y = ved_layout$layout_y,
-#               window = window_poly,
-#               marks = ved_marks)
-ved_pp <- ppp(x = ved_layout$layout_x, y = ved_layout$layout_y, 
-              window = window_exc,
-              marks = ved_marks)
-
-plot(ved_pp, use.marks = FALSE)
-
-# point pattern analysis
-# functions ====================================================================
-
-# function L
-# scaled centered Ripley's K, based on O'Sullivan & Unwin 2010, p. 147
-# @x = ppp object
-estimate_L <- function(x) {
-  stopifnot(verifyclass(x, "ppp"))
-  l <- envelope(x, Kest, nrank = 2, nsim = 99) %>%
[…]

### 2020-08-20 text/article_vedrovice.Rmd
cont. art. text - spatial patterning

@@
 Given statistics for independent random process are estimated using Monte Carlo
-simulation with 99 iterations.
-The most extreme results of simulations are discarded as outliers and the
-second most extreme results are used to construct envelopes encompassing the
-expected range for a given function [@baddeley2016 384-403].
+simulation with 999 iterations.
+Ten most extreme realizations of the simulation are discarded as outliers to
+get a significance level $\alpha = 0.01$ [@baddeley2016 384-403].
 
@@
 
-The quadrat test suggests both random and clustered processes with p values
+The quadrat test shows evidence for both random and clustered processes taking place with p values
 under the 0.05 alpha value but the $\chi^2$ approximation is inaccurate due to small
 expected counts (for the details see supplementary information).
-Fig. \@ref(fig:spat-funs) shows that the curves of $G$ and $F$ functions do not
-deviate too much from the curves expected under independent random process nor
-do they traverse the boundaries of the envelopes.
-This tells us that both the cummulative nearest-neighbor distance (*G*) and
-distance to any random location in the space of the window (*F*) are well
-approximated by random processes and do not show evidence for spatial clustering.
-On the other hand, both *K* and *T* curves have a steep rise and exceed the
-envelope boundaries.
-These functions thus show evidence for clustering process in the observed point
-pattern.
+Inconclusive result of the quadrat test is not suprising because as was noted
+earlier, we do not expect *first-order* properties to play major roles in
+placing of the individual burials.
+Results of functions based on nearest neighbors are reported in Fig. \@ref(fig:spat-funs).
+*G* function does deviate from the curve expected under independent random process
+but does not traverse the boundaries of the envelope.
+Thus the cummulative nearest-neighbor distance (*G*) is well approximated by
+random process and does not show evidence for spatial clustering nor repelent functions
+taking place.
+The empty-space function *F* deviates slightly from the area defined by
+the envelope at around 6 meters.
+This means that the number of nearest neighbours to any empty-space point
+at this and larger distances is lower than it would be expected under complete spatial randomness.
+On the contrary to only slight deviations in case of *G* and *F*
+the *K* and *T* functions show clear deviations from the expected curve and exceed
+the boundaries of given envelopes at the distance of around 5 to 6 meters in
+case of *K* and slightly more in case of *T* function.
+This means that numbers of nearest neighbors at this and larger distances is
+significantly higher than we would exect under complete spatial randomness.
+The overall point pattern thus shows clear signs of clustering process taking place.
+
+This means that the locations of new burials were not chosen at random
+but were affected by the positions of previous burials.
+The pattern begins to cluster at the distance of around 5 to 6 meters, meaning
+that beginning at this distance, the probability of finding a neighboring burial is
+much higher than it would be if the burials were placed randomly.
+The community burying their dead at the cemetery of Vedrovice -- Široká u lesa
+did so in an organized manner taking into account previous burials at the same place,
+meaning their positions were known, probably marked on the surface somehow.
+This would imply either non-perishable marks on the surface of the cemetery or their
+maintenance from time to time.
+
+Spatial pattern was also explored for differently sexed bodies.
 
@@
 
-For relationships (see Fig. \@ref(fig:nb-sex) and Tab. \@ref(tab:nb-tab-sex)) of male-sexed bodies, none of the methods showed any significant
+For relationships (see Fig. \@ref(fig:nb-sex) and Tab. \@ref(tab:nb-tab-sex))
+of male-sexed bodies, none of the methods showed any significant
 differences between expected and observed number of neighbors of a given sex.

### 2020-08-20 text/article_vedrovice.Rmd
cont. spatial org. - multiple K/G

@@
 A polygon of idealized excavation extent is used as a study area (a window).
-Using a window closely approximating locations of the burials was considered,
+<!-- Using a window closely approximating locations of the burials was considered,
 e. g. buffer zone or convex/concave hull but simplified excavation polygon is
 deemed the most useful because it preserves the borders and empty areas
-of the original cemetery.
+of the original cemetery. -->
 
-Overall structure of the burial ground is examined using both density based approach
-and distance based methods. <!--[@osullivan2010, 121-154]-->
+Overall structure of the burial ground can be examined using either density based approaches
+or methods based on distances between neighbors. <!--[@osullivan2010, 121-154]-->
 Density based approaches, for instance quadrat tests,
@@
 The assumption underlying the *first-order effects* is that any point location is
-independent on other point locations and different parameters influence the location.
-On the other hand the methods based on distances between neighbours and triples of neighbors
+independent on other point locations and different parameters influence its location.
+On the other hand the methods based on distances between pairs and triples of neighbours
 characterize the point pattern based on *second- and third-order effects*
@@
 In case of a grave placement in a burial ground we presume that the choice of
-a location is influenced predominantly by other (previous) grave locations and
+a graves location is either random or influenced predominantly by other (previous) grave locations and
 is largely independent on other parameters.
@@
 
-To characterize the pattern based on density, the quadrat test is employed.
+<!-- To characterize the pattern based on density, the quadrat test is employed.
 It divides the area of interest into several cells of fixed sizes
@@
 expected under the null hypothesis that the underlying point process
-is completely spatially random (approximated by a Poisson distribution).
+is completely spatially random (approximated by a Poisson distribution). -->
 Functions $G$, $F$, Ripley's $K$ [@ripley1976; @sayer2013] and $T$ [@schladitz2000] are
-based on the nearest neighbour distances [@nakoinz2016, 135-144].
+based on the nearest neighbour distances and are commonly used
+to characterize point patterns in archaeology [@nakoinz2016, 135-144].
 The $G$ function is a metric of how close events are and summarises
@@
 Because of its dependence on the intensity of events in the study area, it is
-subject to changes in the size of the study area.
+subject to changes in the size of the study area (window).
 The $F$ function, an empty-space function, is based on simulating a point pattern
@@
 certain threshold, not only the nearest neighbours [@nakoinz2016, 138].
-It is widely used in archaeology [@negre2018].
+It is widely used in archaeology cf. [@negre2018].
 Last but not least, the $T$ function is a third-order parallel to the $K$ function [@schladitz2000].
-For formulas of the individual functions see the supplement.
+For formulas and details on individual functions see the supplement.
 
@@
 simulation with 999 iterations.
+Although much smaller number of permutations should be sufficient
+[@baddeley2016, 384] same as @sayer2013 [78] we chose to use 999 permutations.
 Ten most extreme realizations of the simulation are discarded as outliers to
-get a significance level $\alpha = 0.01$ [@baddeley2016 384-403].
+get a significance level $\alpha = 0.01$ [@baddeley2016 410].
 
 To explore whether neighboring individuals are more often of the same sex we use
-three methods to establish neighborhood relations.
+multiple methods to establish neighborhood relations.
 The first two methods are based on constructing neighborhood graphs.
@@
 The Delaunay triangulation can be considered a maximum number of neighbors case
-scenario while the Gabriel graph the exact opposite.
+scenario while the Gabriel graph the opposite.
 For each node, mean number of neighbors of a given sex is determined and recorded.
@@
 
-The third approach of establishing neigborhood is by constructing buffer zones
+Another approach of establishing neigborhood is by constructing buffer zones
 around each of the burials with 6 meters radius.
-The size of a radius is based on the results of nearest neighbors analysis, where
+The size of the radius is based on the results of nearest neighbors analysis, where
 at the given radius there is strong evidence of a clustering process (Fig. \@ref(fig:spat-funs)).
@@
 described for the neighborhood derived from the graphs.
+
+Lastly, a modification of *K* and *G* functions for multitype point patterns
+are used to explore the reationship between individual sex categories.
+These take into account not all of the neighbors of the point pattern, but
+count the statistic for pairs of categories in a given variable.
 
@@
 
-Three methods of neighborhood relationship construction were examined.
-One is distance based where all the neighbors that fall into a given buffer are considered,
+<!-- Methods of neighborhood relationship construction examined include
+distance based approaches where all the neighbors that fall into a given buffer are considered,
 the other two are based on neighborhood graphs, the Delaunay triangulation as a
-maximum number of neighbors scenario and a Gabriel graph as a minimal scenario.
+maximum number of neighbors scenario and a Gabriel graph as a minimal scenario. -->
 

### 2020-09-08 code/exceptionality_index.R
update ei + grouping of ei by hclust

@@ -5,8 +5,8 @@
 # Exceptionality index of individual burials in order to create groups...
 
 # packages =====================================================================
-library(here)
 library(tidyverse)
+library(ggdendro)
 
 # functions ====================================================================
 # root mean square (quadratic mean) for vector x
@@ -15,59 +15,121 @@ rms <- function(x) {
   return(x)
 }
 
-# data =========================================================================
-ved <- read_rds(here("data/temp", "vedrovice_dataset.RDS"))
-# non_rand_vars <- readLines(here("data/temp", "non_random_vars.txt"))
-var_weights <- read_csv(here("data/temp/v_statistics.csv"))
+# # scaling
+# normalize01 <- function(x) {
+#   rg <- range(x, na.rm = TRUE)
+#   (x - rg[1]) / (rg[2] - rg[1])
+# }
 
-normalize01 <- function(x) {
-  rg <- range(x, na.rm = TRUE)
-  (x - rg[1]) / (rg[2] - rg[1])
+# data =========================================================================
+ved <- read_rds(here::here("data/temp/vedrovice_dataset.RDS"))
+# non_rand_vars <- readLines(here("data/temp", "non_random_vars.txt"))
+v_vals <- read_csv(here::here("data/temp/v_statistics.csv")) %>% 
+  select(variable, p) %>% 
+  mutate(weight = 1 - p)
+
+var_weights <- v_vals %>% pull(weight)
+names(var_weights) <- v_vals %>% pull(variable)
+
+
+# weighting variables -----------------------------------------------------
+# m_cont <- matrix(nrow = nrow(ved$bin_vars$count_mat), 
+#                  ncol = ncol(ved$bin_vars$count_mat))
+
+m_bin <- matrix(nrow = nrow(ved$bin_vars$bin_mat), 
+                ncol = ncol(ved$bin_vars$bin_mat))
+
+# for (i in 1:ncol(m_cont)) {
+#   m_cont[, i] <- ved$bin_vars$count_mat[, i] * var_weights[i]
+# }
+
+for (i in 1:ncol(m_bin)) {
[…]

### 2020-09-08 code/randomization.R
add v-stats normalized to range 0-1

@@ -64,6 +64,12 @@ get_p <- function(v_exp, v_obs) {
   mean(v_exp >= v_obs)
 }
 
+# normalize to 0-1 range
+normalize01 <- function(x) {
+  rg <- range(x, na.rm = TRUE)
+  (x - rg[1]) / (rg[2] - rg[1])
+}
+
 # read data ====================================================================
 ved <- readRDS(here("data/temp", "vedrovice_dataset.RDS"))
 
@@ -167,14 +173,42 @@ v_plot
 
 ggsave(here("plots", "v_values.pdf"), plot = v_plot, width = 12, height = 6)
 
+
+# normalizing to range 0-1 ------------------------------------------------
+v_normalized <- bind_rows(exp = v_exp_g, obs = v_obs_g, .id = "orig") %>% 
+  group_by(variable) %>% 
+  nest() %>% 
+  mutate(data = map(data, mutate, value_norm = normalize01(value))) %>% 
+  unnest(cols = data)
+
+v_normalized_plot <- ggplot(filter(v_normalized, orig == "exp"), 
+                            mapping = aes(x = value_norm)) +
+  geom_density(fill = "gray80", color = NA, alpha = 0.6) +
+  geom_rug(alpha = 0.2) +
+  geom_vline(data = filter(v_normalized, orig == "obs"), 
+             mapping = aes(xintercept = value_norm), size = 0.8) +
+  facet_wrap(vars(long), scales = "free", nrow = 4) +
+  xlab("v statistic (normalized to 0 - 1 range)") +
+  geom_text(data = v_annot, aes(x = Inf, y = Inf, label = txt), size = 3,
+            hjust = +1.2, vjust = +1.5) +
+  scale_x_continuous(expand = c(0, 0)) +
+  scale_y_continuous(expand = c(0, 0, 0, 0.2)) +
+  theme_universe
+
+ggsave(here("plots", "v_values_normalized.pdf"), plot = v_normalized_plot, 
+       width = 12, height = 6)
+
+write_csv(v_normalized, here("data/temp", "v_normalized.csv"))
+
+
 # S statistic ==================================================================
 s_observed <- sum(v_observed / ncol_input)
 
-s_p_value <- get_p(s_experimental, s_observed)
-s_p_value < 0.05
-
 s_experimental <- rowSums(apply(v_experimental, 2, 
                                 "/", ncol_input))
+
+s_p_value <- get_p(s_experimental, s_observed)
+s_p_value < 0.05
 
[…]

### 2020-09-08 text/article_vedrovice.Rmd
minor editions in spatial analysis

@@
 social structure. -->
-Here we explore the relations between burials and bodies in terms of biologicall
-sex, age, localness, exceptionality of grave goods and position within the cemetery.
 
@@
 European Neolithic in comparison with other periods.
-This is due to the notion of gender in the Neolihic that is presented as being
+Authors present gender in the Neolihic as being
 of a very different nature than gender in the later prehistory.
-Authors show that gender in the Neolithic is not clearly expressed and
+It is shown that gender in the Neolithic is not clearly expressed and
 biological differences between sexes are not transformed into binary cultural
@@
 
+Here we explore the relations between burials and bodies in terms of biological
+sex, age, local origin, exceptionality of grave goods and location within the
+cemetery.
+We focus predominantly on investigating non-random patterns in artefact
+co-occurences and spatial arragement of the cemetery in relation to the
+proposed *levels* of neolithic identity, i.e. sex, age and localness.
+Our question is thus how are grave good assemblages related to aforementioned
+categories and whether there is a consistent relationship between
+these categories forming the *identity*, grave goods and spatial arrangement
+of the cemetery.
+
 <!-- gender relations in LBK -->
-Although gender relations in the LBK might not be
+<!-- Although gender relations in the LBK might not be -->
 
@@
 
-### Questions
-
+<!-- ### Questions
 * Is there evidence that certain artefacts covary,
@@
 * How are grave good assemblages related to age and sex?
-(... variations in burial rite coinciding with age or sex of the deceased ...)
+(... variations in burial rite coinciding with age or sex of the deceased ...) -->
 
@@
 @masclanslatorre2020.
+Information on localness of different bodies was derived from @hrncir2020.
 
@@
 is completely spatially random (approximated by a Poisson distribution). -->
-Functions $G$, $F$, Ripley's $K$ [@ripley1976; @sayer2013] and $T$ [@schladitz2000] are
+Functions *G*, *F*, Ripley's *K* [@ripley1976; @sayer2013] and *T* [@schladitz2000] are
 based on the nearest neighbour distances and are commonly used
 to characterize point patterns in archaeology [@nakoinz2016, 135-144].
-The $G$ function is a metric of how close events are and summarises
+The *G* function is a metric of how close events are and summarises
 cummulative frequency distribution of the nearest neighbour distances.
@@
 subject to changes in the size of the study area (window).
-The $F$ function, an empty-space function, is based on simulating a point pattern
+The *F* function, an empty-space function, is based on simulating a point pattern
 and calculating of distances between random locations and the observed locations.
-It is a measure of how far are observed events from any arbitrary location [@osullivan2010, 132-135].
-Ripley's $K$ function is in comparison with the $G$ function resilient to
+It is a measure of how far observed events are from any arbitrary location [@osullivan2010, 132-135].
+Ripley's *K* function is in comparison with the *G* function resilient to
 changes in the window size and works with all the proximal events up to a
@@
 It is widely used in archaeology cf. [@negre2018].
-Last but not least, the $T$ function is a third-order parallel to the $K$ function [@schladitz2000].
+Last but not least, the *T* function is a third-order parallel to the *K* function [@schladitz2000].
 For formulas and details on individual functions see the supplement.
@@
 [@baddeley2016, 384] same as @sayer2013 [78] we chose to use 999 permutations.
-Ten most extreme realizations of the simulation are discarded as outliers to
-get a significance level $\alpha = 0.01$ [@baddeley2016 410].
+Ten most extreme realizations of the simulation are discarded as outliers so
+the significance level $\alpha = 0.01$ [@baddeley2016 410].
 
@@
 scenario while the Gabriel graph the opposite.
-For each node, mean number of neighbors of a given sex is determined and recorded.
-To check how common the observed value is, a permutation test
-where sex of burials is randomly assigned while preserving the overall
+For each node, mean number of neighbors of a given category is determined and recorded.
+To check how common the observed value is, a permutation test is performed
+where categories are randomly assigned while preserving the overall
 proportion of given categories and the neighborhood structure determined by the
-graphs is performed.
-The observed values of neighbors sex categories are compared against distributions
-derived from the simulated data.
+graph.
+The observed values of neighbors categories are compared against distributions
+derived from 999 permutations of the simulated data.
 
-Another approach of establishing neigborhood is by constructing buffer zones
-around each of the burials with 6 meters radius.
+Another approach considered for establishing neigborhood relations is by
+constructing buffer zones around each of the burials.
+The question arises how large the buffer zones should be.
+To solve this problem we utilise percolation analysis [@maddison2020].
+Percolation analysis is a method used to identify clusters of points based
+on Euclidean distance.
+
+@maddison2020
+
 The size of the radius is based on the results of nearest neighbors analysis, where
@@
 Data is analyzed in an R environment for statistical computing and graphics
-[@rcore] using functions of many different packages, for the complete list
+[@rcore] using functions of several different packages, for the complete list
 see the supplementary information.

### 2020-09-16 text/article_vedrovice.Rmd
cont analysis + cleaning

@@
 of a very different nature than gender in the later prehistory.
-It is shown that gender in the Neolithic is not clearly expressed and
+Gender in the Neolithic is not clearly expressed and
 biological differences between sexes are not transformed into binary cultural
@@
 Here we explore the relations between burials and bodies in terms of biological
-sex, age, local origin, exceptionality of grave goods and location within the
+sex, age, origin, exceptionality of grave goods and location within the
 cemetery.
@@
 area of the cemetery, amateur excavations during the 19th century or later
-agricultural activities.
-These burials are excluded from the analysis due to missing and/or
+agricultural activities [@podborsky2002].
+These burials are excluded from most of the analysis due to missing and/or
 incomplete data and only a sample of 78 well-preserved burials is analyzed here.
@@
 @masclanslatorre2020.
-Information on localness of different bodies was derived from @hrncir2020.
+Information on localness of different bodies was adapted from @hrncir2020.
 
@@
 thus being useless in establishing relationships between pairs of burials.
-Other variables are removed because they are highly dependent on each other.
+Some other variables are removed because they are highly interdependent.
 
 The variables fall into several categories.
-Firstly there are biological and anthropological characteristics derived from
+Firstly there are biological and anthropological variables derived from
 retrieved human skeletal remains including sex, age simplified into age groups,
-isotope ratios and body height.
+various isotope ratios and body height.
 
-Second large category of variables consists of data on artifacts presence and
-absence in the burials.
+Second large category of variables consists of data on artifacts present in the burials.
 These include variables noting quantities of various types of pottery and
@@
 counts of various raw materials of chipped stone tools,
-presence of polished stone axes or adzes, grinding slabs, pebbles,
+presence of polished and beveled stone axes or adzes, grinding slabs, pebbles,
 bone tools and different kinds of personal ornaments.
-The personal ornaments include beads made of spondylus shell or marble,
+The personal ornament includes beads made of spondylus shell or marble,
 sponylus pendants or oval shaped buckles.
-Presence of ochre pigment is also noted.
-All the variables in this category are simplified by binarization (presence vs
-absence of a trait) in order to minimize effects of only several numerous
-cases on the overall pattern, for instance 20 beads in a grave <!-- 70 -->
-would have much larger effect on the analysis then 1 pottery vessel in the same grave.
+Presence of ochre pigment in the grave is also noted.
 
@@
 I. e. size of the pit, its orientation, position of the body on one of
-the sides and direction of the head.
+the sides and direction of the body and head.
+
+Quantities of various *grave goods* are simplified into a binary matrix indicating
+presence or absence of a given trait. 
+Only grave goods occuring in more than 5% of the graves are used in the analysis.
 
 <!-- methods -->
+
+## Randomness of artefact combinations
+
+<!--
+This study takes advantage of computer-intensive resam-
+pling techniques that are based on simulations of the behavior
+of statistical models through the generation of random samples
+(Baxter, 2003; Efron and Tibshirani, 1993; Manly, 1991). The
+strength of these techniques lies in the potential to effectively
+distinguish between random and non-random patterns and to
+test differences between samples without making assumptions
+about the population from which these samples were drawn
+(Drennan and Peterson, 2004). These techniques proved to
+be very effective for the analysis of patterns in cemeteries
+(Manly, 1996).
+@sosna2008 -->
+
+To determine non-randomness of various artefact combinations we conduct a
+permutation test on the presence/absence data matrix.
+The method based on generalized Monte Carlo tests was first
+introduced to archaeology by @manly1996 and adapted by @sosna2008.
+Our approach is based on both of these works.
+Original (observed) co-occurence matrix is permutated $n$ times with constant
+row and column totals, i.e. both the count of artifacts in individual graves
+and sum of the given type of artifacts across the burials are fixed.
+
+To examine which associations of variables, i.e. artefacts, are infrequent or even not present
+and which tend to appear together more often than it would be expected by random chance only
+the co-occurences of observed and simulated data
+are summarized using two statistics suggested by @manly1996.
+The $S$ statistic is a measure of overall deviation between expected and observed
+number of artefact occurences while
+the $v$ statistic is a measure of difference between co-occurences of
+individual artefact with other artifacts
+[@manly1996, 475, for details see SI].
+To assess relations between groups based on sex, age local origin and exceptionality,
+statistics introduced by @sosna2008 are used.
+
+With $n = 999$ permutations of the original data matrix the results obtained
+are robust+
+
+## Exceptionality index
+
+Exceptionality index ($EI$) is used in order to induce relative exceptionality
+of individual burials based on their attributes.
+This approach is independent on a prior selection of attributes of importance
+or *prestige*.
+It was defined by @demjan2015[367] based on a Root Mean Square (*quadratic mean*) of
+PCA factor scores:
+$$ EI_{attr_g}=\sqrt{\frac{1}{n}\sum_{i=1}^{n}s_i^2} $$
+where $s_i$ are factor scores of individual graves. The mean value is
+normalized to range $[0-1]$:
+
+$$ EI_{attr_g}=\frac{EI_{attr_g}}{max~EI_{attr_g}} $$
 
@@
 count distances between observations used in further analyses. -->
-
-## Randomness of artefact combinations
-
-<!--
-This study takes advantage of computer-intensive resam-
-pling techniques that are based on simulations of the behavior
-of statistical models through the generation of random samples
-(Baxter, 2003; Efron and Tibshirani, 1993; Manly, 1991). The
-strength of these techniques lies in the potential to effectively
-distinguish between random and non-random patterns and to
-test differences between samples without making assumptions
-about the population from which these samples were drawn
-(Drennan and Peterson, 2004). These techniques proved to
-be very effective for the analysis of patterns in cemeteries
-(Manly, 1996).
-@sosna2008 -->
-
-@sosna2013
-
-To determine non-randomness of various artefact combinations we conduct a
-permutation test on the presence/absence data matrix.
-The method based on generalized Monte Carlo tests was first
-introduced to archaeology by @manly1996 and adapted by @sosna2008.
-Our approach is based on both of these works.
-Original (observed) co-occurence matrix is permutated $n$ times with constant
-row and column totals, i. e. both the count of artifacts in individual graves
-and sum of the given type of artifacts across the burials are fixed.
-
-To examine which associations of variables are infrequent or even not present
-and which tend to appear together co-occurences of observed and simulated data
-are summarized using two statistics suggested by @manly1996.
-The $S$ statistic is a measure of overall deviation between expected and observed
-number of artefact occurences while
-the $v$ statistic is a measure of difference between co-occurences of
-individual artefact with other artifacts
-[@manly1996, 475, for details see SI].
-
-With $n = 999$ permutations of the original data matrix the results obtained
-are robust
-
-## Exceptionality index
-
-Exceptionality index ($EI$) is used in order to induce relative exceptionality
-of individual burials based on their attributes.
-This approach is independent on a prior selection of attributes of importance
-or *prestige*.
-It was defined by @demjan2015[367] based on a Root Mean Square (*quadratic mean*) of
-PCA factor scores:
-$$ EI_{attr_g}=\sqrt{\frac{1}{n}\sum_{i=1}^{n}s_i^2} $$
-where $s_i$ are factor scores of individual graves. The mean value is
-normalized to range $[0-1]$:
-
-$$ EI_{attr_g}=\frac{EI_{attr_g}}{max~EI_{attr_g}} $$
 

### 2020-09-16 code/dataset.R

@@ -22,7 +22,9 @@ input <- readr::read_csv(here("data", "data_vedrovice_v01.csv"), skip = 3) %>%
 
 # list to store the data
 ved <- list(bin_vars = list(count_mat = NA,
-                            bin_mat = NA),
+                            bin_mat = NA,
+                            complete_mat = NA,
+                            over5 = NA), # variables (grave goods) present in more than 5% of graves
             cont_vars = NA,
             cat_vars = NA,
             metadata = NA,
@@ -55,6 +57,20 @@ ved$bin_vars$count_mat <- input %>%
 rownames(ved$bin_vars$count_mat) <- ved$id_burials
 
 ved$bin_vars$bin_mat <- apply(ved$bin_vars$count_mat, 2, binarize)
+
+# complete indicator matrix
+ved$bin_vars$complete_mat <- ved$bin_vars$bin_mat %>% 
+  as_tibble() %>% 
+  mutate(across(everything()), abs(. - 1)) %>% 
+  select_all(list(~ paste0(., ".neg"))) %>% 
+  bind_cols(as_tibble(ved$bin_vars$bin_mat)) %>% 
+  select(order(colnames(.))) %>% as.matrix()
+
+rownames(ved$bin_vars$complete_mat) <- ved$id_burials
+
+# grave goods present in more than 5% of graves
+n_burs <- nrow(ved$bin_vars$bin_mat)
+ved$bin_vars$over5 <- (colSums(ved$bin_vars$bin_mat) / n_burs) > 0.05
 
 # continuous variables
 ved$cont_vars <- input %>% 
@@ -138,6 +154,10 @@ ved$var_names$long <- ved$var_names$full$long
 names(ved$var_names$long) <- ved$var_names$full$vnames
 
 # saving processed dataset =====================================================
+if (!dir.exists(here("data", "temp"))) {
+  dir.create(here("data", "temp"))
+}
+
 saveRDS(ved, here("data/temp", "vedrovice_dataset.RDS"))
 
 # ved <- readRDS(here("data/temp", "vedrovice_dataset.RDS"))

### 2020-09-16 code/exceptionality_explore.R

@@ -1,0 +1,69 @@
+# Project "Vedrovice"
+# Script nr. 
+# EXPLORE EXCEPTIONALITY
+# author: Petr Pajdla
+# Explore exceptionality of burials, what does it cause etc...
+
+# libraries
+library(tidyverse)
+library(here)
+
+# data 
+base <- read_rds(here("data/temp", "vedrovice_dataset.RDS"))
+
+ei <- read_csv(here("data/temp", "exceptionality.csv"), col_types = "cdff")
+
+ei_fct_order <- ei %>% group_by(fct) %>% 
+  summarise(mean_ei = mean(ei))
+
+v_vals <- read_csv(here("data/temp", "v_statistics.csv"))
+
+
+# grave goods in various ei  ----------------------------------------------
+ved_gg <- left_join(ei, 
+                    as_tibble(base$bin_vars$count_mat[, base$bin_vars$over5], 
+                              rownames = "burial"),
+                    by = c("burial")) %>% 
+  select(-ei, -ei_cluster)
+
+# proportion of grave goods in a given group
+gg_prop <- function(df, id, group) {
+  temp <- df %>% pivot_longer(-c(!!sym(id), !!sym(group)), 
+                              names_to = "gg", 
+                              values_to = "presence")
+  temp_sums <- temp %>% group_by(gg) %>% 
+    summarise(total = sum(presence)) %>% 
+    ungroup()
+  temp %>% group_by(!!sym(group), gg) %>% 
+    summarise(sum = sum(presence)) %>% 
+    group_map(~ mutate(.x, prop = sum/temp_sums$total), .keep = TRUE) %>% 
+    bind_rows()
+}
+
+# reordering factors
+ved_gg_prop <- gg_prop(ved_gg, "burial", "fct") %>% 
+  left_join(ei_fct_order) %>% 
+  mutate(fct = fct_reorder(fct, desc(mean_ei)),
+         gg_order = as.integer(fct_relevel(gg, v_vals$variable)),
+         gg_long = unname(base$var_names$long[gg]),
+         gg_long = fct_reorder(gg_long, desc(gg_order))) %>% 
+  select(-mean_ei)
+
+
+# plotting
+ved_gg_prop %>% ggplot(aes(y = prop, x = gg_long)) +
+  geom_col(fill = "white", color = "black") +
+  facet_wrap(vars(fct), nrow = 1) +
+  coord_flip() +
[…]

### 2020-09-16 code/exceptionality_index.R

@@ -31,36 +31,37 @@ v_vals <- read_csv(here::here("data/temp/v_statistics.csv")) %>%
 var_weights <- v_vals %>% pull(weight)
 names(var_weights) <- v_vals %>% pull(variable)
 
+ved_in <- ved$bin_vars$bin_mat[, ved$bin_vars$over5]
 
 # weighting variables -----------------------------------------------------
 # variables are weighted by the p value (measure of departure from norm) of v stats
 # m_cont <- matrix(nrow = nrow(ved$bin_vars$count_mat), 
 #                  ncol = ncol(ved$bin_vars$count_mat))
 
-m_bin <- matrix(nrow = nrow(ved$bin_vars$bin_mat), 
-                ncol = ncol(ved$bin_vars$bin_mat))
+m_bin <- matrix(nrow = nrow(ved_in), 
+                ncol = ncol(ved_in))
 
 # for (i in 1:ncol(m_cont)) {
 #   m_cont[, i] <- ved$bin_vars$count_mat[, i] * var_weights[i]
 # }
 
 for (i in 1:ncol(m_bin)) {
-  m_bin[, i] <- ved$bin_vars$bin_mat[, i] * var_weights[i]
+  m_bin[, i] <- ved_in[, i] * var_weights[i]
 }
 
 # colnames(m_cont) <- colnames(ved$bin_vars$count_mat)
-colnames(m_bin) <- colnames(ved$bin_vars$bin_mat)
+colnames(m_bin) <- colnames(ved_in)
 
 
 # exceptionality index =========================================================
 # art_pca_c <- prcomp(m_cont, scale. = TRUE, center = TRUE)
-art_pca_b <- prcomp(m_bin)
+art_pca_b <- prcomp(m_bin, scale. = TRUE, center = TRUE)
 
 # summary(art_pca_c)
 # ggbiplot::ggbiplot(art_pca_c)
 
-summary(art_pca_b)
-# ggbiplot::ggbiplot(art_pca_b)
+# summary(art_pca_b)
+ggbiplot::ggbiplot(art_pca_b)
 
 pc_threshold <- which(summary(art_pca_b)$importance[3, ] >= 0.9)[1]
 
@@ -97,8 +98,8 @@ ei_gg %>% ggplot(aes(ei, n_gg_types)) +
 cor(ei_gg$ei, ei_gg$n_gg)
 cor(ei_gg$ei, ei_gg$n_gg_types)
 # there is a significant correlation between exceptionality and:
[…]

### 2020-09-16 code/explore_exceptionality.R

@@ -1,99 +1,0 @@
-# Project "Vedrovice"
-# Script nr. 5
-# EXPLORE EXCEPTIONALITY
-# author: Petr Pajdla
-# Explore exceptionality of burials, what does it cause etc...
-
-# libraries
-library(tidyverse)
-
-# data 
-base <- read_rds("./data/vedrovice_dataset.RDS")
-
-ei <- read_csv("./data/output/exceptionality.csv")
-
-non_random_vars <- read_lines("./data/output/non_random_vars.txt")
-
-# explore...
-# spatial distribution
-spatial_clustering <- hclust(dist(base$layout, method = "euclidean"))
-spatial_groups <- cutree(spatial_clustering, k = 8)
-
-
-ggplot(as_tibble(base$layout), aes(layout_x, layout_y, color = factor(spatial_groups))) +
-  geom_point() + 
-  ggforce::geom_mark_hull(expand = unit(2.2, "mm"))
-
-ggplot(as_tibble(base$layout), aes(layout_x, layout_y, color = factor(spatial_groups))) +
-  geom_point() + 
-  ggforce::geom_mark_ellipse(expand = unit(2, "mm"))
-
-ved_spatial <- bind_cols(as_tibble(base$layout), ei = ei$group, spat_gr = spatial_groups)
-
-ggplot(ved_spatial, aes(layout_x, layout_y, color = ei)) +
-  geom_point() + 
-  facet_wrap(~factor(spat_gr))
-
-ggplot() +
-  geom_point(ved_spatial[, c("layout_x", "layout_y")], mapping = aes(layout_x, layout_y), alpha = 0.1) +
-  geom_point(ved_spatial, mapping = aes(layout_x, layout_y, color = ei)) +
-  # stat_density2d() +
-  facet_wrap(~ei)
-
-# spat distr. using kmeans (experiment)
-spat_kmeans <- kmeans(base$layout, centers = 8, nstart = 200)
-
-scree1 <- function(df) {
-  x <- vector(mode = "double", length = 12)
-   for (i in 1:12) {
-     x[i] <- kmeans(df, centers = i, nstart = 20)$tot.withinss
-   }
-  return(x)
-}
-
-plot(scree1(base$layout), type = "b")
-
-ggplot(ved_spatial, aes(layout_x, layout_y, color = factor(spat_kmeans$cluster))) +
-  geom_point() +
-  ggforce::geom_mark_ellipse(expand = unit(2, "mm"))
-
-
-# groups items...
[…]

### 2020-09-16 code/local.R

@@ -2,7 +2,8 @@
 # Script nr. 1.2
 # LOCAL AND NON-LOCAL Sr RANGES
 # author: Petr Pajdla
-# Assign bodies into local vs non-local Sr range
+# Simplify and clean age groups 
+# Assign bodies into local vs non-local based on Sr range
 
 library(tidyverse)
 library(here)
@@ -43,11 +44,12 @@ ved_sr <- ved$metadata %>%
          ) %>% 
   filter(!is.na(group), !group %in% c("n. a.", "ind."))
 
-local <- ved_sr %>% group_by(group) %>% 
+local <- ved_sr %>% 
+  filter(group == "juvenile",
+         !is.na(sr)) %>%
   summarise(n(), 
             mean = mean(sr, na.rm = TRUE), 
             sd = sd(sr, na.rm = TRUE)) %>% 
-  filter(group == "juvenile") %>%
   mutate(min = mean - 2 * sd,
          max = mean + 2 * sd)
 
@@ -68,7 +70,7 @@ ved_local %>%
   facet_wrap(~group, nrow = 3) +
   ggthemes::theme_clean()
 
-
+ggsave(here("plots", "local_range.pdf"))
 
 # export ------------------------------------------------------------------
 

### 2020-09-16 code/randomization.R

@@ -74,7 +74,7 @@ normalize01 <- function(x) {
 ved <- readRDS(here("data/temp", "vedrovice_dataset.RDS"))
 
 # counting co-occurrences in observed matrix ===================================
-input_matrix <- ved$bin_vars$bin_mat
+input_matrix <- ved$bin_vars$bin_mat[, ved$bin_vars$over5]
 ncol_input <- ncol(input_matrix)
 
 cooc_obs <- count_cooccurrence(matrix = input_matrix)
@@ -83,26 +83,26 @@ cooc_obs <- cooc_obs %>% filter(var1 != var2)
 
 # randomization of co-occurrence matrix - list of many matrices ----------------
 n_permutations <-  9999
-# # ======
-# rand_mat <- vegan::permatfull(input_matrix,
-#                               fixedmar = "both",
-#                               mtype = "prab",
-#                               times = n_permutations)
-# # ======
+# ======
+rand_mat <- vegan::permatfull(input_matrix,
+                              fixedmar = "both",
+                              mtype = "prab",
+                              times = n_permutations)
+# ======
 
 # going parallel to speed up randomization...
 # detecting cores for parallel
 no_cores <- parallel::detectCores() - 1
-
+  
 # cooccurences on random matrices ==============================================
-# # ======
-# cl <- parallel::makeCluster(no_cores)
-# cooc_rand <- parallel::parLapply(cl, rand_mat$perm, count_cooccurrence)
-# parallel::stopCluster(cl)
-# 
-# # save counts of co-occurences on random matrices
-# readr::write_rds(cooc_rand, here("data/temp", "cooc_random_mat.RDS"))
-# # ======
+# ======
+cl <- parallel::makeCluster(no_cores)
+cooc_rand <- parallel::parLapply(cl, rand_mat$perm, count_cooccurrence)
+parallel::stopCluster(cl)
+
+# save counts of co-occurences on random matrices
+readr::write_rds(cooc_rand, here("data/temp", "cooc_random_mat.RDS"))
+# ======
 
 # load from temporary data
 cooc_rand <- readr::read_rds(here("data/temp", "cooc_random_mat.RDS"))
@@ -332,7 +332,7 @@ plot(g_cooc_positive,
      vertex.label.color = "black",
      edge.width = edge_attr(g_cooc_positive)$weight^1.2,
      edge.color = "black",
-     mark.groups = cluster_louvain(g_cooc_positive),
[…]

### 2020-09-16 code/spat_dataprep.R

@@ -13,6 +13,10 @@ library(tidyverse)
 ved <- read_rds(here("data/temp", "vedrovice_dataset.RDS"))
 origin <- read_rds(here("data/temp/local.rds")) %>% 
   select(id_burial, age_mean, origin)
+ei <- read_csv(here("data/temp", "exceptionality.csv")) %>% 
+  mutate(ei_clust = factor(fct),
+         id_burial = as.character(burial)) %>% 
+  select(-burial, -ei_cluster, -fct)
 
 # data prep ====================================================================
 scale_m <- 4.762 # scale by a factor so 1 unit is 1 meter
@@ -26,8 +30,9 @@ ved_layout <- tibble(layout_x = ved_layout$x, layout_y = ved_layout$y)
 # not a rectangle, but a polygon closely bounding the burials using convex hull
 ved_sf <- st_as_sf(bind_cols(ved_layout, ved$metadata), 
                    coords = c("layout_x", "layout_y")) %>% 
-  left_join(origin) %>% 
-  mutate(sex = fct_relevel(sex, c("n. a.", "F", "M", "ind.")))
+  left_join(origin, by = "id_burial") %>% 
+  mutate(sex = fct_relevel(sex, c("n. a.", "F", "M", "ind."))) %>% 
+  left_join(ei, by = "id_burial")
 # ved_sf_buffer <- st_buffer(ved_sf, dist = 1.6)
 # ved_sf_buffer <- st_simplify(ved_sf_buffer, dTolerance = 1)
 ved_sf_hull <- st_convex_hull(st_union(st_geometry(ved_sf)))
@@ -46,8 +51,8 @@ ved_exc <- st_cast(
         matrix(c(ved_exc$x, ved_exc$y), ncol = 2)))), "POLYGON")
 
 # window_rect <- owin(xrange = c(10, 22), yrange = c(10, 22.7))
-window_poly <- as.owin(st_cast(ved_sf_hull, "POLYGON"))
-window_exc <- as.owin(ved_exc)
+window_poly <- spatstat::as.owin(st_cast(ved_sf_hull, "POLYGON"))
+window_exc <- spatstat::as.owin(ved_exc)
 
 # plot
 gg_baseplan <- ggplot(data = ved_sf) +
@@ -91,6 +96,66 @@ ggplot(data = filter(ved_sf, !is.na(origin))) +
   theme(legend.position = c(0.9, 0.8))
 
[…]

### 2020-09-16 code/spatial_neighbors.R

@@ -29,17 +29,18 @@ theme_universe <- theme(panel.border = element_rect(colour = "black",
 # data --------------------------------------------------------------------
 
 # ved <- read_rds(here("data/temp", "vedrovice_dataset.RDS"))
-ei <- read_csv(here("data/temp", "exceptionality.csv"), col_types = "cdff") %>% 
-  select(id_burial = burial, ei, ei_clust = fct)
+# ei <- read_csv(here("data/temp", "exceptionality.csv"), col_types = "cdff") %>% 
+#   select(id_burial = burial, ei, ei_clust = fct)
 ved_sf <- read_sf(here("data/temp", "layout.shp")) %>% 
   mutate(sex = fct_relevel(sex, c("n. a.", "F", "M", "ind.")),
          age_sim = if_else(is.na(age_sim), "ind.", age_sim),
          age_sim = fct_relevel(age_sim, c("you", "mid", "old", "ind.")),
          origin = if_else(is.na(origin), "ind.", origin),
-         origin = fct_relevel(origin, c("local", "non-local"))) %>% 
-  left_join(ei, by = "id_burial") %>% 
-  mutate(ei_clust = if_else(is.na(ei_clust), "ind.", as.character(ei_clust)),
-         ei_clust = factor(ei_clust, levels = c(letters[1:7], "ind.")))
+         origin = fct_relevel(origin, c("local", "non-local")),
+         ei_clust = if_else(is.na(ei_clust), "ind.", as.character(ei_clust)),
+         ei_clust = factor(ei_clust, 
+                           levels = c(letters[1:length(unique(.$ei_clust))-1], 
+                                      "ind.")))
 ved_layout <- st_coordinates(ved_sf)
 ved_exc <- read_sf(here("data/temp", "window.shp"))
 
@@ -185,30 +186,35 @@ randomize_neighbors_network <- function(g, sf, variable, n_sim = 99) {
 
 # ved_rand_sex_g <- randomize_neighbors_network(ved_g, ved_sf, "sex", n_sim = 999)
 # write_csv(ved_rand_sex_g, here("data/temp", "ved_rand_sex_g.csv"))
[…]

### 2020-09-17 code/randomization.R
create clusters of variables based on co-occurence/co-absence networks

@@ -11,6 +11,7 @@ set.seed(42)
 # packages =====================================================================
 library(here)
 library(dplyr)
+library(purrr)
 library(igraph)
 library(ggplot2)
 
@@ -83,26 +84,26 @@ cooc_obs <- cooc_obs %>% filter(var1 != var2)
 
 # randomization of co-occurrence matrix - list of many matrices ----------------
 n_permutations <-  9999
-# ======
-rand_mat <- vegan::permatfull(input_matrix,
-                              fixedmar = "both",
-                              mtype = "prab",
-                              times = n_permutations)
-# ======
+# # ======
+# rand_mat <- vegan::permatfull(input_matrix,
+#                               fixedmar = "both",
+#                               mtype = "prab",
+#                               times = n_permutations)
+# # ======
 
 # going parallel to speed up randomization...
 # detecting cores for parallel
 no_cores <- parallel::detectCores() - 1
   
 # cooccurences on random matrices ==============================================
-# ======
-cl <- parallel::makeCluster(no_cores)
-cooc_rand <- parallel::parLapply(cl, rand_mat$perm, count_cooccurrence)
-parallel::stopCluster(cl)
-
-# save counts of co-occurences on random matrices
-readr::write_rds(cooc_rand, here("data/temp", "cooc_random_mat.RDS"))
-# ======
+# # ======
+# cl <- parallel::makeCluster(no_cores)
+# cooc_rand <- parallel::parLapply(cl, rand_mat$perm, count_cooccurrence)
+# parallel::stopCluster(cl)
+# 
+# # save counts of co-occurences on random matrices
+# readr::write_rds(cooc_rand, here("data/temp", "cooc_random_mat.RDS"))
+# # ======
 
 # load from temporary data
 cooc_rand <- readr::read_rds(here("data/temp", "cooc_random_mat.RDS"))
@@ -131,7 +132,7 @@ v_experimental <- bind_rows(parallel::parLapply(cl, v_experimental,
 parallel::stopCluster(cl)
 
 # P-val: percentage of experimental v values that are larger than observed v ---
-v_p_values <- purrr::map2(v_experimental, v_observed, get_p) %>% 
[…]

### 2020-09-17 text/article_vedrovice.Rmd
percolation analysis

@@
   distributions of objects from the Early Neolithic cemetery in Vedrovice.
-
   As a result of the investigation we conclude that...  "
@@
 Here we explore the relations between burials and bodies in terms of biological
-sex, age, origin, exceptionality of grave goods and location within the
+sex, age and origin, exceptionality of grave goods and location within the
 cemetery.
@@
 The material analyzed here comes from Vedrovice – Široká u lesa cemetery.
-
 <!-- sample (burials, artefacts) -->
@@
 Information on localness of different bodies was adapted from @hrncir2020.
+Local Sr range is defined as two standard deviations from mean of Sr ranges of
+juveniles.
 
@@
 thus being useless in establishing relationships between pairs of burials.
-Some other variables are removed because they are highly interdependent.
+Only grave goods occuring in more than 5% of the graves are used in the analysis.
+<!-- Some other variables are removed because they are highly interdependent.-->
 
@@
 presence or absence of a given trait.
-Only grave goods occuring in more than 5% of the graves are used in the analysis.
 
@@
 Our approach is based on both of these works.
-Original (observed) co-occurence matrix is permutated $n$ times with constant
+Original co-occurence matrix is permutated *n = 9999* times with constant
 row and column totals, i.e. both the count of artifacts in individual graves
@@
 To examine which associations of variables, i.e. artefacts, are infrequent or even not present
-and which tend to appear together more often than it would be expected by random chance only
+and which tend to appear together more often than it would be expected by random chance
 the co-occurences of observed and simulated data
 are summarized using two statistics suggested by @manly1996.
-The $S$ statistic is a measure of overall deviation between expected and observed
+The *S* statistic is a measure of overall deviation between expected and observed
 number of artefact occurences while
-the $v$ statistic is a measure of difference between co-occurences of
+the *v* statistic is a measure of difference between co-occurences of
 individual artefact with other artifacts
 [@manly1996, 475, for details see SI].
-To assess relations between groups based on sex, age local origin and exceptionality,
+To assess relations between groups based on sex, age and origin,
 statistics introduced by @sosna2008 are used.
 
-With $n = 999$ permutations of the original data matrix the results obtained
-are robust+
+<!-- ## Exceptionality index -->
 
-## Exceptionality index
+<!-- Exceptionality index (*EI*) is used in order to induce relative exceptionality -->
+<!-- of individual burials based on their attributes. -->
+<!-- This approach is independent on a prior selection of attributes of importance -->
+<!-- or *prestige*. -->
+<!-- It was defined by @demjan2015[367] based on a Root Mean Square (*quadratic mean*) of -->
+<!-- PCA factor scores: -->
+<!-- $$ EI_{attr_g}=\sqrt{\frac{1}{n}\sum_{i=1}^{n}s_i^2} $$ -->
+<!-- where $s_i$ are factor scores of individual graves. The mean value is -->
+<!-- normalized to range $[0-1]$: -->
 
-Exceptionality index ($EI$) is used in order to induce relative exceptionality
-of individual burials based on their attributes.
-This approach is independent on a prior selection of attributes of importance
-or *prestige*.
-It was defined by @demjan2015[367] based on a Root Mean Square (*quadratic mean*) of
-PCA factor scores:
-$$ EI_{attr_g}=\sqrt{\frac{1}{n}\sum_{i=1}^{n}s_i^2} $$
-where $s_i$ are factor scores of individual graves. The mean value is
-normalized to range $[0-1]$:
-
-$$ EI_{attr_g}=\frac{EI_{attr_g}}{max~EI_{attr_g}} $$
+<!-- $$ EI_{attr_g}=\frac{EI_{attr_g}}{max~EI_{attr_g}} $$ -->
 
@@
 
+<!-- Percolation analysis -->
 Another approach considered for establishing neigborhood relations is by
@@
 The question arises how large the buffer zones should be.
-To solve this problem we utilise percolation analysis [@maddison2020].
+To avoid this problem we utilise percolation analysis.
 Percolation analysis is a method used to identify clusters of points based
-on Euclidean distance.
+on spatial separation measured by Euclidean distance [@maddison2020, 269].
+The introduction of percolation analysis to archaeology is relatively recent,
+nevertheless its effectivness and usefulness in studies with varied scales
+is already proven [@maddison2020].
+Percolation analysis is based on city clustering algorithm [@rozenfeld2008] and
+its generalization utilizing Euclidean distance [@rozenfeld2011].
+Nodes being in a given distance threshold (radius) are assigned into a cluster.
+This process is repeated for a set of thresholds while recording maximum
+cluster sizes and other statistics.
+This allows to identify distances at which clusters exhibit rapid grow (absorbing
+large number of nodes) and distances at which the clusters remain relatively stable.
+For more details refer to @maddison2020 and associated R package [@schmidt2020].
 
-@maddison2020
-
-The size of the radius is based on the results of nearest neighbors analysis, where
+<!-- The size of the radius is based on the results of nearest neighbors analysis, where
 at the given radius there is strong evidence of a clustering process (Fig. \@ref(fig:spat-funs)).
 The randomness is again tested using a permutation test similar to the one
-described for the neighborhood derived from the graphs.
+described for the neighborhood derived from the graphs.-->
 
@@
 
+## Artifact combinations
+
+By counting co-occurences of analyzed variables on simulated data matrices and comparing
+the distributions of the two studied statistics with the values calculated for
+the original dataset we can establish whether the overall pattern of artefacts
+distribution in the graves is non-random and further identify artefacts that deviate
+from the random pattern.
+The overall deviation from expected co-occurences, i.e. the *S* statistic
+[@manly1996, 475] is significantly large at the 0.05 alpha level.
+This suggests that there is strong evidence that the overall distribution of
+artefacts in different graves is not random (see Fig. \@ref(fig:sstat)).
+
+Examining distributions of individual artefacts it is found that at the 0.05
+significance level only adzes, grinding tools, ceramic bowls and spondylus beads
+are distributed randomly.
+Close to the 0.05 alpha level are distributions of bone tools and ceramic bottles
+(Fig. \@ref(fig:vstat)).
+Among the grave goods that are found to be randomly distributed in the graves
+across the burial ground are ochre pigment, lithics, pebbles, globular pots and
+special types of pottery vessel shapes and spondylus pendants and *buckles*.
+The relationships between artefact types frequently occuring and missing together
+in graves are visualized in Fig. (\@ref(fig:coocnets)) with
+clusters highlighted using fast-greedy algorithm [@newman2004] and strength of the
+given association is expressed by the thickness of the edge connecting the nodes.
+Artifacts most frequently present with other artifacts are with 5 links grinding tools and
+ceramic bowls and with 4 links adzes, non-local lithic tools, ochre pigment and
+spondylus beads.
+The association is strongest between spondylus beads and (a) ceramic bottles and
+(b) pigment; and adzes and (a) non-local lithics and (b) bone tools.
+Artifacts that are frequently absent in the graves together are with 6 links
+local lithic tools, globular pots and ceramic bowls and spondylus buckels with
+5 and 4 links respectively.
+The strongest antagonist relationships are between globular pots and (a) bowls and
+(b) adzes.
+
 ## Spatial patterning
@@
 The quadrat test shows evidence for both random and clustered processes taking place with p values
-under the 0.05 alpha value but the $\chi^2$ approximation is inaccurate due to small
+under the 0.05 alpha level but the $\chi^2$ approximation is inaccurate due to small
 expected counts (for the details see supplementary information).
@@
 the *K* and *T* functions show clear deviations from the expected curve and exceed
-the boundaries of given envelopes at the distance of around 5 to 6 meters in
-case of *K* and slightly more in case of *T* function.
+the boundaries of given envelopes at the distance of around 5 to 6 meters.
+<!-- in case of *K* and slightly more in case of *T* function.-->
 This means that numbers of nearest neighbors at this and larger distances is
@@
 meaning their positions were known, probably marked on the surface somehow.
-This would imply either non-perishable marks on the surface of the cemetery or their
+This would imply existence of non-perishable marks on the surface of the cemetery and/or their
 maintenance from time to time.
 
-Spatial pattern was also explored for differently sexed bodies.
+Results of percolation analysis reveal more detailed view of the clustering process
+taking place.
+The curve in Fig. \@ref(fig:perc-distance) reveals that most of the growth into
+a single cluster happens at distances between approximately 4 and 6 meters.
+The distances where relative stability of clusters emerges are labeled.
+The clusters at these distances are shown in Fig. \@ref(fig:perc-plan).
 
@@
 
-## Randomness of artefact combinations
-
-By counting co-occurences of analyzed variables on simulated data matrices and comparing
-the distributions of the two studied statistics with the values calculated for
-the original dataset we can establish whether the overall pattern of artefacts
-distribution in the graves is non-random and further identify artefacts that deviate
-from the random pattern.
-The value of $S$ statistic is deemed significantly large at the 0.05 alpha level with
-$S \approx 14$ and $p \approx xy$ (Fig. \@ref(fig:f-sstat)).
-This suggests that there is enough evidence that the overall distribution of
-artefacts in the graves is not random.
-
 <!-- ```{r t-vstat} -->
@@
 <!-- FIGURES --><!-- FIGURES --><!-- FIGURES --><!-- FIGURES --><!-- FIGURES -->
-```{r plan-graphs, fig.cap = "Neighborhood graphs: Delaunay triangulation and Gabriel graph"}
-knitr::include_graphics(here("plots/plan_gabriel.pdf"))
-```
 
-```{r label-spat-funs}
-label_spat_funs <- paste("Functions *G*, *F*, *K* and *T*.",
-"*Dashed line* -- theoretical curve under complete spatial randomness, *solid line* -- empirical curve, *gray area* -- envelope.")
-```
-```{r spat-funs, fig.cap=label_spat_funs}
-knitr::include_graphics(here("plots/pointprocess_fun.pdf"))
-```
-
-```{r label-nb-sex}
-label_nb_sex <- paste("Mean neighbors in a given sex category.",
-"*Gray area* -- density estimation for the simulated data, *vertical line* -- observed value.",
-"*F* -- female-sexed body, *M* -- male-sexed body, *n. a.* -- non adult, *ind.* -- indetermined sex.")
-```
-```{r nb-sex, fig.cap=label_nb_sex}
-knitr::include_graphics(here("plots/nb_sex.pdf"))
-```
-
-```{r f-sstat, fig.cap = "Distribution of simulated and observed S statistic"}
+```{r sstat, fig.cap = "Distribution of simulated and observed *S* statistic"}
 knitr::include_graphics(here("plots/s_statistic.pdf"))
@@
 
-```{r f-vstat, fig.cap = "Simulated and observed v statistics for different variables"}
+```{r vstat, fig.cap = "Simulated and observed *V* statistics for different variables"}
 knitr::include_graphics(here("plots", "v_values.pdf"))
@@
 
-```{r f-cooc, fig.cap = "Networks of variables frequently present together and exluding each other"}
+```{r coocnets, fig.cap = "Networks of variables frequently present together and exluding each other"}
 knitr::include_graphics(here("plots", "cooc_networks.pdf"))
@@
 
+<!-- ```{r plan-graphs, fig.cap = "Neighborhood graphs: Delaunay triangulation and Gabriel graph"} -->
+<!-- knitr::include_graphics(here("plots/plan_gabriel.pdf")) -->
+<!-- ``` -->
+
+<!-- ```{r label-spat-funs} -->
+<!-- label_spat_funs <- paste("Functions *G*, *F*, *K* and *T*.", -->
+<!-- "*Dashed line* -- theoretical curve under complete spatial randomness, *solid line* -- empirical curve, *gray area* -- envelope.") -->
+<!-- ``` -->
+<!-- ```{r spat-funs, fig.cap=label_spat_funs} -->
+<!-- knitr::include_graphics(here("plots/pointprocess_fun.pdf")) -->
+<!-- ``` -->
+
+<!-- ```{r label-nb-sex} -->
+<!-- label_nb_sex <- paste("Mean neighbors in a given sex category.", -->
+<!-- "*Gray area* -- density estimation for the simulated data, *vertical line* -- observed value.", -->
+<!-- "*F* -- female-sexed body, *M* -- male-sexed body, *n. a.* -- non adult, *ind.* -- indetermined sex.") -->
+<!-- ``` -->
+<!-- ```{r nb-sex, fig.cap=label_nb_sex} -->
+<!-- knitr::include_graphics(here("plots/nb_sex.pdf")) -->
+<!-- ``` -->
+
 <!-- ======================================================================= -->
 <!-- TABLES --><!-- TABLES --><!-- TABLES --><!-- TABLES --><!-- TABLES -->
-```{r nb-tab-sex}
-tab <- read_csv(here("data/temp/nb_sex_p.csv")) %>%
-         mutate(relation = str_c(from, to, sep = " - ")) %>%
-  select(relation, everything(), -from, -to)
+<!-- ```{r nb-tab-sex} -->
+<!-- tab <- read_csv(here("data/temp/nb_sex_p.csv")) %>% -->
+<!--          mutate(relation = str_c(from, to, sep = " - ")) %>% -->
+<!--   select(relation, everything(), -from, -to) -->
 
-tnames <- c("relation", rep(c("mean", "p"), 3))
+<!-- tnames <- c("relation", rep(c("mean", "p"), 3)) -->
 
-knitr::kable(tab, booktabs = TRUE, col.names = tnames, linesep = "", caption = "Results...") %>%
-  kableExtra::add_header_above(c(" ", "Buffer" = 2, "Delaunay" = 2, "Gabriel" = 2))
-```
+<!-- knitr::kable(tab, booktabs = TRUE, col.names = tnames, linesep = "", caption = "Results...") %>% -->
+<!--   kableExtra::add_header_above(c(" ", "Buffer" = 2, "Delaunay" = 2, "Gabriel" = 2)) -->
+<!-- ``` -->
 

### 2020-09-25 text/article_vedrovice.Rmd
finish spatial analysis

@@
 abstract        :
-  "Spatial patterns in cemeteries and grave goods associations are often explored using various informal methods.
-  We adopt computationaly intensive methods utilizing resampling approaches both to spatial analysis and
-  distributions of objects from the Early Neolithic cemetery in Vedrovice.
+  "Spatial patterns in cemeteries and grave goods associations at the
+  cemetery of Vedrovice - Široká u Lesa are explored using computationaly
+  instensive formal methods.
+  Resampling approaches both to intrasite spatial arrangement and
+  distribution of grave goods are explored in
+  relation to layers of identity represented by sex, age and origin of the
+  buried bodies.
   As a result of the investigation we conclude that...  "
@@
 knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
-library(tidyverse)
+# library(tidyverse)
 library(here)
@@
 sexual inequalities and gender in the Neolithic [@masclanslatorre2020; @bickle2020].
-<!-- Here we focus on detailed analysis of the distribution of the grave goods and
-spatial analysis of the Vedrovice cemetery in respect to gender relations and
-social structure. -->
+Here we focus on exploration of relationship between three proposed *layers* of
+identity, biological sex and age of the body determined from sekeletal remains
+by anthropological methods and origin of the body based on Strontium isotope
+signal.
+We explore whether the structure observed in these three categories is
+mirrored in the grave goods associations and spatial organization within the
+cemetery.
+The analysis is set into a broader discussion of gender relations and identity
+inside Neolithic communities.
 
@@
 
-Here we explore the relations between burials and bodies in terms of biological
+<!-- We focus on relations between burials and bodies in terms of biological
 sex, age and origin, exceptionality of grave goods and location within the
-cemetery.
+cemetery.-->
 We focus predominantly on investigating non-random patterns in artefact
 co-occurences and spatial arragement of the cemetery in relation to the
-proposed *levels* of neolithic identity, i.e. sex, age and localness.
+proposed *layers* of Neolithic identity, i.e. sex, age and localness of the body.
 Our question is thus how are grave good assemblages related to aforementioned
 categories and whether there is a consistent relationship between
-these categories forming the *identity*, grave goods and spatial arrangement
-of the cemetery.
+these categories forming the *identity*, grave goods distributions and spatial
+arrangement of the cemetery.
 
@@
 ability to distinguish random and non-random patterns.
+This is crucial for mortuary analysis because the samples studied, including
+the one presented here, are usually small in size and identifying which
+observed patterns are caused by mere chance associated with small numbers
+and which are actually present.
 
@@
 <!-- site -->
-The material analyzed here comes from Vedrovice – Široká u lesa cemetery.
+Vedrovice site is situated in the south of Moravia (Czech Republic).
+Earliest known finds and archaeological explorations here are from
+the 19th century but the excavation took place in the second half of 20th century.
+There are two known burial areas, *Za Dvorem* and larger one *Široká u Lesa* and
+a settlement area.
+For more details on the site refer to @masclanslatorre2020 or the
+original publication [@podborsky2002].
+The material analyzed here comes from Vedrovice *Široká u lesa* cemetery.
+
 <!-- sample (burials, artefacts) -->
-Originally, the burial ground consisted of at least 108 burials.
+Originally, this burial ground consisted of at least 108 burials.
 Some of the burials were disturbed by later settlement and burial pits in the
@@
 These burials are excluded from most of the analysis due to missing and/or
-incomplete data and only a sample of 78 well-preserved burials is analyzed here.
+incomplete data and only a sample of 78 well-preserved LBK burials is analyzed here.
 Most of the data was obtained from the original publication [@podborsky2002],
-anthropological determination of sex and age were corrected based on
-more recent analysis by @dockalova2008. Carbon, nitrogen and strontium isotope
+anthropological determination of sex and age of the bodies were corrected based on
+analysis by @dockalova2008. Carbon, nitrogen and strontium isotope
 ratios were collected in @whittle2013 and @richards2008.
@@
 Information on localness of different bodies was adapted from @hrncir2020.
-Local Sr range is defined as two standard deviations from mean of Sr ranges of
-juveniles.
 
-Each burial was originally characterized by 55 variables but only subset of
-35 variables was selected for further analysis.
-Some of  the original variables are omitted due to their rare occurrence,
+Each burial was originally characterized by 20 variables recording counts of
+artefacts present in the graves.
+Some of  these original variables are omitted in the analysis due to their rare occurrence,
 for instance spondylus arm ring is present in a single burial only,
-thus being useless in establishing relationships between pairs of burials.
-Only grave goods occuring in more than 5% of the graves are used in the analysis.
-<!-- Some other variables are removed because they are highly interdependent.-->
-
-The variables fall into several categories.
-Firstly there are biological and anthropological variables derived from
-retrieved human skeletal remains including sex, age simplified into age groups,
-various isotope ratios and body height.
-
-Second large category of variables consists of data on artifacts present in the burials.
-These include variables noting quantities of various types of pottery and
-positions of the pottery vessels within the graves,
-counts of various raw materials of chipped stone tools,
-presence of polished and beveled stone axes or adzes, grinding slabs, pebbles,
-bone tools and different kinds of personal ornaments.
-The personal ornament includes beads made of spondylus shell or marble,
-sponylus pendants or oval shaped buckles.
-Presence of ochre pigment in the grave is also noted.
-
-Last category describes physical characteristics of the grave pit and the
-position of the body.
-I. e. size of the pit, its orientation, position of the body on one of
-the sides and direction of the body and head.
-
+thus being of no use in establishing relationships between pairs of burials.
+Only grave goods occuring in more than 5% of the graves are used in the analysis,
+this step reduces the count of the variables to 14.
+These include variables noting quantities of various types of pottery vessels in
+the graves, counts of chipped stone tools made of either local or non-local raw materials,
+presence of polished and beveled stone adzes, grinding slabs, pebbles,
+bone tools and different kinds of personal ornament.
+The personal ornament includes beads made of spondylus shell,
+spondylus pendants or oval shaped buckles.
+Last but not least presence of ochre pigment in the grave is also noted.
 Quantities of various *grave goods* are simplified into a binary matrix indicating
 presence or absence of a given trait.
+
+Another category of variables is derived from retrieved human skeletal remains.
+These include biological sex, age simplified into age groups and indicator
+variable whether the Strontium signal in the remains is of local origin.
+The variable representing the recorded sex is simplified into the following
+categories: female-sexed body (abbreviated *F*), male-sexed body (*M*),
+non adult (*n. a.*) and indetermined sex (*ind.*).
+Age of the body is also simplified into three categories: juveniles (*juv.*)
+comprising of groups Infans and Juvenis, adults (*ad.*, all bodies aged as Adultus)
+and mature bodies (*mat.*, anthropological categories Adultus-Maturus, Maturus and
+Senilis) and bodies of indetermined age (*ind.*).
+Last category is the indicator for the origin of the buried person.
+Local Strontium range is defined as two standard deviations from mean of Sr ranges of
+bodies with age category determined as Infans.
+Bodies faling within this range are labeled as local, outside the range as non-local
+and where data on Sr values are missing origin is not determined (*ind.*).
 
@@
 
-To determine non-randomness of various artefact combinations we conduct a
+To determine non-randomness of various artefact combinations in burials we conduct a
 permutation test on the presence/absence data matrix.
-The method based on generalized Monte Carlo tests was first
+The method based on generalized Monte Carlo tests was
 introduced to archaeology by @manly1996 and adapted by @sosna2008.
@@
 row and column totals, i.e. both the count of artifacts in individual graves
-and sum of the given type of artifacts across the burials are fixed.
+and sum of the given type of artifact across the burials are fixed.
 
@@
 [@manly1996, 475, for details see SI].
-To assess relations between groups based on sex, age and origin,
+To assess relations between groups based on sex, age and origin
 statistics introduced by @sosna2008 are used.
@@
 <!-- spatial analysis -->
-## Spatial patterning
+## Spatial analysis
 
@@
 
-Overall structure of the burial ground can be examined using either density based approaches
-or methods based on distances between neighbors. <!--[@osullivan2010, 121-154]-->
-Density based approaches, for instance quadrat tests,
+Overall structure of the cemeter is examined using
+<!-- either density based approaches or -->
+methods based on distances between neighboring burials. <!--[@osullivan2010, 121-154]-->
+<!-- Density based approaches, for instance quadrat tests,
 characterize the point pattern in terms of its *first-order properties* [@osullivan2010, 126].
 The assumption underlying the *first-order effects* is that any point location is
-independent on other point locations and different parameters influence its location.
-On the other hand the methods based on distances between pairs and triples of neighbours
-characterize the point pattern based on *second- and third-order effects*
+independent on other point locations and different parameters influence its location.-->
+These methods are based on distances between pairs and triples of neighbours and
+characterize the point pattern based on *second- and third-order properties*
 [@nakoinz2016, 130-135; @osullivan2010, 107].
+These are in contrast to the assumption underlying *first-order effects* that
+any point location is independent on other point locations and
+defferent parameters influence its location [@osullivan2010, 126].
 <!-- [@nakoinz2016, 135-144; @osullivan2010, 130-139] -->
-
 In spatial analysis of settlement patterns individual locations of
@@
 In case of a grave placement in a burial ground we presume that the choice of
-a graves location is either random or influenced predominantly by other (previous) grave locations and
-is largely independent on other parameters.
+a graves location is either random or influenced predominantly by other (previous)
+grave locations and
+is largely independent on the outside parameters.
 In other words *first-order properties* are of little to no importance in
 our point of view.
-The *second-order properties*, and possibly *third-order properties*
+The *second-order properties* and possibly *third-order properties*
 are thus considered major factors in locating of the graves.
@@
 is completely spatially random (approximated by a Poisson distribution). -->
-Functions *G*, *F*, Ripley's *K* [@ripley1976; @sayer2013] and *T* [@schladitz2000] are
-based on the nearest neighbour distances and are commonly used
+We utilize *G*, *F*, Ripley's *K* [@ripley1976; @sayer2013] and *T* [@schladitz2000] functions.
+These functions are based on nearest neighbour distances and are commonly used
 to characterize point patterns in archaeology [@nakoinz2016, 135-144].
@@
 certain threshold, not only the nearest neighbours [@nakoinz2016, 138].
-It is widely used in archaeology cf. [@negre2018].
+We find it the most useful in most of archaeology use cases, cf. [@negre2018].
 Last but not least, the *T* function is a third-order parallel to the *K* function [@schladitz2000].
-For formulas and details on individual functions see the supplement.
+For formulas and details on individual functions see the SI.
 
@@
 simulation with 999 iterations.
-Although much smaller number of permutations should be sufficient
-[@baddeley2016, 384] same as @sayer2013 [78] we chose to use 999 permutations.
+Although smaller number of permutations should be sufficient to achieve
+credible results
+[@baddeley2016, 384] we follow the approach taken by @sayer2013 [78].
 Ten most extreme realizations of the simulation are discarded as outliers so
-the significance level $\alpha = 0.01$ [@baddeley2016 410].
+the significance level *alpha* is 0.01 [@baddeley2016 410].
 
-To explore whether neighboring individuals are more often of the same sex we use
-multiple methods to establish neighborhood relations.
-The first two methods are based on constructing neighborhood graphs.
+To explore frequency of neighboring burials falling into same categories we
+explore several methods.
+A modification of *K* function for multitype point patterns
+does not take into account all of the
+neighboring individuals in the point pattern but pairs of observations of the
+same value within the given category are explored.
+Same as *K* function is indicative of the distances at which clustering or
+repulsive processes take place, the multitype *K* function shows these distances
+for given pairs of levels within a given categorical variable.
+The envelopes are constructed using the same procedure as described for other
+functions.
+
+To further establish the neighborhood structure we construct neighborhood graphs.
 We use Delaunay triangulation and its subgraph, a Gabriel graph [@gabriel1969]
 with burial locations as nodes (Fig. \@ref(fig:plan-graphs)).
-The Delaunay triangulation can be considered a maximum number of neighbors case
+The Delaunay triangulation is considered as a maximum number of neighbors per individual case
 scenario while the Gabriel graph the opposite.
@@
 To check how common the observed value is, a permutation test is performed
-where categories are randomly assigned while preserving the overall
+where the categories are randomly assigned while preserving the overall
 proportion of given categories and the neighborhood structure determined by the
 graph.
-The observed values of neighbors categories are compared against distributions
+The observed values of neighbor categories are compared against distributions
 derived from 999 permutations of the simulated data.
@@
 <!-- Percolation analysis -->
-Another approach considered for establishing neigborhood relations is by
+Another approach we consider for establishing neigborhood relations is by
 constructing buffer zones around each of the burials.
-The question arises how large the buffer zones should be.
-To avoid this problem we utilise percolation analysis.
-Percolation analysis is a method used to identify clusters of points based
+To avoid determinism in designating the radius of the buffer zone around each of
+the individuals we utilize percolation analysis.
+Percolation analysis is a data centric method used to identify clusters of points based
 on spatial separation measured by Euclidean distance [@maddison2020, 269].
 The introduction of percolation analysis to archaeology is relatively recent,
-nevertheless its effectivness and usefulness in studies with varied scales
-is already proven [@maddison2020].
+nevertheless we consider its effectiveness and usefulness in studies with varied scales
+well proven [@maddison2020].
 Percolation analysis is based on city clustering algorithm [@rozenfeld2008] and
-its generalization utilizing Euclidean distance [@rozenfeld2011].
+its generalization employing Euclidean distance [@rozenfeld2011].
 Nodes being in a given distance threshold (radius) are assigned into a cluster.
 This process is repeated for a set of thresholds while recording maximum
-cluster sizes and other statistics.
-This allows to identify distances at which clusters exhibit rapid grow (absorbing
+cluster sizes and possibly other statistics.
+This allows to identify distances at which clusters exhibit rapid growth (absorbing
 large number of nodes) and distances at which the clusters remain relatively stable.
 For more details refer to @maddison2020 and associated R package [@schmidt2020].
+
+By means of percolation analysis we identify threshold distances at which cluster
+sizes are growing and clusters formed at these distances are further explored.
+For each cluster number of individual bodies of a given category is recorded.
+The same procedure is repeated for 999 resamples with the given
+categories assigned at random with constant probabilities.
+Observed numbers of occurences deviating by two standard deviations from the mean of the
+resampled results are regarded as strong evidence of non-random patterns being
+present in given clusters.
+To achieve stability of results and eliminate observations occuring due to
+chance in small clusters with low numbers of categories of interest,
+the whole resampling procedure is repeated three times and only observations present in
+all the iterations are reported with expected numbers of occurence and their
+deviations from observed values summarized as mean values.
 
@@
 described for the neighborhood derived from the graphs.-->
-
-Lastly, a modification of *K* and *G* functions for multitype point patterns
-are used to explore the reationship between individual sex categories.
-These take into account not all of the neighbors of the point pattern, but
-count the statistic for pairs of categories in a given variable.
 
@@
 
+<!--
 ## Distance between burials
@@
 variables is treated as nominal data.
+-->
 
@@
 
-The quadrat test shows evidence for both random and clustered processes taking place with p values
+<!-- The quadrat test shows evidence for both random and clustered processes taking place with p values
 under the 0.05 alpha level but the $\chi^2$ approximation is inaccurate due to small
@@
 earlier, we do not expect *first-order* properties to play major roles in
-placing of the individual burials.
+placing of the individual burials. -->
 Results of functions based on nearest neighbors are reported in Fig. \@ref(fig:spat-funs).
-*G* function does deviate from the curve expected under independent random process
-but does not traverse the boundaries of the envelope.
-Thus the cummulative nearest-neighbor distance (*G*) is well approximated by
-random process and does not show evidence for spatial clustering nor repelent functions
+Both *G* and  *F* functions do slightly deviate from the curve expected
+under independent random process
+but neither traverses the boundaries of the envelope.
+Thus the cummulative nearest-neighbor distance (*G*) as well as the
+empty-space function *F* are well approximated by
+random process and do not show evidence for spatial clustering nor repelent functions
 taking place.
-The empty-space function *F* deviates slightly from the area defined by
-the envelope at around 6 meters.
-This means that the number of nearest neighbours to any empty-space point
-at this and larger distances is lower than it would be expected under complete spatial randomness.
-On the contrary to only slight deviations in case of *G* and *F*
-the *K* and *T* functions show clear deviations from the expected curve and exceed
-the boundaries of given envelopes at the distance of around 5 to 6 meters.
+On the contrary
+the *K* and *T* functions show clear deviations from the expected curve and both exceed
+the boundaries of given envelopes at distances around or smaller than 5 meters.
 <!-- in case of *K* and slightly more in case of *T* function.-->
 This means that numbers of nearest neighbors at this and larger distances is
-significantly higher than we would exect under complete spatial randomness.
+significantly higher than we would expect under complete spatial randomness.
 The overall point pattern thus shows clear signs of clustering process taking place.
@@
 but were affected by the positions of previous burials.
-The pattern begins to cluster at the distance of around 5 to 6 meters, meaning
+The pattern begins to cluster at the distance of around 5 meters, meaning
 that beginning at this distance, the probability of finding a neighboring burial is
-much higher than it would be if the burials were placed randomly.
-The community burying their dead at the cemetery of Vedrovice -- Široká u lesa
+much higher than it would be if the burials were placed randomly in space.
+The community burying their dead at the cemetery of Vedrovice *Široká u lesa*
 did so in an organized manner taking into account previous burials at the same place,
@@
 Results of percolation analysis reveal more detailed view of the clustering process
-taking place.
-The curve in Fig. \@ref(fig:perc-distance) reveals that most of the growth into
-a single cluster happens at distances between approximately 4 and 6 meters.
+taking place, see Fig. \@ref(fig:perc-distance).
 The distances where relative stability of clusters emerges are labeled.
-The clusters at these distances are shown in Fig. \@ref(fig:perc-plan).
+The curve reveals a steady growth up until
+radius of 5 meters and then a large skip towards a single cluster
+between 5.5 and 6.5 meters.
+This point of cluster growth is revealed by the point pattern analysis as well.
+The clusters at some of the distances are shown in Fig. \@ref(fig:perc-plan).
 
 ## Neighborhood structure
+
+The results of multitype *K* function show deviations from random processes for
+most of the observed categories in Fig. \@ref(fig:pp-crossk).
+The distances at which these deviations start occuring are in most cases rather large,
+well over 10 meters and considering the fact that at the 10.4 meters radius
+around individual points the percolation analysis proved existence
+of single large cluster encompassing almost all of the points, these
+observations correspond with the deviations observed for single *K* function.
+The only notable deviation we consider worth mentioning is in case of age
+category in pairs of juveniles and in sex category between pairs of non-adults.
+Here the deviation from random processes is observed at smaller distances than
+in case of other variables at distances around 5 meters.
+This proves tighter bound cluster of juveniles in south-eastern part of the cemetery.
+
+In case of neighborhood structure derived from Delaunay triangulation (maximum
+number of neighbors scenario) and
+Gabriel graph (minimum number of neighbors scenario, Fig. \@ref(fig:plan-graphs))
+most of the observed relationships are not significantly different from those
+derived by randomization experiment.
+In the Delaunay triangulation the significant relations are observed in two cases
+between (1) pairs of non-adults and (2) female-sexed bodies and non-adults.
+The first observation can be again given into relation with higher concentration of
+non-adults in south-eastern space of the burial ground.
+In regard to sex category Gabriel graph shows an association
+between female-sexed and male-sexed bodies.
+In only one case are the observations from both of the graphs significant and
+that is between pairs of bodies of mature age category.
+
+In case of clusters at different percolation distances
+there are some deviations from random number of
+occurences of given categories at different scales.
+Regarding the determination of sex of the bodies, at the threshold of 3.9 meters
+we observe that in the largest cluster (nr. 3, Fig. \@ref(fig:perc-plan))
+there are more non adults and less male-sexed bodies and in the second largest
+cluster (nr. 10) there are slightly more male-sexed bodies than would
+be expected under random process.
+The overrepresentation of non-adult bodies is repeated in the following
+threshold distance of 4.6 meters in the largest cluster (nr. 3).
+Last but not least, in the cluster nr. 8 (north of the cemetery) at the distance
+of 5.5 meters female-sexed bodies are more frequent than we would expect.
+The age categories correspond to the previous obsevations where
+most evident is the overrepresentaion of juvenile bodies at the 3.9 meters threshold in
+cluster nr. 3.
+Exploring the origin of the bodies based on Strontium isotope signature we
+find that cluster in the south-western part at threshold distances 3.9 and 5.5 meters
+(nr. 10 and 5 respectively) consists of less non-local bodies than would be expected.
+Lastly at the distance radius 2.8 meters there are more local bodies in the cenral
+cluster of the cemetery (nr. 8).
 
@@
 maximum number of neighbors scenario and a Gabriel graph as a minimal scenario. -->
-
+<!--
 For relationships (see Fig. \@ref(fig:nb-sex) and Tab. \@ref(tab:nb-tab-sex))
@@
 indeterminate category $p \approx 0.02$.
+-->
 

### 2020-09-25 code/common_art_cat.R
archived unneccessary scripts, renaming acc. to conventions

@@ -1,74 +1,0 @@
-# Project "Vedrovice"
-# Script nr. 6
-# CREATE GROUPS BASED ON COMMON ARTEFACTS
-# author: Petr Pajdla
-# Create groups based on common artefacts
-
-# pckg
-library(tidyverse)
-
-# data
-ved <- read_rds("./data/vedrovice_dataset.RDS")
-
-non_rand_vars <- read_lines("./data/output/non_random_vars.txt")
-
-ei <- read_csv("./data/output/exceptionality.csv")
-
-# experiment
-
-ved_bin_dist_nr <- dist(ved$bin_vars$bin_mat[, non_rand_vars], method = "binary")
-ved_bin_dist <- dist(ved$bin_vars$bin_mat, method = "binary")
-
-ved_hclust_nr <- hclust(ved_bin_dist_nr, method = "ward.D2")
-ved_hclust <- hclust(ved_bin_dist, method = "ward.D2")
-
-par(mfcol = c(1, 2))
-plot(ved_hclust)
-plot(ved_hclust_nr)
-
-groups_common_items <- cutree(ved_hclust_nr, k = 6)
-
-ved_gathered <- bind_cols(id = ved$metadata$id_burial, 
-                          as_tibble(ved$bin_vars$bin_mat), 
-                          gr_item = groups_common_items,
-                          gr_ei = ei$group) %>% 
-  gather(-gr_item, -gr_ei, -id, key = "item", value = "presence")
-
-# ad hoc function
-prop_item <- function(df) {
-  temp_total <- df %>% group_by(item) %>% 
-    summarise(tot = sum(presence))
-  temp_groups <- df %>% group_by(gr_item, item) %>% 
-    summarise(sum = sum(presence)) %>% 
-    group_map(~ mutate(.x, prop = sum/temp_total$tot))
-  return(temp_groups)
-}
-
-item_prop <- prop_item(ved_gathered) %>% 
-  bind_rows(.id = "gr_item") %>% 
-  mutate(gr_item = str_c("group", gr_item))
-
-# randomly distributed items
-ggplot(item_prop, aes(y = prop, x = item)) +
-  geom_col(fill = "white", color = "black") +
-  facet_wrap(~ gr_item, nrow = 1) +
-  coord_flip() +
-  labs(x = "artefact category", y = "proportion")
-
-item_prop %>% group_by(item) %>% summarise(sumx = sum(prop))
-
-# non-randomly distributed items
[…]

### 2020-09-25 code/exceptionality_explore.R

@@ -1,69 +1,0 @@
-# Project "Vedrovice"
-# Script nr. 
-# EXPLORE EXCEPTIONALITY
-# author: Petr Pajdla
-# Explore exceptionality of burials, what does it cause etc...
-
-# libraries
-library(tidyverse)
-library(here)
-
-# data 
-base <- read_rds(here("data/temp", "vedrovice_dataset.RDS"))
-
-ei <- read_csv(here("data/temp", "exceptionality.csv"), col_types = "cdff")
-
-ei_fct_order <- ei %>% group_by(fct) %>% 
-  summarise(mean_ei = mean(ei))
-
-v_vals <- read_csv(here("data/temp", "v_statistics.csv"))
-
-
-# grave goods in various ei  ----------------------------------------------
-ved_gg <- left_join(ei, 
-                    as_tibble(base$bin_vars$count_mat[, base$bin_vars$over5], 
-                              rownames = "burial"),
-                    by = c("burial")) %>% 
-  select(-ei, -ei_cluster)
-
-# proportion of grave goods in a given group
-gg_prop <- function(df, id, group) {
-  temp <- df %>% pivot_longer(-c(!!sym(id), !!sym(group)), 
-                              names_to = "gg", 
-                              values_to = "presence")
-  temp_sums <- temp %>% group_by(gg) %>% 
-    summarise(total = sum(presence)) %>% 
-    ungroup()
-  temp %>% group_by(!!sym(group), gg) %>% 
-    summarise(sum = sum(presence)) %>% 
-    group_map(~ mutate(.x, prop = sum/temp_sums$total), .keep = TRUE) %>% 
-    bind_rows()
-}
-
-# reordering factors
-ved_gg_prop <- gg_prop(ved_gg, "burial", "fct") %>% 
-  left_join(ei_fct_order) %>% 
-  mutate(fct = fct_reorder(fct, desc(mean_ei)),
-         gg_order = as.integer(fct_relevel(gg, v_vals$variable)),
-         gg_long = unname(base$var_names$long[gg]),
-         gg_long = fct_reorder(gg_long, desc(gg_order))) %>% 
-  select(-mean_ei)
-
-
-# plotting
-ved_gg_prop %>% ggplot(aes(y = prop, x = gg_long)) +
-  geom_col(fill = "white", color = "black") +
-  facet_wrap(vars(fct), nrow = 1) +
-  coord_flip() +
[…]

### 2020-09-25 code/exceptionality_index.R

@@ -1,145 +1,0 @@
-# Project "Vedrovice"
-# Script nr. 2.2
-# EXCEPTIONAITY INDEX FOR BURIALS
-# author: Petr Pajdla
-# Exceptionality index of individual burials in order to create groups...
-
-# packages =====================================================================
-library(tidyverse)
-library(ggdendro)
-
-# functions ====================================================================
-# root mean square (quadratic mean) for vector x
-rms <- function(x) {
-  x <- sqrt(mean(x^2))
-  return(x)
-}
-
-# # scaling
-# normalize01 <- function(x) {
-#   rg <- range(x, na.rm = TRUE)
-#   (x - rg[1]) / (rg[2] - rg[1])
-# }
-
-# data =========================================================================
-ved <- read_rds(here::here("data/temp/vedrovice_dataset.RDS"))
-# non_rand_vars <- readLines(here("data/temp", "non_random_vars.txt"))
-v_vals <- read_csv(here::here("data/temp/v_statistics.csv")) %>% 
-  select(variable, p) %>% 
-  mutate(weight = 1 - p)
-
-var_weights <- v_vals %>% pull(weight)
-names(var_weights) <- v_vals %>% pull(variable)
-
-ved_in <- ved$bin_vars$bin_mat[, ved$bin_vars$over5]
-
-# weighting variables -----------------------------------------------------
-# variables are weighted by the p value (measure of departure from norm) of v stats
-# m_cont <- matrix(nrow = nrow(ved$bin_vars$count_mat), 
-#                  ncol = ncol(ved$bin_vars$count_mat))
-
-m_bin <- matrix(nrow = nrow(ved_in), 
-                ncol = ncol(ved_in))
-
-# for (i in 1:ncol(m_cont)) {
-#   m_cont[, i] <- ved$bin_vars$count_mat[, i] * var_weights[i]
-# }
-
-for (i in 1:ncol(m_bin)) {
-  m_bin[, i] <- ved_in[, i] * var_weights[i]
-}
-
-# colnames(m_cont) <- colnames(ved$bin_vars$count_mat)
-colnames(m_bin) <- colnames(ved_in)
-
-
-# exceptionality index =========================================================
[…]

### 2020-09-25 code/exp_mixsiar.R

@@ -1,47 +1,0 @@
-library(tidyverse)
-library(MixSIAR)
-
-x <- read_rds("./data/vedrovice_dataset.RDS")
-
-write_csv(bind_cols(x$metadata[, c("id_burial", "sex")],
-                    x$cont_vars$cont_vars[, c("d13c", "d15n")]), 
-            "./data/output/mixing.csv")
-          
-
-source_data <- tribble(
-  ~spec, ~Meand13c, ~SDd13c, ~Meand15n, ~SDd15n, ~n,
-  "cattle",-20.2,  0.29, 6.2,    1.04, 3,
-  "sheep", -19.8,  0.18, 5.9,    0.38, 3,
-  "pig",   -20.4,  0.53, 8.2,    0.86, 2,
-  "fishy", -18, 0.3, 13, 2.5, 2
-)
-
-write_csv(source_data, "./data/output/source_data.csv")
-
-discr <- tribble(
-  ~spec, ~Meand13c, ~SDd13c, ~Meand15n, ~SDd15n,
-  "cattle", 0, 0, 0, 0,
-  "sheep", 0, 0, 0, 0,
-  "pig", 0, 0, 0, 0,
-  "fishy", 0, 0, 0, 0
-)
-
-write_csv(discr, "./data/output/discr_data.csv")
-
-mix <- load_mix_data(filename = "./data/output/mixing.csv", 
-              iso_names = c("d13c", "d15n"), factors = c("sex"),
-              fac_random = FALSE, fac_nested = FALSE, cont_effects = NULL)
-
-srcs <- load_source_data("./data/output/source_data.csv", source_factors = NULL,
-                        conc_dep = FALSE, data_type = "means", mix = mix)
-
-discr <- load_discr_data("./data/output/discr_data.csv", mix = mix)
-
-plot_data(mix = mix, source = srcs, discr = discr)
-calc_area(srcs, mix, discr)
-plot_prior(alpha.prior = 1, srcs)
-write_JAGS_model("./data/output/mixsiar_model.txt", resid_err = TRUE, process_err = FALSE, mix = mix, source = srcs)
-
-jags.1 <- run_model(run="test", mix, srcs, discr, "./data/output/mixsiar_model.txt", 
-                    alpha.prior = 1, TRUE, FALSE)
-

### 2020-09-25 code/exp_simmr.R

@@ -1,79 +1,0 @@
-library(tidyverse)
-library(simmr)
-
-# vignette("quick_start", package = "simmr")
-# vignette("simmr", package = "simmr")
-
-# load data - vedrovice
-x <- read_rds("./data/vedrovice_dataset.RDS")
-
-# prepare data (matrix, grouping factor)
-mix_in <- cbind(x$metadata$sex, x$cont_vars$cont_vars[, c("d13c", "d15n")]) %>% 
-  drop_na()
-
-mix <- as.matrix(mix_in[, 2:3])
-grp <- mix_in$`x$metadata$sex`
-
-# create sources data (Bickle, Whittle 2013, p. 121)
-source_data <- tribble(
-  ~spec, ~Meand13c, ~SDd13c, ~Meand15n, ~SDd15n, ~n,
-  "cattle",-20.2,  0.29, 6.2,    1.04, 3,
-  "sheep", -19.8,  0.18, 5.9,    0.38, 3,
-  "pig",   -20.4,  0.53, 8.2,    0.86, 2
-)
-
-s_names <- source_data$spec
-s_means <- source_data[, c("Meand13c", "Meand15n")]
-s_sd <- source_data[, c("SDd13c", "SDd15n")]
-
-# create simmr object
-simmr_obj <- simmr_load(mix, s_names, s_means, s_sd, group = grp)
-
-# overall plot
-plot(simmr_obj, group = 1:3)
-
-# fit model
-simmr_out <- simmr_mcmc(simmr_obj)
-
-# validate - values should be close to 1
-summary(simmr_out, type='diagnostics', group = 1:3)
-
-# blue y values should fall within light blue y_rep intervals
-posterior_predictive(simmr_out, group = 1)
-posterior_predictive(simmr_out, group = 2)
-posterior_predictive(simmr_out, group = 3)
-
-# prior prob. distr. viz
-prior_viz(simmr_out, group = 1)
-prior_viz(simmr_out, group = 2)
-prior_viz(simmr_out, group = 3)
-
-summary(simmr_out, type='statistics', group = 1:3)
-summary(simmr_out, type='quantiles', group = 1:3)
-
-# plots
-plot(simmr_out, type = 'density', group = 1)
-plot(simmr_out, type = 'density', group = 2)
-plot(simmr_out, type = 'density', group = 3)
-
-# matrices
-# large  negative correlations - model is not able to differentiate the two groups
-plot(simmr_out, type = 'matrix', group = 1)
[…]

### 2020-09-25 code/local.R

@@ -1,82 +1,0 @@
-# Project "Vedrovice"
-# Script nr. 1.2
-# LOCAL AND NON-LOCAL Sr RANGES
-# author: Petr Pajdla
-# Simplify and clean age groups 
-# Assign bodies into local vs non-local based on Sr range
-
-library(tidyverse)
-library(here)
-
-
-# data --------------------------------------------------------------------
-
-ved <- read_rds(here("data/temp/vedrovice_dataset.RDS"))
-
-
-# edit age ----------------------------------------------------------------
-
-age <- ved$orig_dataset %>% 
-  select(id_burial, age_interval) %>% 
-  mutate(age = if_else(age_interval == "Adult", "20-40", age_interval),
-         age = if_else(age_interval == "±1", "0-2", age),
-         age = if_else(age_interval == "20+", "20-30", age),
-         age = if_else(age_interval == "50+", "50-60", age),
-         age = if_else(age_interval == "60+", "60-70", age),
-         age = if_else(age_interval == "9", "8-10", age),
-         age = if_else(age_interval == "30", "25-35", age),
-         age = if_else(age_interval == "20", "15-25", age),
-         age_min = as.numeric(str_extract(age, "^[:digit:]*(?=-|–)")),
-         age_max = as.numeric(str_extract(age, "(?<=-|–)[:digit:]*$")),
-         id_burial = as.character(id_burial)) %>% 
-  select(-age_interval, -age)
-
-
-# group by age ------------------------------------------------------------
-
-ved_sr <- ved$metadata %>% 
-  select(id_burial, sex, age_cat) %>% 
-  left_join(tibble(rownames_to_column(ved$cont_vars)), 
-            by = c("id_burial" = "rowname")) %>% 
-  left_join(age) %>% 
-  mutate(age_mean = (age_min + age_max) / 2,
-         group = if_else(age_mean < 18, "juvenile", as.character(sex))
-         ) %>% 
-  filter(!is.na(group), !group %in% c("n. a.", "ind."))
-
-local <- ved_sr %>% 
-  filter(group == "juvenile",
-         !is.na(sr)) %>%
-  summarise(n(), 
-            mean = mean(sr, na.rm = TRUE), 
-            sd = sd(sr, na.rm = TRUE)) %>% 
-  mutate(min = mean - 2 * sd,
-         max = mean + 2 * sd)
-
-
-# local range of sr -------------------------------------------------------
-
[…]

### 2020-09-25 code/recoding_vars.R

@@ -1,26 +1,0 @@
-# recoding data, recode and reimport to spreadsheet...
-
-input <- read_csv("./input/data_vedrovice.csv")
-
-# recoding head_orient =========================================================
-# output <- input %>% transmute(id_burial, 
-#                     head_orient_new = ifelse(input$head_orient > 90 - (45/2) & input$head_orient < 90 + (45/2), "E", 
-#                                           ifelse(input$head_orient > 135 - (45/2) & input$head_orient < 135 + (45/2), "SE", 
-#                                                  ifelse(input$head_orient > 180 - (45/2) & input$head_orient < 180 + (45/2), "S", 
-#                                                         ifelse(input$head_orient > 315 - (45/2) - 2 & input$head_orient < 315 + (45/2), "NW", "NOPE")))))
-# 
-# write_csv(output, "./output/head_orient.csv")
-
-# recoding pit_orient ==========================================================
-# # CHECK DATA !!! 
-# input %>% group_by(pit_orient_cat) %>% summarise(min = min(pit_orient, na.rm = T), max = max(pit_orient, na.rm = T), n())
-# hist(input$pit_orient)
-# ggplot(input, aes(pit_orient, factor(pit_orient_cat))) + geom_jitter()
-# ggplot(input, aes(pit_orient)) + geom_histogram(binwidth = 2)
-
-# binarizing values ============================================================
-
-# binary <- biclust::binarize(input[, 14:36], threshold = 0)
-# 
-# write_csv(binary, "./output/binary.csv")
-

### 2020-09-25 code/spat_functions.R

@@ -33,7 +33,8 @@ theme_universe <- theme(panel.border = element_rect(colour = "black",
 # read data and layouts --------------------------------------------------------
 ved <- read_rds(here("data/temp", "vedrovice_dataset.RDS"))
 
-ved_sf <- read_sf(here("data/temp", "layout.shp"))
+ved_sf <- read_sf(here("data/temp", "layout.shp")) %>% 
+  filter(pres != "dist.")
 ved_exc <- read_sf(here("data/temp", "window.shp")) %>% 
   as.owin(ved_exc)
 
@@ -42,10 +43,12 @@ ved_sf <- bind_cols(ved_sf,
                     y = st_coordinates(ved_sf)[, 2])
 
 # marks ------------------------------------------------------------------------
-ved_marks <- ved$metadata %>% 
-  select(sex) %>% 
-  mutate(sex = if_else(is.na(sex), "ind.", as.character(sex)),
-         sex = as_factor(sex))
+ved_marks <- ved_sf %>% 
+  select(sex, age = age_sim, origin) %>% 
+  st_drop_geometry() %>% 
+  mutate(sex = factor(sex, levels = c("n. a.", "F", "M", "ind.")),
+         age = factor(age, levels = c("juv.", "ad.", "mat.", "ind.")),
+         origin = factor(origin, levels = c("local", "non-local", "ind.")))
 
 # create ppp object ------------------------------------------------------------
 # ved_pp <- ppp(x = ved_layout$layout_x, y = ved_layout$layout_y,
@@ -55,7 +58,7 @@ ved_pp <- ppp(x = ved_sf$x, y = ved_sf$y,
               window = ved_exc,
               marks = ved_marks)
 
-plot(ved_pp, use.marks = FALSE)
+# plot(ved_pp, use.marks = FALSE)
 
 # point pattern analysis
 # functions ====================================================================
@@ -186,115 +189,115 @@ grid_fns <- bind_rows("G(r)" = ved_g,
 
 grid_fns
 
-ggsave(here("plots", "pointprocess_fun.pdf"), width = 12, height = 3)
+ggsave(here("plots", "pp_funs.pdf"), width = 12, height = 3)
 
[…]

### 2020-09-25 code/spat_neighbors.R

@@ -33,17 +33,15 @@ theme_universe <- theme(panel.border = element_rect(colour = "black",
 #   select(id_burial = burial, ei, ei_clust = fct)
 ved_sf <- read_sf(here("data/temp", "layout.shp")) %>% 
   mutate(sex = fct_relevel(sex, c("n. a.", "F", "M", "ind.")),
-         age_sim = if_else(is.na(age_sim), "ind.", age_sim),
-         age_sim = fct_relevel(age_sim, c("you", "mid", "old", "ind.")),
-         origin = if_else(is.na(origin), "ind.", origin),
-         origin = fct_relevel(origin, c("local", "non-local")),
-         ei_clust = if_else(is.na(ei_clust), "ind.", as.character(ei_clust)),
-         ei_clust = factor(ei_clust, 
-                           levels = c(letters[1:length(unique(.$ei_clust))-1], 
-                                      "ind.")))
-ved_layout <- st_coordinates(ved_sf)
+         age_sim = fct_relevel(age_sim, c("juv.", "ad.", "mat.", "ind.")),
+         origin = fct_relevel(origin, c("local", "non-local", "ind.")))
+ved_layout <- ved_sf %>% 
+  filter(pres != "dist.") %>% 
+  st_coordinates()
 ved_exc <- read_sf(here("data/temp", "window.shp"))
 
+ved_sf_pres <- filter(ved_sf, pres != "dist.") %>% 
+  mutate(age_sim = forcats::fct_drop(age_sim))
 
 # create delaunay + gabriel neighborhoods ---------------------------------
 ved_gabriel <- gabrielneigh(ved_layout)
@@ -51,7 +49,7 @@ ved_delaunay <- tri2nb(ved_layout)
 
 # plot delaunay triangulation and gabriel graph ---------------------------
 ved_gabriel_lines <- ved_gabriel %>% 
-  graph2nb(row.names = rownames(ved_sf)) %>% 
+  graph2nb(row.names = ved_sf_pres$id_burial) %>% 
   nb2lines(coords = ved_layout) %>% 
   st_as_sf()
 
@@ -63,8 +61,8 @@ g_gabriel <- ggplot() +
   geom_sf(data = ved_exc, fill = NA, color = "gray90", size = 4) +
   # geom_sf(data = ved_delaunay_lines, linetype = 1, size = 0.2) +
[…]

### 2020-09-25 code/spat_perc.R

@@ -17,9 +17,13 @@
 library(igraph)
 library(dplyr)
 library(tidyr)
+library(purrr)
 library(ggplot2)
 library(ggforce)
 library(ggspatial)
+library(concaveman)
+library(sf)
+library(forcats)
 
 # theme for ggplot graphics
 theme_universe <- theme(panel.border = element_rect(colour = "black", 
@@ -36,8 +40,13 @@ theme_universe <- theme(panel.border = element_rect(colour = "black",
 # data --------------------------------------------------------------------
 
 ved_sf <- sf::read_sf(here::here("data/temp", "layout.shp")) %>% 
-  mutate(sex = forcats::fct_relevel(sex, c("n. a.", "F", "M", "ind.")))
-ved_coord <- sf::st_coordinates(ved_sf)
+  mutate(sex = fct_relevel(sex, c("n. a.", "F", "M", "ind.")),
+         age_sim = fct_relevel(age_sim, c("juv.", "ad.", "mat.", "ind.")),
+         origin = fct_relevel(origin, c("local", "non-local", "ind."))) %>% 
+  filter(pres != "dist.")
+ved_coord <- bind_cols(id = as.character(ved_sf$id_burial), 
+                       sf::st_coordinates(ved_sf)) %>% 
+  rename(x = X, y = Y)
 ved_exc <- sf::read_sf(here::here("data/temp", "window.shp"))
 
 # experiments -------------------------------------------------------------
@@ -63,21 +72,23 @@ ved_exc <- sf::read_sf(here::here("data/temp", "window.shp"))
 # percolation -------------------------------------------------------------
 
 percolation <- function(x, lower = NULL, upper = NULL, step = NULL) {
-  stopifnot(is.matrix(x))
+  # stopifnot(is.matrix(x))
+  m <- matrix(c(x$x, x$y), ncol = 2)
+  rownames(m) <- x$id
   
-  dist <- as.matrix(stats::dist(x, method = "euclidean"))
+  dist <- as.matrix(stats::dist(m, method = "euclidean"))
   
   # radius
-  triangle <- dist[lower.tri(dist)]
-  if (is.null(lower)) {
-    lower <- min(triangle)
-  }
-  if (is.null(upper)) {
-    upper <- max(triangle) / 4
-  }
-  if (is.null(step)) {
-    step <- (upper - lower) / 10
-  }
+  # triangle <- dist[lower.tri(dist)]
+  # if (is.null(lower)) {
+  #   lower <- min(triangle)
+  # }
[…]

### 2020-09-25 code/spat_perc_clusters.R

@@ -1,0 +1,175 @@
+# Project "Vedrovice"
+# Script nr. 
+# SPATIAL ARRANGEMENT - Grouping of vars into clusters derived from percolation
+# author: Petr Pajdla
+# Identify deviations from random for different percolation clusters 
+
+set.seed(42)
+
+# number of simulations
+nsim <- 999
+
+# libs --------------------------------------------------------------------
+
+library(tidyverse)
+library(sf)
+
+
+# data --------------------------------------------------------------------
+
+explored_rad <- paste0(c(2.8, 3.9, 4.6, 5.5, 6.5, 7.8))
+
+perc_cl <- read_csv(here::here("data/temp", "perc_clusters.csv")) %>% 
+  mutate(id_burial = as.character(id)) %>% 
+  select(-id) %>% 
+  filter(radius %in% explored_rad)
+
+ved_sf <- sf::st_read(here::here("data/temp", "layout.shp")) %>% 
+  mutate(sex = fct_relevel(sex, c("n. a.", "F", "M", "ind.")),
+         age_sim = fct_relevel(age_sim, c("juv.", "ad.", "mat.", "ind.")),
+         origin = fct_relevel(origin, c("local", "non-local", "ind."))) %>% 
+  filter(pres != "dist.")
+
+# prep ---------------------------------------------------------------------
+
+input <- ved_sf %>% select(id_burial, sex, age_sim, origin) %>% 
+  st_drop_geometry() %>% 
+  left_join(perc_cl, by = c("id_burial"))
+
+
+# functions ---------------------------------------------------------------
+
+count_var_per_radius <- function(input, variable) {
+  res <- input %>% 
+    filter(!is.na(radius), !is.na(cluster)) %>%
+    group_by(radius) %>% 
+    nest() %>% 
+    mutate(data = map(data, group_by, cluster, !!sym(variable)),
+           data = map(data, count)) %>% 
+    unnest(data) %>% 
+    ungroup(radius) %>% 
+    setNames(c("radius", "cluster", "variable", "n"))
+  return(res)
+}
+
+randomize_var_per_radius <- function(input, variable, n_sim) {
+  n_bur <- nrow(input)
+  variable0 <- as.symbol(variable)
+  
+  prob <- input %>%
+    select(!!variable0) %>%
+    group_by(!!variable0) %>%
+    count() %>%
+    ungroup(!!variable0) %>% 
[…]

### 2020-09-25 code/spatial_kde.R

@@ -1,98 +1,0 @@
-# Project "Vedrovice"
-# Script nr. 3.2
-# SPATIAL ARRANGEMENT: Kernel density estimations
-# author: Petr Pajdla
-# KDE
-
-library(here)
-library(sf)
-# library(raster)
-library(tidyverse)
-
-# data --------------------------------------------------------------------
-
-ved_sf <- read_sf(here("data/temp/", "layout.shp")) %>% 
-  mutate(sex = fct_relevel(sex, c("n. a.", "F", "M", "ind.")))
-ved_exc <- read_sf(here("data/temp/", "window.shp"))
-
-# kernel density est. -----------------------------------------------------
-
-get_kde <- function(x, win = NA, variable = NA, value = NA) {
-  if (any(is.na(win))) {
-    win <- x
-  }
-  ncells <- nrow(x) * 2
-  if (!is.na(variable) & !is.na(value)) {
-    x <- filter(x, !!as.symbol(variable) == value)
-  }
-  coord <- st_coordinates(x)
-  bbox <- st_bbox(win)
-  bbox <- c(bbox[1], bbox[3], bbox[2], bbox[4])
-  kde <- MASS::kde2d(coord[, 1], coord[, 2], lims = bbox, n = ncells)
-  raster::raster(kde)
-}
-
-# plot(get_kde(ved_sf, ved_exc, "sex", "M"))
-# plot(get_kde(ved_sf, ved_exc, "sex", "F"))
-# plot(get_kde(ved_sf, ved_exc, "sex", "n. a."))
-# plot(get_kde(ved_sf, ved_exc, "sex", "ind."))
-# 
-# plot(cut(get_kde(ved_sf, ved_exc, "sex", "M"), breaks = 4))
-# 
-# contour(get_kde(ved_sf, ved_exc, "sex", "M"))
-# contour(get_kde(ved_sf, ved_exc, "sex", "F"))
-
-# raster to tibble for computations...
-# as_tibble(rasterToPoints(get_kde(ved_sf, ved_exc, "sex", "M")))
-
-# plot kde ----------------------------------------------------------------
-
-ved <- bind_cols(ved_sf, 
-                 x = st_coordinates(ved_sf)[, 1],
-                 y = st_coordinates(ved_sf)[, 2])
-
-ved_bg <- ved %>% transmute(s = sex, x, y)
-
-ved_fg <- ved %>% filter(sex != "ind.")
-
[…]

### 2020-09-25 code/dataset.R
unify data preparation tasks

@@ -75,39 +75,44 @@ ved$bin_vars$over5 <- (colSums(ved$bin_vars$bin_mat) / n_burs) > 0.05
 # continuous variables
 ved$cont_vars <- input %>% 
   filter(undisturbed == TRUE) %>% 
-  select(pit_len, pit_wid, pit_dep,
-         d13c, d15n, sr, sr_ppm, 
-         body_height) %>% 
+  select(
+    # pit_len, pit_wid, pit_dep,
+    # d13c, d15n, 
+    sr, sr_ppm, 
+    # body_height
+  ) %>% 
   as.data.frame()
 
 rownames(ved$cont_vars) <- ved$id_burials
 
 # factor variables
-ved$cat_vars <- input %>% 
-  filter(undisturbed == TRUE) %>% 
-  select(pit_orient_cat, head_orient_cat, body_side) %>%
-  mutate_all(factor) %>% 
-  as.data.frame()
-
-rownames(ved$cat_vars) <- ved$id_burials
+# ved$cat_vars <- input %>% 
+#   filter(undisturbed == TRUE) %>% 
+#   select(pit_orient_cat, head_orient_cat, body_side) %>%
+#   mutate_all(factor) %>% 
+#   as.data.frame()
+# 
+# rownames(ved$cat_vars) <- ved$id_burials
 
 # metadata
-lookup_sex <- c("n. a." = "gold", "F" = "tomato", "M" = "steelblue")
+# lookup_sex <- c("n. a." = "gold", "F" = "tomato", "M" = "steelblue")
 
-ved$metadata <- input %>% select(id_burial, sex, age_cat, dat) %>% 
+ved$metadata <- input %>% 
+  select(id_burial, pres = undisturbed, sex, age_cat, dat) %>% 
   mutate(id_burial = as.character(id_burial),
+         pres = if_else(pres, "pres.", "dist."),
          sex = factor(sex),
-         sex_col = unname(lookup_sex[input$sex]),
+         # sex_col = unname(lookup_sex[input$sex]),
          age_cat = ordered(age_cat, 
                            levels = c("I1", "I", "I2", "J",
                                       "A1", "A", "A2", "AM", 
                                       "M1", "M", "M2", "S")),
-         age_sim = ordered(if_else(age_cat %in% c("I1", "I", "I2", "J"), "you",
-                                   if_else(age_cat %in% c("A1", "A", "A2"), "mid",
[…]

### 2020-09-25 code/spat_dataprep.R

@@ -11,12 +11,11 @@ library(tidyverse)
 
 # read data
 ved <- read_rds(here("data/temp", "vedrovice_dataset.RDS"))
-origin <- read_rds(here("data/temp/local.rds")) %>% 
-  select(id_burial, age_mean, origin)
-ei <- read_csv(here("data/temp", "exceptionality.csv")) %>% 
-  mutate(ei_clust = factor(fct),
-         id_burial = as.character(burial)) %>% 
-  select(-burial, -ei_cluster, -fct)
+
+# ei <- read_csv(here("data/temp", "exceptionality.csv")) %>% 
+#   mutate(ei_clust = factor(fct),
+#          id_burial = as.character(burial)) %>% 
+#   select(-burial, -ei_cluster, -fct)
 
 # data prep ====================================================================
 scale_m <- 4.762 # scale by a factor so 1 unit is 1 meter
@@ -30,12 +29,21 @@ ved_layout <- tibble(layout_x = ved_layout$x, layout_y = ved_layout$y)
 # not a rectangle, but a polygon closely bounding the burials using convex hull
 ved_sf <- st_as_sf(bind_cols(ved_layout, ved$metadata), 
                    coords = c("layout_x", "layout_y")) %>% 
-  left_join(origin, by = "id_burial") %>% 
-  mutate(sex = fct_relevel(sex, c("n. a.", "F", "M", "ind."))) %>% 
-  left_join(ei, by = "id_burial")
+  # left_join(ei, by = "id_burial") %>% 
+  mutate(sex = fct_relevel(sex, c("n. a.", "F", "M", "ind.")),
+         # age_sim = if_else(is.na(age_sim), "ind.", as.character(age_sim)),
+         age_sim = fct_relevel(age_sim, c("juv.", "ad.", "mat.", "ind.")),
+         # origin = if_else(is.na(origin), "ind.", origin),
+         origin = fct_relevel(origin, c("local", "non-local", "ind.")),
+         # ei_clust = if_else(is.na(ei_clust), "ind.", as.character(ei_clust)),
+         # ei_clust = factor(ei_clust,
+         #                   levels = c(letters[1:length(unique(.$ei_clust))-1],
+         #                              "ind."))
+  )
+
 # ved_sf_buffer <- st_buffer(ved_sf, dist = 1.6)
 # ved_sf_buffer <- st_simplify(ved_sf_buffer, dTolerance = 1)
[…]

### 2020-09-25 text/article_vedrovice.Rmd
correct figures

@@
 <!-- The size of the radius is based on the results of nearest neighbors analysis, where
-at the given radius there is strong evidence of a clustering process (Fig. \@ref(fig:spat-funs)).
+at the given radius there is strong evidence of a clustering process (Fig. \@ref(fig:pp-funs)).
 The randomness is again tested using a permutation test similar to the one
@@
 placing of the individual burials. -->
-Results of functions based on nearest neighbors are reported in Fig. \@ref(fig:spat-funs).
+Results of functions based on nearest neighbors are reported in Fig. \@ref(fig:pp-funs).
 Both *G* and  *F* functions do slightly deviate from the curve expected
@@
 
+<!-- randomization -->
+<!--
 ```{r sstat, fig.cap = "Distribution of simulated and observed *S* statistic"}
-knitr::include_graphics(here("plots/s_statistic.pdf"))
+# knitr::include_graphics(here("plots/s_statistic.pdf"))
 ```
@@
 ```{r vstat, fig.cap = "Simulated and observed *V* statistics for different variables"}
-knitr::include_graphics(here("plots", "v_values.pdf"))
+# knitr::include_graphics(here("plots", "v_values.pdf"))
 ```
@@
 ```{r coocnets, fig.cap = "Networks of variables frequently present together and exluding each other"}
-knitr::include_graphics(here("plots", "cooc_networks.pdf"))
+# knitr::include_graphics(here("plots", "cooc_networks.pdf"))
+```
+-->
+
+<!-- plans -->
+
+<!-- point pattern analysis -->
+
+```{r label-pp-funs}
+label_pp_funs <- paste("Functions *G*, *F*, *K* and *T*.",
+"*Dashed line* -- theoretical curve under complete spatial randomness, *solid line* -- empirical curve, *gray area* -- envelope")
+```
+```{r pp-funs, fig.cap=label_pp_funs}
+knitr::include_graphics(here("text/figs/pp_funs.pdf"))
 ```
 
-<!-- ```{r plan-graphs, fig.cap = "Neighborhood graphs: Delaunay triangulation and Gabriel graph"} -->
-<!-- knitr::include_graphics(here("plots/plan_gabriel.pdf")) -->
-<!-- ``` -->
+```{r label-pp-crossk}
+label_crossk <- paste("Multitype K function for different categorical variables, for legend see previous Fig.")
+```
+```{r pp-crossk, fig.cap=label_crossk}
+knitr::include_graphics(here("text/figs/pp_crossk.pdf"))
+```
 
-<!-- ```{r label-spat-funs} -->
-<!-- label_spat_funs <- paste("Functions *G*, *F*, *K* and *T*.", -->
-<!-- "*Dashed line* -- theoretical curve under complete spatial randomness, *solid line* -- empirical curve, *gray area* -- envelope.") -->
-<!-- ``` -->
-<!-- ```{r spat-funs, fig.cap=label_spat_funs} -->
-<!-- knitr::include_graphics(here("plots/pointprocess_fun.pdf")) -->
-<!-- ``` -->
+<!-- neighborhood -->
+```{r plan-graphs, fig.cap="Neighborhood graphs: Delaunay triangulation and Gabriel graph"}
+knitr::include_graphics(here("text/figs/plan_graphs.pdf"))
+```
 
-<!-- ```{r label-nb-sex} -->
-<!-- label_nb_sex <- paste("Mean neighbors in a given sex category.", -->
-<!-- "*Gray area* -- density estimation for the simulated data, *vertical line* -- observed value.", -->
-<!-- "*F* -- female-sexed body, *M* -- male-sexed body, *n. a.* -- non adult, *ind.* -- indetermined sex.") -->
-<!-- ``` -->
-<!-- ```{r nb-sex, fig.cap=label_nb_sex} -->
-<!-- knitr::include_graphics(here("plots/nb_sex.pdf")) -->
-<!-- ``` -->
+<!-- to SI
+```{r label-nb-sex}
+label_nb_sex <- paste("Mean neighbors in a given sex category.",
+"*Gray area* -- density estimation for the simulated data, *vertical line* -- observed value;",
+"*F* -- female-sexed body, *M* -- male-sexed body, *n. a.* -- non-adult body")
+```
+```{r nb-sex, fig.cap=label_nb_sex}
+knitr::include_graphics(here("text/figs/nb_sex.pdf"))
+```
+
+```{r label-nb-age}
+label_nb_age <- paste("Mean neighbors in a given age category.",
+"*Gray area* -- density estimation for the simulated data, *vertical line* -- observed value;",
+"*F* -- female-sexed body, *M* -- male-sexed body, *n. a.* -- non-adult body")
+```
+```{r nb-age, fig.cap=label_nb_age}
+knitr::include_graphics(here("text/figs/nb_age.pdf"))
+```
+-->
+
+<!-- percolation analysis -->
+
+```{r perc-distance, fig.cap="Percolation thresholds"}
+knitr::include_graphics(here("text/figs/perc_distance.pdf"))
+```
+
+```{r perc-plan, fig.cap="Clusters formed at given percolation threshold distances"}
+knitr::include_graphics(here("text/figs/perc_plan_ids.pdf"))
+```
 

### 2020-09-26 code/knit_text_si.R
update makefile procedure, split text, si and word doc

@@ -1,11 +1,11 @@
 # Project "Vedrovice"
 # Script nr. 00
-# KNIT OUTPUT FILE
+# KNIT OUTPUT FILE 
 # author: Petr Pajdla
-# Create the text of the article in pdf
+# Create the SI
 
-# knit main text
-rmarkdown::render(input = here::here("text", "article_vedrovice.Rmd"), 
+# knit si
+rmarkdown::render(input = here::here("text", "si_vedrovice.Rmd"),
                   output_format = bookdown::pdf_document2(
                     toc = FALSE,
                     latex = "pdflatex",

### 2020-09-26 code/knit_text_word.R

@@ -1,0 +1,13 @@
+# Project "Vedrovice"
+# Script nr. 00
+# KNIT OUTPUT FILE
+# author: Petr Pajdla
+# Create the text of the article in word
+
+# knit main text
+rmarkdown::render(input = here::here("text", "article_vedrovice.Rmd"), 
+                  output_format = bookdown::word_document2(
+                    toc = FALSE,
+                    fig_caption = TRUE,
+                    number_sections = FALSE
+                  ))

### 2020-09-26 text/article_vedrovice.Rmd
spatial analysis - combination of vars

@@
         latex: pdflatex
+        toc: false
         fig_caption: true
@@
 csl             : "european-journal-of-archaeology.csl"
+header-includes:
+  - \usepackage[left]{lineno}
+  - \linenumbers
 ---
@@
 
+We consider there three variables, sex, age and origin to be indicative of
+ones identity in the Neolithic.
+To explore the relations between the variables we use them both as individual
+categories and combined into following indicator variables -- (1) sex/age,
+(2) origin/sex, (3) origin/age and finally (4) origin/sex/age.
+Due to frequent missing or indetermined values we decided to limit the studied
+sample only to bodies with all of the categories determined and only categories
+that occure in more than 5% of burials.
+
 <!-- methods -->
@@
 the *v* statistic is a measure of difference between co-occurences of
-individual artefact with other artifacts
+individual artefact with other artefacts
 [@manly1996, 475, for details see SI].
+
 To assess relations between groups based on sex, age and origin
@@
 Ten most extreme realizations of the simulation are discarded as outliers so
-the significance level *alpha* is 0.01 [@baddeley2016 410].
+the significance level is 0.01 [@baddeley2016 410].
 
@@
 
-## Artifact combinations
+## Artefact combinations
 
 By counting co-occurences of analyzed variables on simulated data matrices and comparing
-the distributions of the two studied statistics with the values calculated for
+the distributions of the statistics proposed by @manly1996 with the values calculated for
 the original dataset we can establish whether the overall pattern of artefacts
@@
 This suggests that there is strong evidence that the overall distribution of
-artefacts in different graves is not random (see Fig. \@ref(fig:sstat)).
+artefacts in different graves is not random. <!--(see Fig. \@ref(fig:sstat))-->
 
-Examining distributions of individual artefacts it is found that at the 0.05
+Examining distributions of individual artefacts we find that at the 0.05
 significance level only adzes, grinding tools, ceramic bowls and spondylus beads
@@
 Close to the 0.05 alpha level are distributions of bone tools and ceramic bottles
-(Fig. \@ref(fig:vstat)).
+(Fig. \@ref(fig:cooc-vstat)).
 Among the grave goods that are found to be randomly distributed in the graves
 across the burial ground are ochre pigment, lithics, pebbles, globular pots and
-special types of pottery vessel shapes and spondylus pendants and *buckles*.
+special types of pottery vessel shapes and spondylus pendants and buckles.
+
 The relationships between artefact types frequently occuring and missing together
-in graves are visualized in Fig. (\@ref(fig:coocnets)) with
-clusters highlighted using fast-greedy algorithm [@newman2004] and strength of the
-given association is expressed by the thickness of the edge connecting the nodes.
+in graves are visualized in Fig. (\@ref(fig:cooc-nets)) with
+clusters established using fast-greedy algorithm [@newman2004] and strength of the
+given association expressed by the thickness of the edge connecting the nodes.
 Artifacts most frequently present with other artifacts are with 5 links grinding tools and
@@
 
+Multitype *K* for combination of age and sex categories suggests clustering process
+between bodies determined as adult females and juveniles and vice versa.
+The relationship between pairs of non-adults/juveniles is also confirmed.
+Several other observed deviations occure at large distances and are same as
+in case of multitype *K* for single categories.
+
 In case of neighborhood structure derived from Delaunay triangulation (maximum
@@
 
-In case of clusters at different percolation distances
-there are some deviations from random number of
+The combinations of categories in Delaunay graph show significant relations for
+pairs of bodies labeled as (1) females from maturus/senilis category with
+juvenile females and (2) unsexed juveniles in general
+and as already noted at several places, between (3) pairs of bodies labeled as
+juveniles/non-adults.
+Gabriel graph in the sex/age category shows relations between (1) adultus female and male bodies,
+(2) maturus female and male bodies and (3) maturus female bodies.
+The origin/sex combination of categories produces significant deviation from
+expected frequencies for (1) non-local female bodies and local non-adults in
+Delaunay triangulation and in Gabriel graph for (1) non-local female and local male
+bodies and vice versa.
+The origin/age combination shows significant departures only in Gabriel graph for
+local adultus bodies with (1) non-local maturus bodies and (2) local juvenile bodies.
+Last but not least, incorporating all of the categories into the networks we observe
+significant results in both of the graphs (1) non-local maturus female bodies and
+local juvenile (non-adult) bodies and  in case of Gabrie graph relation
+between (2) non-local adultus female and local maturus male bodies.
+
+Clusters formed at different percolation distances
+show following deviations from expected numbers of
 occurences of given categories at different scales.
@@
 cluster of the cemetery (nr. 8).
+
+In the combined categories most of the already observed patterns are repeated and
+highlighted.
+In the largest cluster across percolation thresholds at 3.9 and 4.6 meters
+(nr. 3) there is abundance of (1) local non-adults/juveniles and (2) non-local
+female sexed bodies of maturus age category and on the other hand
+lesser numbers of (3) juvenile female-sexed bodies.
+The central cluster across distances 2.8, 3.9 and 4.6 (nr. 8, 10 and 7 respectively)
+shows higher numbers of local male- and female-sexed bodies of adultus category and in case of
+female bodies maturus age categories.
+The south-western cluster (nr. 10, 7 and 5) at percolation distances 3.9, 4.6 and 5.5
+is clearly dominated by local male-sexed bodies of maturus age category.
+Last but not least, in case of all the categories combined together a cluster in
+the western part of the cemetery at percolation threshold 4.6 and 5.5 (nr. 8 and 6 respectively)
+reveals higher counts of non-local female-sexed bodies of adultus age category.
 
@@
 
+# Conclusions
+
+Our goal was to identify non-random differences between burials both in terms
+of the artefact contents of the graves and position within the cemetery.
+The burials or more precisely the buried bodies were divided into categories
+based on three variables commonly connected to the identity in the Neolithic period.
+These categories are biological sex, age and origin.
+The exploration was done both for the categories as single variables and for
+all of their combinations.
+
+<!-- Artefact combinations -->
+
+<!-- Spatial pattern -->
+Spatial organization of the cemetery was explored using point pattern analysis
+techniques.
+The simpler methods based on cumulative frequency distribution of nearest neighbors
+(*G* function) and its generalization for empty space points (*F* function)
+did not show deviations from complete spatial random processes.
+On the other hand the methods including all the distances between events in the
+given area  (*K* and *T* functions) indicated clustering processes are taking place.
+*K* function for multivariate point patterns was also used in order to identify
+clustering processes between events of different categories.
+We consider this not useful in most cases because the distances at which interesting
+clustering processes are suggested imply existence of singular cluster across
+the cemetery and a finer scale is of much higher interest for our endeavour.
+
+Thus to identify whether close burials are more similar in terms of categories
+outlined above two neighborhood graphs, Delaunay triangulation and Gabriel
+graph were used.
+They are examples of high and low neighbors case scenario respectively.
+Most of the observations from the graphs proved not significantly different from
+randomization experiment.
+There are associations between non-local female-sexed
+bodies of maturus age category and (local) juveniles in both of the graphs.
+This observation goes in hand with association of female-sexed bodies
+and non-adults in Delaunay triangulation.
+This is not trivial because in general non-adult bodies are
+concentrated in the south-eastern part within the cemetery whereas female-sexed
+bodies are scattered all around the area of the cemetery.
+Similarly individuals determined as of maturus/senilis age category are in contrast to
+juvenile bodies concentrated in the western half of the cemetery.
+Furthermore in Gabriel graph there is a clear association between non-local adultus
+female-sexed bodies and local maturus male-bodies.
+This agrees with general association of female- and male-sexed bodies of
+adultus and maturus/senilis age categories in the Gabriel graph.
+
+Most detailed view on the clustering within the cemetery and what kinds of
+bodies are prevalent in the given clusters is achieved by the percolation analysis.
+The largest cluster across several percolation distances consists of more
+non-adult/juvenile bodies and non-local female-sexed bodies of maturus category
+on one hand and on the other the male-sexed bodies are underrepresented here.
+This is in full accordance with previous observations.
+The central cluster visible at small percolation distances showes overrepresentation of
+local male- and female-sexed bodies.
+Also the south-western cluster has higher counts of local male-sexed bodies of maturus age category and
+low amount of non-local bodies.
+On the other hand non-local female-sexed bodies of adultus age category are
+overrepresented in the western cluster neighboring with the previously mentioned one.
+
+The spatial analysis clearly proved that there is a clustering process taking place
+in the cemetery and what more, age, sex and origin of the buried bodies play
+structuring roles in the arrangement of the burials.
+The patterns are not too strict but some appear to be quite constant across a
+range of methods used.
+
+<!-- ======================================================================= -->
+
 # Discussion
@@
 <!-- randomization -->
-<!--
-```{r sstat, fig.cap = "Distribution of simulated and observed *S* statistic"}
+
+<!--```{r sstat, fig.cap = "Distribution of simulated and observed *S* statistic"}
 # knitr::include_graphics(here("plots/s_statistic.pdf"))
+```-->
+
+```{r cooc-vstat, fig.cap = "Simulated and observed *V* statistics for different variables"}
+knitr::include_graphics(here("plots", "cooc_vstats.pdf"))
 ```
 
-```{r vstat, fig.cap = "Simulated and observed *V* statistics for different variables"}
-# knitr::include_graphics(here("plots", "v_values.pdf"))
+```{r label-cooc-nets}
+label_cooc_nets <- "Networks of variables frequently present together and exluding each other, non-randomly occuring artefacts in bold"
+```
+```{r cooc-nets, fig.cap=label_cooc_nets }
+knitr::include_graphics(here("text/figs", "cooc_nets.pdf"))
+```
+<!-- plans -->
+
+```{r plan-sex, fig.cap="Plan of the cemetery, kernel density estimation for differently sexed bodies"}
+knitr::include_graphics(here("text/figs/plan_sex.pdf"))
 ```
 
-```{r coocnets, fig.cap = "Networks of variables frequently present together and exluding each other"}
-# knitr::include_graphics(here("plots", "cooc_networks.pdf"))
+```{r plan-age, fig.cap="Plan of the cemetery, kernel density estimation for bodies of different ages"}
+knitr::include_graphics(here("text/figs/plan_age.pdf"))
 ```
--->
 
-<!-- plans -->
+```{r plan-orig, fig.cap="Plan of the cemetery, kernel density estimation for bodies of different origins"}
+knitr::include_graphics(here("text/figs/plan_origin.pdf"))
+```
 
@@
 ```{r label-pp-funs}
-label_pp_funs <- paste("Functions *G*, *F*, *K* and *T*.",
-"*Dashed line* -- theoretical curve under complete spatial randomness, *solid line* -- empirical curve, *gray area* -- envelope")
+label_pp_funs <- paste(
+  "Functions *G*, *F*, *K* and *T*.",
+  "*Dashed line* -- theoretical curve under complete spatial randomness,
+  *solid line* -- empirical curve, *gray area* -- envelope")
 ```
@@
 
-<!-- to SI
 ```{r label-nb-sex}
-label_nb_sex <- paste("Mean neighbors in a given sex category.",
-"*Gray area* -- density estimation for the simulated data, *vertical line* -- observed value;",
-"*F* -- female-sexed body, *M* -- male-sexed body, *n. a.* -- non-adult body")
+label_nb_sex <- paste(
+  "Mean neighbors in a given category of sex;",
+  "*Gray area* -- density estimation for the simulated data, *vertical line* -- observed value"
+  # "*F* -- female-sexed body, *M* -- male-sexed body, *n. a.* -- non-adult body"
+)
 ```
@@
 ```{r label-nb-age}
-label_nb_age <- paste("Mean neighbors in a given age category.",
-"*Gray area* -- density estimation for the simulated data, *vertical line* -- observed value;",
-"*F* -- female-sexed body, *M* -- male-sexed body, *n. a.* -- non-adult body")
+label_nb_age <- paste(
+  "Mean neighbors in a given simplified age category"
+  # "*Gray area* -- density estimation for the simulated data, *vertical line* -- observed value;",
+)
 ```
@@
 ```
--->
+
+```{r label-nb-orig}
+label_nb_orig <- paste(
+  "Mean neighbors in a given category of origin based on Sr isotopes"
+  # "*Gray area* -- density estimation for the simulated data, *vertical line* -- observed value;",
+  # "*F* -- female-sexed body, *M* -- male-sexed body, *n. a.* -- non-adult body"
+)
+```
+```{r nb-orig, fig.cap=label_nb_orig}
+knitr::include_graphics(here("text/figs/nb_orig.pdf"))
+```
 
@@
 
-```{r perc-distance, fig.cap="Percolation thresholds"}
+```{r label-perc-dist}
+label_perc_dist <- "Normalized sizes of largest clusters at given distances, percolation thresholds labeled"
+```
+```{r perc-distance, fig.cap=label_perc_dist}
 knitr::include_graphics(here("text/figs/perc_distance.pdf"))
@@
 
+\pagebreak
+
 # References

### 2020-09-26 code/dataset.R

@@ -7,6 +7,7 @@
 # packages =====================================================================
 library(dplyr)
 library(here)
+library(stringr)
 
 # functions ====================================================================
 # continuous vector to dichotomous (binarization)
@@ -49,9 +50,10 @@ ved$bin_vars$count_mat <- input %>%
          lit_local, lit_nonlocal, 
          pot_special, pot_bowl, pot_globular, pot_bottle, 
          # pot_head, pot_mid, pot_leg, 
-         bone_tool, 
+         bone_tool, bone_awl,
          pendant_L, pendant_U, pendant_I, buckle_O, 
-         bracelet_spond, beads_spond, beads_marble, antler) %>% 
+         bracelet_spond, beads_spond, beads_marble, 
+         deer_teeth, antler, shell) %>% 
   as.matrix()
 
 rownames(ved$bin_vars$count_mat) <- ved$id_burials
@@ -180,6 +182,44 @@ ved$metadata <- left_join(ved$metadata,
          origin = factor(origin, levels = c("local", "non-local", "ind."))) %>% 
   select(-sr, -sr_ppm)
 
+
+# mixed categories --------------------------------------------------------
+detect_ind <- function(x) {
+  if_else(str_detect(x, "ind."), "ind.", x)
+}
+
+# only categories that are common for at least 5% of burials are kept
+threshold <- 0.05 * nrow(ved$bin_vars$count_mat)
+
+vec_origin <- c("local" = "loc.", "non-local" ="non.", "ind." = "ind.")
+
+comb_cats <- ved$metadata %>% 
+  filter(pres != "dist.") %>% 
+  mutate(
+    origin = vec_origin[origin],
+    sex = str_remove(sex, "\\s"),
+    cat_sa = str_c(sex, age_sim, sep = "/"),
+    cat_os = str_c(origin, sex, sep = "/"),
+    cat_oa = str_c(origin, age_sim, sep = "/"),
+    cat_osa = str_c(origin, sex, age_sim, sep = "/"),
+    across(starts_with("cat_"), detect_ind)
+  ) %>% 
+  select(id_burial, starts_with("cat"))
+
+get_perv_lvls <- function(x) {
+  lvls <- table(x)
+  names(lvls[lvls >= threshold])
+}
+
+comb_cats_over5 <- comb_cats %>% mutate(
[…]

### 2020-09-26 code/randomization.R

@@ -94,7 +94,7 @@ n_permutations <-  9999
 # going parallel to speed up randomization...
 # detecting cores for parallel
 no_cores <- parallel::detectCores() - 1
-  
+
 # cooccurences on random matrices ==============================================
 # # ======
 # cl <- parallel::makeCluster(no_cores)
@@ -137,12 +137,16 @@ v_p_values <- map2(v_experimental, v_observed, get_p) %>%
 
 v_statistic <- tibble(variable = names(v_observed),
                       v = t(unname(v_observed))[, 1],
-                      p = unname(v_p_values),
-                      signif = unname(v_p_values < 0.05)) %>% 
+                      p = unname(v_p_values)) %>% 
+  mutate(
+    # across(is.numeric, round, 2),
+    signif = if_else(p <= 0.05, "*", ""),
+    signif = if_else(p <= 0.01, "**", signif),
+    signif = if_else(p <= 0.001, "***", signif)) %>% 
   mutate(abbrv = unname(ved$var_names$short[variable]),
          long = unname(ved$var_names$long[variable]))
 
-readr::write_csv(v_statistic, here("data/temp", "v_statistics.csv"))
+readr::write_csv(v_statistic, here("data/temp", "cooc_v_stats.csv"))
 
 # visualizing of v values by small multiples -----------------------------------
 v_exp_g <- v_experimental %>% tidyr::gather(variable, value) %>% 
@@ -155,7 +159,6 @@ v_obs_g <- v_observed %>% tidyr::gather(variable, value) %>%
 v_annot <- v_statistic %>% select(long, p, signif) %>%
   mutate(p = round(p, 2),
          long = forcats::as_factor(long),
-         signif = if_else(signif == TRUE, "*", ""),
          txt = paste0(p, signif))
 
 v_plot <- ggplot(v_exp_g, mapping = aes(x = value)) +
@@ -172,7 +175,7 @@ v_plot <- ggplot(v_exp_g, mapping = aes(x = value)) +
 
 v_plot
 
-ggsave(here("plots", "v_values.pdf"), plot = v_plot, width = 6, height = 6)
+ggsave(here("plots", "cooc_vstats.pdf"), plot = v_plot, width = 6, height = 6)
 
 
 # normalizing to range 0-1 ------------------------------------------------
[…]

### 2020-09-26 code/spat_functions.R

@@ -44,11 +44,12 @@ ved_sf <- bind_cols(ved_sf,
 
 # marks ------------------------------------------------------------------------
 ved_marks <- ved_sf %>% 
-  select(sex, age = age_sim, origin) %>% 
+  select(sex, age = age_sim, origin, starts_with("cat")) %>% 
   st_drop_geometry() %>% 
   mutate(sex = factor(sex, levels = c("n. a.", "F", "M", "ind.")),
          age = factor(age, levels = c("juv.", "ad.", "mat.", "ind.")),
-         origin = factor(origin, levels = c("local", "non-local", "ind.")))
+         origin = factor(origin, levels = c("local", "non-local", "ind.")),
+         across(starts_with("cat_"), factor))
 
 # create ppp object ------------------------------------------------------------
 # ved_pp <- ppp(x = ved_layout$layout_x, y = ved_layout$layout_y,
@@ -135,9 +136,9 @@ plot_estimate_facet <- function(x) {
 # overall structure ============================================================
 # density based ----------------------------------------------------------------
 # quadrat test
-quadrat.test(ved_pp)
-quadrat.test(ved_pp, alternative = "clustered")
-quadrat.test(ved_pp, alternative = "regular")
+# quadrat.test(ved_pp)
+# quadrat.test(ved_pp, alternative = "clustered")
+# quadrat.test(ved_pp, alternative = "regular")
 # quadrat test suggests clustering
 
 # distance based ---------------------------------------------------------------
@@ -218,15 +219,15 @@ multiplecrossfunction <- function(pp, var, crossfun, nsim, nrank) {
 
 # sex
 mppk_sex <- multiplecrossfunction(ved_pp, "sex", "Kcross", nsim, nrank)
-mppg_sex <- multiplecrossfunction(ved_pp, "sex", "Gcross", nsim, nrank)
+# mppg_sex <- multiplecrossfunction(ved_pp, "sex", "Gcross", nsim, nrank)
 
 # age
 mppk_age <- multiplecrossfunction(ved_pp, "age", "Kcross", nsim, nrank)
-mppg_age <- multiplecrossfunction(ved_pp, "age", "Gcross", nsim, nrank)
[…]

### 2020-09-26 code/spat_neighbors.R

@@ -116,6 +116,18 @@ names(ved_age) <- ved_sf_pres$id_burial
 ved_origin <- ved_sf_pres$origin
 names(ved_origin) <- ved_sf_pres$id_burial
 
+ved_sa <- ved_sf_pres$cat_sa
+names(ved_sa) <- ved_sf_pres$id_burial
+
+ved_os <- ved_sf_pres$cat_os
+names(ved_os) <- ved_sf_pres$id_burial
+
+ved_oa <- ved_sf_pres$cat_oa
+names(ved_oa) <- ved_sf_pres$id_burial
+
+ved_osa <- ved_sf_pres$cat_osa
+names(ved_osa) <- ved_sf_pres$id_burial
+
 # ved_ei <- ved_sf_pres$ei_clust
 # names(ved_ei) <- ved_sf_pres$id_burial
 
@@ -148,6 +160,18 @@ nb_age_d <- mean_neighbors_networks(ved_d, ved_age)
 
 nb_orig_g <- mean_neighbors_networks(ved_g, ved_origin)
 nb_orig_d <- mean_neighbors_networks(ved_d, ved_origin)
+
+nb_sa_g <- mean_neighbors_networks(ved_g, ved_sa)
+nb_sa_d <- mean_neighbors_networks(ved_d, ved_sa)
+
+nb_os_g <- mean_neighbors_networks(ved_g, ved_os)
+nb_os_d <- mean_neighbors_networks(ved_d, ved_os)
+
+nb_oa_g <- mean_neighbors_networks(ved_g, ved_oa)
+nb_oa_d <- mean_neighbors_networks(ved_d, ved_oa)
+
+nb_osa_g <- mean_neighbors_networks(ved_g, ved_osa)
+nb_osa_d <- mean_neighbors_networks(ved_d, ved_osa)
 
 # nb_ei_g <- mean_neighbors_networks(ved_g, ved_ei)
 # nb_ei_d <- mean_neighbors_networks(ved_d, ved_ei)
@@ -205,6 +229,35 @@ randomize_neighbors_network <- function(g, sf, variable, n_sim = 99) {
 # ved_rand_orig_d <- randomize_neighbors_network(ved_d, ved_sf_pres, "origin", 
 #                                                n_sim = 999)
 # write_csv(ved_rand_orig_d, here("data/temp", "ved_rand_orig_d.csv"))
+
+ved_rand_sa_g <- randomize_neighbors_network(ved_g, ved_sf_pres, "cat_sa",
[…]

### 2020-09-26 code/spat_perc.R

@@ -239,7 +239,7 @@ g_long <- left_join(p$membership_long, ved_coord, by = "id") %>%
 # clust_col <- rep("gray", 19)
 
 scale_facet <- tibble::tibble(
-  label_radius = factor(c("dist. 8.8 m")),
+  label_radius = factor(c("dist. 7.8 m")),
   location = c("br"),
 )
 
@@ -293,9 +293,12 @@ polygs <- ved_sf %>%
   mutate(label_radius = paste("dist.", radius, "m"),
          label_radius = forcats::fct_reorder(label_radius, radius))
 
+disp_radii <- p_breaks[1:6]
+
 ggplot() +
   geom_sf(data = ved_exc, fill = NA, color = "black", size = 0.2) +
-  geom_sf(data = polygs, fill = "gray", color = NA, alpha = 0.4) +
+  geom_sf(data = filter(polygs, radius %in% disp_radii), 
+          fill = "gray", color = NA, alpha = 0.4) +
   geom_sf(data = ved_sf, 
           aes(shape = sex), 
           size = 0.6) +
@@ -315,21 +318,23 @@ ggplot() +
         legend.position = "bottom",
         legend.direction = "horizontal")
 
-ggsave(here::here("plots", "perc_plan.pdf"), width = 7, height = 14)
+ggsave(here::here("plots", "perc_plan.pdf"), width = 7, height = 10)
 
 # numbers of clusters
-polygs_ids <- polygs %>% group_by(radius) %>%
+polygs_ids <- filter(polygs, radius %in% disp_radii) %>% 
+  group_by(radius) %>%
   mutate(cluster = 1:n())
 
 ggplot() +
   geom_sf(data = ved_exc, fill = NA, color = "black", size = 0.1) +
   geom_sf(data = ved_sf, aes(shape = sex), size = 0.4, alpha = 0.2) +
-  geom_sf(data = polygs, fill = "gray", color = NA, alpha = 0.4) +
+  geom_sf(data = filter(polygs, radius %in% disp_radii), 
+          fill = "gray", color = NA, alpha = 0.4) +
   # ggspatial::geom_spatial_label_repel(data = polygs_ids, aes(label = cluster)) +
   ggrepel::geom_text_repel(data = polygs_ids, 
                            aes(label = cluster, geometry = polygons),
                            stat = "sf_coordinates", 
-                           size = 1.6, nudge_x = 1, nudge_y = 1) +
[…]

### 2020-09-26 code/spat_perc_clusters.R

@@ -32,7 +32,7 @@ ved_sf <- sf::st_read(here::here("data/temp", "layout.shp")) %>%
 
 # prep ---------------------------------------------------------------------
 
-input <- ved_sf %>% select(id_burial, sex, age_sim, origin) %>% 
+input <- ved_sf %>% select(id_burial, sex, age_sim, origin, starts_with("cat_")) %>% 
   st_drop_geometry() %>% 
   left_join(perc_cl, by = c("id_burial"))
 
@@ -164,6 +164,63 @@ devs_loc <- inner_join(deviates_from_random(input, "origin", nsim),
             expected = (expected.x + expected.y)/2,
             deviation = (deviation.x + deviation.y)/2)
 
+# multicategorical variables
+devs_sa <- inner_join(deviates_from_random(input, "cat_sa", nsim),
+                      deviates_from_random(input, "cat_sa", nsim), 
+                      by = c("radius", "cluster", "variable")) %>% 
+  transmute(radius, cluster, variable,
+            observed = observed.x,
+            expected = (expected.x + expected.y)/2,
+            deviation = (deviation.x + deviation.y)/2) %>% 
+  inner_join(deviates_from_random(input, "cat_sa", nsim),
+             by = c("radius", "cluster", "variable")) %>% 
+  transmute(radius, cluster, variable,
+            observed = observed.x,
+            expected = (expected.x + expected.y)/2,
+            deviation = (deviation.x + deviation.y)/2)
+
+devs_os <- inner_join(deviates_from_random(input, "cat_os", nsim),
+                      deviates_from_random(input, "cat_os", nsim), 
+                      by = c("radius", "cluster", "variable")) %>% 
+  transmute(radius, cluster, variable,
+            observed = observed.x,
+            expected = (expected.x + expected.y)/2,
+            deviation = (deviation.x + deviation.y)/2) %>% 
+  inner_join(deviates_from_random(input, "cat_os", nsim),
+             by = c("radius", "cluster", "variable")) %>% 
+  transmute(radius, cluster, variable,
+            observed = observed.x,
+            expected = (expected.x + expected.y)/2,
+            deviation = (deviation.x + deviation.y)/2)
+
+devs_oa <- inner_join(deviates_from_random(input, "cat_oa", nsim),
+                      deviates_from_random(input, "cat_oa", nsim), 
+                      by = c("radius", "cluster", "variable")) %>% 
+  transmute(radius, cluster, variable,
+            observed = observed.x,
[…]

### 2020-09-28 code/04-art_ed.R
edit code, finish analysis

@@ -1,0 +1,309 @@
+# Project "Vedrovice"
+# Script nr. 
+# 
+# author: Petr Pajdla
+# 
+
+set.seed(42)
+
+n_sim <- 9999
+
+# libs --------------------------------------------------------------------
+
+library(tidyverse)
+library(here)
+library(infer)
+
+theme_set(theme_minimal())
+
+# data --------------------------------------------------------------------
+
+ved <- readRDS(here("data/temp", "vedrovice_dataset.RDS"))
+
+# v_statistic <- read_csv(here("data/temp", "cooc_v_stats.csv"))
+# v_normalized <- read_csv(here("data/temp", "cooc_v_normal.csv"))
+
+non_rand_vars_abbrv <- read_lines(here("data/temp", "cooc_non_rand_vars.txt"))
+non_rand_vars <- ved$var_names$full %>% 
+  filter(abbrv %in% non_rand_vars_abbrv) %>% 
+  pull(vnames)
+
+# var_clusters <- read_csv(here("data/temp", "cooc_var_clusters.csv"))
+
+
+# modified input ----------------------------------------------------------
+
+# goal: count mean euclidean distance for groups of burials 
+ved_input <- ved$metadata %>% 
+  filter(pres != "dist.") %>% 
+  select(id_burial, sex, age = age_sim, orig = origin, starts_with("cat_")) %>% 
+  left_join(as_tibble(ved$bin_vars$bin_mat[, non_rand_vars], 
+                      rownames = "id_burial"), by = c("id_burial"))
+
+
+# functions ---------------------------------------------------------------
+
+# euclidean distance between burials in groups
+edistance <- function(df, grouping_var, distance_vars) {
+  var <- sym(grouping_var)
+  
+  mat <- df %>% 
+    select(id_burial, !!var, all_of(distance_vars)) %>% 
+    group_by(!!var) %>% 
+    nest() %>% 
+    mutate(data = map(data, column_to_rownames, "id_burial"),
+           data = map(data, as.matrix)) %>% 
+    filter(!!var != "ind.")
+  
+  lvls <- as.character(pull(mat, !!var))
+  comb <- gtools::combinations(n = length(lvls), 
+                               r = 2, 
+                               v = lvls, 
[…]

### 2020-09-28 code/05-art_ed_pvals.R

@@ -1,0 +1,265 @@
+# Project "Vedrovice"
+# Script nr. 
+# 
+# author: Petr Pajdla
+# 
+
+set.seed(42)
+
+n_sim <- 9999
+
+# libs --------------------------------------------------------------------
+
+library(tidyverse)
+library(here)
+library(infer)
+
+theme_set(theme_minimal())
+
+# data --------------------------------------------------------------------
+
+ved <- readRDS(here("data/temp", "vedrovice_dataset.RDS"))
+
+# v_statistic <- read_csv(here("data/temp", "cooc_v_stats.csv"))
+# v_normalized <- read_csv(here("data/temp", "cooc_v_normal.csv"))
+
+non_rand_vars_abbrv <- read_lines(here("data/temp", "cooc_non_rand_vars.txt"))
+non_rand_vars <- ved$var_names$full %>% 
+  filter(abbrv %in% non_rand_vars_abbrv) %>% 
+  pull(vnames)
+
+# var_clusters <- read_csv(here("data/temp", "cooc_var_clusters.csv"))
+
+
+# modified input ----------------------------------------------------------
+
+# goal: count mean euclidean distance for groups of burials 
+ved_input <- ved$metadata %>% 
+  filter(pres != "dist.") %>% 
+  select(id_burial, sex, age = age_sim, orig = origin, starts_with("cat_")) %>% 
+  left_join(as_tibble(ved$bin_vars$bin_mat[, non_rand_vars], 
+                      rownames = "id_burial"), by = c("id_burial"))
+
+
+# functions ---------------------------------------------------------------
+
+# euclidean distance between burials in groups
+edistance <- function(df, grouping_var, distance_vars) {
+  var <- sym(grouping_var)
+  
+  mat <- df %>% 
+    select(id_burial, !!var, all_of(distance_vars)) %>% 
+    group_by(!!var) %>% 
+    nest() %>% 
+    mutate(data = map(data, column_to_rownames, "id_burial"),
+           data = map(data, as.matrix)) %>% 
+    filter(!!var != "ind.")
+  
+  lvls <- as.character(pull(mat, !!var))
+  comb <- gtools::combinations(n = length(lvls), 
+                               r = 2, 
+                               v = lvls, 
+                               repeats.allowed = FALSE) %>% 
+    as_tibble() %>% 
[…]

### 2020-09-28 text/article_vedrovice.Rmd
finish text

@@
 ---
-title           : "Spatial Analysis and Grave Goods Distribution at the Cemetery of Vedrovice"
+title           : "Spatial Patterns and grave goods differences at the Cemetery of Vedrovice"
 # "A resampling approach to spatial arrangement and contents of burials in the Vedrovice cemetery"
@@
 abstract        :
-  "Spatial patterns in cemeteries and grave goods associations at the
+  "Spatial pattern and grave goods associations at the
   cemetery of Vedrovice - Široká u Lesa are explored using computationaly
@@
   distribution of grave goods are explored in
-  relation to layers of identity represented by sex, age and origin of the
-  buried bodies.
-  As a result of the investigation we conclude that...  "
+  relation to layers of identity represented by different buried bodies
+  attributes -- sex, age, origin and their combinations.
+  We find that among non-randomly distributed grave goods
+  are polished stone adzes, grinding tools, ceramic bowls and spondylus beads.
+  As a result of the investigation of differences of the non-randomly distributed
+  artefacts and attributes of burials we conclude that the most eminent are
+  dichotomies between genders, although accentuated by both origin and age.
+  Spatial analysis shows clear trends towards clustering between burials and
+  the emerging clusters are further explored for similarities in burial attributes."
 
@@
 Our approach is based on both of these works.
-Original co-occurence matrix is permutated *n = 9999* times with constant
+Original co-occurence matrix is permutated 9999 times with constant
 row and column totals, i.e. both the count of artifacts in individual graves
@@
 
+To assess whether differences between artefacts that are non-randomly distributed
+are reflected in relations between groups based on sex, age, origin and their combinations
+we calculate mean Euclidean distance (ED, for details see SI) between pairs of burials in multidimensional
+space settings defined by the non-randomly distributed artefacts.
+The observed mean ED is compared against mean ED distribution  derived by random
+reshuffling of the category labels between the burials 9999 times with constant
+probabilities of appearance of the given category label.
+<!-- To infer whether the observed and randomly created mean ED
+distributions are samples of the same population we use non-parametric
+Kruskal-Wallis test [@kruskal1952].-->
+For each of the categories the deviation of the mean ED from its randomly generated
+distributions is expressed as the fraction of random values of ED larger than
+the one observed.
+The deviation is deemed significant if less than 5% of
+random values are larger than the observed value.
+The process is a variation of the one described for resampling in case of
+artefact combinations and similar approach is taken by @sosna2008 [346-347].
+
+<!--
 To assess relations between groups based on sex, age and origin
-statistics introduced by @sosna2008 are used.
+statistics introduced by @sosna2008 are used.-->
 
@@
 
-Overall structure of the cemeter is examined using
+Overall structure of the cemetery is examined using
 <!-- either density based approaches or -->
@@
 any point location is independent on other point locations and
-defferent parameters influence its location [@osullivan2010, 126].
+different parameters influence its location [@osullivan2010, 126].
 <!-- [@nakoinz2016, 135-144; @osullivan2010, 130-139] -->
@@
 significance level only adzes, grinding tools, ceramic bowls and spondylus beads
-are distributed randomly.
+are distributed non-randomly.
 Close to the 0.05 alpha level are distributions of bone tools and ceramic bottles
@@
 (b) adzes.
+
+Artefacts non-randomly distributed across the cemetery are further used to
+check whether significant differences exist between their occurences in
+different groups defined based on origin, sex, age and their combinations.
+Exploring single attributes the only differences are in the categorization to
+differently sexed bodies, these are significant differences
+between male- and female-sexed bodies.
+In case of age or origin there are no significant differences observed.
+If we combine sex and age categories the distinction between juvenile, adult and mature
+female-sexed bodies and adult male-sexed bodies is evident.
+Another observed dichotomy is present between adult males and juveniles (non-adults).
+Moreless the same relations are present in case of combinations of sex and origin attributes.
+Local male-sexed bodies have significantly different combinations of non-randomly
+distributed artefacts from both local and non-local female-sexed bodies and non-adults.
+If all the categories are combined the divisions are even more evident.
+Local male-sexed bodies of adultus age category deviate in the artefact distributions
+from adult local and non-local females, maturus/senilis non-local female-sexed bodies
+and juveniles.
 
@@
 significant results in both of the graphs (1) non-local maturus female bodies and
-local juvenile (non-adult) bodies and  in case of Gabrie graph relation
+local juvenile (non-adult) bodies and  in case of Gabriel graph relation
 between (2) non-local adultus female and local maturus male bodies.
@@
 
-# Conclusions
+# Conclusions and discussion
 
@@
 <!-- Artefact combinations -->
+In case of grave goods analysis we explored the associations between artefacts and
+different categories of the bodies not to establish or reveal any new relations
+but merely to explore whether the known associations between sexes and specific grave goods
+are influenced by factors such as age and local or non-local origin of the buried person.
+We did not explore the relations between grave goods and burial attributes at the
+individual artefact scale so to say but used an overall measure (Euclidean distance)
+of between burial distance in terms of non-randomly distributed artefacts.
+By establishing whether the mean ED between categories defined by the grouping attributes
+mentioned above falls within or defies the expected distribution derived by
+randomization experiment we identified deviations that are not random.
+
+The most often reappearing is the difference between local adult male-sexed bodies and
+both local and non-local female-sexed ones and unsexed juveniles.
+This relation shows that sex classification of the body, although being
+accentuated by local origin and age category, eminently structures the artefact assemblage
+in the grave.
 
@@
 This is in full accordance with previous observations.
-The central cluster visible at small percolation distances showes overrepresentation of
+The central cluster visible at small percolation distances shows overrepresentation of
 local male- and female-sexed bodies.
@@
 
-# Discussion
+<!-- # Discussion -->
+
+We have successfully shown that all of the categories considered
+to be somehow related to identity in the Neolithic and especially their combinations
+play substantive roles in structuring the assemblage of grave goods deposited
+within the burials.
+Some of the relations between the explored categories and
+artefacts are well known across the LBK burial grounds,
+for instance the association between
+male burials and polished stone adzes [e.g. @bickle2020; @bentley2012] and some
+are known specifically from Vedrovice - *Široká u lesa* cemetery, for
+example the dichotomy
+between globular pottery shapes being associated with female-sexed bodies and
+on the other hand bowls with male-sexed bodies [@kvetina2004].
+Our goal was not to establish any new connections between genders or other burial
+attributes and grave good categories, but to use these associations
+to see which groups of burials are well separated based on the grave goods.
+
+First of all, we have shown that overall distribution of artefacts in the graves
+was not random, the grave goods distributed non-randomly were identified and
+subsequently only these were used in the further analysis effectively removing any
+possible *noise* from artefacts distributed randomly.
+Our main premise here is that the identity of the dead is based mostly on the attributes like sex,
+age and origin defined on the basis of Strontium isotope signal.
+We view these *layers* of identity as responsible for differences in
+artefact compositions of different graves as well as to a certain degree a structuring
+factor for the overall spatial organization of the cemetery.
+In case of the grave goods this relationship is largely dominated by a category
+of adult local male-sexed bodies that are distinctly different from both female-sexed bodies
+of all the age categories and juveniles.
+This is caused by already mentioned artefact characteristic of male burials, the stone adze.
+@bickle2020 removes polished and beveled stone tools from her analysis completely
+in order to remove their dominance over other patterns.
+In our case, adzes are one of four non-randomly occuring artefacts and removing
+them from the analysis completely would render it difficult to note any
+of the visible patterns.
+
+In case of spatial analysis we explore several approaches.
+The nearest-neighbor methods are a great tool in analysing the overall cemetery but
+in the small scale they do not prove useful to discover any interesting relationships
+between studied *layers* of identity.
+By far the most useful method in analysing the intra-site spatial arrangement
+is the percolation analysis.
+Another important point is the use of resampling methods that allows for identification
+of random and non random patterns both in grave goods analysis and spatial organization.
+
+<!--Traditional GIS based methods were already applied to study the cemetery
+[@kvetina2004] the techniques applied here reveal some trends that were previously-->
+
+We have shown that not only sex that is traditionally percieved
+as the most important structuring proxy for identity differentiation but also
+age and quite importantly also origin of the deceased played substantial
+roles in equiping the graves with varius types of artefacts by the Vedrovice community.
+In regard to spatial arrangement of the cemetery we demonstrated that there are
+pevalent clustering procesess taking place across the cemetery and also groups
+based on certain identities tend to be buried in clusters.
+These observations are based on extensive use of resampling approaches that
+allowed us to distinguish between patterns occuring solely by chance and
+patterns significantly deviating from randomness.
 
@@
 <!-- randomization -->
-
-<!--```{r sstat, fig.cap = "Distribution of simulated and observed *S* statistic"}
-# knitr::include_graphics(here("plots/s_statistic.pdf"))
-```-->
+```{r sstat, fig.cap = "Distribution of simulated and observed *S* statistic"}
+knitr::include_graphics(here("plots/cooc_sstat.pdf"))
+```
 
@@
 ```
+
 <!-- plans -->
-
 ```{r plan-sex, fig.cap="Plan of the cemetery, kernel density estimation for differently sexed bodies"}
@@
 
+<!--
 ```{r label-nb-sex}
@@
 ```{r nb-sex, fig.cap=label_nb_sex}
-knitr::include_graphics(here("text/figs/nb_sex.pdf"))
+# knitr::include_graphics(here("text/figs/nb_sex.pdf"))
 ```
@@
 ```{r nb-age, fig.cap=label_nb_age}
-knitr::include_graphics(here("text/figs/nb_age.pdf"))
+# knitr::include_graphics(here("text/figs/nb_age.pdf"))
 ```
@@
 ```{r nb-orig, fig.cap=label_nb_orig}
-knitr::include_graphics(here("text/figs/nb_orig.pdf"))
+# knitr::include_graphics(here("text/figs/nb_orig.pdf"))
 ```
+-->
 

### 2021-02-22 code/knit_text_word.R
update code to knit the resulting files

@@ -2,7 +2,7 @@
 # Script nr. 00
 # KNIT OUTPUT FILE
 # author: Petr Pajdla
-# Create the text of the article in word
+# Create the text of the article in word for further processing
 
 # knit main text
 rmarkdown::render(input = here::here("text", "article_vedrovice.Rmd"), 

### 2021-02-22 code/spat_functions.R
rm unused code

@@ -1,322 +1,0 @@
-# Project "Vedrovice"
-# Script nr. 3.2
-# SPATIAL ARRANGEMENT
-# author: Petr Pajdla
-# Spatial organization within the cemetery
-
-library(here)
-library(spatstat)
-library(sf)
-library(tidyverse)
-
-# number of simulations ---------------------------------------------------
-nsim = 999
-nrank = 5
-
-# k...nrank
-# M...nsim
-# probability 2 * k / M + 1
-# thus for 999 iterations 2 * 25 / 999 + 1 = 0.05 (5% probability)
-
-# theme for ggplot graphics
-theme_universe <- theme(panel.border = element_rect(colour = "black", 
-                                                    fill = NA, 
-                                                    size = 0.8),
-                        panel.background = element_blank(),
-                        line = element_blank(),
-                        strip.background = element_blank(), 
-                        strip.text = element_text(face = "italic"),
-                        axis.text.y = element_blank(), 
-                        axis.title.y = element_blank(), 
-                        panel.spacing = unit(1, "lines"))
-
-# read data and layouts --------------------------------------------------------
-ved <- read_rds(here("data/temp", "vedrovice_dataset.RDS"))
-
-ved_sf <- read_sf(here("data/temp", "layout.shp")) %>% 
-  filter(pres != "dist.")
-ved_exc <- read_sf(here("data/temp", "window.shp")) %>% 
-  as.owin(ved_exc)
-
-ved_sf <- bind_cols(ved_sf, 
-                    x = st_coordinates(ved_sf)[, 1], 
-                    y = st_coordinates(ved_sf)[, 2])
-
-# marks ------------------------------------------------------------------------
-ved_marks <- ved_sf %>% 
-  select(sex, age = age_sim, origin, starts_with("cat")) %>% 
-  st_drop_geometry() %>% 
-  mutate(sex = factor(sex, levels = c("n. a.", "F", "M", "ind.")),
-         age = factor(age, levels = c("juv.", "ad.", "mat.", "ind.")),
[…]

### 2021-02-22 code/spat_neighbors.R

@@ -1,785 +1,0 @@
-# Project "Vedrovice"
-# Script nr. 3.3
-# SPATIAL NEIGHBORS: Gabriel graph
-# author: Petr Pajdla
-# Mean number of neighbours with a given sex based on Gabriel graph
-
-library(here)
-library(tidyverse)
-library(broom)
-library(sf)
-library(spdep)
-library(igraph)
-
-set.seed(42)
-
-# theme -------------------------------------------------------------------
-
-theme_universe <- theme(panel.border = element_rect(colour = "black", 
-                                                    fill = NA, 
-                                                    size = 0.8),
-                        panel.background = element_blank(),
-                        line = element_blank(),
-                        strip.background = element_blank(), 
-                        strip.text = element_text(face = "italic"),
-                        # axis.text.y = element_blank(), 
-                        # axis.title.y = element_blank(), 
-                        panel.spacing = unit(1, "lines"))
-
-# data --------------------------------------------------------------------
-
-# ved <- read_rds(here("data/temp", "vedrovice_dataset.RDS"))
-# ei <- read_csv(here("data/temp", "exceptionality.csv"), col_types = "cdff") %>% 
-#   select(id_burial = burial, ei, ei_clust = fct)
-ved_sf <- read_sf(here("data/temp", "layout.shp")) %>% 
-  mutate(sex = fct_relevel(sex, c("n. a.", "F", "M", "ind.")),
-         age_sim = fct_relevel(age_sim, c("juv.", "ad.", "mat.", "ind.")),
-         origin = fct_relevel(origin, c("local", "non-local", "ind.")))
-ved_layout <- ved_sf %>% 
-  filter(pres != "dist.") %>% 
-  st_coordinates()
-ved_exc <- read_sf(here("data/temp", "window.shp"))
-
-ved_sf_pres <- filter(ved_sf, pres != "dist.") %>% 
-  mutate(age_sim = forcats::fct_drop(age_sim))
-
-# create delaunay + gabriel neighborhoods ---------------------------------
-ved_gabriel <- gabrielneigh(ved_layout)
[…]

### 2021-02-22 code/spat_perc.R

@@ -1,358 +1,0 @@
-# Project "Vedrovice"
-# Script nr. 
-# SPATIAL ARRANGEMENT - Percolation analysis
-# author: Petr Pajdla
-# Spatial organization within the cemetery
-
-# package percopackage
-# devtools::install_github("SCSchmidt/percopackage")
-# percolation analysis
-# uses in archaeology: Sophie C. Schmid 
-# https://osf.io/7extc/
-# https://github.com/SCSchmidt/percopackage
-
-# libs --------------------------------------------------------------------
-
-# library(percopackage)
-library(igraph)
-library(dplyr)
-library(tidyr)
-library(purrr)
-library(ggplot2)
-library(ggforce)
-library(ggspatial)
-library(concaveman)
-library(sf)
-library(forcats)
-
-# theme for ggplot graphics
-theme_universe <- theme(panel.border = element_rect(colour = "black", 
-                                                    fill = NA, 
-                                                    size = 0.8),
-                        panel.background = element_blank(),
-                        line = element_blank(),
-                        strip.background = element_blank(), 
-                        strip.text = element_text(face = "italic"),
-                        # axis.text.y = element_blank(), 
-                        # axis.title.y = element_blank(), 
-                        panel.spacing = unit(1, "lines"))
-
-# data --------------------------------------------------------------------
-
-ved_sf <- sf::read_sf(here::here("data/temp", "layout.shp")) %>% 
-  mutate(sex = fct_relevel(sex, c("n. a.", "F", "M", "ind.")),
-         age_sim = fct_relevel(age_sim, c("juv.", "ad.", "mat.", "ind.")),
-         origin = fct_relevel(origin, c("local", "non-local", "ind."))) %>% 
-  filter(pres != "dist.")
-ved_coord <- bind_cols(id = as.character(ved_sf$id_burial), 
-                       sf::st_coordinates(ved_sf)) %>% 
-  rename(x = X, y = Y)
-ved_exc <- sf::read_sf(here::here("data/temp", "window.shp"))
-
-# experiments -------------------------------------------------------------
-
[…]

### 2021-02-22 code/spat_perc_clusters.R

@@ -1,238 +1,0 @@
-# Project "Vedrovice"
-# Script nr. 
-# SPATIAL ARRANGEMENT - Grouping of vars into clusters derived from percolation
-# author: Petr Pajdla
-# Identify deviations from random for different percolation clusters 
-
-set.seed(42)
-
-# number of simulations
-nsim <- 999
-
-# libs --------------------------------------------------------------------
-
-library(tidyverse)
-library(sf)
-
-
-# data --------------------------------------------------------------------
-
-explored_rad <- paste0(c(2.8, 3.9, 4.6, 5.5, 6.5, 7.8))
-
-perc_cl <- read_csv(here::here("data/temp", "perc_clusters.csv")) %>% 
-  mutate(id_burial = as.character(id)) %>% 
-  select(-id) %>% 
-  filter(radius %in% explored_rad)
-
-ved_sf <- sf::st_read(here::here("data/temp", "layout.shp")) %>% 
-  mutate(sex = fct_relevel(sex, c("n. a.", "F", "M", "ind.")),
-         age_sim = fct_relevel(age_sim, c("juv.", "ad.", "mat.", "ind.")),
-         origin = fct_relevel(origin, c("local", "non-local", "ind."))) %>% 
-  filter(pres != "dist.")
-
-# prep ---------------------------------------------------------------------
-
-input <- ved_sf %>% select(id_burial, sex, age_sim, origin, starts_with("cat_")) %>% 
-  st_drop_geometry() %>% 
-  left_join(perc_cl, by = c("id_burial"))
-
-
-# functions ---------------------------------------------------------------
-
-count_var_per_radius <- function(input, variable) {
-  res <- input %>% 
-    filter(!is.na(radius), !is.na(cluster)) %>%
-    group_by(radius) %>% 
-    nest() %>% 
-    mutate(data = map(data, group_by, cluster, !!sym(variable)),
-           data = map(data, count)) %>% 
-    unnest(data) %>% 
-    ungroup(radius) %>% 
-    setNames(c("radius", "cluster", "variable", "n"))
-  return(res)
-}
-
-randomize_var_per_radius <- function(input, variable, n_sim) {
-  n_bur <- nrow(input)
-  variable0 <- as.symbol(variable)
-  
-  prob <- input %>%
-    select(!!variable0) %>%
-    group_by(!!variable0) %>%
-    count() %>%
[…]

### 2021-02-22 text/article_vedrovice.Rmd
update figures

@@
 ---
-title           : "Spatial Patterns and grave goods differences at the Cemetery of Vedrovice"
+title           : "Spatial Patterns and grave goods differences at the Cemetery of Vedrovice (Czech Republic): A Resampling Approach to Identity Markers in the Early Neolihic."
 # "A resampling approach to spatial arrangement and contents of burials in the Vedrovice cemetery"
@@
     # name          : "Petr Pajdla"
-    # affiliation   : "Department of Archaeology and Museology,
-    #                  Masaryk University, Brno, Czech Republic"
-    # email         : "petr.pajdla@gmail.com"
+    # affiliation   : "Department of Archaeology and Museology, Masaryk University, Brno, Czech Republic"
+    # email         : "petr.pajdla@protonmail.com"
 date            : "`r format(Sys.time(), '%d %B, %Y')`"
@@
   relation to layers of identity represented by different buried bodies
-  attributes -- sex, age, origin and their combinations.
+  attributes -- sex, age, geographical origin and their combinations.
   We find that among non-randomly distributed grave goods
@@
 Here we focus on exploration of relationship between three proposed *layers* of
-identity, biological sex and age of the body determined from sekeletal remains
+identity, biological sex and age of the body determined from skeletal remains
 by anthropological methods and origin of the body based on Strontium isotope
@@
 The analysis is set into a broader discussion of gender relations and identity
-inside Neolithic communities.
+in Neolithic communities.
 
@@
 European Neolithic in comparison with other periods.
-Authors present gender in the Neolihic as being
+Authors present gender in the Neolithic as being
 of a very different nature than gender in the later prehistory.
-Gender in the Neolithic is not clearly expressed and
+It is not clearly expressed and
 biological differences between sexes are not transformed into binary cultural
@@
 cemetery.-->
-We focus predominantly on investigating non-random patterns in artefact
-co-occurences and spatial arragement of the cemetery in relation to the
+We focus predominantly on investigating non-random patterns in artifact
+co-occurrences and spatial arrangement in the cemetery in relation to the
 proposed *layers* of Neolithic identity, i.e. sex, age and localness of the body.
-Our question is thus how are grave good assemblages related to aforementioned
+Our question is thus how grave goods relate to aforementioned
 categories and whether there is a consistent relationship between
-these categories forming the *identity*, grave goods distributions and spatial
+these categories forming the *identity* and spatial
 arrangement of the cemetery.
@@
 observed patterns are caused by mere chance associated with small numbers
-and which are actually present.
+and which are actually present is problematic.
 
 <!-- ### Questions
-* Is there evidence that certain artefacts covary,
+* Is there evidence that certain artifacts covary,
 either positively or negatively?
-Do certain artefacts tend to occur together, or are certain combinations
+Do certain artifacts tend to occur together, or are certain combinations
 absent or rare?
 * How can the presence and absence data best display which burials
-contain similar artefacts,
-and which artefacts have similar distributions over burials?
-* *How can the data be analysed so as to show most clearly the relationships*
+contain similar artifacts,
+and which artifacts have similar distributions over burials?
+* *How can the data be analyzed so as to show most clearly the relationships*
 *between the different* **burials** in terms of their **contents**?
-* *How can the data be analysed so as to show most clearly the relationship*
-*between the different* **artefacts** in terms of their
+* *How can the data be analyzed so as to show most clearly the relationship*
+*between the different* **artifacts** in terms of their
 **distributions in graves**?
@@
 
-<!-- sample (burials, artefacts) -->
+<!-- sample (burials, artifacts) -->
 Originally, this burial ground consisted of at least 108 burials.
@@
 area of the cemetery, amateur excavations during the 19th century or later
-agricultural activities [@podborsky2002].
+agricultural activities.
 These burials are excluded from most of the analysis due to missing and/or
@@
 ratios were collected in @whittle2013 and @richards2008.
-Further details on use and function of the artefacts were gathered from
+Further details on use and function of the artifacts were gathered from
 @masclanslatorre2020.
@@
 Each burial was originally characterized by 20 variables recording counts of
-artefacts present in the graves.
+artifacts present in the graves.
 Some of  these original variables are omitted in the analysis due to their rare occurrence,
@@
 thus being of no use in establishing relationships between pairs of burials.
-Only grave goods occuring in more than 5% of the graves are used in the analysis,
+Only grave goods occurring in more than 5% of the graves are used in the analysis,
 this step reduces the count of the variables to 14.
@@
 spondylus pendants or oval shaped buckles.
-Last but not least presence of ochre pigment in the grave is also noted.
+Last but not least presence of ocher pigment in the grave is also noted.
 Quantities of various *grave goods* are simplified into a binary matrix indicating
@@
 categories: female-sexed body (abbreviated *F*), male-sexed body (*M*),
-non adult (*n. a.*) and indetermined sex (*ind.*).
+non adult (*n. a.*) and undetermined sex (*ind.*).
 Age of the body is also simplified into three categories: juveniles (*juv.*)
@@
 and mature bodies (*mat.*, anthropological categories Adultus-Maturus, Maturus and
-Senilis) and bodies of indetermined age (*ind.*).
+Senilis) and bodies of undetermined age (*ind.*).
 Last category is the indicator for the origin of the buried person.
@@
 bodies with age category determined as Infans.
-Bodies faling within this range are labeled as local, outside the range as non-local
+Bodies falling within this range are labeled as local, outside the range as non-local
 and where data on Sr values are missing origin is not determined (*ind.*).
 
-We consider there three variables, sex, age and origin to be indicative of
+As mentioned previously we consider these three variables, sex, age and origin to be indicative of
 ones identity in the Neolithic.
@@
 (2) origin/sex, (3) origin/age and finally (4) origin/sex/age.
-Due to frequent missing or indetermined values we decided to limit the studied
+Due to frequent missing or undetermined values we decided to limit the studied
 sample only to bodies with all of the categories determined and only categories
-that occure in more than 5% of burials.
+that occur in more than 5% of burials.
 
@@
 
-To determine non-randomness of various artefact combinations in burials we conduct a
+To determine non-randomness of various artifact combinations in burials we conduct a
 permutation test on the presence/absence data matrix.
@@
 Our approach is based on both of these works.
-Original co-occurence matrix is permutated 9999 times with constant
+Original co-occurrence matrix is permuted 9999 times with constant
 row and column totals, i.e. both the count of artifacts in individual graves
@@
 
-To examine which associations of variables, i.e. artefacts, are infrequent or even not present
+To examine which associations of variables, i.e. artifacts, are infrequent or even not present
 and which tend to appear together more often than it would be expected by random chance
-the co-occurences of observed and simulated data
-are summarized using two statistics suggested by @manly1996.
+the co-occurrences of observed and simulated data
+are summarized using two statistics suggested by @manly1996 [475].
 The *S* statistic is a measure of overall deviation between expected and observed
-number of artefact occurences while
-the *v* statistic is a measure of difference between co-occurences of
-individual artefact with other artefacts
-[@manly1996, 475, for details see SI].
+number of artifact occurrences while
+the *v* statistic is a measure of difference between co-occurrences of
+individual artifact with other artifacts
+(for details see SI).
 
-To assess whether differences between artefacts that are non-randomly distributed
+To assess whether differences between artifacts that are non-randomly distributed
 are reflected in relations between groups based on sex, age, origin and their combinations
 we calculate mean Euclidean distance (ED, for details see SI) between pairs of burials in multidimensional
-space settings defined by the non-randomly distributed artefacts.
+space settings defined by the non-randomly distributed artifacts.
 The observed mean ED is compared against mean ED distribution  derived by random
@@
 The process is a variation of the one described for resampling in case of
-artefact combinations and similar approach is taken by @sosna2008 [346-347].
+artifact combinations and similar approach is taken by @sosna2008 [346-347].
 
@@
 <!-- either density based approaches or -->
-methods based on distances between neighboring burials. <!--[@osullivan2010, 121-154]-->
+methods utilizing distances between neighboring burials. <!--[@osullivan2010, 121-154]-->
 <!-- Density based approaches, for instance quadrat tests,
@@
 independent on other point locations and different parameters influence its location.-->
-These methods are based on distances between pairs and triples of neighbours and
+These methods are based on distances between pairs and triples of neighbors and
 characterize the point pattern based on *second- and third-order properties*
@@
 We utilize *G*, *F*, Ripley's *K* [@ripley1976; @sayer2013] and *T* [@schladitz2000] functions.
-These functions are based on nearest neighbour distances and are commonly used
+These functions are based on nearest neighbor distances and are commonly used
 to characterize point patterns in archaeology [@nakoinz2016, 135-144].
-The *G* function is a metric of how close events are and summarises
-cummulative frequency distribution of the nearest neighbour distances.
+The *G* function is a metric of how close events are and summarizes
+cumulative frequency distribution of the nearest neighbor distances.
 Because of its dependence on the intensity of events in the study area, it is
@@
 changes in the window size and works with all the proximal events up to a
-certain threshold, not only the nearest neighbours [@nakoinz2016, 138].
+certain threshold, not only the nearest neighbors [@nakoinz2016, 138].
 We find it the most useful in most of archaeology use cases, cf. [@negre2018].
@@
 Ten most extreme realizations of the simulation are discarded as outliers so
-the significance level is 0.01 [@baddeley2016 410].
+the significance level is 0.01 [cf. @baddeley2016 410].
+The remaining realizations of given point processes are used to construct envelopes 
+around function curves derived from observed data.
 
@@
 explore several methods.
-A modification of *K* function for multitype point patterns
+A modification of *K* function for multitype point pattern
 does not take into account all of the
@@
 
-To further establish the neighborhood structure we construct neighborhood graphs.
+To further explore the neighborhood structure we construct neighborhood graphs.
 We use Delaunay triangulation and its subgraph, a Gabriel graph [@gabriel1969]
@@
 <!-- Percolation analysis -->
-Another approach we consider for establishing neigborhood relations is by
+Another approach we consider for establishing neighborhood relations is by
 constructing buffer zones around each of the burials.
@@
 its generalization employing Euclidean distance [@rozenfeld2011].
-Nodes being in a given distance threshold (radius) are assigned into a cluster.
+Points being within a given distance threshold (radius) are assigned into a cluster.
 This process is repeated for a set of thresholds while recording maximum
@@
 
-By means of percolation analysis we identify threshold distances at which cluster
-sizes are growing and clusters formed at these distances are further explored.
+By means of percolation analysis we identify threshold distances at which maximum cluster
+size is growing and clusters formed at these distances are further explored.
 For each cluster number of individual bodies of a given category is recorded.
@@
 categories assigned at random with constant probabilities.
-Observed numbers of occurences deviating by two standard deviations from the mean of the
+Observed numbers of occurrences deviating by more than two standard deviations from the mean of the
 resampled results are regarded as strong evidence of non-random patterns being
 present in given clusters.
-To achieve stability of results and eliminate observations occuring due to
+To achieve stability of results and eliminate observations occurring due to
 chance in small clusters with low numbers of categories of interest,
 the whole resampling procedure is repeated three times and only observations present in
-all the iterations are reported with expected numbers of occurence and their
+all the iterations are reported with expected numbers of occurrence and their
 deviations from observed values summarized as mean values.
@@
 
-By counting co-occurences of analyzed variables on simulated data matrices and comparing
+By counting co-occurrences of analyzed variables on simulated data matrices and comparing
 the distributions of the statistics proposed by @manly1996 with the values calculated for
-the original dataset we can establish whether the overall pattern of artefacts
-distribution in the graves is non-random and further identify artefacts that deviate
+the original data set we can establish whether the overall pattern of artifacts
+distribution in the graves is non-random and further identify artifacts that deviate
 from the random pattern.
-The overall deviation from expected co-occurences, i.e. the *S* statistic
+The overall deviation from expected co-occurrences, i.e. the *S* statistic
 [@manly1996, 475] is significantly large at the 0.05 alpha level.
 This suggests that there is strong evidence that the overall distribution of
-artefacts in different graves is not random. <!--(see Fig. \@ref(fig:sstat))-->
+artifacts in different graves is not random. <!--(see Fig. \@ref(fig:sstat))-->
 
-Examining distributions of individual artefacts we find that at the 0.05
+Examining distributions of individual artifacts we find that at the 0.05
 significance level only adzes, grinding tools, ceramic bowls and spondylus beads
@@
 Among the grave goods that are found to be randomly distributed in the graves
-across the burial ground are ochre pigment, lithics, pebbles, globular pots and
-special types of pottery vessel shapes and spondylus pendants and buckles.
+across the burial ground are ocher pigment, lithics, pebbles, globular pots and
+special types of pottery vessel shapes, spondylus pendants and buckles.
 
-The relationships between artefact types frequently occuring and missing together
-in graves are visualized in Fig. (\@ref(fig:cooc-nets)) with
+The relationships between artifact types frequently occurring and missing together
+in graves are visualized in Fig. \@ref(fig:cooc-nets) with
 clusters established using fast-greedy algorithm [@newman2004] and strength of the
@@
 Artifacts most frequently present with other artifacts are with 5 links grinding tools and
-ceramic bowls and with 4 links adzes, non-local lithic tools, ochre pigment and
+ceramic bowls and with 4 links adzes, non-local lithic tools, ocher pigment and
 spondylus beads.
@@
 Artifacts that are frequently absent in the graves together are with 6 links
-local lithic tools, globular pots and ceramic bowls and spondylus buckels with
+local lithic tools, globular pots, ceramic bowls and spondylus buckles with
 5 and 4 links respectively.
@@
 
-Artefacts non-randomly distributed across the cemetery are further used to
-check whether significant differences exist between their occurences in
+Artifacts non-randomly distributed across the cemetery are further used to
+check whether significant differences exist between their occurrences in
 different groups defined based on origin, sex, age and their combinations.
@@
 Another observed dichotomy is present between adult males and juveniles (non-adults).
-Moreless the same relations are present in case of combinations of sex and origin attributes.
+More or less the same relations are present in case of combinations of sex and origin attributes.
 Local male-sexed bodies have significantly different combinations of non-randomly
-distributed artefacts from both local and non-local female-sexed bodies and non-adults.
+distributed artifacts from both local and non-local female-sexed bodies and non-adults.
 If all the categories are combined the divisions are even more evident.
-Local male-sexed bodies of adultus age category deviate in the artefact distributions
-from adult local and non-local females, maturus/senilis non-local female-sexed bodies
+Local male-sexed bodies of Adultus age category deviate in the artifact distributions
+from adult local and non-local females, Maturus/Senilis non-local female-sexed bodies
 and juveniles.
@@
 expected counts (for the details see supplementary information).
-Inconclusive result of the quadrat test is not suprising because as was noted
+Inconclusive result of the quadrat test is not surprising because as was noted
 earlier, we do not expect *first-order* properties to play major roles in
@@
 but neither traverses the boundaries of the envelope.
-Thus the cummulative nearest-neighbor distance (*G*) as well as the
-empty-space function *F* are well approximated by
-random process and do not show evidence for spatial clustering nor repelent functions
+Thus the cumulative nearest-neighbor distance (*G*) as well as the
+empty-space function (*F*) are well approximated by
+random process and do not show evidence for spatial clustering nor repellent functions
 taking place.
@@
 <!-- in case of *K* and slightly more in case of *T* function.-->
-This means that numbers of nearest neighbors at this and larger distances is
+This means that number of nearest neighbors at this and larger distances is
 significantly higher than we would expect under complete spatial randomness.
@@
 This point of cluster growth is revealed by the point pattern analysis as well.
-The clusters at some of the distances are shown in Fig. \@ref(fig:perc-plan).
+The clusters at chosen distances are shown in Fig. \@ref(fig:perc-plan).
 
@@
 most of the observed categories in Fig. \@ref(fig:pp-crossk).
-The distances at which these deviations start occuring are in most cases rather large,
+The distances at which these deviations start occurring are in most cases rather large,
 well over 10 meters and considering the fact that at the 10.4 meters radius
@@
 The relationship between pairs of non-adults/juveniles is also confirmed.
-Several other observed deviations occure at large distances and are same as
-in case of multitype *K* for single categories.
+Several other observed deviations occur at large distances and the explanation for
+these is same as in case of multitype *K* for single categories.
 
@@
 The combinations of categories in Delaunay graph show significant relations for
-pairs of bodies labeled as (1) females from maturus/senilis category with
-juvenile females and (2) unsexed juveniles in general
-and as already noted at several places, between (3) pairs of bodies labeled as
-juveniles/non-adults.
-Gabriel graph in the sex/age category shows relations between (1) adultus female and male bodies,
-(2) maturus female and male bodies and (3) maturus female bodies.
+pairs of bodies labeled as Maturus/Senilis category females with
+(1) juvenile females and with (2) unsexed juveniles.
+The relationship between pairs of juvenile/non-adult bodies is again confirmed.
+Gabriel graph in the sex/age category shows relations between (1) Adultus female and male bodies,
+(2) Maturus female and male bodies and (3) pairs of Maturus female bodies.
 The origin/sex combination of categories produces significant deviation from
@@
 The origin/age combination shows significant departures only in Gabriel graph for
-local adultus bodies with (1) non-local maturus bodies and (2) local juvenile bodies.
+local Adultus bodies neighboring with (1) non-local Maturus bodies and (2) local juvenile bodies.
 Last but not least, incorporating all of the categories into the networks we observe
-significant results in both of the graphs (1) non-local maturus female bodies and
+significant results in both of the graphs (1) non-local Maturus female bodies and
 local juvenile (non-adult) bodies and  in case of Gabriel graph relation
-between (2) non-local adultus female and local maturus male bodies.
+between (2) non-local Adultus female and local Maturus male bodies.
 
@@
 show following deviations from expected numbers of
-occurences of given categories at different scales.
+occurrences of given categories at different scales.
 Regarding the determination of sex of the bodies, at the threshold of 3.9 meters
@@
 of 5.5 meters female-sexed bodies are more frequent than we would expect.
-The age categories correspond to the previous obsevations where
+The age categories correspond to the previous observations where
 most evident is the overrepresentaion of juvenile bodies at the 3.9 meters threshold in
@@
 (nr. 10 and 5 respectively) consists of less non-local bodies than would be expected.
-Lastly at the distance radius 2.8 meters there are more local bodies in the cenral
+Lastly at the distance radius 2.8 meters there are more local bodies in the central
 cluster of the cemetery (nr. 8).
@@
 (nr. 3) there is abundance of (1) local non-adults/juveniles and (2) non-local
-female sexed bodies of maturus age category and on the other hand
+female sexed bodies of Maturus age category and on the other hand
 lesser numbers of (3) juvenile female-sexed bodies.
 The central cluster across distances 2.8, 3.9 and 4.6 (nr. 8, 10 and 7 respectively)
-shows higher numbers of local male- and female-sexed bodies of adultus category and in case of
-female bodies maturus age categories.
+shows higher numbers of local male- and female-sexed bodies of Adultus category and in case of
+female bodies Maturus age categories.
 The south-western cluster (nr. 10, 7 and 5) at percolation distances 3.9, 4.6 and 5.5
-is clearly dominated by local male-sexed bodies of maturus age category.
+is clearly dominated by local male-sexed bodies of Maturus age category.
 Last but not least, in case of all the categories combined together a cluster in
 the western part of the cemetery at percolation threshold 4.6 and 5.5 (nr. 8 and 6 respectively)
-reveals higher counts of non-local female-sexed bodies of adultus age category.
+reveals higher counts of non-local female-sexed bodies of Adultus age category.
 
@@
 but never significant ($p > 0.07$).
-In case of bodies unsexed due to early age (non-adults), the neigborhood based
-on buffer zones show a signficant realtionship with other non-adults ($p \approx 0.04$).
+In case of bodies unsexed due to early age (non-adults), the neighborhood based
+on buffer zones show a significant relationship with other non-adults ($p \approx 0.04$).
 Finally burials with female-sexed bodies show a significantly higher number of
@@
 Two further relationships showed non-random deviations.
-In case of Delaunay trinagulation for burials with bodies with indeterminate sex with
+In case of Delaunay triangulation for burials with bodies with indeterminate sex with
 non-adults ($p \approx 0.02$) and in case of Gabriel graph between non-adults and
@@
 Our goal was to identify non-random differences between burials both in terms
-of the artefact contents of the graves and position within the cemetery.
+of the artifact contents of the graves and position within the cemetery.
 The burials or more precisely the buried bodies were divided into categories
@@
 
-<!-- Artefact combinations -->
-In case of grave goods analysis we explored the associations between artefacts and
+<!-- Artifact combinations -->
+In case of grave goods analysis we explored the associations between artifacts and
 different categories of the bodies not to establish or reveal any new relations
@@
 We did not explore the relations between grave goods and burial attributes at the
-individual artefact scale so to say but used an overall measure (Euclidean distance)
-of between burial distance in terms of non-randomly distributed artefacts.
+individual artifact scale so to say but used an overall measure (Euclidean distance)
+of between burial distance in terms of non-randomly distributed artifacts.
 By establishing whether the mean ED between categories defined by the grouping attributes
@@
 This relation shows that sex classification of the body, although being
-accentuated by local origin and age category, eminently structures the artefact assemblage
+accentuated by local origin and age category, eminently structures the artifact assemblage
 in the grave.
@@
 clustering processes are suggested imply existence of singular cluster across
-the cemetery and a finer scale is of much higher interest for our endeavour.
+the cemetery and a finer scale is of much higher interest for our endeavor.
 
@@
 There are associations between non-local female-sexed
-bodies of maturus age category and (local) juveniles in both of the graphs.
+bodies of Maturus age category and (local) juveniles in both of the graphs.
 This observation goes in hand with association of female-sexed bodies
@@
 bodies are scattered all around the area of the cemetery.
-Similarly individuals determined as of maturus/senilis age category are in contrast to
+Similarly individuals determined as of Maturus/Senilis age category are in contrast to
 juvenile bodies concentrated in the western half of the cemetery.
-Furthermore in Gabriel graph there is a clear association between non-local adultus
-female-sexed bodies and local maturus male-bodies.
+Furthermore in Gabriel graph there is a clear association between non-local Adultus
+female-sexed bodies and local Maturus male-bodies.
 This agrees with general association of female- and male-sexed bodies of
-adultus and maturus/senilis age categories in the Gabriel graph.
+Adultus and Maturus/Senilis age categories in the Gabriel graph.
 
@@
 The largest cluster across several percolation distances consists of more
-non-adult/juvenile bodies and non-local female-sexed bodies of maturus category
+non-adult/juvenile bodies and non-local female-sexed bodies of Maturus category
 on one hand and on the other the male-sexed bodies are underrepresented here.
@@
 local male- and female-sexed bodies.
-Also the south-western cluster has higher counts of local male-sexed bodies of maturus age category and
+Also the south-western cluster has higher counts of local male-sexed bodies of Maturus age category and
 low amount of non-local bodies.
-On the other hand non-local female-sexed bodies of adultus age category are
+On the other hand non-local female-sexed bodies of Adultus age category are
 overrepresented in the western cluster neighboring with the previously mentioned one.
@@
 Some of the relations between the explored categories and
-artefacts are well known across the LBK burial grounds,
+artifacts are well known across the LBK burial grounds,
 for instance the association between
@@
 
-First of all, we have shown that overall distribution of artefacts in the graves
+First of all, we have shown that overall distribution of artifacts in the graves
 was not random, the grave goods distributed non-randomly were identified and
 subsequently only these were used in the further analysis effectively removing any
-possible *noise* from artefacts distributed randomly.
+possible *noise* from artifacts distributed randomly.
 Our main premise here is that the identity of the dead is based mostly on the attributes like sex,
@@
 We view these *layers* of identity as responsible for differences in
-artefact compositions of different graves as well as to a certain degree a structuring
+artifact compositions of different graves as well as to a certain degree a structuring
 factor for the overall spatial organization of the cemetery.
@@
 of all the age categories and juveniles.
-This is caused by already mentioned artefact characteristic of male burials, the stone adze.
+This is caused by already mentioned artifact characteristic of male burials, the stone adze.
 @bickle2020 removes polished and beveled stone tools from her analysis completely
 in order to remove their dominance over other patterns.
-In our case, adzes are one of four non-randomly occuring artefacts and removing
+In our case, adzes are one of four non-randomly occurring artifacts and removing
 them from the analysis completely would render it difficult to note any
@@
 In case of spatial analysis we explore several approaches.
-The nearest-neighbor methods are a great tool in analysing the overall cemetery but
+The nearest-neighbor methods are a great tool in analyzing the overall cemetery but
 in the small scale they do not prove useful to discover any interesting relationships
 between studied *layers* of identity.
-By far the most useful method in analysing the intra-site spatial arrangement
+By far the most useful method in analyzing the intra-site spatial arrangement
 is the percolation analysis.
@@
 
-We have shown that not only sex that is traditionally percieved
+We have shown that not only sex that is traditionally perceived
 as the most important structuring proxy for identity differentiation but also
 age and quite importantly also origin of the deceased played substantial
-roles in equiping the graves with varius types of artefacts by the Vedrovice community.
+roles in equipping the graves with various types of artifacts by the Vedrovice community.
 In regard to spatial arrangement of the cemetery we demonstrated that there are
-pevalent clustering procesess taking place across the cemetery and also groups
+prevalent clustering processes taking place across the cemetery and also groups
 based on certain identities tend to be buried in clusters.
 These observations are based on extensive use of resampling approaches that
-allowed us to distinguish between patterns occuring solely by chance and
+allowed us to distinguish between patterns occurring solely by chance and
 patterns significantly deviating from randomness.
@@
 ```{r label-cooc-nets}
-label_cooc_nets <- "Networks of variables frequently present together and exluding each other, non-randomly occuring artefacts in bold"
+label_cooc_nets <- paste("Networks of variables frequently present together and exluding each other.",
+                        "1 – bottle, 2 – bowl, 3 – globular pot, 4 – polished stone tool, 5 – local lithics,",
+                        "6 – non-local lithics, 7 – pebble, 8 – grinding slab, 9 – bone tool,",
+                        "10 – U-shaped pendant, 11 – spondylus beads, 12 – O-shaped buckle.",
+                        "Images adapted from Podborský 2002")
 ```
 ```{r cooc-nets, fig.cap=label_cooc_nets }
-knitr::include_graphics(here("text/figs", "cooc_nets.pdf"))
+knitr::include_graphics(here("text/figs", "cooc_nets_fin.png"))
 ```

### 2021-06-03 text/article_vedrovice.Rmd
update text after proofreading

--- 2020-09-26 text/article_vedrovice.Rmd
@@
 ---
-title           : "Spatial Analysis and Grave Goods Distribution at the Cemetery of Vedrovice"
-# "A resampling approach to spatial arrangement and contents of burials in the Vedrovice cemetery"
-
+title           : "Spatial Patterns and Grave Goods Differences at the Cemetery of Vedrovice (Czech Republic): A Resampling Approach to Identity Markers in the Early Neolithic"
 author          : "Petr Pajdla"
-    # name          : "Petr Pajdla"
-    # affiliation   : "Department of Archaeology and Museology,
-    #                  Masaryk University, Brno, Czech Republic"
-    # email         : "petr.pajdla@gmail.com"
 date            : "`r format(Sys.time(), '%d %B, %Y')`"
 abstract        :
-  "Spatial patterns in cemeteries and grave goods associations at the
-  cemetery of Vedrovice - Široká u Lesa are explored using computationaly
-  instensive formal methods.
-  Resampling approaches both to intrasite spatial arrangement and
-  distribution of grave goods are explored in
-  relation to layers of identity represented by sex, age and origin of the
-  buried bodies.
-  As a result of the investigation we conclude that...  "
-
+  "The spatial pattern and grave good associations at the cemetery of Vedrovice - Široká u Lesa are explored using computationally intensive formal methods.
+Resampling approaches to both intra-site spatial arrangement and the distribution of grave goods are explored in relation to layers of identity represented by different buried body attributes – sex, age, geographical origin and their combinations.
+We found that polished stone adzes, grinding tools, ceramic bowls and spondylus beads occur among non-randomly distributed grave goods.
+As a result of the investigation of differences among the non-randomly distributed artifacts and attributes of burials, we conclude that the most significant are dichotomies between genders, accentuated by both origin and age.
+Spatial analysis shows clear trends towards clustering between burials, and the emerging clusters are further explored for similarities in burial attributes."
 output:
-    bookdown::pdf_document2:
-    # bookdown::word_document2:
+    bookdown::word_document2:
         latex: pdflatex
@@
 bibliography    : "bib_vedrovice.bib"
-csl             : "european-journal-of-archaeology.csl"
+csl             : "apa.csl"
 header-includes:
@@
 *Linearbandkeramik* (LBK) cemeteries.
-Recent contributions use the evidence from the cemetery in a broader discussion
-of post-marital residence patterns [@hrncir2020] and
+Recent articles use evidence from the cemetery in a broader discussion
+of post-marital residence patterns [@hrncir2020],
 sexual inequalities and gender in the Neolithic [@masclanslatorre2020; @bickle2020].
-Here we focus on exploration of relationship between three proposed *layers* of
-identity, biological sex and age of the body determined from sekeletal remains
-by anthropological methods and origin of the body based on Strontium isotope
+Here we focus on an exploration of the relationship between three proposed *layers* of
+identity, biological sex and age of the body determined from skeletal remains
+by anthropological methods and the origin of the body based on the strontium isotope
 signal.
 We explore whether the structure observed in these three categories is
-mirrored in the grave goods associations and spatial organization within the
+mirrored in the grave good associations and spatial organization within the
 cemetery.
-The analysis is set into a broader discussion of gender relations and identity
-inside Neolithic communities.
+The analysis is set into a broader discussion of gender relationships and identity
+in Neolithic communities.
 
 <!-- view of gender in the Neolithic, archaeological context -->
-@robb2017 [128] noted lack of gender archaeology debate for the
-European Neolithic in comparison with other periods.
-Authors present gender in the Neolihic as being
-of a very different nature than gender in the later prehistory.
-Gender in the Neolithic is not clearly expressed and
+@robb2017 [p. 128] noted a lack of debate in archaeology on gender for the
+European Neolithic compared to other periods.
+The authors present gender in the Neolithic as being
+of a very different nature than gender in later prehistory.
+It is argued that gender is not clearly expressed and
 biological differences between sexes are not transformed into binary cultural
-categories (ibid. 140-142).
+categories [@robb2017, p. 140--142].
 Considering the Neolithic burial evidence across Europe, authors find that it
-'shows very little gender distinction; when it does, it is rarely a clear-cut
+"shows very little gender distinction; when it does, it is rarely a clear-cut
 binary categorization but a matter of degree and overlap, sometimes polythetically
-combined with factors such as age and localness' (ibid. 138).
-In response, @bickle2020 [14] argues that 'biological sex may not have defined gender
-during the LBK' although the 'sexed bodies did matter'.
+combined with factors such as age and localness" [@robb2017, p. 138].
+In response, @bickle2020 [p. 14] argues that "biological sex may not have defined gender
+during the LBK" although the "sexed bodies did matter".
 
@@
 cemetery.-->
-We focus predominantly on investigating non-random patterns in artefact
-co-occurences and spatial arragement of the cemetery in relation to the
-proposed *layers* of Neolithic identity, i.e. sex, age and localness of the body.
-Our question is thus how are grave good assemblages related to aforementioned
+We focus mainly on investigating non-random patterns in artifact
+co-occurrences and spatial arrangement in the cemetery in relation to the
+proposed *layers* of Neolithic identity, i.e., sex, age and localness of the body.
+Our question is thus how grave goods relate to the aforementioned
 categories and whether there is a consistent relationship between
-these categories forming the *identity*, grave goods distributions and spatial
+these categories forming the *identity* and spatial
 arrangement of the cemetery.
@@
 <!-- resampling and computer-intensive techniques-->
-The approach taken is a quantitative one embracing resampling techniques and
+The chosen approach is a quantitative one embracing resampling techniques and
 computer-intensive methods in general.
-The resampling techniques were introduced to mortuary studies by
-@manly1996 but the application is still limited [@sosna2008, 343].
-As noted by Sosna (ibid.), the most important benefit of resampling is in its
+The resampling techniques were introduced in mortuary studies by
+@manly1996 but the application is still limited [@sosna2008, p. 343].
+As noted by @sosna2008, the most important benefit of resampling is its
 ability to distinguish random and non-random patterns.
-This is crucial for mortuary analysis because the samples studied, including
+This is crucial for mortuary analysis because the studied samples, including
 the one presented here, are usually small in size and identifying which
 observed patterns are caused by mere chance associated with small numbers
-and which are actually present.
+and which are actually present is problematic.
 
 <!-- ### Questions
-* Is there evidence that certain artefacts covary,
+* Is there evidence that certain artifacts covary,
 either positively or negatively?
-Do certain artefacts tend to occur together, or are certain combinations
+Do certain artifacts tend to occur together, or are certain combinations
 absent or rare?
 * How can the presence and absence data best display which burials
-contain similar artefacts,
-and which artefacts have similar distributions over burials?
-* *How can the data be analysed so as to show most clearly the relationships*
+contain similar artifacts,
+and which artifacts have similar distributions over burials?
+* *How can the data be analyzed so as to show most clearly the relationships*
 *between the different* **burials** in terms of their **contents**?
-* *How can the data be analysed so as to show most clearly the relationship*
-*between the different* **artefacts** in terms of their
+* *How can the data be analyzed so as to show most clearly the relationship*
+*between the different* **artifacts** in terms of their
 **distributions in graves**?
@@
 <!-- site -->
-Vedrovice site is situated in the south of Moravia (Czech Republic).
-Earliest known finds and archaeological explorations here are from
-the 19th century but the excavation took place in the second half of 20th century.
-There are two known burial areas, *Za Dvorem* and larger one *Široká u Lesa* and
+The Vedrovice site is situated in south Moravia (Czech Republic).
+The earliest known finds and archaeological explorations here are from
+the 19th century, but excavations were conducted in the second half of 20th century.
+There are two known burial areas, *Za Dvorem* and the larger *Široká u Lesa* and
 a settlement area.
-For more details on the site refer to @masclanslatorre2020 or the
-original publication [@podborsky2002].
-The material analyzed here comes from Vedrovice *Široká u lesa* cemetery.
+For more details on the site, see @masclanslatorre2020 or the
+original publication by @podborsky2002.
+The material analyzed here comes from the *Široká u lesa* cemetery in Vedrovice.
 
-<!-- sample (burials, artefacts) -->
+<!-- sample (burials, artifacts) -->
 Originally, this burial ground consisted of at least 108 burials.
 Some of the burials were disturbed by later settlement and burial pits in the
-area of the cemetery, amateur excavations during the 19th century or later
-agricultural activities [@podborsky2002].
+area of the cemetery, amateur excavations during the 19th century and later
+agricultural activities.
 These burials are excluded from most of the analysis due to missing and/or
@@
 Most of the data was obtained from the original publication [@podborsky2002],
-anthropological determination of sex and age of the bodies were corrected based on
+and the anthropological determination of the sex and age of the bodies was corrected based on an
 analysis by @dockalova2008. Carbon, nitrogen and strontium isotope
 ratios were collected in @whittle2013 and @richards2008.
-Further details on use and function of the artefacts were gathered from
+Further details on the use and function of the artifacts were gathered from
 @masclanslatorre2020.
-Information on localness of different bodies was adapted from @hrncir2020.
+Information on the localness of various bodies was adapted from @hrncir2020.
 
-Each burial was originally characterized by 20 variables recording counts of
-artefacts present in the graves.
-Some of  these original variables are omitted in the analysis due to their rare occurrence,
-for instance spondylus arm ring is present in a single burial only,
-thus being of no use in establishing relationships between pairs of burials.
-Only grave goods occuring in more than 5% of the graves are used in the analysis,
-this step reduces the count of the variables to 14.
+Each burial was originally characterized by 20 variables recording the number of
+artifacts present in the graves.
+Some of  these original variables are omitted in the analysis due to their rare occurrence;
+for instance, a spondylus arm ring is present in only a single burial,
+rendering it useless in establishing relationships between pairs of burials.
+Only grave goods occurring in more than 5% of the graves are used in the analysis;
+this step reduces the number of variables to 14.
 These include variables noting quantities of various types of pottery vessels in
 the graves, counts of chipped stone tools made of either local or non-local raw materials,
-presence of polished and beveled stone adzes, grinding slabs, pebbles,
-bone tools and different kinds of personal ornament.
-The personal ornament includes beads made of spondylus shell,
-spondylus pendants or oval shaped buckles.
-Last but not least presence of ochre pigment in the grave is also noted.
+the presence of polished and beveled stone adzes, grinding slabs, pebbles,
+bone tools and different kinds of personal ornaments.
+The personal ornaments include beads made of spondylus shells,
+spondylus pendants and oval shaped buckles.
+Last but not least, the presence of ocher pigment in the grave is also noted.
 Quantities of various *grave goods* are simplified into a binary matrix indicating
-presence or absence of a given trait.
+the presence or absence of a given trait.
 
 Another category of variables is derived from retrieved human skeletal remains.
-These include biological sex, age simplified into age groups and indicator
-variable whether the Strontium signal in the remains is of local origin.
+These include biological sex, age simplified into age groups and an indicator
+variable as to whether the strontium signal in the remains is of local origin.
 The variable representing the recorded sex is simplified into the following
 categories: female-sexed body (abbreviated *F*), male-sexed body (*M*),
-non adult (*n. a.*) and indetermined sex (*ind.*).
-Age of the body is also simplified into three categories: juveniles (*juv.*)
-comprising of groups Infans and Juvenis, adults (*ad.*, all bodies aged as Adultus)
-and mature bodies (*mat.*, anthropological categories Adultus-Maturus, Maturus and
-Senilis) and bodies of indetermined age (*ind.*).
-Last category is the indicator for the origin of the buried person.
-Local Strontium range is defined as two standard deviations from mean of Sr ranges of
-bodies with age category determined as Infans.
-Bodies faling within this range are labeled as local, outside the range as non-local
-and where data on Sr values are missing origin is not determined (*ind.*).
+non-adult (*n. a.*) and undetermined sex (*ind.*).
+The age of the body is also simplified into three categories: juveniles (*juv.*)
+comprised of the groups *Infans* and *Juvenis*, adults (*ad.*, all bodies aged as *Adultus*)
+and mature bodies (*mat.*, anthropological categories *Adultus-Maturus*, *Maturus* and
+*Senilis*) and bodies of an undetermined age (*ind.*).
+The last category is the indicator for the origin of the buried person.
+The local strontium range is defined as two standard deviations from the mean of Sr ranges of
+bodies with an age category determined as *Infans*.
+Bodies falling within this range are labeled as local, outside the range as non-local;
+where data on Sr values is missing, origin is not determined (*ind.*).
 
-We consider there three variables, sex, age and origin to be indicative of
-ones identity in the Neolithic.
-To explore the relations between the variables we use them both as individual
-categories and combined into following indicator variables -- (1) sex/age,
-(2) origin/sex, (3) origin/age and finally (4) origin/sex/age.
-Due to frequent missing or indetermined values we decided to limit the studied
-sample only to bodies with all of the categories determined and only categories
-that occure in more than 5% of burials.
+As mentioned above, we consider these three variables -- sex, age and origin -- to be indicative of
+an individual's identity in the Neolithic.
+To explore the relationships between the variables, we use them both as individual
+categories and combined into the following indicator variables -- (1) sex/age;
+(2) origin/sex; (3) origin/age; and (4) origin/sex/age.
+Due to frequent missing or undetermined values, we decided to limit the studied
+sample only to bodies with all of the categories determined (and only categories
+that occur in more than 5% of burials).
 
@@
 
-## Randomness of artefact combinations
+## Randomness of artifact combinations
 
@@
 
-To determine non-randomness of various artefact combinations in burials we conduct a
+To determine the non-randomness of various artifact combinations in burials, we conducted a
 permutation test on the presence/absence data matrix.
 The method based on generalized Monte Carlo tests was
-introduced to archaeology by @manly1996 and adapted by @sosna2008.
+introduced into archaeology by @manly1996 and adapted by @sosna2008.
 Our approach is based on both of these works.
-Original co-occurence matrix is permutated *n = 9999* times with constant
-row and column totals, i.e. both the count of artifacts in individual graves
-and sum of the given type of artifact across the burials are fixed.
+The original co-occurrence matrix was permuted 9,999 times with constant
+row and column totals, i.e., both the count of artifacts in individual graves
+and the sum of the given type of artifact across the burials are fixed.
 
-To examine which associations of variables, i.e. artefacts, are infrequent or even not present
-and which tend to appear together more often than it would be expected by random chance
-the co-occurences of observed and simulated data
-are summarized using two statistics suggested by @manly1996.
-The *S* statistic is a measure of overall deviation between expected and observed
-number of artefact occurences while
-the *v* statistic is a measure of difference between co-occurences of
-individual artefact with other artefacts
-[@manly1996, 475, for details see SI].
+To examine which associations of variables, i.e., artifacts, are infrequent or even not present
+and which tend to appear together more often than it would be expected by random chance,
+the co-occurrences of observed and simulated data
+are summarized using two statistics suggested by @manly1996 [475].
+The *S* statistic is a measure of overall deviation between the expected and observed
+number of artifact occurrences, while
+the *v* statistic is a measure of difference between co-occurrences of
+individual artifacts with other artifacts.
 
+To assess whether differences between artifacts that are non-randomly distributed
+are reflected in relationships between groups based on sex, age, origin and their combinations,
+we calculate mean Euclidean distance (ED) between pairs of burials in multidimensional
+space settings defined by the non-randomly distributed artifacts.
+The observed mean ED is compared to the mean ED distribution  derived by random
+reshuffling of the category labels between the burials 9,999 times with the constant
+probabilities of appearance of the given category label.
+<!-- To infer whether the observed and randomly created mean ED
+distributions are samples of the same population we use non-parametric
+Kruskal-Wallis test [@kruskal1952].-->
+For each of the categories, the deviation of the mean ED from its randomly generated
+distributions is expressed as the fraction of random values of ED larger than
+the one observed.
+The deviation is deemed significant if less than 5 % of
+random values are larger than the observed value.
+The process is a variation of the one described for resampling in the case of
+artifact combinations, and a similar approach is taken by @sosna2008 [346-347].
+
+<!--
 To assess relations between groups based on sex, age and origin
-statistics introduced by @sosna2008 are used.
+statistics introduced by @sosna2008 are used.-->
 
@@
 We approach the task as a point pattern analysis problem given that
-a burial (event) is simplified as its centroid location in a two dimensional space.
-A polygon of idealized excavation extent is used as a study area (a window).
+a burial (event) is simplified as its centroid location in two dimensional space.
+A polygon of the idealized excavation extent is used as the study area (a window).
 <!-- Using a window closely approximating locations of the burials was considered,
@@
 
-Overall structure of the cemeter is examined using
-<!-- either density based approaches or -->
-methods based on distances between neighboring burials. <!--[@osullivan2010, 121-154]-->
 <!-- Density based approaches, for instance quadrat tests,
@@
 independent on other point locations and different parameters influence its location.-->
-These methods are based on distances between pairs and triples of neighbours and
+The overall structure of the cemetery is examined using
+methods utilizing distances between neighboring burials.<!--[@osullivan2010, 121-154]-->
+These methods are based on distances between pairs and threesomes of neighbors and
 characterize the point pattern based on *second- and third-order properties*
@@
 These are in contrast to the assumption underlying *first-order effects* that
-any point location is independent on other point locations and
-defferent parameters influence its location [@osullivan2010, 126].
-<!-- [@nakoinz2016, 135-144; @osullivan2010, 130-139] -->
-In spatial analysis of settlement patterns individual locations of
-settlements (points) can depend on a number of parameters, for instance
-elevation, soil type etc.
-In case of a grave placement in a burial ground we presume that the choice of
-a graves location is either random or influenced predominantly by other (previous)
+any point location is independent of other point locations and
+different parameters influence its location [@osullivan2010, 126].
+In the spatial analysis of settlement patterns, individual locations of
+settlements (points) can depend on a number of parameters, e.g.,
+elevation, soil type, etc.
+In the case of a grave placement in a burial ground, we presume that the choice of
+the grave location is either random or influenced predominantly by other (previous)
 grave locations and
-is largely independent on the outside parameters.
-In other words *first-order properties* are of little to no importance in
+is largely independent of outside parameters.
+In other words, *first-order properties* are of little to no importance in
 our point of view.
 The *second-order properties* and possibly *third-order properties*
-are thus considered major factors in locating of the graves.
+are thus considered major factors in the location of the graves.
 <!-- It is possible that the position of burials was dependent on parameters that were not
@@
 We utilize *G*, *F*, Ripley's *K* [@ripley1976; @sayer2013] and *T* [@schladitz2000] functions.
-These functions are based on nearest neighbour distances and are commonly used
+These functions are based on the nearest neighbor distances and are commonly used
 to characterize point patterns in archaeology [@nakoinz2016, 135-144].
-The *G* function is a metric of how close events are and summarises
-cummulative frequency distribution of the nearest neighbour distances.
+The *G* function is a metric of how close events are and summarizes
+cumulative frequency distribution of the nearest neighbor distances.
 Because of its dependence on the intensity of events in the study area, it is
@@
 The *F* function, an empty-space function, is based on simulating a point pattern
-and calculating of distances between random locations and the observed locations.
+and calculating the distances between random locations and the observed locations.
 It is a measure of how far observed events are from any arbitrary location [@osullivan2010, 132-135].
-Ripley's *K* function is in comparison with the *G* function resilient to
+Compared to the *G* function, Ripley's *K* function is resilient to
 changes in the window size and works with all the proximal events up to a
-certain threshold, not only the nearest neighbours [@nakoinz2016, 138].
-We find it the most useful in most of archaeology use cases, cf. [@negre2018].
+certain threshold, not only the nearest neighbors [@nakoinz2016, 138].
+We find it the most useful in most archaeology cases, cf. @negre2018.
 Last but not least, the *T* function is a third-order parallel to the *K* function [@schladitz2000].
-For formulas and details on individual functions see the SI.
 
-Given statistics for independent random process are estimated using Monte Carlo
+The given statistics for the independent random process are estimated using Monte Carlo
 simulation with 999 iterations.
-Although smaller number of permutations should be sufficient to achieve
+Although a smaller number of permutations should be sufficient to achieve
 credible results
-[@baddeley2016, 384] we follow the approach taken by @sayer2013 [78].
-Ten most extreme realizations of the simulation are discarded as outliers so
-the significance level is 0.01 [@baddeley2016 410].
+[@baddeley2016, 384], we follow the approach taken by @sayer2013 [78].
+The ten most extreme realizations of the simulation are discarded as outliers, so
+the significance level is 0.01 [cf. @baddeley2016 410].
+The remaining realizations of given point processes are used to construct envelopes
+around function curves derived from observed data.
 
-To explore frequency of neighboring burials falling into same categories we
+To explore the frequency of neighboring burials falling into the same categories, we
 explore several methods.
-A modification of *K* function for multitype point patterns
+A modification of the *K* function for a multitype point pattern
 does not take into account all of the
@@
 same value within the given category are explored.
-Same as *K* function is indicative of the distances at which clustering or
+In the same way that the *K* function is indicative of the distances at which clustering or
 repulsive processes take place, the multitype *K* function shows these distances
-for given pairs of levels within a given categorical variable.
+for the given pairs of levels within a given categorical variable.
 The envelopes are constructed using the same procedure as described for other
@@
 
-To further establish the neighborhood structure we construct neighborhood graphs.
+To further explore the neighborhood structure, we construct neighborhood graphs.
 We use Delaunay triangulation and its subgraph, a Gabriel graph [@gabriel1969]
 with burial locations as nodes (Fig. \@ref(fig:plan-graphs)).
-The Delaunay triangulation is considered as a maximum number of neighbors per individual case
-scenario while the Gabriel graph the opposite.
-For each node, mean number of neighbors of a given category is determined and recorded.
+The Delaunay triangulation is regarded as the maximum number of neighbors per individual case
+scenario, while the Gabriel graph is the opposite.
+For each node, the mean number of neighbors of a given category is determined and recorded.
 To check how common the observed value is, a permutation test is performed
@@
 graph.
-The observed values of neighbor categories are compared against distributions
+The observed values of neighbor categories are compared to distributions
 derived from 999 permutations of the simulated data.
@@
 <!-- Percolation analysis -->
-Another approach we consider for establishing neigborhood relations is by
+Another approach we consider for establishing neighborhood relationships is
 constructing buffer zones around each of the burials.
 To avoid determinism in designating the radius of the buffer zone around each of
-the individuals we utilize percolation analysis.
-Percolation analysis is a data centric method used to identify clusters of points based
+the individuals, we utilize percolation analysis,
+a data centric method used to identify clusters of points based
 on spatial separation measured by Euclidean distance [@maddison2020, 269].
-The introduction of percolation analysis to archaeology is relatively recent,
-nevertheless we consider its effectiveness and usefulness in studies with varied scales
-well proven [@maddison2020].
-Percolation analysis is based on city clustering algorithm [@rozenfeld2008] and
+Although the introduction of percolation analysis in archaeology is relatively recent,
+we consider its effectiveness and usefulness in studies with varied scales
+well-proven [@maddison2020].
+Percolation analysis is based on a city clustering algorithm [@rozenfeld2008] and
 its generalization employing Euclidean distance [@rozenfeld2011].
-Nodes being in a given distance threshold (radius) are assigned into a cluster.
+Points falling within a given distance threshold (radius) are added to a cluster.
 This process is repeated for a set of thresholds while recording maximum
 cluster sizes and possibly other statistics.
-This allows to identify distances at which clusters exhibit rapid growth (absorbing
-large number of nodes) and distances at which the clusters remain relatively stable.
-For more details refer to @maddison2020 and associated R package [@schmidt2020].
+This makes it possible to identify distances at which clusters exhibit rapid growth (absorbing
+a large number of nodes) and distances at which the clusters remain relatively stable.
+For more details, refer to @maddison2020 and the associated R package [@schmidt2020].
 
-By means of percolation analysis we identify threshold distances at which cluster
-sizes are growing and clusters formed at these distances are further explored.
-For each cluster number of individual bodies of a given category is recorded.
+By means of percolation analysis, we identify threshold distances at which maximum cluster
+size is growing, and clusters formed at these distances are further explored.
+For each cluster, the number of individual bodies of a given category is recorded.
 The same procedure is repeated for 999 resamples with the given
 categories assigned at random with constant probabilities.
-Observed numbers of occurences deviating by two standard deviations from the mean of the
+The observed numbers of occurrences deviating by more than two standard deviations from the mean of the
 resampled results are regarded as strong evidence of non-random patterns being
 present in given clusters.
-To achieve stability of results and eliminate observations occuring due to
+To achieve the stability of results and eliminate observations occurring by
 chance in small clusters with low numbers of categories of interest,
 the whole resampling procedure is repeated three times and only observations present in
-all the iterations are reported with expected numbers of occurence and their
+all the iterations are reported with the expected numbers of occurrence and their
 deviations from observed values summarized as mean values.
@@
 Data is analyzed in an R environment for statistical computing and graphics
-[@rcore] using functions of several different packages, for the complete list
+[@rcoreteam2021] using functions of several different packages; for the complete list,
 see the supplementary information.
@@
 
-## Artefact combinations
+## Artifact combinations
 
-By counting co-occurences of analyzed variables on simulated data matrices and comparing
+By counting the co-occurrences of analyzed variables on simulated data matrices and comparing
 the distributions of the statistics proposed by @manly1996 with the values calculated for
-the original dataset we can establish whether the overall pattern of artefacts
-distribution in the graves is non-random and further identify artefacts that deviate
+the original data set, we can establish whether the overall pattern of artifact
+distribution in the graves is non-random and further identify artifacts that deviate
 from the random pattern.
-The overall deviation from expected co-occurences, i.e. the *S* statistic
+The overall deviation from expected co-occurrences, i.e., the *S* statistic
 [@manly1996, 475] is significantly large at the 0.05 alpha level.
 This suggests that there is strong evidence that the overall distribution of
-artefacts in different graves is not random. <!--(see Fig. \@ref(fig:sstat))-->
+artifacts in different graves is not random. <!--(see Fig. \@ref(fig:sstat))-->
 
-Examining distributions of individual artefacts we find that at the 0.05
+Examining distributions of individual artifacts, we find that at the 0.05
 significance level only adzes, grinding tools, ceramic bowls and spondylus beads
-are distributed randomly.
+are distributed non-randomly.
 Close to the 0.05 alpha level are distributions of bone tools and ceramic bottles
 (Fig. \@ref(fig:cooc-vstat)).
-Among the grave goods that are found to be randomly distributed in the graves
-across the burial ground are ochre pigment, lithics, pebbles, globular pots and
-special types of pottery vessel shapes and spondylus pendants and buckles.
+Among the grave goods found to be randomly distributed in the graves
+across the burial ground are ocher pigment, lithics, pebbles, globular pots and
+special types of pottery vessel shapes, spondylus pendants and buckles.
 
-The relationships between artefact types frequently occuring and missing together
-in graves are visualized in Fig. (\@ref(fig:cooc-nets)) with
-clusters established using fast-greedy algorithm [@newman2004] and strength of the
+The relationships between artifact types frequently occurring and missing together
+in graves are visualized in Fig. \@ref(fig:cooc-nets) with
+clusters established using a fast-greedy algorithm [@newman2004] and the strength of the
 given association expressed by the thickness of the edge connecting the nodes.
-Artifacts most frequently present with other artifacts are with 5 links grinding tools and
-ceramic bowls and with 4 links adzes, non-local lithic tools, ochre pigment and
-spondylus beads.
+Artifacts most frequently present with other artifacts are grinding tools and
+ceramic bowls with five links and adzes, non-local lithic tools, ocher pigment and
+spondylus beads with four links.
 The association is strongest between spondylus beads and (a) ceramic bottles and
 (b) pigment; and adzes and (a) non-local lithics and (b) bone tools.
-Artifacts that are frequently absent in the graves together are with 6 links
-local lithic tools, globular pots and ceramic bowls and spondylus buckels with
-5 and 4 links respectively.
+Artifacts frequently absent with other artifacts in the graves are  
+globular pots, local lithic tools, ceramic bowls and spondylus buckles
+with seven to four links respectively.
 The strongest antagonist relationships are between globular pots and (a) bowls and
 (b) adzes.
+
+Artifacts non-randomly distributed across the cemetery are further used to
+check whether significant differences exist between their occurrences in
+different groups defined based on origin, sex, age and their combinations.
+Exploring single attributes, the only differences are in the categorization to
+differently sexed bodies; there are significant differences
+between male- and female-sexed bodies.
+In the case of age or origin, no significant differences are observed.
+If we combine sex and age categories, the distinction between juvenile, adult and mature
+female-sexed bodies and adult male-sexed bodies is evident.
+Another observed dichotomy is present between adult males and juveniles (non-adults).
+The same relationships are essentially present in the case of combinations of sex and origin attributes.
+Local male-sexed bodies have significantly different combinations of non-randomly
+distributed artifacts from both local and non-local female-sexed bodies and non-adults.
+If all the categories are combined the divisions are even more evident.
+Local male-sexed bodies in the *Adultus* age category deviate in the artifact distributions
+from adult local and non-local females, *Maturus/Senilis* non-local female-sexed bodies
+and juveniles.
 
@@
 expected counts (for the details see supplementary information).
-Inconclusive result of the quadrat test is not suprising because as was noted
+Inconclusive result of the quadrat test is not surprising because as was noted
 earlier, we do not expect *first-order* properties to play major roles in
 placing of the individual burials. -->
-Results of functions based on nearest neighbors are reported in Fig. \@ref(fig:pp-funs).
-Both *G* and  *F* functions do slightly deviate from the curve expected
-under independent random process
-but neither traverses the boundaries of the envelope.
-Thus the cummulative nearest-neighbor distance (*G*) as well as the
-empty-space function *F* are well approximated by
-random process and do not show evidence for spatial clustering nor repelent functions
+The results of functions based on nearest neighbors are reported in Fig. \@ref(fig:pp-funs).
+Both *G* and  *F* functions slightly deviate from the curve expected
+under an independent random process,
+but neither crosses the boundaries of the envelope.
+Thus, the cumulative nearest-neighbor distance (*G*) as well as the
+empty-space function (*F*) are well approximated by
+the random process and do not show evidence for spatial clustering or repellent functions
 taking place.
-On the contrary
+On the contrary,
 the *K* and *T* functions show clear deviations from the expected curve and both exceed
-the boundaries of given envelopes at distances around or smaller than 5 meters.
+the boundaries of the given envelopes at distances around or smaller than 5 meters.
 <!-- in case of *K* and slightly more in case of *T* function.-->
-This means that numbers of nearest neighbors at this and larger distances is
+This means that the number of nearest neighbors at this and larger distances is
 significantly higher than we would expect under complete spatial randomness.
-The overall point pattern thus shows clear signs of clustering process taking place.
+The overall point pattern thus shows a clear signs of clustering process taking place.
 
-This means that the locations of new burials were not chosen at random
-but were affected by the positions of previous burials.
-The pattern begins to cluster at the distance of around 5 meters, meaning
-that beginning at this distance, the probability of finding a neighboring burial is
+This suggests that the locations for new burials were not chosen at random
+but were influenced by the positions of previous burials.
+The pattern begins to cluster at a distance of around 5 meters, i.e.,
+beginning at this distance, the probability of finding a neighboring burial is
 much higher than it would be if the burials were placed randomly in space.
-The community burying their dead at the cemetery of Vedrovice *Široká u lesa*
+The community burying their dead at the *Široká u lesa* cemetery in Vedrovice
 did so in an organized manner taking into account previous burials at the same place,
-meaning their positions were known, probably marked on the surface somehow.
-This would imply existence of non-perishable marks on the surface of the cemetery and/or their
-maintenance from time to time.
+meaning their positions were known, probably marked somehow on the surface.
+This would imply the existence of non-perishable marks on the surface of the cemetery and/or their maintenance from time to time.
 
-Results of percolation analysis reveal more detailed view of the clustering process
-taking place, see Fig. \@ref(fig:perc-distance).
-The distances where relative stability of clusters emerges are labeled.
-The curve reveals a steady growth up until
+The results of a percolation analysis reveal a more detailed view of the clustering process
+taking place; see Fig. \@ref(fig:perc-distance).
+The distances at which the relative stability of clusters emerges are labeled.
+The curve reveals a steady growth up until a
 radius of 5 meters and then a large skip towards a single cluster
@@
 
-The results of multitype *K* function show deviations from random processes for
+The results of the multitype *K* function show deviations from random processes for
 most of the observed categories in Fig. \@ref(fig:pp-crossk).
-The distances at which these deviations start occuring are in most cases rather large,
-well over 10 meters and considering the fact that at the 10.4 meters radius
-around individual points the percolation analysis proved existence
-of single large cluster encompassing almost all of the points, these
-observations correspond with the deviations observed for single *K* function.
-The only notable deviation we consider worth mentioning is in case of age
-category in pairs of juveniles and in sex category between pairs of non-adults.
+The distances at which these deviations start occurring are in most cases rather large,
+well over 10 meters and, considering the fact that at the 10.4 meters radius
+around individual points the percolation analysis proved the existence
+of a single large cluster encompassing almost all of the points, these
+observations correspond to the deviations observed for the single *K* function.
+The only notable deviation we consider noteworthy is in the case of age
+category in pairs of juveniles and in the sex category between pairs of non-adults.
 Here the deviation from random processes is observed at smaller distances than
-in case of other variables at distances around 5 meters.
-This proves tighter bound cluster of juveniles in south-eastern part of the cemetery.
+in the case of other variables at distances around 5 meters.
+This proves a tighter bound cluster of juveniles in the southeast part of the cemetery.
 
-Multitype *K* for combination of age and sex categories suggests clustering process
+Multitype *K* for the combination of age and sex categories suggests a clustering process
 between bodies determined as adult females and juveniles and vice versa.
 The relationship between pairs of non-adults/juveniles is also confirmed.
-Several other observed deviations occure at large distances and are same as
-in case of multitype *K* for single categories.
+Several other observed deviations occur at large distances and the explanation for
+these is the same as in the case of multitype *K* for single categories.
 
-In case of neighborhood structure derived from Delaunay triangulation (maximum
+In the case of neighborhood structure derived from Delaunay triangulation (maximum
 number of neighbors scenario) and
-Gabriel graph (minimum number of neighbors scenario, Fig. \@ref(fig:plan-graphs))
+Gabriel graph (minimum number of neighbors scenario, Fig. \@ref(fig:plan-graphs)),
 most of the observed relationships are not significantly different from those
 derived by randomization experiment.
-In the Delaunay triangulation the significant relations are observed in two cases
+In the Delaunay triangulation the significant relationships are observed in two cases
 between (1) pairs of non-adults and (2) female-sexed bodies and non-adults.
-The first observation can be again given into relation with higher concentration of
-non-adults in south-eastern space of the burial ground.
-In regard to sex category Gabriel graph shows an association
+The first observation can again be connected with a higher concentration of
+non-adults in the southeast space of the burial ground.
+In regard to the sex category, the Gabriel graph shows an association
 between female-sexed and male-sexed bodies.
-In only one case are the observations from both of the graphs significant and
-that is between pairs of bodies of mature age category.
+In only one case are the observations from both of the graphs significant, and
+that is among pairs of bodies in the mature age category.
 
-The combinations of categories in Delaunay graph show significant relations for
-pairs of bodies labeled as (1) females from maturus/senilis category with
-juvenile females and (2) unsexed juveniles in general
-and as already noted at several places, between (3) pairs of bodies labeled as
-juveniles/non-adults.
-Gabriel graph in the sex/age category shows relations between (1) adultus female and male bodies,
-(2) maturus female and male bodies and (3) maturus female bodies.
-The origin/sex combination of categories produces significant deviation from
+The combinations of categories in the Delaunay graph show significant relationships for
+pairs of bodies labeled as *Maturus/Senilis* category females with
+(1) juvenile females and with (2) unsexed juveniles.
+The relationship between pairs of juvenile/non-adult bodies is again confirmed.
+The Gabriel graph in the sex/age category shows relationships between (1) *Adultus* female and male bodies;
+(2) *Maturus* female and male bodies; and (3) pairs of *Maturus* female bodies.
+The origin/sex combination of categories produces a significant deviation from
 expected frequencies for (1) non-local female bodies and local non-adults in
-Delaunay triangulation and in Gabriel graph for (1) non-local female and local male
+Delaunay triangulation and in a Gabriel graph for (1) non-local female and local male
 bodies and vice versa.
-The origin/age combination shows significant departures only in Gabriel graph for
-local adultus bodies with (1) non-local maturus bodies and (2) local juvenile bodies.
-Last but not least, incorporating all of the categories into the networks we observe
-significant results in both of the graphs (1) non-local maturus female bodies and
-local juvenile (non-adult) bodies and  in case of Gabrie graph relation
-between (2) non-local adultus female and local maturus male bodies.
+The origin/age combination shows significant departures only in the Gabriel graph for
+local *Adultus* bodies neighboring with (1) non-local *Maturus* bodies and (2) local juvenile bodies.
+Last but not least, incorporating all of the categories into the networks, we observe
+significant results in both of the graphs for non-local *Maturus* female bodies and
+local juvenile (non-adult) bodies and in the case of the Gabriel graph a relationship
+between non-local *Adultus* female and local *Maturus* male bodies.
 
 Clusters formed at different percolation distances
-show following deviations from expected numbers of
-occurences of given categories at different scales.
-Regarding the determination of sex of the bodies, at the threshold of 3.9 meters
-we observe that in the largest cluster (nr. 3, Fig. \@ref(fig:perc-plan))
-there are more non adults and less male-sexed bodies and in the second largest
-cluster (nr. 10) there are slightly more male-sexed bodies than would
-be expected under random process.
-The overrepresentation of non-adult bodies is repeated in the following
-threshold distance of 4.6 meters in the largest cluster (nr. 3).
-Last but not least, in the cluster nr. 8 (north of the cemetery) at the distance
-of 5.5 meters female-sexed bodies are more frequent than we would expect.
-The age categories correspond to the previous obsevations where
-most evident is the overrepresentaion of juvenile bodies at the 3.9 meters threshold in
-cluster nr. 3.
-Exploring the origin of the bodies based on Strontium isotope signature we
-find that cluster in the south-western part at threshold distances 3.9 and 5.5 meters
-(nr. 10 and 5 respectively) consists of less non-local bodies than would be expected.
-Lastly at the distance radius 2.8 meters there are more local bodies in the cenral
-cluster of the cemetery (nr. 8).
+show the following deviations from the expected number of
+occurrences of given categories at different scales.
+Regarding the determination of the sex of the bodies, at the threshold of 3.9 meters
+we observe that in the largest cluster (no. 3, Fig. \@ref(fig:perc-plan))
+there are more non-adults and less male-sexed bodies and in the second largest
+cluster (no. 10) there are slightly more male-sexed bodies than would
+be expected under a random process.
+The over-representation of non-adult bodies is repeated in the following
+threshold distance of 4.6 meters in the largest cluster (no. 3).
+Last but not least, in cluster no. 8 (north of the cemetery) at the distance
+of 5.5 meters, female-sexed bodies are more frequent than we would expect.
+The age categories correspond to the previous observations where
+the over-representation of juvenile bodies at the 3.9-meter threshold in
+cluster no. 3 is most evident.
+Exploring the origin of the bodies based on the strontium isotope signature, we
+find that the cluster in the southwest part at threshold distances of 3.9 and 5.5 meters
+(no. 10 and 5 respectively) consists of less non-local bodies than would be expected.
+Finally, at a radius of 2.8 meters there are more local bodies in the central
+cluster of the cemetery (no. 8).
 
@@
 In the largest cluster across percolation thresholds at 3.9 and 4.6 meters
-(nr. 3) there is abundance of (1) local non-adults/juveniles and (2) non-local
-female sexed bodies of maturus age category and on the other hand
-lesser numbers of (3) juvenile female-sexed bodies.
-The central cluster across distances 2.8, 3.9 and 4.6 (nr. 8, 10 and 7 respectively)
-shows higher numbers of local male- and female-sexed bodies of adultus category and in case of
-female bodies maturus age categories.
-The south-western cluster (nr. 10, 7 and 5) at percolation distances 3.9, 4.6 and 5.5
-is clearly dominated by local male-sexed bodies of maturus age category.
-Last but not least, in case of all the categories combined together a cluster in
-the western part of the cemetery at percolation threshold 4.6 and 5.5 (nr. 8 and 6 respectively)
-reveals higher counts of non-local female-sexed bodies of adultus age category.
+(no. 3), there is an abundance of (1) local non-adults/juveniles and (2) non-local
+female-sexed bodies of the *Maturus* age category and, on the other hand,
+lower numbers of (3) juvenile female-sexed bodies.
+The central cluster across the distances of 2.8, 3.9 and 4.6 meters (no. 8, 10 and 7 respectively)
+shows higher numbers of local male- and female-sexed bodies of the *Adultus* category
+and, in the case of female bodies, the *Maturus* age categories.
+The southwest cluster (no. 10, 7 and 5) at percolation distances 3.9, 4.6 and 5.5 meters
+is clearly dominated by local male-sexed bodies of the *Maturus* age category.
+Last but not least, in the case of all the categories combined together, a cluster in
+the western part of the cemetery at the percolation threshold of 4.6 and 5.5 meters (no. 8 and 6 respectively)
+reveals higher counts of non-local female-sexed bodies of the *Adultus* age category.
 
@@
 
-# Conclusions
+# Conclusions and discussion
 
 Our goal was to identify non-random differences between burials both in terms
-of the artefact contents of the graves and position within the cemetery.
-The burials or more precisely the buried bodies were divided into categories
-based on three variables commonly connected to the identity in the Neolithic period.
+of the artifact contents of the graves and their position within the cemetery.
+The burials or, more precisely, the buried bodies were divided into categories
+based on three variables commonly connected to identity in the Neolithic period.
 These categories are biological sex, age and origin.
-The exploration was done both for the categories as single variables and for
+The exploration was conducted both for the categories as single variables and for
 all of their combinations.
 
-<!-- Artefact combinations -->
+<!-- Artifact combinations -->
+In the case of the grave good analysis, we explored the associations between artifacts and
+different categories of the bodies not to establish or reveal any new relationships
+but merely to explore whether the known associations between sexes and specific grave goods
+are influenced by factors such as age and the local or non-local origin of the buried person.
+We did not explore the relationships between grave goods and burial attributes at the
+individual artifact scale, so to say; instead, we used an overall measure (Euclidean distance)
+of between burial distance in terms of non-randomly distributed artifacts.
+By establishing whether the mean ED between categories defined by the grouping attributes
+mentioned above falls within or defies the expected distribution derived by
+randomization experiment, we identified deviations that are not random.
+
+The most frequently reappearing is the difference between local adult male-sexed bodies and
+both local and non-local female-sexed ones and unsexed juveniles.
+This relationship shows that sex classification of the body, although being
+accentuated by local origin and age category, structures the artifact assemblage
+in the grave very well.
 
 <!-- Spatial pattern -->
-Spatial organization of the cemetery was explored using point pattern analysis
+The spatial organization of the cemetery was explored using point pattern analysis
 techniques.
@@
 did not show deviations from complete spatial random processes.
-On the other hand the methods including all the distances between events in the
-given area  (*K* and *T* functions) indicated clustering processes are taking place.
-*K* function for multivariate point patterns was also used in order to identify
+On the other hand, the methods, including all the distances between events in the
+given area  (*K* and *T* functions), indicated that clustering processes are taking place.
+The *K* function for multivariate point patterns was also used in order to identify
 clustering processes between events of different categories.
 We consider this not useful in most cases because the distances at which interesting
-clustering processes are suggested imply existence of singular cluster across
-the cemetery and a finer scale is of much higher interest for our endeavour.
+clustering processes are suggested imply the existence of a single cluster across
+the cemetery and a finer scale is of much greater interest for our work.
 
-Thus to identify whether close burials are more similar in terms of categories
-outlined above two neighborhood graphs, Delaunay triangulation and Gabriel
-graph were used.
-They are examples of high and low neighbors case scenario respectively.
+Thus, to identify whether close burials are more similar in terms of the categories
+outlined above, two neighborhood graphs – Delaunay triangulation and a Gabriel
+graph – were used.
+They are examples of high and low neighbor case scenario, respectively.
 Most of the observations from the graphs proved not significantly different from
-randomization experiment.
+the randomization experiment.
 There are associations between non-local female-sexed
-bodies of maturus age category and (local) juveniles in both of the graphs.
-This observation goes in hand with association of female-sexed bodies
-and non-adults in Delaunay triangulation.
+bodies of the *Maturus* age category and (local) juveniles in both of the graphs.
+This observation goes hand in hand with the association of female-sexed bodies
+and non-adults in the Delaunay triangulation.
 This is not trivial because in general non-adult bodies are
-concentrated in the south-eastern part within the cemetery whereas female-sexed
+concentrated in the southeast part of the cemetery, whereas female-sexed
 bodies are scattered all around the area of the cemetery.
-Similarly individuals determined as of maturus/senilis age category are in contrast to
+Similarly, individuals determined as belonging to the *Maturus/Senilis* age category are in contrast to
 juvenile bodies concentrated in the western half of the cemetery.
-Furthermore in Gabriel graph there is a clear association between non-local adultus
-female-sexed bodies and local maturus male-bodies.
-This agrees with general association of female- and male-sexed bodies of
-adultus and maturus/senilis age categories in the Gabriel graph.
+Furthermore, the Gabriel graph shows a clear association between non-local *Adultus*
+female-sexed bodies and local *Maturus* male-bodies.
+This corresponds to the general association of female- and male-sexed bodies of
+the *Adultus* and *Maturus/Senilis* age categories in the Gabriel graph.
 
-Most detailed view on the clustering within the cemetery and what kinds of
+The most detailed view of the clustering in the cemetery and what kinds of
 bodies are prevalent in the given clusters is achieved by the percolation analysis.
 The largest cluster across several percolation distances consists of more
-non-adult/juvenile bodies and non-local female-sexed bodies of maturus category
-on one hand and on the other the male-sexed bodies are underrepresented here.
+non-adult/juvenile bodies and non-local female-sexed bodies of the *Maturus* category
+on the one hand while, on the other, the male-sexed bodies are underrepresented here.
 This is in full accordance with previous observations.
-The central cluster visible at small percolation distances showes overrepresentation of
+The central cluster visible at small percolation distances shows the overrepresentation of
 local male- and female-sexed bodies.
-Also the south-western cluster has higher counts of local male-sexed bodies of maturus age category and
-low amount of non-local bodies.
-On the other hand non-local female-sexed bodies of adultus age category are
+Also, the southwest cluster has higher counts of local male-sexed bodies of the *Maturus* age category and
+a low number of non-local bodies.
+On the other hand, non-local female-sexed bodies of the *Adultus* age category are
 overrepresented in the western cluster neighboring with the previously mentioned one.
@@
 The spatial analysis clearly proved that there is a clustering process taking place
-in the cemetery and what more, age, sex and origin of the buried bodies play
+in the cemetery and what more, the age, sex and origin of the buried bodies play
 structuring roles in the arrangement of the burials.
-The patterns are not too strict but some appear to be quite constant across a
-range of methods used.
+The patterns are not too strict, but some appear to be quite constant across the a
+range of applied methods.
 
@@
 
-# Discussion
+<!-- # Discussion -->
+
+We have successfully shown that all of the categories considered
+to be somehow related to identity in the Neolithic and especially their combinations
+play substantive roles in structuring the assemblage of grave goods deposited
+within the burials.
+Some of the relationships between the explored categories and
+artifacts are well known at the LBK burial grounds,
+for instance, the association between
+male burials and polished stone adzes [e.g., @bickle2020; @bentley2012], and some
+are known specifically from the *Široká u lesa* cemetery in Vedrovice, e.g.,  the dichotomy
+between globular pottery shapes being associated with female-sexed bodies and,
+on the other hand, bowls with male-sexed bodies [@kvetina2004].
+Our goal was not to establish any new connections between genders or other burial
+attributes and grave good categories, but to use these associations
+to see which groups of burials are well separated based on the grave goods.
+
+First of all, we have shown that the overall distribution of the artifacts in the graves
+was not random; the grave goods distributed non-randomly were identified and
+subsequently only these were used in the further analysis, effectively removing any
+possible *noise* from artifacts distributed randomly.
+Our main premise here is that the identity of the dead is based mostly on attributes like sex,
+age and origin defined on the basis of the strontium isotope signal.
+We view these *layers* of identity as responsible for differences in
+artifact compositions of different graves as well as, to a certain degree, a structuring
+factor for the overall spatial organization of the cemetery.
+In the case of the grave goods, this relationship is largely dominated by a category
+of adult local male-sexed bodies that are distinctly different from both female-sexed bodies
+of all age categories and juveniles.
+This is caused by the aforementioned artifact characteristic of male burials, the stone adze.
+@bickle2020 completely removes polished and beveled stone tools from her analysis
+in order to remove their dominance over other patterns.
+In our case, adzes are one of four non-randomly occurring artifacts and removing
+them from the analysis completely would render it difficult to note any
+of the visible patterns.
+
+In the case of spatial analysis, we explore several approaches.
+The nearest-neighbor methods are a great tool in analyzing the overall cemetery, but
+on a small scale they do not prove useful for discovering any interesting relationships
+between studied *layers* of identity.
+By far the most useful method in analyzing the intra-site spatial arrangement
+is percolation analysis.
+Another important point is the use of resampling methods that allow for the identification
+of random and non-random patterns, both in grave good analysis and spatial organization.
+
+<!--Traditional GIS based methods were already applied to study the cemetery
+[@kvetina2004] the techniques applied here reveal some trends that were previously-->
+
+We have shown that not only sex, which is traditionally perceived
+as the most important structuring proxy for identity differentiation, but also
+age and, quite importantly, also the origin of the deceased played substantial
+roles in equipping the graves with various types of artifacts by the Vedrovice community.
+In regard to the spatial arrangement of the cemetery, we demonstrated that there are
+prevalent clustering processes taking place across the cemetery and also groups
+based on certain identities tend to be buried in clusters.
+These observations are based on the extensive use of resampling approaches that
+allowed us to distinguish between patterns occurring solely by chance and
+patterns significantly deviating from randomness.
 
@@
 
+```{r fig-widths}
+fig_wid <- c(".3\\textwidth",
+             ".6\\textwidth",
+             ".8\\textwidth",
+             "\\textwidth")
+```
+
 <!-- randomization -->
+```{r sstat, fig.cap="Distribution of simulated and observed *S* statistic", out.width=fig_wid[1]}
+knitr::include_graphics(here("plots/cooc_sstat.pdf"))
+```
 
-<!--```{r sstat, fig.cap = "Distribution of simulated and observed *S* statistic"}
-# knitr::include_graphics(here("plots/s_statistic.pdf"))
-```-->
-
-```{r cooc-vstat, fig.cap = "Simulated and observed *V* statistics for different variables"}
+```{r cooc-vstat, fig.cap="Simulated and observed *V* statistics for different variables", out.width=fig_wid[4]}
 knitr::include_graphics(here("plots", "cooc_vstats.pdf"))
@@
 ```{r label-cooc-nets}
-label_cooc_nets <- "Networks of variables frequently present together and exluding each other, non-randomly occuring artefacts in bold"
+label_cooc_nets <- paste("Networks of variables frequently present together and excluding each other.",
+                         "1 – bottle, 2 – bowl, 3 – globular pot, 4 – polished stone tool, 5 – local lithics,",
+                         "6 – non-local lithics, 7 – pebble, 8 – grinding slab, 9 – bone tool,",
+                         "10 – U-shaped pendant, 11 – spondylus beads, 12 – O-shaped buckle.",
+                         "Images adapted from Podborský 2002")
 ```
-```{r cooc-nets, fig.cap=label_cooc_nets }
-knitr::include_graphics(here("text/figs", "cooc_nets.pdf"))
+```{r cooc-nets, fig.cap=label_cooc_nets, out.width=fig_wid[4]}
+knitr::include_graphics(here::here("plots", "cooc_nets_fin.jpeg"))
 ```
+
 <!-- plans -->
-
-```{r plan-sex, fig.cap="Plan of the cemetery, kernel density estimation for differently sexed bodies"}
+```{r plan-sex, fig.cap="Plan of the cemetery, kernel density estimation for differently sexed bodies", out.width=fig_wid[4]}
 knitr::include_graphics(here("text/figs/plan_sex.pdf"))
@@
 
-```{r plan-age, fig.cap="Plan of the cemetery, kernel density estimation for bodies of different ages"}
+```{r plan-age, fig.cap="Plan of the cemetery, kernel density estimation for bodies of different ages", out.width=fig_wid[4]}
 knitr::include_graphics(here("text/figs/plan_age.pdf"))
@@
 
-```{r plan-orig, fig.cap="Plan of the cemetery, kernel density estimation for bodies of different origins"}
+```{r plan-orig, fig.cap="Plan of the cemetery, kernel density estimation for bodies of different origins", out.width=fig_wid[2]}
 knitr::include_graphics(here("text/figs/plan_origin.pdf"))
@@
 ```
-```{r pp-funs, fig.cap=label_pp_funs}
+```{r pp-funs, fig.cap=label_pp_funs, out.width=fig_wid[4]}
 knitr::include_graphics(here("text/figs/pp_funs.pdf"))
@@
 ```{r label-pp-crossk}
-label_crossk <- paste("Multitype K function for different categorical variables, for legend see previous Fig.")
+label_crossk <- paste("Multitype K function for different categorical variables; for legend, see previous Figure")
 ```
-```{r pp-crossk, fig.cap=label_crossk}
+```{r pp-crossk, fig.cap=label_crossk, out.width=fig_wid[3]}
 knitr::include_graphics(here("text/figs/pp_crossk.pdf"))
@@
 <!-- neighborhood -->
-```{r plan-graphs, fig.cap="Neighborhood graphs: Delaunay triangulation and Gabriel graph"}
+```{r plan-graphs, fig.cap="Neighborhood graphs: Delaunay triangulation and Gabriel graph", out.width=fig_wid[4]}
 knitr::include_graphics(here("text/figs/plan_graphs.pdf"))
-```
-
-```{r label-nb-sex}
-label_nb_sex <- paste(
-  "Mean neighbors in a given category of sex;",
-  "*Gray area* -- density estimation for the simulated data, *vertical line* -- observed value"
-  # "*F* -- female-sexed body, *M* -- male-sexed body, *n. a.* -- non-adult body"
-)
-```
-```{r nb-sex, fig.cap=label_nb_sex}
-knitr::include_graphics(here("text/figs/nb_sex.pdf"))
-```
-
-```{r label-nb-age}
-label_nb_age <- paste(
-  "Mean neighbors in a given simplified age category"
-  # "*Gray area* -- density estimation for the simulated data, *vertical line* -- observed value;",
-)
-```
-```{r nb-age, fig.cap=label_nb_age}
-knitr::include_graphics(here("text/figs/nb_age.pdf"))
-```
-
-```{r label-nb-orig}
-label_nb_orig <- paste(
-  "Mean neighbors in a given category of origin based on Sr isotopes"
-  # "*Gray area* -- density estimation for the simulated data, *vertical line* -- observed value;",
-  # "*F* -- female-sexed body, *M* -- male-sexed body, *n. a.* -- non-adult body"
-)
-```
-```{r nb-orig, fig.cap=label_nb_orig}
-knitr::include_graphics(here("text/figs/nb_orig.pdf"))
 ```
@@
 ```
-```{r perc-distance, fig.cap=label_perc_dist}
+```{r perc-distance, fig.cap=label_perc_dist, out.width=fig_wid[2]}
 knitr::include_graphics(here("text/figs/perc_distance.pdf"))
@@
 
-```{r perc-plan, fig.cap="Clusters formed at given percolation threshold distances"}
+```{r perc-plan, fig.cap="Clusters formed at given percolation threshold distances", out.width=fig_wid[3]}
 knitr::include_graphics(here("text/figs/perc_plan_ids.pdf"))
 ```
-
-<!-- ======================================================================= -->
-<!-- TABLES --><!-- TABLES --><!-- TABLES --><!-- TABLES --><!-- TABLES -->
-<!-- ```{r nb-tab-sex} -->
-<!-- tab <- read_csv(here("data/temp/nb_sex_p.csv")) %>% -->
-<!--          mutate(relation = str_c(from, to, sep = " - ")) %>% -->
-<!--   select(relation, everything(), -from, -to) -->
-
-<!-- tnames <- c("relation", rep(c("mean", "p"), 3)) -->
-
-<!-- knitr::kable(tab, booktabs = TRUE, col.names = tnames, linesep = "", caption = "Results...") %>% -->
-<!--   kableExtra::add_header_above(c(" ", "Buffer" = 2, "Delaunay" = 2, "Gabriel" = 2)) -->
-<!-- ``` -->
 

### 2021-06-04 text/article_vedrovice.Rmd
cite packages

@@
 credible results
-[@baddeley2016, 384], we follow the approach taken by @sayer2013 [78].
+[@baddeley2015, 384], we follow the approach taken by @sayer2013 [78].
 The ten most extreme realizations of the simulation are discarded as outliers, so
-the significance level is 0.01 [cf. @baddeley2016 410].
+the significance level is 0.01 [cf. @baddeley2015 410].
 The remaining realizations of given point processes are used to construct envelopes
@@
 Data is analyzed in an R environment for statistical computing and graphics
-[@rcoreteam2021] using functions of several different packages; for the complete list,
-see the supplementary information.
+[@rcoreteam2021] using functions of several different packages; 
+the most frequently used are packages from the *tidyverse* family [@wickham2019a].
+Packages *sf* [@pebesma2018], *spatstat* [@baddeley2015], *spdep* [@bivand2018] and 
+*concaveman* [@gombin2020] are used to perform various spatial analysis tasks; 
+*igraph* [@csardi2006a] to construct and analyze networks;
+*vegan* [@oksanen2020] and *randomizr* [@coppock2019] in the randomization experiments.
 
@@
              ".8\\textwidth",
-             "\\textwidth")
+             ".98\\textwidth")
 ```
@@
 ```{r sstat, fig.cap="Distribution of simulated and observed *S* statistic", out.width=fig_wid[1]}
-knitr::include_graphics(here("plots/cooc_sstat.pdf"))
+knitr::include_graphics(here("plots", "cooc_sstat.pdf"))
 ```
@@
 ```{r cooc-nets, fig.cap=label_cooc_nets, out.width=fig_wid[4]}
-knitr::include_graphics(here::here("plots", "cooc_nets_fin.jpeg"))
+knitr::include_graphics(here("plots", "cooc_nets_fin.jpeg"))
 ```
@@
 ```{r plan-sex, fig.cap="Plan of the cemetery, kernel density estimation for differently sexed bodies", out.width=fig_wid[4]}
-knitr::include_graphics(here("text/figs/plan_sex.pdf"))
+knitr::include_graphics(here("plots", "plan_sex.pdf"))
 ```
@@
 ```{r plan-age, fig.cap="Plan of the cemetery, kernel density estimation for bodies of different ages", out.width=fig_wid[4]}
-knitr::include_graphics(here("text/figs/plan_age.pdf"))
+knitr::include_graphics(here("plots", "plan_age.pdf"))
 ```
@@
 ```{r plan-orig, fig.cap="Plan of the cemetery, kernel density estimation for bodies of different origins", out.width=fig_wid[2]}
-knitr::include_graphics(here("text/figs/plan_origin.pdf"))
+knitr::include_graphics(here("plots", "plan_origin.pdf"))
 ```
@@
   "Functions *G*, *F*, *K* and *T*.",
-  "*Dashed line* -- theoretical curve under complete spatial randomness,
-  *solid line* -- empirical curve, *gray area* -- envelope")
+  "*Dashed line* - theoretical curve under complete spatial randomness;
+  *solid line* - empirical curve; *gray area* - envelope")
 ```
 ```{r pp-funs, fig.cap=label_pp_funs, out.width=fig_wid[4]}
-knitr::include_graphics(here("text/figs/pp_funs.pdf"))
+knitr::include_graphics(here("plots", "pp_funs.pdf"))
 ```
@@
 ```{r pp-crossk, fig.cap=label_crossk, out.width=fig_wid[3]}
-knitr::include_graphics(here("text/figs/pp_crossk.pdf"))
+knitr::include_graphics(here("plots", "pp_crossk.pdf"))
 ```
@@
 ```{r plan-graphs, fig.cap="Neighborhood graphs: Delaunay triangulation and Gabriel graph", out.width=fig_wid[4]}
-knitr::include_graphics(here("text/figs/plan_graphs.pdf"))
+knitr::include_graphics(here("plots", "plan_graphs.pdf"))
 ```
@@
 ```{r perc-distance, fig.cap=label_perc_dist, out.width=fig_wid[2]}
-knitr::include_graphics(here("text/figs/perc_distance.pdf"))
+knitr::include_graphics(here("plots", "perc_distance.pdf"))
 ```
@@
 ```{r perc-plan, fig.cap="Clusters formed at given percolation threshold distances", out.width=fig_wid[3]}
-knitr::include_graphics(here("text/figs/perc_plan_ids.pdf"))
+knitr::include_graphics(here("plots", "perc_plan_ids.pdf"))
 ```

### 2021-06-04 code/02-spat_dataprep.R
EPS figures export

@@ -51,7 +51,7 @@ ved_exc <- st_cast(
 
 # window_rect <- owin(xrange = c(10, 22), yrange = c(10, 22.7))
 # window_poly <- spatstat::as.owin(st_cast(ved_sf_hull, "POLYGON"))
-window_exc <- spatstat::as.owin(ved_exc)
+window_exc <- spatstat.geom::as.owin(ved_exc)
 
 # plot
 # g_exc <- ggplot() + 
@@ -115,6 +115,8 @@ g_exc +
   facet_wrap(~origin)
 
 ggsave(here("plots", "plan_origin.pdf"), width = 12, height = 6, units = "cm")
+ggsave(here("plots", "plan_origin.eps"), width = 12, height = 6, units = "cm")
+
 
 # plan age
 g_exc + 
@@ -126,6 +128,7 @@ g_exc +
   facet_wrap(~age_sim)
 
 ggsave(here("plots", "plan_age.pdf"), width = 19, height = 6, units = "cm")
+ggsave(here("plots", "plan_age.eps"), width = 19, height = 6, units = "cm")
 
 # plan sex
 g_exc +
@@ -137,6 +140,7 @@ g_exc +
   facet_wrap(~sex)
 
 ggsave(here("plots", "plan_sex.pdf"), width = 19, height = 6, units = "cm")
+ggsave(here("plots", "plan_sex.eps"), width = 19, height = 6, units = "cm")
 
 # write layouts -----------------------------------------------------------
 

### 2021-06-04 code/03-randomization.R

@@ -183,6 +183,22 @@ ggsave(here("plots", "cooc_vstats.pdf"),
        plot = v_plot, 
        width = 19, height = 20, units = "cm")
 
+# EPS does not allow transparency...
+v_plot_eps <- ggplot(v_exp_g, mapping = aes(x = value)) +
+  geom_density(fill = "gray80", color = NA) +
+  geom_vline(data = v_obs_g, mapping = aes(xintercept = value), size = 0.8) +
+  facet_wrap(vars(long), scales = "free", ncol = 3) +
+  xlab("v statistic") +
+  geom_text(data = v_annot, aes(x = Inf, y = Inf, label = txt), size = 3,
+            hjust = +1.2, vjust = +1.5) +
+  scale_x_continuous(expand = c(0, 0)) +
+  scale_y_continuous(expand = c(0, 0, 0, 0.2)) +
+  theme_universe
+
+ggsave(here("plots", "cooc_vstats.eps"), 
+       plot = v_plot_eps, 
+       width = 19, height = 20, units = "cm")
+
 
 # normalizing to range 0-1 ------------------------------------------------
 v_normalized <- bind_rows(exp = v_exp_g, obs = v_obs_g, .id = "orig") %>% 
@@ -236,6 +252,22 @@ s_plot <- ggplot(as_tibble(s_experimental), aes(value)) +
 s_plot
 
 ggsave(here("plots", "cooc_sstat.pdf"), plot = s_plot, 
+       width = 6, height = 4, units = "cm")
+
+# EPS
+s_plot_eps <- ggplot(as_tibble(s_experimental), aes(value)) +
+  geom_density(fill = "gray80", color = NA) +
+  geom_vline(xintercept = s_observed, size = 0.8) +
+  geom_text(data = tibble(p = paste0(round(s_p_value, 3), "**")), 
+            aes(x = Inf, y = Inf, label = p), 
+            size = 3,
+            hjust = +1.1, vjust = +1.4) +
+  xlab("S statistic") +
+  scale_x_continuous(expand = c(0, 0)) +
+  scale_y_continuous(expand = c(0, 0, 0, 0.2)) +
+  theme_universe
+
+ggsave(here("plots", "cooc_sstat.eps"), plot = s_plot_eps, 
        width = 6, height = 4, units = "cm")
 
 # The S statistic is significantly larger than experimental S and

### 2021-06-04 code/04-spat_neighbors.R

@@ -95,6 +95,9 @@ g <- gridExtra::grid.arrange(g_delaunay, g_gabriel, nrow = 1)
 ggsave(plot = g, filename = here("plots/plan_graphs.pdf"), 
        width = 19, height = 8, units = "cm")
 
+# EPS
+ggsave(plot = g, filename = here("plots/plan_graphs.eps"), 
+       width = 19, height = 8, units = "cm")
 
 # graph objects for delaunay and gabriel ----------------------------------
 # gabriel

### 2021-06-04 code/05-spat_functions.R

@@ -119,7 +119,7 @@ plot_estimate_facet <- function(x) {
   stopifnot(all(names(x) %in% c("fun", "r", "obs", "theo", "lo", "hi")))
   p1 <- x %>% 
     ggplot() +
-    geom_ribbon(aes(r, ymin = lo, ymax = hi), fill = "gray80", alpha = 0.6) +
+    geom_ribbon(aes(r, ymin = lo, ymax = hi), fill = "gray80") +
     geom_line(aes(r, theo), linetype = 2) +
     geom_line(aes(r, obs), size = 0.8) + 
     facet_wrap(vars(fun), nrow = 1, scales = "free") +
@@ -189,6 +189,9 @@ grid_fns
 
 ggsave(here("plots", "pp_funs.pdf"), width = 19, height = 5, units = "cm")
 
+# EPS
+ggsave(here("plots", "pp_funs.eps"), width = 19, height = 5, units = "cm")
+
 # for different marks ==========================================================
 multiplecrossfunction <- function(pp, var, crossfun, nsim, nrank) {
   input <- subset(pp, select = var, drop = TRUE)
@@ -229,7 +232,7 @@ mppk_orig <- multiplecrossfunction(ved_pp, "origin", "Kcross", nsim, nrank)
 # K funs
 g1 <- mppk_sex %>% 
   ggplot(aes(x = r)) +
-  geom_ribbon(aes(ymin = lo, ymax = hi), fill = "gray80", alpha = 0.6) +
+  geom_ribbon(aes(ymin = lo, ymax = hi), fill = "gray80") +
   geom_line(aes(y = theo), linetype = 2) +
   geom_line(aes(y = obs), size = 0.8) + 
   scale_x_continuous(expand = c(0, 0)) +
@@ -240,7 +243,7 @@ g1 <- mppk_sex %>%
 
 g2 <- mppk_age %>% 
   ggplot(aes(x = r)) +
-  geom_ribbon(aes(ymin = lo, ymax = hi), fill = "gray80", alpha = 0.6) +
+  geom_ribbon(aes(ymin = lo, ymax = hi), fill = "gray80") +
   geom_line(aes(y = theo), linetype = 2) +
   geom_line(aes(y = obs), size = 0.8) + 
   scale_x_continuous(expand = c(0, 0)) +
@@ -251,7 +254,7 @@ g2 <- mppk_age %>%
 
 g3 <- mppk_orig %>% 
   ggplot(aes(x = r)) +
[…]

### 2021-06-04 code/06-spat_perc.R

@@ -177,6 +177,9 @@ p$stats %>%
 
 ggsave(here::here("plots", "perc_distance.pdf"), width = 10, height = 7, units = "cm")
 
+# EPS
+ggsave(here::here("plots", "perc_distance.eps"), width = 10, height = 7, units = "cm")
+
 # mean and median nr of nodes
 # p$stats %>% 
 #   ggplot(aes(x = radius)) +
@@ -236,7 +239,7 @@ polygs <- ved_sf %>%
   select(id_burial, radius, clust) %>% 
   filter(radius %in% c(1.6, p_breaks[1:7])) %>% 
   group_by(radius) %>% 
-  nest() %>% 
+  nest() %>%
   mutate(data = map(data, group_by, clust),
          data = map(data, nest),
          data = map(data, rename, pts = data),
@@ -280,6 +283,9 @@ ggplot() +
 ggsave(here::here("plots", "perc_plan.pdf"), 
        width = 14, height = 21, units = "cm")
 
+ggsave(here::here("plots", "perc_plan.eps"), 
+       width = 14, height = 21, units = "cm")
+
 # numbers of clusters
 polygs_ids <- filter(polygs, radius %in% disp_radii) %>% 
   group_by(radius) %>%
@@ -297,7 +303,7 @@ ggplot() +
                             label.size = NA,
                             label.padding = .2,
                             na.rm = TRUE,
-                            fill = alpha(c("white"), 0.8)) +
+                            fill = "white") + # alpha(c("white"), 0.8)
   scale_shape_manual(values = c(22, 21, 24, 4)) +
   facet_wrap(vars(label_radius), ncol = 2) +
   coord_sf() +
@@ -315,6 +321,9 @@ ggplot() +
         legend.direction = "horizontal")
 
 ggsave(here::here("plots", "perc_plan_ids.pdf"),
+       width = 14, height = 21, units = "cm")
+
+ggsave(here::here("plots", "perc_plan_ids.eps"),
        width = 14, height = 21, units = "cm")
 
 # cluster assignment ------------------------------------------------------

### 2021-06-04 text/article_vedrovice.Rmd
add figure placeholders

@@
 categories: female-sexed body (abbreviated *F*), male-sexed body (*M*),
-non-adult (*n. a.*) and undetermined sex (*ind.*).
+non-adult (*n. a.*) and undetermined sex (*ind.*); see Fig. \@ref(fig:plan-sex).
 The age of the body is also simplified into three categories: juveniles (*juv.*)
@@
 and mature bodies (*mat.*, anthropological categories *Adultus-Maturus*, *Maturus* and
-*Senilis*) and bodies of an undetermined age (*ind.*).
+*Senilis*) and bodies of an undetermined age (*ind.*); see Fig. \@ref(fig:plan-age).
 The last category is the indicator for the origin of the buried person.
@@
 Bodies falling within this range are labeled as local, outside the range as non-local;
-where data on Sr values is missing, origin is not determined (*ind.*).
+where data on Sr values is missing, origin is not determined (*ind.*); see Fig. \@ref(fig:plan-orig).
+
+[Figure_04]
+
+[Figure_05]
+
+[Figure_06]
 
@@
 
+[Figure_09]
+
 <!-- Percolation analysis -->
@@
 This suggests that there is strong evidence that the overall distribution of
-artifacts in different graves is not random. <!--(see Fig. \@ref(fig:sstat))-->
+artifacts in different graves is not random (see Fig. \@ref(fig:sstat)).
+
+[Figure_01]
 
@@
 special types of pottery vessel shapes, spondylus pendants and buckles.
+
+[Figure_02]
 
@@
 (b) adzes.
+
+[Figure_03]
 
@@
 
+[Figure_07]
+
 This suggests that the locations for new burials were not chosen at random
@@
 
+[Figure_10]
+
 The results of a percolation analysis reveal a more detailed view of the clustering process
@@
 The clusters at chosen distances are shown in Fig. \@ref(fig:perc-plan).
+
+[Figure_11]
 
@@
 This proves a tighter bound cluster of juveniles in the southeast part of the cemetery.
+
+[Figure_08]
 

### 2021-08-24 code/02-spat_dataprep.R
update eps exports

@@ -88,8 +88,8 @@ gg_baseplan <- g_exc +
 
 gg_baseplan
 
-gg_ids <- gg_baseplan + 
-  ggsflabel::geom_sf_text_repel(data = ved_sf, aes(label = id_burial))
+# gg_ids <- gg_baseplan + 
+#   ggsflabel::geom_sf_text_repel(data = ved_sf, aes(label = id_burial))
 
 # ggsave(here("plots", "plan_base.pdf"), gg_baseplan, scale = 2)
 # ggsave(here("plots", "plan_base_ids.pdf"), gg_ids, scale = 2)
@@ -115,7 +115,7 @@ g_exc +
   facet_wrap(~origin)
 
 ggsave(here("plots", "plan_origin.pdf"), width = 12, height = 6, units = "cm")
-ggsave(here("plots", "plan_origin.eps"), width = 12, height = 6, units = "cm")
+ggsave(here("plots", "plan_origin.eps"), width = 15, height = 7, units = "cm")
 
 
 # plan age
@@ -128,7 +128,7 @@ g_exc +
   facet_wrap(~age_sim)
 
 ggsave(here("plots", "plan_age.pdf"), width = 19, height = 6, units = "cm")
-ggsave(here("plots", "plan_age.eps"), width = 19, height = 6, units = "cm")
+ggsave(here("plots", "plan_age.eps"), width = 22, height = 7, units = "cm")
 
 # plan sex
 g_exc +
@@ -140,7 +140,7 @@ g_exc +
   facet_wrap(~sex)
 
 ggsave(here("plots", "plan_sex.pdf"), width = 19, height = 6, units = "cm")
-ggsave(here("plots", "plan_sex.eps"), width = 19, height = 6, units = "cm")
+ggsave(here("plots", "plan_sex.eps"), width = 22, height = 7, units = "cm")
 
 # write layouts -----------------------------------------------------------
 

### 2021-08-24 code/05-spat_functions.R

@@ -280,7 +280,7 @@ ggsave(plot = grob, here::here("plots", "pp_crossk.pdf"),
 
 # EPS
 ggsave(plot = grob, here::here("plots", "pp_crossk.eps"), 
-       width = 14, height = 21, units = "cm")
+       width = 18, height = 26, units = "cm")
 
 # # G funs
 # mppg_sex %>% 

### 2021-08-24 code/06-spat_perc.R
explicitly ignore two-point clusters (ie. geometry line)

@@ -234,6 +234,8 @@ get_concaveman <- function(pts) {
   map(pts, concaveman)
 }
 
+# correct geometries - polygon of length 3 == line
+# groups of two point are omitted for the analysis
 polygs <- ved_sf %>% 
   left_join(p$membership_long, by = c("id_burial" = "id")) %>% 
   select(id_burial, radius, clust) %>% 
@@ -243,13 +245,18 @@ polygs <- ved_sf %>%
   mutate(data = map(data, group_by, clust),
          data = map(data, nest),
          data = map(data, rename, pts = data),
-         data = map(data, ~map(.x$pts, concaveman))) %>% 
-  unnest(cols = data) %>% 
-  unnest(cols = data) %>% 
+         data = map(data, ~map(.x$pts, concaveman))) %>%
+  unnest(cols = data) %>%
+  mutate(len = map_int(data, \(x) nrow(st_coordinates(x))),
+         data = if_else(len > 0 & len < 4, map(data, st_cast, "LINESTRING"), data)) %>% 
+  unnest(cols = data) %>%
+  ungroup("radius") %>% 
+  filter(len > 3) %>% 
+  select(-len) %>% 
   st_as_sf()
 
 polygs <- polygs %>% 
-  subset(st_is_valid(polygs)) %>% 
+  # subset(st_is_valid(polygs)) %>% 
   st_make_valid() %>% 
   st_buffer(1.6) %>% 
   mutate(label_radius = paste("dist.", radius, "m"),
@@ -280,11 +287,11 @@ ggplot() +
         legend.position = "bottom",
         legend.direction = "horizontal")
 
-ggsave(here::here("plots", "perc_plan.pdf"), 
-       width = 14, height = 21, units = "cm")
-
-ggsave(here::here("plots", "perc_plan.eps"), 
-       width = 14, height = 21, units = "cm")
+# ggsave(here::here("plots", "perc_plan.pdf"), 
+#        width = 14, height = 21, units = "cm")
+# 
+# ggsave(here::here("plots", "perc_plan.eps"), 
+#        width = 16, height = 24, units = "cm")
 
 # numbers of clusters
 polygs_ids <- filter(polygs, radius %in% disp_radii) %>% 
@@ -324,8 +331,8 @@ ggsave(here::here("plots", "perc_plan_ids.pdf"),
        width = 14, height = 21, units = "cm")
 
 ggsave(here::here("plots", "perc_plan_ids.eps"),
-       width = 14, height = 21, units = "cm")
[…]

### 2021-08-24 text/article_vedrovice.Rmd
update text before submission

--- 2021-06-04 text/article_vedrovice.Rmd
@@
 title           : "Spatial Patterns and Grave Goods Differences at the Cemetery of Vedrovice (Czech Republic): A Resampling Approach to Identity Markers in the Early Neolithic"
-author          : "Petr Pajdla"
+author          : ""
 date            : "`r format(Sys.time(), '%d %B, %Y')`"
@@
         number_sections: false
+        global_numbering: false
         keep_tex: false
 bibliography    : "bib_vedrovice.bib"
-csl             : "apa.csl"
-header-includes:
-  - \usepackage[left]{lineno}
-  - \linenumbers
+csl             : "journal-of-computer-applications-in-archaeology.csl"
 ---
@@
 
-# Introduction
+# Introduction {-}
 
@@
 Considering the Neolithic burial evidence across Europe, authors find that it
-"shows very little gender distinction; when it does, it is rarely a clear-cut
+'shows very little gender distinction; when it does, it is rarely a clear-cut
 binary categorization but a matter of degree and overlap, sometimes polythetically
-combined with factors such as age and localness" [@robb2017, p. 138].
-In response, @bickle2020 [p. 14] argues that "biological sex may not have defined gender
-during the LBK" although the "sexed bodies did matter".
+combined with factors such as age and localness' [@robb2017, p. 138].
+In response, @bickle2020 [p. 14] argues that 'biological sex may not have defined gender
+during the LBK' although the 'sexed bodies did matter'.
 
@@
 
-# Material and methods
+# Material and Methods
 
@@
 categories: female-sexed body (abbreviated *F*), male-sexed body (*M*),
-non-adult (*n. a.*) and undetermined sex (*ind.*).
+non-adult (*n. a.*) and undetermined sex (*ind.*); see Figure \@ref(fig:plan-sex).
 The age of the body is also simplified into three categories: juveniles (*juv.*)
@@
 and mature bodies (*mat.*, anthropological categories *Adultus-Maturus*, *Maturus* and
-*Senilis*) and bodies of an undetermined age (*ind.*).
+*Senilis*) and bodies of an undetermined age (*ind.*); see Figure \@ref(fig:plan-age).
 The last category is the indicator for the origin of the buried person.
@@
 Bodies falling within this range are labeled as local, outside the range as non-local;
-where data on Sr values is missing, origin is not determined (*ind.*).
+where data on Sr values is missing, origin is not determined (*ind.*); see Figure \@ref(fig:plan-orig).
+
+<!--
+[Figure_04]
+
+[Figure_05]
+
+[Figure_06]
+-->
 
@@
 
-## Randomness of artifact combinations
+## Randomness of Artifact Combinations
 
@@
 <!-- spatial analysis -->
-## Spatial analysis
+## Spatial Analysis
 
@@
 We use Delaunay triangulation and its subgraph, a Gabriel graph [@gabriel1969]
-with burial locations as nodes (Fig. \@ref(fig:plan-graphs)).
+with burial locations as nodes (Figure \@ref(fig:plan-graphs)).
 The Delaunay triangulation is regarded as the maximum number of neighbors per individual case
@@
 derived from 999 permutations of the simulated data.
+
+<!--[Figure_09]-->
 
@@
 <!-- The size of the radius is based on the results of nearest neighbors analysis, where
-at the given radius there is strong evidence of a clustering process (Fig. \@ref(fig:pp-funs)).
+at the given radius there is strong evidence of a clustering process (Figure \@ref(fig:pp-funs)).
 The randomness is again tested using a permutation test similar to the one
@@
 [@rcoreteam2021] using functions of several different packages; 
-the most frequently used are packages from the *tidyverse* family [@wickham2019a].
+the most frequently used are packages from the *tidyverse* family [@wickham2019].
 Packages *sf* [@pebesma2018], *spatstat* [@baddeley2015], *spdep* [@bivand2018] and 
@@
 
-## Artifact combinations
+## Artifact Combinations
 
@@
 This suggests that there is strong evidence that the overall distribution of
-artifacts in different graves is not random. <!--(see Fig. \@ref(fig:sstat))-->
+artifacts in different graves is not random (see Figure \@ref(fig:sstat)).
+
+<!--[Figure_01]-->
 
@@
 Close to the 0.05 alpha level are distributions of bone tools and ceramic bottles
-(Fig. \@ref(fig:cooc-vstat)).
+(Figure \@ref(fig:cooc-vstat)).
 Among the grave goods found to be randomly distributed in the graves
@@
 
+<!--[Figure_02]-->
+
 The relationships between artifact types frequently occurring and missing together
-in graves are visualized in Fig. \@ref(fig:cooc-nets) with
+in graves are visualized in Figure \@ref(fig:cooc-nets) with
 clusters established using a fast-greedy algorithm [@newman2004] and the strength of the
@@
 (b) adzes.
+
+<!--[Figure_03]-->
 
@@
 
-## Spatial patterning
+## Spatial Patterning
 
@@
 placing of the individual burials. -->
-The results of functions based on nearest neighbors are reported in Fig. \@ref(fig:pp-funs).
+The results of functions based on nearest neighbors are reported in Figure \@ref(fig:pp-funs).
 Both *G* and  *F* functions slightly deviate from the curve expected
@@
 
+<!--[Figure_07]-->
+
 This suggests that the locations for new burials were not chosen at random
@@
 
+<!--[Figure_10]-->
+
 The results of a percolation analysis reveal a more detailed view of the clustering process
-taking place; see Fig. \@ref(fig:perc-distance).
+taking place; see Figure \@ref(fig:perc-distance).
 The distances at which the relative stability of clusters emerges are labeled.
@@
 This point of cluster growth is revealed by the point pattern analysis as well.
-The clusters at chosen distances are shown in Fig. \@ref(fig:perc-plan).
+The clusters at chosen distances are shown in Figure \@ref(fig:perc-plan).
 
-## Neighborhood structure
+<!--[Figure_11]-->
+
+## Neighborhood Structure
 
 The results of the multitype *K* function show deviations from random processes for
-most of the observed categories in Fig. \@ref(fig:pp-crossk).
+most of the observed categories in Figure \@ref(fig:pp-crossk).
 The distances at which these deviations start occurring are in most cases rather large,
@@
 
+<!--[Figure_08]-->
+
 Multitype *K* for the combination of age and sex categories suggests a clustering process
@@
 number of neighbors scenario) and
-Gabriel graph (minimum number of neighbors scenario, Fig. \@ref(fig:plan-graphs)),
+Gabriel graph (minimum number of neighbors scenario, Figure \@ref(fig:plan-graphs)),
 most of the observed relationships are not significantly different from those
@@
 Regarding the determination of the sex of the bodies, at the threshold of 3.9 meters
-we observe that in the largest cluster (no. 3, Fig. \@ref(fig:perc-plan))
+we observe that in the largest cluster (no. 3, Figure \@ref(fig:perc-plan))
 there are more non-adults and less male-sexed bodies and in the second largest
@@
 <!--
-For relationships (see Fig. \@ref(fig:nb-sex) and Tab. \@ref(tab:nb-tab-sex))
+For relationships (see Figure ref(fig:nb-sex) and Tab. ref(tab:nb-tab-sex))
 of male-sexed bodies, none of the methods showed any significant
@@
 
-# Conclusions and discussion
+# Conclusions and Discussion {-}
 
@@
 
+<!-- Acknowledgement / Reproducibility -->
+
+# Reproducibility {-}
+
+To enable re-use and reproducibility, a compendium containing data and code accompanying this paper is released at Zenodo data repository under DOI [10.5281/zenodo.4899652)](https://doi.org/10.5281/zenodo.4899652).
+
+# Acknowledgements {-}
+
+The research described in this paper was accomplished with support from the project ‘Lifestyle as an unintentional identity in the Neolithic’ (Project 19-16304S), financed by the Czech Science Foundation.
+
 <!-- ======================================================================= -->
@@
              ".8\\textwidth",
-             ".98\\textwidth")
+             ".9\\textwidth")
 ```
@@
 ```{r label-pp-funs}
-label_pp_funs <- paste(
-  "Functions *G*, *F*, *K* and *T*.",
-  "*Dashed line* - theoretical curve under complete spatial randomness;
-  *solid line* - empirical curve; *gray area* - envelope")
+label_pp_funs <- paste("Functions G, F, K and T. Dashed line – theoretical curve under complete spatial randomness; solid line – empirical curve; gray area – envelope")
 ```

### 2021-08-24 code/knit_text.R

@@ -10,7 +10,8 @@ rmarkdown::render(input = here::here("text", "article_vedrovice.Rmd"),
                     toc = FALSE,
                     latex = "pdflatex",
                     fig_caption = TRUE,
-                    number_sections = FALSE,
+                    number_sections = TRUE,
+                    global_numbering = FALSE,
                     keep_tex = FALSE
                   ))
 

### 2021-08-24 code/knit_text_word.R

@@ -9,5 +9,6 @@ rmarkdown::render(input = here::here("text", "article_vedrovice.Rmd"),
                   output_format = bookdown::word_document2(
                     toc = FALSE,
                     fig_caption = TRUE,
-                    number_sections = FALSE
+                    number_sections = FALSE,
+                    global_numbering = FALSE
                   ))

### 2022-03-31 code/01-dataset.R
update code

@@ -18,8 +18,8 @@ binarize <- function(x, threshold = 0) {
 }
 
 # data input and manipulation ==================================================
-input <- readr::read_csv(here("data", "data_vedrovice_v01.csv"), skip = 3) %>% 
-  filter(lbk == TRUE)
+input <- readr::read_csv(here("data", "data_vedrovice_v02.csv"), skip = 3) %>% 
+  filter(space)
 
 # list to store the data
 ved <- list(bin_vars = list(count_mat = NA,
@@ -40,18 +40,19 @@ ved <- list(bin_vars = list(count_mat = NA,
 
 # # burial id
 ved$id_burials <- input %>%
-  filter(undisturbed == TRUE) %>%
+  filter(analysis == TRUE) %>%
   pull(id_burial)
 
 # binary variables - artefact coocurences
 ved$bin_vars$count_mat <- input %>% 
-  filter(undisturbed == TRUE) %>% 
+  filter(analysis == TRUE) %>% 
   select(pigment, pol_adze, pol_axe, grinding, pebble, 
          lit_local, lit_nonlocal, 
          pot_special, pot_bowl, pot_globular, pot_bottle, 
          # pot_head, pot_mid, pot_leg, 
-         bone_tool, bone_awl,
-         pendant_L, pendant_U, pendant_I, buckle_O, 
+         bone_tool, # bone_awl,
+         pendant, # pendant_L, pendant_U, pendant_I, 
+         buckle_O, 
          bracelet_spond, beads_spond, beads_marble, 
          deer_teeth, antler, shell) %>% 
   as.matrix()
@@ -71,12 +72,13 @@ ved$bin_vars$complete_mat <- ved$bin_vars$bin_mat %>%
 rownames(ved$bin_vars$complete_mat) <- ved$id_burials
 
 # grave goods present in more than 5% of graves
-n_burs <- nrow(ved$bin_vars$bin_mat)
+n_burs <- nrow(filter(input, analysis))
 ved$bin_vars$over5 <- (colSums(ved$bin_vars$bin_mat) / n_burs) > 0.05
+table(ved$bin_vars$over5)
 
 # continuous variables
 ved$cont_vars <- input %>% 
-  filter(undisturbed == TRUE) %>% 
+  filter(analysis == TRUE) %>% 
   select(
     # pit_len, pit_wid, pit_dep,
     # d13c, d15n, 
[…]

### 2022-03-31 code/02-spat_dataprep.R

@@ -4,17 +4,17 @@
 # author: Petr Pajdla
 # Data prep for spatial analysis and basic plans
 
-library(here)
 # library(spatstat)
 library(sf)
 library(tidyverse)
+library(patchwork)
 
 # read data
-ved <- read_rds(here("data/temp", "vedrovice_dataset.RDS"))
+ved <- read_rds(here::here("data/temp", "vedrovice_dataset.RDS"))
 
 # data prep ====================================================================
 scale_m <- 4.762 # scale by a factor so 1 unit is 1 meter
-angle <- 40 / 360 * 100 * 1/2 # rotate by an angle of 40° so north faces upwards
+angle <- 40 / 360 * 100 * 1/2 # rotate by an angle of 51° so north faces upwards
 
 ved_layout <- DescTools::Rotate(ved$layout * scale_m, 
                                 theta = angle, mx = 0, my = 0)
@@ -38,9 +38,9 @@ ved_sf <- st_as_sf(bind_cols(ved_layout, ved$metadata),
 
 # excavation polygon
 ved_exc <- matrix(c(c(0.5, 0.5, 0, 0, -1, -1, 1.6, 1.6, 7.6, 7.6, 9.1, 9.1, 
-                      12.3, 12.3, 13.5, 13.5, 12.3, 12.3, 8.6, 8.6),
+                      12.3, 12.3, 13.5, 13.5, 14, 14, 13.5, 13.5, 12.3, 12.3, 8.6, 8.6),
                     c(-1, 2.5, 2.5, 10.3, 10.3, 14.3, 14.3, 13.6, 13.6, 
-                      12.8, 12.8, 13.4, 13.4, 12.4, 12.4, 4.5, 4.5, 1.5, 1.5, -1)), 
+                      12.8, 12.8, 13.4, 13.4, 12.4, 12.4, 7.5, 7.5, 6.5, 6.5, 4.5, 4.5, 1.5, 1.5, -1)), 
                   ncol = 2, byrow = FALSE)
 ved_exc <- DescTools::Rotate(ved_exc * scale_m, theta = angle, mx = 0, my = 0)
 ved_exc <- st_cast(
@@ -66,27 +66,33 @@ window_exc <- spatstat.geom::as.owin(ved_exc)
 #   theme(legend.position = c(0.9, 0.8))
 
 g_exc <- ggplot() + 
-  geom_sf(data = ved_exc, fill = "gray90", color = NA) +
-  ggspatial::annotation_north_arrow(style = ggspatial::north_arrow_minimal(),
[…]

### 2022-03-31 code/03-randomization.R

@@ -9,7 +9,7 @@
 set.seed(42)
 
 # packages =====================================================================
-library(here)
+
 library(dplyr)
 library(purrr)
 library(igraph)
@@ -72,7 +72,7 @@ normalize01 <- function(x) {
 }
 
 # read data ====================================================================
-ved <- readRDS(here("data/temp", "vedrovice_dataset.RDS"))
+ved <- readRDS(here::here("data/temp", "vedrovice_dataset.RDS"))
 
 # counting co-occurrences in observed matrix ===================================
 input_matrix <- ved$bin_vars$bin_mat[, ved$bin_vars$over5]
@@ -97,7 +97,7 @@ no_cores <- parallel::detectCores() - 1
 
 # cooccurences on random matrices ==============================================
 # ======
-# note to self: on Debian without everything was smooth even without specifying 
+# note to self: on Debian everything was smooth even without specifying 
 # setup_strategy, but apparently, this is Rstudio vs R version problem
 # see: https://github.com/rstudio/rstudio/issues/6692
 cl <- parallel::makeCluster(no_cores, setup_strategy = "sequential")
@@ -105,11 +105,11 @@ cooc_rand <- parallel::parLapply(cl, rand_mat$perm, count_cooccurrence)
 parallel::stopCluster(cl)
 
 # save counts of co-occurences on random matrices
-readr::write_rds(cooc_rand, here("data/temp", "cooc_random_mat.RDS"))
+readr::write_rds(cooc_rand, here::here("data/temp", "cooc_random_mat.RDS"))
 # ======
 
 # load from temporary data
-cooc_rand <- readr::read_rds(here("data/temp", "cooc_random_mat.RDS"))
+# cooc_rand <- readr::read_rds(here::here("data/temp", "cooc_random_mat.RDS"))
 
 # summarization of random occurences into expected cooccurences ----------------
 cooc_exp <- cooc_rand %>% bind_rows() %>% 
@@ -149,14 +149,16 @@ v_statistic <- tibble(variable = names(v_observed),
   mutate(abbrv = unname(ved$var_names$short[variable]),
          long = unname(ved$var_names$long[variable]))
 
[…]

### 2022-03-31 code/04-spat_neighbors.R

@@ -4,7 +4,6 @@
 # author: Petr Pajdla
 # Mean number of neighbours with a given sex based on Gabriel graph
 
-library(here)
 library(tidyverse)
 library(broom)
 library(sf)
@@ -29,25 +28,30 @@ theme_universe <- theme(panel.border = element_rect(colour = "black",
 # data --------------------------------------------------------------------
 
 # ved <- read_rds(here("data/temp", "vedrovice_dataset.RDS"))
-ved_sf <- read_sf(here("data/temp", "layout.shp")) %>% 
+ved_sf <- read_sf(here::here("data/temp", "layout.geojson")) %>% 
   mutate(sex = fct_relevel(sex, c("n. a.", "F", "M", "ind.")),
          age_sim = fct_relevel(age_sim, c("juv.", "ad.", "mat.", "ind.")),
-         origin = fct_relevel(origin, c("local", "non-local", "ind.")))
-ved_layout <- ved_sf %>% 
-  filter(pres != "dist.") %>% 
-  st_coordinates()
-ved_exc <- read_sf(here("data/temp", "window.shp"))
+         origin = fct_relevel(origin, c("local", "non-local", "ind."))) %>% 
+  st_set_crs(NA)
 
-ved_sf_pres <- filter(ved_sf, pres != "dist.") %>% 
-  mutate(age_sim = forcats::fct_drop(age_sim))
+ved_layout <- ved_sf %>% 
+  st_coordinates()
+
+ved_exc <- read_sf(here::here("data/temp", "window.geojson")) %>% 
+  st_set_crs(NA)
+
+# ved_sf_pres <- filter(ved_sf, pres != "dist.") %>% 
+#   mutate(age_sim = forcats::fct_drop(age_sim))
 
 # create delaunay + gabriel neighborhoods ---------------------------------
+
 ved_gabriel <- gabrielneigh(ved_layout)
 ved_delaunay <- tri2nb(ved_layout)
 
 # plot delaunay triangulation and gabriel graph ---------------------------
+
 ved_gabriel_lines <- ved_gabriel %>% 
-  graph2nb(row.names = ved_sf_pres$id_burial) %>% 
+  graph2nb(row.names = ved_sf$id_burial) %>% 
   nb2lines(coords = ved_layout) %>% 
   st_as_sf()
 
@@ -56,9 +60,9 @@ ved_delaunay_lines <- ved_delaunay %>%
   st_as_sf()
 
 g_gabriel <- ggplot() +
[…]

### 2022-03-31 code/05-spat_functions.R

@@ -5,6 +5,7 @@
 # Spatial organization within the cemetery - point patters analysis
 
 library(here)
+library(patchwork)
 library(spatstat)
 library(sf)
 library(tidyverse)
@@ -23,19 +24,21 @@ theme_universe <- theme(panel.border = element_rect(colour = "black",
                                                     fill = NA, 
                                                     size = 0.8),
                         panel.background = element_blank(),
-                        line = element_blank(),
+                        # line = element_blank(),
                         strip.background = element_blank(), 
-                        strip.text = element_text(face = "italic"),
-                        axis.text.y = element_blank(), 
-                        axis.title.y = element_blank(), 
-                        panel.spacing = unit(1, "lines"))
+                        # strip.text = element_text(face = "italic"),
+                        # axis.text.y = element_blank(), 
+                        # axis.title.y = element_blank(), 
+                        # panel.spacing = unit(0.6, "lines")
+)
 
 # read data and layouts --------------------------------------------------------
 ved <- read_rds(here("data/temp", "vedrovice_dataset.RDS"))
 
-ved_sf <- read_sf(here("data/temp", "layout.shp")) %>% 
-  filter(pres != "dist.")
-ved_exc <- read_sf(here("data/temp", "window.shp")) %>% 
+ved_sf <- read_sf(here("data/temp", "layout.geojson")) %>% 
+  st_set_crs(NA)
+ved_exc <- read_sf(here("data/temp", "window.geojson")) %>% 
+  st_set_crs(NA) %>% 
   as.owin(ved_exc)
 
 ved_sf <- bind_cols(ved_sf, 
@@ -97,38 +100,41 @@ plot_estimate <- function(x, fun) {
   stopifnot(verifyclass(x, "tbl_df"))
   stopifnot(all(names(x) %in% c("r", "obs", "theo", "lo", "hi")))
   ylabel = paste0(fun, "(r)")
-  p1 <- x %>% 
+  p1 <- x %>%
     ggplot() +
[…]

### 2022-03-31 code/06-spat_perc.R

@@ -25,28 +25,33 @@ library(sf)
 library(forcats)
 
 # theme for ggplot graphics
-theme_universe <- theme(panel.border = element_rect(colour = "black", 
-                                                    fill = NA, 
-                                                    size = 0.8),
-                        panel.background = element_blank(),
-                        line = element_blank(),
-                        strip.background = element_blank(), 
-                        strip.text = element_text(face = "italic"),
-                        # axis.text.y = element_blank(), 
-                        # axis.title.y = element_blank(), 
-                        panel.spacing = unit(1, "lines"))
+theme_universe <- theme(panel.border = element_rect(
+  colour = "black", 
+  fill = NA, 
+  size = 0.8
+),
+panel.background = element_blank(),
+line = element_blank(),
+strip.background = element_blank(), 
+# strip.text = element_text(face = "italic"),
+# axis.text.y = element_blank(), 
+# axis.title.y = element_blank(), 
+panel.spacing = unit(1, "lines"))
 
 # data --------------------------------------------------------------------
 
-ved_sf <- sf::read_sf(here::here("data/temp", "layout.shp")) %>% 
+ved_sf <- sf::read_sf(here::here("data/temp", "layout.geojson")) %>% 
+  st_set_crs(NA) %>% 
   mutate(sex = fct_relevel(sex, c("n. a.", "F", "M", "ind.")),
          age_sim = fct_relevel(age_sim, c("juv.", "ad.", "mat.", "ind.")),
-         origin = fct_relevel(origin, c("local", "non-local", "ind."))) %>% 
-  filter(pres != "dist.")
+         origin = fct_relevel(origin, c("local", "non-local", "ind.")))
+
 ved_coord <- bind_cols(id = as.character(ved_sf$id_burial), 
                        sf::st_coordinates(ved_sf)) %>% 
   rename(x = X, y = Y)
-ved_exc <- sf::read_sf(here::here("data/temp", "window.shp"))
+
+ved_exc <- sf::read_sf(here::here("data/temp", "window.geojson")) %>% 
+  st_set_crs(NA)
 
 
 # percolation -------------------------------------------------------------
[…]

### 2022-03-31 code/07-spat_perc_clusters.R

@@ -17,22 +17,24 @@ library(sf)
 
 # data --------------------------------------------------------------------
 
-explored_rad <- paste0(c(2.8, 3.9, 4.6, 5.5, 6.5, 7.8))
+explored_rad <- paste0(c(2.1, 2.8, 3.4, 4.1, 4.6, 5.6, 6.3, 8.8, 9.2))
 
 perc_cl <- read_csv(here::here("data/temp", "perc_clusters.csv")) %>% 
   mutate(id_burial = as.character(id)) %>% 
   select(-id) %>% 
   filter(radius %in% explored_rad)
 
-ved_sf <- sf::st_read(here::here("data/temp", "layout.shp")) %>% 
+ved_sf <- sf::st_read(here::here("data/temp", "layout.geojson")) %>% 
+  st_set_crs(NA) %>% 
   mutate(sex = fct_relevel(sex, c("n. a.", "F", "M", "ind.")),
          age_sim = fct_relevel(age_sim, c("juv.", "ad.", "mat.", "ind.")),
          origin = fct_relevel(origin, c("local", "non-local", "ind."))) %>% 
-  filter(pres != "dist.")
+  filter(analysis)
 
 # prep ---------------------------------------------------------------------
 
-input <- ved_sf %>% select(id_burial, sex, age_sim, origin, starts_with("cat_")) %>% 
+input <- ved_sf %>% 
+  select(id_burial, sex, age_sim, origin, starts_with("cat_")) %>% 
   st_drop_geometry() %>% 
   left_join(perc_cl, by = c("id_burial"))
 
@@ -164,62 +166,62 @@ devs_loc <- inner_join(deviates_from_random(input, "origin", nsim),
             expected = (expected.x + expected.y)/2,
             deviation = (deviation.x + deviation.y)/2)
 
-# multicategorical variables
-devs_sa <- inner_join(deviates_from_random(input, "cat_sa", nsim),
-                      deviates_from_random(input, "cat_sa", nsim), 
-                      by = c("radius", "cluster", "variable")) %>% 
-  transmute(radius, cluster, variable,
-            observed = observed.x,
-            expected = (expected.x + expected.y)/2,
-            deviation = (deviation.x + deviation.y)/2) %>% 
-  inner_join(deviates_from_random(input, "cat_sa", nsim),
-             by = c("radius", "cluster", "variable")) %>% 
-  transmute(radius, cluster, variable,
-            observed = observed.x,
[…]

### 2022-03-31 text/article_vedrovice.Rmd
text and references before review
update texts

@@
 The Cemetery of Vedrovice is one of the earliest and most intensively studied
-*Linearbandkeramik* (LBK) cemeteries.
+*Linearbandkeramik* (LBK) cemeteries in the Czech Republic.
 Recent articles use evidence from the cemetery in a broader discussion
@@
 <!-- site -->
-The Vedrovice site is situated in south Moravia (Czech Republic).
+The Vedrovice site is situated in south Moravia, Czech Republic.
 The earliest known finds and archaeological explorations here are from
@@
 <!-- sample (burials, artifacts) -->
-Originally, this burial ground consisted of at least 108 burials.
-Some of the burials were disturbed by later settlement and burial pits in the
-area of the cemetery, amateur excavations during the 19th century and later
-agricultural activities.
-These burials are excluded from most of the analysis due to missing and/or
-incomplete data and only a sample of 78 well-preserved LBK burials is analyzed here.
-Most of the data was obtained from the original publication [@podborsky2002],
+Originally, this burial ground consisted of at least 96 LBK burials and
+one Lengyel culture cremation [@ondrus2002a, 21].
+Some of the LBK burials were disturbed by later settlement pits in the
+area of the cemetery (4 burials), amateur excavations during the 19th century
+(11 burial pits with recent fills) and later agricultural activities
+(at least 6 burials disturbed by deep deep plowing) or combinations of these.
+These burials are excluded from the analysis of artifact combinations due to missing and/or
+incomplete data and only included in the spatial analysis.
+This leaves us with a sample of of 76 well-preserved LBK burials in case of
+the artifact combinations (79 % of the LBK burials) and
+Most of the data was obtained from the original publication [@ondrus2002a],
 and the anthropological determination of the sex and age of the bodies was corrected based on an
